---
title: "Canopy & light-environment models"
---
::: {.eyebrow}
implementation
:::

::: {.callout-note}
## Prerequisites
This page builds on [The big picture](../overview.qmd) and
[the FF16 model](../models/ff16.qmd). Read those first if the notation here is
unfamiliar.
:::

## The idea first

Imagine looking down on a forest. Tall trees intercept light first, leaving
shorter trees in their shade. A leaf lower in the canopy has more leaf area
above it, so less open-sky light reaches it. This accumulation of shade with
depth is the central idea behind a canopy, or light-environment, model.

Rather than following every leaf, the model constructs a vertical light profile:
at every height, it records the fraction of open-sky light that remains. Near
the canopy top, that fraction is close to one; lower down, crowns above have
filtered more of the light. All plants in a patch use the same profile. Each
plant reads it to determine the light reaching its leaves, while also changing
it by shading the plants below. Growth therefore depends not only on a plant's
own size and strategy, but also on the light environment created by its
neighbours.

`plant` provides several ways to represent this process. Some resolve shading
throughout each crown; others use faster approximations that sacrifice some
detail. This page explains the six implemented canopy and shading models, how
each calculates the light environment and assimilation, the trade-off between
realism and speed, and why the perfect-plasticity approximation (PPA) must be
smoothed for `plant`'s adaptive solver. It is an implementation reference for
the current code.

<!-- out-of-scope: the full model comparison — demographic equilibrium, invasion
     (mutant) fitness, the box-competition fitness comparison, the evolutionary
     attractor / singular-strategy / selection-gradient analysis from
     DunFalster 2026 — belongs in the paper reproduction, not this reference
     page. See the pointer at the foot of this page. -->

## How canopy shading affects the model

### Light interception, and why it matters

Light is the currency of forest competition. A plant's carbon gain depends on
the light intercepted by its leaves. That light has already been filtered by
taller neighbours' crowns and by leaves higher in the plant's own crown
(self-shading). `plant` represents this with one shared patch **light profile**,
$E(z)$: the fraction of open-sky light reaching height $z$. Every plant both
*reads* this profile to photosynthesise and *contributes* to it, because its
leaves lower $E$ for plants beneath them.

This profile is consequential, not a minor implementation detail. It determines
plant growth, which determines stand size structure and competitive hierarchy,
and ultimately affects which strategies persist and which traits selection
favours. Errors in shading can therefore propagate to the ecological and
evolutionary predictions that the model is designed to make.

### Realism versus speed

A fully detailed model would resolve the light field throughout every crown and
integrate photosynthesis across it. This is costly, particularly when leaf-level
photosynthesis is itself expensive: `TF24` uses coupled stomatal/hydraulic
optimisation rather than a simple light-response curve. A common alternative is
to simplify the crown to a point, or the canopy to a small number of layers.
That saves computation but reduces fidelity.

`plant` lets you choose a point on this spectrum. The important question is not
which simplification is universally best, but **what each one changes**. As the
model comparison shows, an approximation can be suitable for one purpose and
seriously misleading for another.

### The models

The `FF16` strategy offers six shading models, selected with
`control$shading_model`. Start by separating two ideas: the shade a plant
*casts* (competition) and how it turns the light it *receives* into assimilation.
Four models use the same smooth Yokozawa leaf-area profile for competition.
They differ only in how they calculate a plant's own assimilation, except that
PPA also changes how the light profile is assembled:

* **`"deep-crown"`** (the default) -- the original `plant` behaviour. Leaf area
  is distributed continuously through the crown according to the Yokozawa (1995)
  foliage profile. Gross assimilation is calculated by *integrating*
  photosynthesis over crown depth against the light profile. The resulting patch
  light profile is smooth with height.

* **`"crown-centre"`** -- builds the light profile exactly as `"deep-crown"` does,
  but approximates assimilation with a *single evaluation* at the crown centre
  rather than an integral over crown depth. Skipping the per-plant quadrature
  makes it faster, but removes within-crown self-shading.

* **`"mean-light"`** -- sits between deep-crown and crown-centre. It integrates
  the *light* over crown depth, weighted by leaf-area density, to obtain one
  leaf-area-weighted mean light level, then evaluates photosynthesis once at
  that mean. It therefore retains the changing within-crown light environment
  as the canopy develops, while replacing many photosynthesis calculations with
  *one* per plant. The saving is modest for FF16, whose leaf photosynthesis is a
  cheap saturating function. It matters more for models with expensive leaf
  submodels: `TF24`'s coupled stomatal/hydraulic optimisation makes integrating
  light and optimising once the natural choice. The trade-off is bias. Because
  photosynthesis saturates with light, the rate evaluated at mean light exceeds
  the depth-averaged rate (Jensen's inequality). `"mean-light"` therefore always
  gives at least as much assimilation as `"deep-crown"`; they agree only when
  light is uniform within the crown.

* **`"ppa"`** -- the perfect-plasticity approximation used by large models such
  as FATES. It divides the patch light profile into discrete canopy *layers*:
  light decreases by a fixed optical depth
  (`control$ppa_layer_optical_depth`) in each layer. Assimilation is evaluated
  at the crown centre, as in `"crown-centre"`, but reads this stepped profile.
  A hard step is not differentiable and causes the ODE solver to fail, so layer
  boundaries are smoothed with a $C^1$-continuous ramp whose width is set by
  `control$ppa_layer_smoothing`. This creates two versions of the single
  `"ppa"` model, paralleling the Flat Top pair below: **PPA (hard step)**
  (`ppa_layer_smoothing = 0`) is the literal field discretisation (e.g. Strigul
  2008, FATES). It is discontinuous and does not run in `plant`'s adaptive
  solver. **PPA (smoothed)** (`ppa_layer_smoothing > 0`, default `0.3`) is the
  runnable version.

The remaining two models change *competition* itself: the shade a plant casts.
Instead of spreading leaf area through the crown, they collapse it towards a
thin layer at the crown centre. Many large vegetation models, including
LPJ-GUESS, use a similar discrete-layer view of canopy competition, so these
are useful approximations to examine rather than dismiss:

* **`"flat-top-box"`** -- casts a hard *step* of shade: full below the crown
  centre and none above. Across cohorts this makes the patch light profile
  discontinuous, so `plant` cannot solve the stand at all.
* **`"flat-top-soft-box"`** -- uses the same box competition but smooths the step
  into a continuous drop, allowing the light environment to be built. It runs,
  but its incorrectly shaped shade biases predictions.

The selected model is resolved once when a strategy is prepared, so offering the
choice has no per-call cost. `TF24` supports `"deep-crown"`, `"mean-light"`
(its default), and `"crown-centre"`, but not the PPA or box variants.

```{r}
#| label: libraries
#| message: false
library(plant)
library(dplyr)
library(ggplot2)
library(patchwork)
```

## A patch simulated under each model

We grow one `FF16` species in a patch and run the identical patch under each of
the four smooth-competition models. Because everything except
`control$shading_model` is fixed, differences among runs can be attributed to
the shading model alone.

```{r}
#| label: run-models
params <- scm_base_parameters("FF16")
patch  <- add_strategies(params, trait_matrix(0.0825, "lma"), birth_rate = 1)

models <- c("deep-crown", "mean-light", "crown-centre", "ppa")

run_model <- function(model) {
  ctrl <- Control()
  ctrl$shading_model <- model
  run_scm(patch, ctrl = ctrl, collect = TRUE)
}

results <- lapply(models, run_model)
names(results) <- models
```

Each run returns its full collected state (`run_scm(collect = TRUE)`): a tidy
`species` table, with one row for every node at every step, and the light
environment (`env$light_availability`). We label each `species` table with its
shading model, then combine the tables so that `plant`'s tidy helpers can work
across all four runs.

```{r}
#| label: bind-species
species_all <- dplyr::bind_rows(
  lapply(models, function(m) dplyr::mutate(results[[m]]$species, method = m))
)
species_all$method <- factor(species_all$method, levels = models)

# Light-extinction coefficient, used throughout to convert canopy openness to
# cumulative leaf area: E = exp(-k_I * LAI), so LAI_above = -log(E) / k_I.
k_I <- results[[1]]$p$strategies[[1]]$pars$k_I
```

## What happens to the stand

The first and least demanding test is whether a stand develops sensibly. We
inspect the same patch in three complementary ways as it grows and self-thins:
its size distribution, leaf area index, and vertical light profile.

### Size distribution

The size distribution consists of node height trajectories. Each line follows a
cohort introduced at a particular patch age; line opacity is scaled by cohort
density, so sparse cohorts appear faint. `plant`'s `plot_size_distribution()`
builds this view directly from the tidy `species` table, and facets by `method`
make the shading models easy to compare.

```{r}
#| label: fig-sizedist
#| fig-width: 10
#| fig-height: 3.4
#| fig-cap: "Size-distribution trajectories under each shading model."
p_size <- plot_size_distribution(species_all) +
  facet_wrap(~method) +
  theme(legend.position = "bottom")
p_size
```

The discrete layers in `"ppa"` create visibly different trajectories. Cohorts
grow under nearly constant light within a layer, then accelerate as they emerge
into a brighter one.

### Leaf area index

Leaf area index (LAI) is total one-sided leaf area per unit ground area. It is
the cumulative leaf area above the ground, and the recorded light profile
already contains it. By Beer's law, $E = \exp(-k_I\,\mathrm{LAI})$, so patch LAI
is $-\log E(0) / k_I$, where $E(0)$ is canopy openness at ground level. We can
therefore read it directly from `env$light_availability`, without integrating
over the size distribution. LAI is real leaf area, so the smooth recorded
profile gives its true value for all four models: PPA layering changes the light
plants *experience*, not the leaf area that is present.

```{r}
#| label: lai-table
lai_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    results[[m]]$env$light_availability |>
      dplyr::group_by(time) |>
      dplyr::slice_min(height, n = 1, with_ties = FALSE) |>
      dplyr::ungroup() |>
      dplyr::transmute(method = m, time, lai = -log(light_availability) / k_I)
  })
)
lai_all$method <- factor(lai_all$method, levels = models)
```

```{r}
#| label: fig-lai
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Leaf area index over patch age under each shading model."
p_lai <- ggplot(lai_all, aes(time, lai, colour = method)) +
  geom_line(linewidth = 1) +
  labs(x = "Patch age (years)", y = "Leaf area index", colour = "Shading model") +
  theme_classic()
p_lai
```

`"deep-crown"`, `"mean-light"`, and `"crown-centre"` closely track one another
and settle to a steady LAI. Under a nearly uniform light profile, their three
ways of summarising within-crown light produce similar carbon gain.
`"mean-light"` and `"crown-centre"` are slightly richer because they do not
fully resolve self-shading. In contrast, `"ppa"` has large, persistent
oscillations: discrete light layers create sharp thresholds for cohort success,
leading to successive waves of recruitment and self-thinning. The pulses of
emerging cohorts are also visible in the size-distribution panel above.

### Vertical light profile

Next, we look *within* the canopy at selected patch ages. The vertical profile
of cumulative leaf area above each height is $\mathrm{LAI}_{above}(z) =
-\log E(z) / k_I$, where $E$ is canopy openness. This is the quantity that
determines how strongly each individual is shaded.

`env$light_availability` records the smooth underlying profile shared by all
models. PPA plants actually *experience* a stepped version of that profile, so
the `"ppa"` panel applies the same internal layering transform (see
`FF16_Environment::step_light`).

```{r}
#| label: profile-data
d <- Control()$ppa_layer_optical_depth   # layer thickness (optical depth)
w <- Control()$ppa_layer_smoothing       # boundary smoothing fraction

# C1-smoothed staircase: flat over the lower (1 - w) of each layer, cubic ramp
# over the top w. Mirrors FF16_Environment::smooth_floor() / step_light().
smooth_floor <- function(u, w) {
  n <- floor(u)
  f <- u - n
  t <- pmax(0, f - (1 - w)) / w
  n + ifelse(f <= 1 - w, 0, t^2 * (3 - 2 * t))
}
ppa_step_light <- function(E, d, w) {
  ifelse(E >= 1 | E <= 0, E,
         exp(-d * smooth_floor(-log(E) / d, w)))
}

target_ages <- c(5, 10, 20, 60)

profile_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    light <- results[[m]]$env$light_availability
    if (m == "ppa") {
      light$light_availability <- ppa_step_light(light$light_availability, d, w)
    }
    # nearest recorded step to each target age
    steps <- results[[m]]$steps
    keep <- sapply(target_ages, function(a) steps$step[which.min(abs(steps$time - a))])
    light |>
      dplyr::filter(step %in% keep) |>
      dplyr::mutate(method = m,
                    age = target_ages[match(step, keep)],
                    lai_above = -log(light_availability) / k_I)
  })
)
profile_all$method <- factor(profile_all$method, levels = models)
```

```{r}
#| label: fig-vprofile
#| fig-width: 10
#| fig-height: 3.6
#| fig-cap: "Vertical profile of cumulative leaf area above each height, at several patch ages."
p_profile <- ggplot(profile_all,
                    aes(lai_above, height, colour = factor(age), group = age)) +
  geom_path() +
  facet_wrap(~method) +
  labs(x = expression(LAI[above]~(m^2/m^2)), y = "Height (m)",
       colour = "Patch age\n(years)") +
  theme_classic()
p_profile
```

`"deep-crown"`, `"mean-light"`, and `"crown-centre"` have the same smooth
profile because they construct the light environment identically and differ only
in their assimilation calculation. `"ppa"` has its characteristic staircase:
cumulative leaf area is constant within a layer and jumps at layer boundaries.

## Computational cost: work per step and solver steps

Speed is why the simpler models exist. One patch is inexpensive, but `plant` is
often run thousands of times for trait sweeps, model fitting, or fitness
landscapes. At that scale, reducing the cost of an inner calculation matters.

The saving depends on two competing effects. Unlike many vegetation models,
`plant` integrates a patch with an **adaptive** ODE solver. The solver chooses
large steps where dynamics are smooth and small steps where they are not, aiming
to achieve the requested accuracy efficiently. A model's total cost therefore
depends both on the work *per step* and on the *number of steps* the solver
chooses:

* **Work per step:** `"deep-crown"` is most expensive because it performs a
  Gauss-Kronrod quadrature over crown depth for every individual. `"crown-centre"`
  and `"ppa"` replace this with one light evaluation.
* **Number of steps:** a smooth light profile lets the solver stride. PPA's
  near-stepped profile is much stiffer, forcing many more and smaller steps.

Which effect dominates depends on how the system is stepped. We therefore report
both adaptive cost (the cost normally paid) and per-step cost on a fixed
schedule.

```{r}
#| label: speed-helper
# Median wall-clock over a few repeats. `times = NULL` uses the adaptive solver;
# supplying a fixed schedule pins every model to the same steps.
time_model <- function(model, times = NULL, reps = 5) {
  p <- patch
  use_fixed <- !is.null(times)
  if (use_fixed) p$ode_times <- times
  ctrl <- Control()
  ctrl$shading_model <- model
  run1 <- function() run_scm(p, ctrl = ctrl, use_ode_times = use_fixed)
  run1()                                   # warm up
  median(replicate(reps, system.time(run1())[["elapsed"]]))
}

speed_table <- function(timings) {
  data.frame(
    model = names(timings),
    `median time (s)` = round(unname(timings), 4),
    `relative to deep-crown` = round(unname(timings / timings[["deep-crown"]]), 2),
    check.names = FALSE,
    row.names = NULL
  )
}
```

### Adaptive solver: real-world cost

This is `plant`'s default mode and the cost normally paid. Adaptive stepping is
part of what makes `plant` efficient: the solver spends steps only where the
dynamics need them. It also means that cost is not determined solely by work per
step. A model that is harder to integrate pays an additional cost through extra
steps. Most large vegetation models instead use a *fixed* timestep, where this
second consideration does not arise; we return to that case below.

```{r}
#| label: speed-adaptive
timings_adaptive <- vapply(models, time_model, numeric(1))
knitr::kable(speed_table(timings_adaptive))
```

`"crown-centre"` is fastest. It shares deep-crown's smooth light profile, and
therefore its solver-step count, but skips the per-plant crown-depth integral.
`"mean-light"` costs about as much as `"deep-crown"`: it still integrates over
crown depth, although it integrates light rather than rate, so both per-step
work and step count are similar. Despite cheaper per-plant assimilation,
`"ppa"` is *slower* overall because its near-stepped profile makes the adaptive
solver take many more steps. Finer layers (smaller `ppa_layer_optical_depth`) or
sharper boundaries (smaller `ppa_layer_smoothing`) make this effect stronger.

### Fixed ODE schedule: per-step cost

When all models are pinned to the *same* fixed schedule
(`run_scm(..., use_ode_times = TRUE)`), the step-count effect disappears and
only per-step cost remains. PPA's advantage is then clear: like
`"crown-centre"`, it skips crown-depth quadrature, so both are markedly faster
than `"deep-crown"`.

```{r}
#| label: speed-fixed
times <- seq(0, params$max_patch_lifetime, length.out = 4000)
timings_fixed <- vapply(models, time_model, numeric(1), times = times)
knitr::kable(speed_table(timings_fixed))
```

PPA is therefore not inherently slow: on a fixed schedule it is *faster* than
`"deep-crown"`. Its apparent adaptive-solver cost comes entirely from the extra
steps required by its stiff profile. This also explains why discrete-layer
schemes are unproblematic in large models that use fixed timesteps. There, a
stepped, stiff profile costs no additional steps, so the difficulty does not
appear. `plant`'s adaptive solver makes stiffness visible and, as the next
section shows, makes a *hard* step fail outright. `"deep-crown"` remains the
default because it is the most faithful representation of within-crown light
capture and has no speed penalty relative to earlier versions of `plant`.

## Why PPA must be smoothed

A literal PPA discretisation is a *hard* step function. When a growing plant's
crown centre crosses a layer boundary, the light it receives jumps
discontinuously, and so does its growth rate. `ppa_layer_smoothing` exists to
remove this problem. The issue is specifically **adaptive ODE stepping**, not
PPA itself.

### Why adaptive stepping fails at a discontinuity

`plant` solves patch dynamics with an adaptive, error-controlled embedded
Runge-Kutta scheme. At each step, it makes two state estimates of different
orders. Their difference estimates local truncation error. If that error is
larger than the requested tolerance, the solver *rejects* the step and retries
with a smaller one. This relies on a smooth solution: halving the step should
roughly quarter the error.

That assumption fails at a discontinuity. However small the step becomes, a
step that straddles the jump has an O(jump) error that never falls below the
tolerance. The solver keeps rejecting and shrinking the step until it gives up.
This is exactly what happens with a hard step (`ppa_layer_smoothing = 0`):

```{r}
#| label: ppa-hard
ctrl_hard <- Control()
ctrl_hard$shading_model <- "ppa"
ctrl_hard$ppa_layer_smoothing <- 0      # hard step, no smoothing

tryCatch(
  run_scm(patch, ctrl = ctrl_hard),
  error = function(e) conditionMessage(e)
)
```

A *fixed* schedule (`run_scm(..., use_ode_times = TRUE)`) avoids the
error-control loop because it simply takes its prescribed steps. However, it
then has no protection against a large explicit step overshooting the jump,
which sends growth to non-finite values unless the schedule is very fine. A hard
step is therefore not robustly handled by either approach.

Changing the rounding direction (using light at the top rather than the bottom
of a layer) does not solve this. The problem is the discontinuity, not its
location.

### The smoothed profile

`ppa_layer_smoothing` replaces each hard boundary with a $C^1$-continuous cubic
ramp over the upper fraction of the layer (see `FF16_Environment::smooth_floor`).
The light profile, and therefore each plant's growth rate, now has a continuous
first derivative. The adaptive error estimate behaves as the solver expects, so
steps are accepted normally. Even a small amount of smoothing restores a
well-behaved integration; with the default, adaptive and fixed-step solutions
agree.

Two parameters control the discretisation:

* `ppa_layer_optical_depth` -- the optical-depth thickness of a layer (default
  `0.5`, meaning one unit of leaf area index per layer at the default extinction
  coefficient `k_I = 0.5`). Smaller values produce finer layers.
* `ppa_layer_smoothing` -- the fraction of each layer over which its boundary is
  smoothed (default `0.3`). Values approaching `0` recover a hard step and its
  instability; `1` removes the flat region and approaches a smooth
  `"deep-crown"`-style profile. This creates a trade-off: smaller values more
  closely represent a true PPA step but are stiffer and slower, as the speed
  table shows; larger values integrate faster but blur the layers.

## How the models represent competition

The four smooth-competition models above construct the *same* light environment.
They differ only in how a plant summarises its own assimilation. The two box
models instead alter the shade a plant *casts*, concentrating leaf area near a
thin layer at the crown centre. Because this mirrors the discrete-layer
competition used by large vegetation models such as LPJ-GUESS, it is useful to
inspect the difference directly.

```{r}
#| label: fig-compbox
#| fig-width: 7
#| fig-height: 3.6
#| fig-cap: "Shade cast (competition profile) by a 10 m plant under the correct crown-centre profile versus the hard and soft box approximations."
comp_profile <- function(model, height = 10) {
  s <- FF16_Strategy()
  s$control$shading_model <- model
  ind <- FF16_Individual(s)
  ind$set_state("height", height)
  z <- seq(0, height, length.out = 400)
  data.frame(model = model, z = z,
             competition = sapply(z, function(zz) ind$compute_competition(zz)))
}
comp <- dplyr::bind_rows(
  comp_profile("crown-centre"),
  comp_profile("flat-top-box"),
  comp_profile("flat-top-soft-box")
)

ggplot(comp, aes(competition, z, colour = model)) +
  geom_path() +
  labs(x = "Shade cast (competition) by a 10 m plant", y = "Height (m)",
       colour = "Model") +
  theme_classic()
```

The correct crown-centre profile tapers continuously to zero through the crown.
The box profile is a step; the soft box replaces that step with a
continuous-but-incorrect drop concentrated near the crown centre.

**The hard step does not even run.** When summed across a population of discrete
cohorts, step competition produces a discontinuous, piecewise-constant patch
light profile. The adaptive light-environment spline cannot represent it, so the
model fails before a stand develops:

```{r}
#| label: box-fail
ctrl_box <- Control()
ctrl_box$shading_model <- "flat-top-box"
tryCatch(
  run_scm(patch, ctrl = ctrl_box),
  error = function(e) conditionMessage(e)
)
```

The soft box runs because its continuous profile allows the light environment to
be built. However, it distributes shade incorrectly through the canopy and so
biases predictions relative to the deep-crown reference.

<!-- out-of-scope: the demographic-equilibrium iteration, the invasion (mutant)
     fitness landscapes, the box-vs-correct fitness comparison, and the
     evolutionary-attractor / singular-strategy / selection-gradient analysis
     (DunFalster 2026) were dropped here — they are the substance of the paper
     reproduction, not the implementation reference. The continuity requirement
     they establish is summarised below; the full analysis is reproduced in the
     reproduction post linked at the foot of this page. -->

## What this comparison shows

The full model comparison (see the reproduction below) tests these models in
increasingly demanding settings: stand dynamics, demographic equilibrium,
invasion fitness, and the evolutionary endpoint. Its recurring result is that
cheap models are often adequate for stand dynamics and acceptable for
equilibrium, whereas invasion fitness is much less forgiving. The key
requirement is **continuity of the shade each plant casts as a function of
height**. `"deep-crown"`, `"mean-light"`, and `"crown-centre"` all construct a
continuous light profile and behave well; PPA's stepped profile and the box
models' incorrectly shaped competition do not.

This does not mean that large models are wrong to use discrete layers. For
stand-level simulation on a fixed timestep, that representation is fast,
robust, and adequate. The caution applies specifically to the eco-evolutionary
questions that `plant` is built to answer.

The full model comparison and evolutionary analysis -- demographic equilibrium,
invasion-fitness landscapes, and singular-strategy / work-precision results --
is reproduced in the DunFalster 2026 canopy reproduction post.
<!-- TODO: link this once the DunFalster 2026 reproduction is published. The
     draft lives in posts/_drafts/2026-06-20-dunfalster2026-canopy/ and is not
     rendered, so no link target exists yet. Restore as
     [DunFalster 2026 canopy reproduction](../../posts/<slug>.qmd) when it moves
     out of _drafts/. -->
