---
title: "Canopy & light-environment models"
---
::: {.eyebrow}
implementation
:::

This page documents the canopy / shading models that `plant` implements: what
the six models are, how each computes the light environment a plant experiences,
the realism-versus-speed tradeoff between them, and why the perfect-plasticity
approximation (PPA) requires smoothing to run in `plant`'s adaptive solver. It
is an implementation reference for the current code.

<!-- out-of-scope: the full model comparison — demographic equilibrium, invasion
     (mutant) fitness, the box-competition fitness comparison, the evolutionary
     attractor / singular-strategy / selection-gradient analysis from
     DunFalster 2026 — belongs in the paper reproduction, not this reference
     page. See the pointer at the foot of this page. -->

# Background

## Light interception, and why it matters

Light is the currency of competition in a forest. A plant's carbon gain depends
on the light its leaves intercept, and that light has been filtered twice: by
the crowns of taller neighbours above it, and by the plant's own leaves higher
in its crown (self-shading). `plant` captures this through a single shared
object -- the patch **light profile** $E(z)$, the fraction of open-sky light
reaching height $z$. Every plant both *reads* this profile (to photosynthesise)
and *contributes* to it (the shade its leaves cast lowers $E$ below them).

Getting the light profile right is not a detail. It sets each plant's growth
rate, which sets the size structure of the stand, which sets the competitive
hierarchy, which ultimately decides which strategies persist and which traits
selection favours. Errors here propagate all the way up to the ecological and
evolutionary predictions the model exists to make.

## Realism versus speed

A fully realistic treatment would resolve the light field through each crown and
integrate photosynthesis over it. That is expensive -- especially when the
leaf-level photosynthesis model is itself costly (as in `TF24`, whose coupled
stomatal/hydraulic optimisation is far heavier than a simple light-response
curve). The classic response is to *simplify*: treat the crown as a single
point, or the canopy as a few discrete layers -- trading fidelity for speed.
`plant` lets you choose where to sit on that spectrum. The job of this page is
to describe **what each simplification does**, because, as the model comparison
shows, a model can be perfectly adequate for one question and badly wrong for
another.

## The models

The `FF16` strategy offers six shading models, selected with
`control$shading_model`. Four of them share the *same* per-plant competition --
the smooth Yokozawa leaf-area profile -- and differ only in how a plant's own
assimilation is computed (and, for PPA, how the light profile is assembled):

* **`"deep-crown"`** (the default) -- the original `plant` behaviour. Leaf area
  is distributed continuously through the crown following the Yokozawa (1995)
  foliage profile, and a plant's gross assimilation is computed by *integrating*
  photosynthesis over crown depth against the light profile. The patch light
  profile is a smooth function of height.

* **`"crown-centre"`** -- the light profile is built exactly as for `"deep-crown"`,
  but a plant's assimilation is approximated by a *single evaluation* of the
  light at its crown centre rather than an integral over depth. This skips the
  per-plant quadrature, so it is faster, and removes within-crown self-shading.

* **`"mean-light"`** -- partway between deep-crown and crown-centre. Rather than
  integrating the *photosynthetic rate* over depth, it integrates the *light*
  over depth (weighted by leaf-area density) to a single leaf-area-weighted mean
  light, then takes one photosynthesis evaluation of that mean. It therefore
  still feels the within-crown change in light environment -- and how that environment
  shifts as the canopy develops -- but collapses the expensive photosynthesis
  calculation to a *single* call per plant rather than one per crown depth. For
  FF16 that saving is modest (its leaf photosynthesis is a cheap saturating
  function), but it becomes important for models with an expensive leaf
  submodel: the `TF24` strategy's coupled stomatal/hydraulic optimisation is far
  more costly to evaluate, so integrating the light and optimising once is the
  natural choice there. The cost is bias: because photosynthesis saturates with
  light, evaluating the rate at the mean light overestimates the depth-averaged
  rate (Jensen's inequality). So `"mean-light"` always gives at least as much
  assimilation as `"deep-crown"`, the two agreeing only when the within-crown
  light is uniform.

* **`"ppa"`** -- the perfect-plasticity approximation, the scheme that underlies
  large models such as FATES. The patch light profile is discretised into
  discrete canopy *layers*: light steps down by a fixed amount of optical depth
  (`control$ppa_layer_optical_depth`) per layer.
  Assimilation is evaluated at the crown centre as in `"crown-centre"`, but now
  reads this stepped profile. Because a hard step is not differentiable (and
  makes the ODE solver fail), the layer boundaries are smoothed with a
  C^1^-continuous ramp whose width is set by `control$ppa_layer_smoothing`.
  This gives two variants of the *one* `"ppa"` model, exactly paralleling the
  Flat Top pair below: **PPA (hard step)** (`ppa_layer_smoothing = 0`) is the
  literal field discretisation (e.g. Strigul 2008, FATES) -- discontinuous, and
  it does not run in `plant`'s adaptive solver -- while **PPA (smoothed)**
  (`ppa_layer_smoothing > 0`, default `0.3`) is the runnable version.

Two further models instead change the *competition* itself -- the shade a plant
casts -- collapsing its leaf area towards a thin crown-centre layer rather than
spreading it through the crown. This discrete-layer view is close to how many
large vegetation models (e.g. LPJ-GUESS) represent canopy competition, so it is
worth taking seriously rather than dismissing:

* **`"flat-top-box"`** -- the shade cast is a hard *step*: full below the crown
  centre, none above. Summed over cohorts this makes the patch light profile
  discontinuous, and in `plant` the stand then cannot be solved at all.
* **`"flat-top-soft-box"`** -- the same box competition with the step smoothed to
  a continuous drop, so the light environment can be built. It runs, but the
  mis-shaped shade biases its predictions.

The model is resolved once when a strategy is prepared, so there is no per-call
cost to having the choice available. The `TF24` strategy supports `"deep-crown"`,
`"mean-light"` (its default) and `"crown-centre"`, but not the PPA or box
variants.

```{r}
#| label: libraries
#| message: false
library(plant)
library(dplyr)
library(ggplot2)
library(patchwork)
```

# Simulating a patch under each model

We grow a single `FF16` species in a patch, and run the same patch under each of
the four smooth-competition models. Everything except `control$shading_model`
is held fixed, so any differences are due to the shading model alone.

```{r}
#| label: run-models
params <- scm_base_parameters("FF16")
patch  <- expand_parameters(trait_matrix(0.0825, "lma"), params)

models <- c("deep-crown", "mean-light", "crown-centre", "ppa")

run_model <- function(model) {
  ctrl <- Control()
  ctrl$shading_model <- model
  run_scm(patch, ctrl = ctrl, collect = TRUE)
}

results <- lapply(models, run_model)
names(results) <- models
```

Each run returns the full collected state (`run_scm(collect = TRUE)`): a tidy
`species` table (one row per node per step) and the light environment
(`env$light_availability`). We tag each run's `species` table with its shading
model and bind them together so we can use `plant`'s tidy helpers across all
four.

```{r}
#| label: bind-species
species_all <- dplyr::bind_rows(
  lapply(models, function(m) dplyr::mutate(results[[m]]$species, method = m))
)
species_all$method <- factor(species_all$method, levels = models)

# Light-extinction coefficient, used throughout to convert canopy openness to
# cumulative leaf area: E = exp(-k_I * LAI), so LAI_above = -log(E) / k_I.
k_I <- results[[1]]$p$strategies[[1]]$k_I
```

# Stand dynamics

The least demanding test of a shading model is whether a stand develops
sensibly. We look at three views of the same patch as it grows and self-thins:
the size distribution, the leaf area index, and the vertical light profile.

## Size distribution

The size distribution is the set of node height trajectories: each line is a
cohort introduced at a given patch age and followed as it grows, with line
opacity scaled by the cohort's density (faint lines are sparse cohorts).
`plant`'s `plot_size_distribution()` builds this directly from the tidy
`species` table; faceting by `method` compares the shading models.

```{r}
#| label: fig-sizedist
#| fig-width: 10
#| fig-height: 3.4
#| fig-cap: "Size-distribution trajectories under each shading model."
p_size <- plot_size_distribution(species_all) +
  facet_wrap(~method) +
  theme(legend.position = "bottom")
p_size
```

The discrete layering in `"ppa"` produces visibly different growth trajectories:
cohorts grow at near-constant light within a layer, then accelerate as they
emerge into a brighter layer.

## Leaf area index

Leaf area index (LAI) -- the total one-sided leaf area per unit ground -- is the
cumulative leaf area above the ground, which is already encoded in the recorded
light profile: with Beer's law $E = \exp(-k_I\,\mathrm{LAI})$, the patch LAI is
just $-\log E(0) / k_I$, where $E(0)$ is canopy openness at ground level. This
reads straight off `env$light_availability` and needs no integration over the
size distribution. (LAI is real leaf area, so the smooth recorded profile gives
the true value for all four models -- the PPA layering only affects the light
plants *experience*, not the leaf area present.)

```{r}
#| label: lai-table
lai_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    results[[m]]$env$light_availability |>
      dplyr::group_by(time) |>
      dplyr::slice_min(height, n = 1, with_ties = FALSE) |>
      dplyr::ungroup() |>
      dplyr::transmute(method = m, time, lai = -log(light_availability) / k_I)
  })
)
lai_all$method <- factor(lai_all$method, levels = models)
```

```{r}
#| label: fig-lai
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Leaf area index over patch age under each shading model."
p_lai <- ggplot(lai_all, aes(time, lai, colour = method)) +
  geom_line(linewidth = 1) +
  labs(x = "Patch age (years)", y = "Leaf area index", colour = "Shading model") +
  theme_classic()
p_lai
```

`"deep-crown"`, `"mean-light"` and `"crown-centre"` track each other closely and
settle to a steady LAI: under a near-uniform light profile, all three ways of
summarising within-crown light give similar carbon gain (with
`"mean-light"` and `"crown-centre"` running slightly richer, since they do not
fully resolve self-shading). The `"ppa"` model instead shows large, sustained
oscillations -- successive waves of recruitment and self-thinning -- because the
discrete light layers create sharp thresholds for cohort success (visible as the
pulses of emerging cohorts in the size-distribution panel above).

## Vertical light profile

We can also look *within* the canopy at a few patch ages, at the vertical
profile of cumulative leaf area above each height -- $\mathrm{LAI}_{above}(z) =
-\log E(z) / k_I$, where $E$ is canopy openness. This is the quantity that sets
how much each individual is shaded.

The recorded `env$light_availability` is the smooth underlying profile shared by
all models. The light that plants actually *experience* under PPA is the stepped
version, so for the `"ppa"` panel we apply the same layering transform used
internally (see `FF16_Environment::step_light`).

```{r}
#| label: profile-data
d <- Control()$ppa_layer_optical_depth   # layer thickness (optical depth)
w <- Control()$ppa_layer_smoothing       # boundary smoothing fraction

# C1-smoothed staircase: flat over the lower (1 - w) of each layer, cubic ramp
# over the top w. Mirrors FF16_Environment::smooth_floor() / step_light().
smooth_floor <- function(u, w) {
  n <- floor(u)
  f <- u - n
  t <- pmax(0, f - (1 - w)) / w
  n + ifelse(f <= 1 - w, 0, t^2 * (3 - 2 * t))
}
ppa_step_light <- function(E, d, w) {
  ifelse(E >= 1 | E <= 0, E,
         exp(-d * smooth_floor(-log(E) / d, w)))
}

target_ages <- c(5, 10, 20, 60)

profile_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    light <- results[[m]]$env$light_availability
    if (m == "ppa") {
      light$light_availability <- ppa_step_light(light$light_availability, d, w)
    }
    # nearest recorded step to each target age
    steps <- results[[m]]$steps
    keep <- sapply(target_ages, function(a) steps$step[which.min(abs(steps$time - a))])
    light |>
      dplyr::filter(step %in% keep) |>
      dplyr::mutate(method = m,
                    age = target_ages[match(step, keep)],
                    lai_above = -log(light_availability) / k_I)
  })
)
profile_all$method <- factor(profile_all$method, levels = models)
```

```{r}
#| label: fig-vprofile
#| fig-width: 10
#| fig-height: 3.6
#| fig-cap: "Vertical profile of cumulative leaf area above each height, at several patch ages."
p_profile <- ggplot(profile_all,
                    aes(lai_above, height, colour = factor(age), group = age)) +
  geom_path() +
  facet_wrap(~method) +
  labs(x = expression(LAI[above]~(m^2/m^2)), y = "Height (m)",
       colour = "Patch age\n(years)") +
  theme_classic()
p_profile
```

`"deep-crown"`, `"mean-light"` and `"crown-centre"` share an identical smooth
profile (they build the light environment the same way and differ only in how
assimilation is evaluated); `"ppa"` shows the characteristic staircase, with
cumulative leaf area held constant within a layer and jumping at each layer
boundary.

# Computational cost

Speed is the whole reason the simpler models exist. A single patch is cheap, but
`plant` is usually run thousands of times over -- sweeping traits, fitting to
data, or computing fitness landscapes -- so the per-run cost compounds, and
shaving the most expensive inner calculation is worthwhile. This is the payoff
side of the realism-versus-speed trade.

How much you save depends on two competing effects, and on a feature that is
relatively unusual in vegetation models: `plant` integrates the patch with an
**adaptive** ODE solver, which chooses its own step sizes (large where the
dynamics are smooth, small only where they are not) to hit a requested accuracy
cheaply. So the cost of a model is set both by the work *per step* and by the
*number of steps* the solver decides to take:

* **Per step**, `"deep-crown"` is the most expensive -- it performs a
  Gauss-Kronrod quadrature over crown depth for every individual -- while
  `"crown-centre"` and `"ppa"` replace that with a single light evaluation.
* **Number of steps**: a model whose light profile is smooth lets the solver
  stride; PPA's near-stepped profile is much stiffer, forcing many more, smaller
  steps.

Which effect dominates depends on how the system is stepped, so we report both
the adaptive cost (what you actually pay) and the per-step cost on a fixed
schedule.

```{r}
#| label: speed-helper
# Median wall-clock over a few repeats. `times = NULL` uses the adaptive solver;
# supplying a fixed schedule pins every model to the same steps.
time_model <- function(model, times = NULL, reps = 5) {
  p <- patch
  use_fixed <- !is.null(times)
  if (use_fixed) p$ode_times <- times
  ctrl <- Control()
  ctrl$shading_model <- model
  run1 <- function() run_scm(p, ctrl = ctrl, use_ode_times = use_fixed)
  run1()                                   # warm up
  median(replicate(reps, system.time(run1())[["elapsed"]]))
}

speed_table <- function(timings) {
  data.frame(
    model = names(timings),
    `median time (s)` = round(unname(timings), 4),
    `relative to deep-crown` = round(unname(timings / timings[["deep-crown"]]), 2),
    check.names = FALSE,
    row.names = NULL
  )
}
```

## Adaptive solver (real-world cost)

This is how `plant` runs by default, and the cost you actually pay. The adaptive
solver is part of what makes `plant` efficient: it spends steps only where the
dynamics demand them. But it also means a model's cost is not just its per-step
work -- a model that makes the dynamics harder to integrate pays for it in extra
steps. (Most large vegetation models instead march on a *fixed* timestep, where
this consideration simply does not arise; we return to that below.)

```{r}
#| label: speed-adaptive
timings_adaptive <- vapply(models, time_model, numeric(1))
knitr::kable(speed_table(timings_adaptive))
```

`"crown-centre"` is fastest -- it shares deep-crown's smooth light profile (so the
same number of solver steps) but skips the per-plant crown-depth integral.
`"mean-light"` costs about the same as `"deep-crown"`: it also integrates over
crown depth (just the light rather than the rate), so the per-step work and the
step count are both similar. `"ppa"` is *slower* overall despite the cheaper
per-plant assimilation, because its near-stepped profile forces the adaptive
solver to take many more steps; finer layers (smaller `ppa_layer_optical_depth`)
or sharper boundaries (smaller `ppa_layer_smoothing`) make this worse.

## Fixed ODE schedule (per-step cost)

If instead every model is pinned to the *same* fixed schedule
(`run_scm(..., use_ode_times = TRUE)`), the step-count effect is removed and only
the per-step cost remains. Now PPA's advantage shows: like `"crown-centre"`, it
skips the crown-depth quadrature, so both are markedly faster than
`"deep-crown"`.

```{r}
#| label: speed-fixed
times <- seq(0, params$max_patch_lifetime, length.out = 4000)
timings_fixed <- vapply(models, time_model, numeric(1), times = times)
knitr::kable(speed_table(timings_fixed))
```

So PPA is not inherently slow -- on a fixed schedule it is *faster* than
`"deep-crown"`. Its apparent cost under the adaptive solver comes entirely from
the extra steps the stiff profile demands. This is also why discrete-layer
schemes are unproblematic in the large models that use them: those models march
on a fixed timestep, so a stepped, stiff light profile costs nothing extra and
the difficulty never surfaces. It is `plant`'s adaptive solver that makes the
stiffness visible -- and, as the next section shows, a *hard* step visible as
outright failure. `"deep-crown"` nonetheless remains the default here: it is the
most faithful representation of within-crown light capture and carries no speed
penalty relative to earlier versions of `plant`.

# Why PPA needs smoothing

A literal PPA discretisation is a *hard* step function: as a plant grows and its
crown centre crosses a layer boundary, the light it experiences jumps
discontinuously, and so does its growth rate. This is what `ppa_layer_smoothing`
exists to tame, and it is worth being explicit about *why* -- because the
problem is specifically one of **adaptive ODE stepping**, not of the model
itself.

## Adaptive stepping cannot integrate a discontinuity

`plant` solves the patch dynamics with an adaptive, error-controlled solver
(an embedded Runge-Kutta scheme). At each step it forms two estimates of the
state at different orders, and uses their difference as an estimate of the local
truncation error. If that error exceeds the requested tolerance, the step is
*rejected* and retried with a smaller size; the method assumes the solution is
smooth, so halving the step should roughly quarter the error.

Across a discontinuity that assumption breaks. No matter how small the step, a
step that straddles the jump carries an O(jump) error that never falls below
tolerance, so the solver rejects and shrinks the step indefinitely until it
gives up. With a hard step (`ppa_layer_smoothing = 0`) this is exactly what
happens:

```{r}
#| label: ppa-hard
ctrl_hard <- Control()
ctrl_hard$shading_model <- "ppa"
ctrl_hard$ppa_layer_smoothing <- 0      # hard step, no smoothing

tryCatch(
  run_scm(patch, ctrl = ctrl_hard),
  error = function(e) conditionMessage(e)
)
```

A *fixed* schedule (`run_scm(..., use_ode_times = TRUE)`) sidesteps the
error-control loop -- it simply takes the prescribed steps without rejecting any
-- but then nothing protects against a large explicit step overshooting the jump,
which sends growth to non-finite values unless the schedule is very fine. Either
way, a hard step is not something the solver can handle robustly.

Switching the rounding direction (light at the top vs the bottom of a layer)
does not help: the difficulty is the discontinuity itself, not where the step
sits.

## The smoothed profile

`ppa_layer_smoothing` replaces each hard boundary with a C^1^-continuous cubic
ramp over the top fraction of the layer (see `FF16_Environment::smooth_floor`).
The light profile -- and hence each plant's growth rate -- is now continuous with
a continuous first derivative, so the adaptive error estimate behaves as the
solver assumes and steps are accepted normally. Even a small amount of smoothing
is enough to recover a well-behaved integration, and with the default the
adaptive and fixed-step solutions agree.

Two parameters control the discretisation:

* `ppa_layer_optical_depth` -- the optical-depth thickness of one layer
  (default `0.5`, i.e. one unit of leaf area index per layer at the default
  extinction coefficient `k_I = 0.5`). Smaller values give finer layering.
* `ppa_layer_smoothing` -- the fraction of each layer over which the boundary is
  smoothed (default `0.3`). Values approaching `0` recover a hard step (and the
  instability above); `1` removes the flat region and approaches the smooth
  `"deep-crown"`-style profile. There is a tension here: smaller values are
  closer to a true PPA step but stiffer (slower, per the speed table above),
  while larger values integrate faster but blur the layer structure.

# The competition profile

The four smooth-competition models above all build the *same* light environment
-- they differ only in how a plant summarises its own assimilation. The two box
models instead change the shade a plant *casts*, collapsing its leaf area towards
a thin crown-centre layer. This mirrors the discrete-layer competition of large
vegetation models (e.g. LPJ-GUESS), so it is worth inspecting directly.

```{r}
#| label: fig-compbox
#| fig-width: 7
#| fig-height: 3.6
#| fig-cap: "Shade cast (competition profile) by a 10 m plant under the correct crown-centre profile versus the hard and soft box approximations."
comp_profile <- function(model, height = 10) {
  s <- FF16_Strategy()
  s$control$shading_model <- model
  ind <- FF16_Individual(s)
  ind$set_state("height", height)
  z <- seq(0, height, length.out = 400)
  data.frame(model = model, z = z,
             competition = sapply(z, function(zz) ind$compute_competition(zz)))
}
comp <- dplyr::bind_rows(
  comp_profile("crown-centre"),
  comp_profile("flat-top-box"),
  comp_profile("flat-top-soft-box")
)

ggplot(comp, aes(competition, z, colour = model)) +
  geom_path() +
  labs(x = "Shade cast (competition) by a 10 m plant", y = "Height (m)",
       colour = "Model") +
  theme_classic()
```

The correct profile (crown-centre) tapers continuously to zero through the crown;
the box profile is a step, and the soft box a continuous-but-wrong drop
concentrated near the crown centre.

**The hard step does not even run.** Summed over a population of discrete
cohorts, a step competition makes the patch light profile a discontinuous,
piecewise-constant function of height, which the adaptive light-environment
spline cannot represent -- the model fails before a stand can develop:

```{r}
#| label: box-fail
ctrl_box <- Control()
ctrl_box$shading_model <- "flat-top-box"
tryCatch(
  run_scm(patch, ctrl = ctrl_box),
  error = function(e) conditionMessage(e)
)
```

The soft box runs (its continuous profile lets the light environment build), but
the shade is mis-distributed through the canopy, which biases its predictions
relative to the deep-crown reference.

<!-- out-of-scope: the demographic-equilibrium iteration, the invasion (mutant)
     fitness landscapes, the box-vs-correct fitness comparison, and the
     evolutionary-attractor / singular-strategy / selection-gradient analysis
     (DunFalster 2026) were dropped here — they are the substance of the paper
     reproduction, not the implementation reference. The continuity requirement
     they establish is summarised below; the full analysis is reproduced in the
     reproduction post linked at the foot of this page. -->

# What the comparison establishes

The full model comparison (see the reproduction below) puts these models through
a ladder of increasingly demanding tests -- stand dynamics, demographic
equilibrium, invasion fitness, and the evolutionary endpoint. The recurring
finding is that the cheap models are often fine for stand dynamics and
acceptable for equilibrium, but invasion fitness is unforgiving. The key
requirement it exposes is **continuity of the shade each plant casts as a
function of height**: `"deep-crown"`, `"mean-light"` and `"crown-centre"` all
build a continuous light profile and behave well; PPA's stepped profile and the
box models' mis-shaped competition do not. None of this means the large models
are wrong to use discrete layers -- for stand-level simulation on a fixed
timestep the representation is fast, robust, and adequate. The caution is
specific to the eco-evolutionary questions `plant` is built for.

The full model comparison and evolutionary analysis -- demographic equilibrium,
invasion-fitness landscapes, and the singular-strategy / work-precision results
-- is reproduced in the DunFalster 2026 canopy reproduction post.
<!-- TODO: link this once the DunFalster 2026 reproduction is published. The
     draft lives in posts/_drafts/2026-06-20-dunfalster2026-canopy/ and is not
     rendered, so no link target exists yet. Restore as
     [DunFalster 2026 canopy reproduction](../../posts/<slug>.qmd) when it moves
     out of _drafts/. -->
