---
title: "ODE stepping & control"
---
::: {.eyebrow}
implementation
:::

::: {.callout-note}
## Prerequisites
This page builds on [The size-structured PDE](../size-structured-pde.qmd) and
[the method of characteristics](characteristic-method.qmd). Read those first if
the notation here is unfamiliar.
:::

## The idea first

Once the model has been expressed as ordinary differential equations (ODEs),
the solver's task is straightforward in principle: begin with the current
state and move it forward through time in a sequence of small steps. The
important practical question is how large those steps should be. Large steps
can let unobserved numerical error accumulate; very small steps spend a great
deal of computation on periods when little is changing.

The solver balances these two risks by adapting as it goes. For each proposed
step, it also constructs a more careful estimate of that same step. The
difference between the estimates is used as an error estimate. A small
difference permits a longer next step, while a large difference triggers a
shorter, more cautious one. In this way, the calculation concentrates effort
where growth and mortality change quickly and takes larger strides through
quieter periods, while keeping the estimated error per step below a chosen
tolerance. The sections below explain how `plant` does this and how to control
it.

`plant` casts the entire patch as a system of ordinary differential equations
(ODEs) and steps it through time with classic ODE methods. (For details on the
equations themselves see the [size-structured PDE](../size-structured-pde.qmd)
page.) This page first describes the Runge–Kutta solver and the `Control()`
parameters that govern it, then works through the trade-off between the default
**adaptive** stepper and a **fixed-step** forward-Euler mode using a
work–precision diagram.

## The adaptive Runge–Kutta 45 stepper

`plant` integrates its system of ODEs with an explicit Runge-Kutta solver: the
adaptive Runge-Kutta Cash-Karp 4-5 (RK45) algorithm.

Cash-Karp RK45 is an *embedded* method: the same six evaluations of the rates of
change produce both a fourth-order and a fifth-order estimate of the next state.
Their disagreement estimates the local numerical error. If that estimate is too
large, the solver rejects the proposed step and retries with a smaller one. If
it is comfortably below the target, the solver attempts a larger step next
time. Because both estimates share the same rate evaluations, error control is
much cheaper than solving the step twice with separate methods. For general
background, see [Runge-Kutta methods](https://en.wikipedia.org/wiki/Runge-Kutta_methods).

RK45 is an *explicit* method: it calculates the next state directly from the
current state and its rates of change. An implicit method instead solves an
equation involving both the current and future states. Explicit methods are
usually simpler and cheaper per step, but can require impractically small steps
for a *stiff* system, in which important processes operate on very different
time scales. Implicit methods cost more per step but are often more stable for
such problems. See [explicit and implicit
methods](https://en.wikipedia.org/wiki/Explicit_and_implicit_methods) for more
background.

In practice, RK45 has been extremely stable and fast for the size-density
distribution. It can nevertheless struggle when an extension makes the system
stiff. For example:

- When implementing soil water into plant via the Richards Equation, Isaac
  Towers found the system became stiff.
- When implementing the FF16 model in another solver, Joshi et al 2023
  <https://doi.org/10.1101/2023.08.04.551683> reported that the system became
  stiff when enabling a continuous feedback of seeds into germination.

When extending the model, therefore, check carefully for stiff behaviour.

## How the templated C++ solver is set up

The characteristic method that solves density regularly introduces new nodes.
That means the ODE system must be resized as a simulation progresses. To support
this, Rich FitzJohn designed a custom RK45-based ODE solver class. The code is
built into the model, with some details ported from the
[GNU Scientific Library](http://www.gnu.org/software/gsl/) [@Galassi-2009].

The main interface is `odelia::ode::Solver`. It was extracted from `plant` into
the standalone `odelia` library, so it no longer lives under
`inst/include/plant/`; `plant` includes it with
`#include <odelia/ode_solver.hpp>` (see `scm.h`). The class is parameterised by
a template `System`, allowing any class that provides the required interface to
be treated as an ODE system. This is the same broad C++ idea that allows a
vector to hold a chosen data type.

From odelia's `ode_solver.hpp`:

```cpp
template <class System>
class Solver {
public:
  Solver(const System& system, OdeControl control_);
  ...
};
```

### Stepping an entire patch

The model's main solver is `SCM` (in `scm.h`). It creates an ODE solver for the
patch type and owns the patch instance containing the species and environment.

```cpp
patch_type patch;                        # Instance of patch
odelia::ode::Solver<patch_type> solver;  # ODE solver (owns the system)
```

To advance the system, call one of the following methods. They use the patch to
evaluate rates of change. One accepts specified fixed times; the other lets the
solver select an adaptive spacing. Because the solver now *owns* the system,
neither call receives the patch as an argument:

```cpp
solver.advance_fixed(e.times);
solver.advance_adaptive({solver.time(), e.time_end()});
```

A third method, `solver.advance_euler(...)`, backs the fixed-step forward-Euler
mode (`fixed_time_step`, discussed below).

Behind these calls, the solver exchanges states and rates of change with the
patch. The patch supplies the biological information that determines the rates;
the ODE solver applies RK45 to those rates to calculate future states.

The following function resizes the ODE system after a node is introduced.

```cpp
solver.set_state_from_system()
```

For this arrangement to work, the patch class (see `patch.h`) supplies the
essential functions required by the ODE system.

```cpp
// * ODE interface
// Caluclate size of ode system (number of equations).
size_t ode_size() const;
// How many auxiallary variables are we tracking. These are being collected but
// are not part of core ode system
size_t aux_size() const;
double ode_time() const;

// Retrieve ode state from patch and save into the ode solver
ode::iterator ode_state(ode::iterator it) const;
// Retrieve ode rates from patch and save into the ode solver
ode::iterator ode_rates(ode::iterator it) const;
// Retrieve auxillary variables and save into the ode solver
ode::iterator ode_aux(ode::iterator it) const;

// Set state of patch, based on estimate of future state estimated by the solver
ode::const_iterator set_ode_state(ode::const_iterator it, double time);
```

The iterators are important here: they translate the state held in a complicated
structure such as a patch into the vector required by the ODE solver, and then
translate it back again.

### Stepping individuals and other systems

Because this is templated C++, `odelia::ode::Solver` can step any "system" that
provides the required functions. As well as patches, it is used to run an
individual plant in a given environment. The wrapper class
`plant::tools::IndividualRunner` (in `individual_runner.h`) manages the
ownership relationship between the solver and the object it steps.

Like the patch system, `IndividualRunner` is templated on a strategy `T` and an
environment `E`, so it can represent a variety of systems.

```cpp
template <typename T, typename E>
class IndividualRunner {
public:
  IndividualRunner(Individual<T,E> individual_, E environment_)
    : individual(individual_), environment(environment_) {
    individual.compute_rates(environment);
  }
  // ... ode_size(), ode_time(), ode_state(), ode_rates(), set_ode_state() ...
};
```

`IndividualRunner` provides exactly the `ode_size()`/`ode_state()`/`ode_rates()`/
`set_ode_state()` interface expected by the solver. It can therefore be passed
to `odelia::ode::Solver` to step an individual plant to a given size. The R
functions `grow_individual_...` expose this capability and use
`IndividualRunner` internally.

The test suite also shows how to drive the solver directly on the classic Lorenz
system (see `tests/testthat/test-ode-euler.R`).

## Controlling ODE stepping

The RK45 solver is controlled by parameters that determine how it advances
through time.

For a first example, run a system with the default ODE controls:

```{r}
library(plant)

# configure a patch containing a single species
p0 <- scm_base_parameters("FF16")
p0$max_patch_lifetime <- 5

p <- add_strategies(p0, trait_matrix(0.0825, "lma"))

# Controls on numerical methods
ctrl <- Control()

# run
out <- run_scm(p, ctrl = ctrl)

# number of steps
length(out$ode_times)
```

The exact times used by the solver are stored in `out$ode_times`; its length is
therefore the number of recorded solver times.

The controls change both the number and timing of steps. The following extracts
the fields of `ctrl` beginning with `ode_`:

```{r}
ctrl %>% subset(grepl("(?<!n)ode_", names(.), perl = TRUE))
```

These fields have the following effects:

- `ode_step_size_initial`: The initial step size used by the solver in the first
  step.
- `ode_step_size_min`, `ode_step_size_max`: The permitted minimum and maximum
  step sizes. An attempted step outside these bounds is limited to the relevant
  bound.
- `ode_tol_abs`, `ode_tol_rel`: The allowed absolute and relative error for an
  individual equation in a step. Excess error causes a smaller attempted step;
  error well below the target permits expansion at the next step.
- `ode_a_y`, `ode_a_dydt`: Weights on y and dy/dt when calculating allowable
  relative error (see `OdeControl::errlevel`).

Changing these values can change the number of steps taken.

Unlike a conventional ODE solver, however, node introduction also affects the
step count. The stepper cannot extend a step beyond the next node-introduction
event for any species. Node scheduling therefore sets an additional upper bound
on step size. In the default setup it mainly determines time steps during the
first months of a simulation; later, the ODE stepper has more control.

### Limiting the minimum and maximum step size

First, change the maximum step size and compare the result with the default.

```{r}
#| eval: false
ctrl2 <- ctrl
ctrl2$ode_step_size_max <- 0.1

# run
out2 <- run_scm(p, ctrl = ctrl2)

# compare
length(out$ode_times)
length(out2$ode_times)

# maximum
f <- function(ode_times, title = "
") {
diff(ode_times) %>% hist(main = title, xlim = c(0,2), breaks  =seq(0, 2, length.out=20), freq = FALSE)}

f(out$ode_times, "original")
f(out2$ode_times, "max 0.1")
```

Next, change the minimum step size and make the same comparison.

```{r}
#| eval: false
ctrl3 <- ctrl
ctrl3$ode_step_size_min <- 0.01
ctrl3$ode_step_size_initial <- 0.01

# run
out3 <- run_scm(p, ctrl = ctrl3)

# compare
length(out$ode_times)
length(out3$ode_times)

f(out$ode_times, "original")
f(out3$ode_times, "min 0.01")
```

### Fixing step sizes: bounds (option 1)

Setting identical minimum and maximum sizes requests an exact step size. Other
times can still occur because of the cohort schedule.

```{r}
#| eval: false
ctrl <- Control()
ctrl$ode_step_size_min <- 0.1
ctrl$ode_step_size_max <- 0.1
ctrl$ode_step_size_initial <- 0.1
```

### Fixing step sizes: supplied times (option 2)

Alternatively, supply selected exact steps in a vector of `ode_times`.

```{r}
pc <- p
pc$node_schedule_times[[1]] <- c(0, 0.1, 0.4, 0.8, 1.2)
pc$ode_times <- seq(0, p$max_patch_lifetime, by=0.1)

result2 <- run_scm(pc, ctrl = ctrl, use_ode_times = TRUE)

result2$ode_times
```

### Fixing step sizes: forward Euler (option 3)

The two options above still invoke the full adaptive RK45 machinery at each
step. For operational-realism comparisons with daily-step DGVMs, `plant` can
instead use plain **forward (explicit) Euler** on a uniform grid. Euler uses one
derivative evaluation per step and makes no error estimate. Enable it by setting
`fixed_time_step` (in years) in `Control`:

```{r}
#| eval: false
ctrl <- Control(fixed_time_step = 1 / 365)  # daily step; 0 = adaptive (default)
run_scm(p, ctrl = ctrl)
```

This mode is not a route to either an accurate or an inexpensive solution:
forward Euler is first-order, so matching the adaptive solver's accuracy needs
many more evaluations. The [next section](#sec-adaptive-vs-fixed-step) shows
this trade-off with a work–precision diagram. `fixed_time_step` is also
incompatible with the mutant-fitness replay path
(`save_RK45_cache` / `use_ode_times`) and errors clearly if combined.

### Possible future changes

The current interface still leaves several implementation questions open:

- `NodeSchedule` interleaves `ode_times` with `node_schedule_times`. Simplifying
  that code may not justify the regression risk without a clear use case.
- Should supplied `ode_times` be stored in the parameters and used
  automatically?
- If they are present, should they always take precedence over adaptive times?
- When is reusing the times from an earlier run scientifically or
  computationally useful?

## Adaptive versus fixed-step integration {#sec-adaptive-vs-fixed-step}

By default, `plant` integrates a patch with the **adaptive, error-controlled
Cash–Karp Runge–Kutta (RK45)** stepper described above. It chooses each step
size to keep the local error within a tolerance: very small steps early in a
run, when newly introduced cohorts diverge rapidly, and much coarser steps
later, once the stand has settled.

Many widely used **dynamic global vegetation models (DGVMs)** instead move the
state forward on a *fixed* grid, most often with a **daily** time step, using
plain forward (explicit) Euler. `plant` can do this too through
`fixed_time_step` (option 3 above). This section shows how to enable it, then
uses a **work–precision diagram** to make the key comparison concrete: at
*matched accuracy*, adaptive stepping performs far less work.

```{r}
#| label: load-libraries
library(plant)
library(dplyr)
library(ggplot2)
```

### Enabling fixed-step forward Euler

One control field selects the integration mode: `fixed_time_step`, measured in
**years**:

- `fixed_time_step = 0` (the default) → adaptive RK45.
- `fixed_time_step > 0` → forward Euler on a uniform grid of that spacing.

A *daily* step is therefore `fixed_time_step = 1 / 365`.

```{r}
p0 <- scm_base_parameters("FF16")
p0$max_patch_lifetime <- 60

p <- add_strategies(p0, trait_matrix(0.0825, "lma"), birth_rate = 1.0)
env <- Environment("FF16")

ctrl_adaptive <- Control()                       # fixed_time_step = 0
ctrl_daily    <- Control(fixed_time_step = 1 / 365)

out_adaptive <- run_scm(p, env, ctrl_adaptive)
out_daily    <- run_scm(p, env, ctrl_daily)
```

Forward Euler performs **one** derivative evaluation per step
(`y <- y + h * dy/dt`); each RK45 step evaluates derivatives six times. The two
runs therefore use very different numbers of steps and report slightly
different fitness:

```{r}
tibble::tibble(
  mode    = c("adaptive RK45", "daily Euler"),
  R0      = c(out_adaptive$net_reproduction_ratios,
              out_daily$net_reproduction_ratios),
  n_steps = c(length(out_adaptive$ode_times),
              length(out_daily$ode_times))
)
```

> **Caution — don't read this as "Euler is faster".** A *single* daily step is
> cheaper than a single adaptive step (1 derivative evaluation versus 6), so a
> naive timing race makes Euler look quick. But that comparison holds the *step
> count* fixed, not the *accuracy*. Daily Euler here is visibly less accurate
> than the adaptive run. The honest question is: *how much work does each method
> need to reach the same accuracy?* That is what the work–precision diagram
> below answers — and there adaptive wins comfortably.

### A work–precision diagram

To compare methods fairly, sweep the adaptive solver across error tolerances and
forward Euler across step sizes. For every run, record:

- **work**: total derivative evaluations — `6 × n_steps` for RK45, `1 × n_steps`
  for Euler (this is the dominant cost of the inner loop);
- **error**: relative error in the net reproduction ratio $R_0$ against a
  tight-tolerance adaptive **reference**.

```{r}
## Tight-tolerance adaptive reference.
ctrl_ref <- control_accurate()
ctrl_ref$ode_tol_rel <- 1e-8
ctrl_ref$ode_tol_abs <- 1e-8
R0_ref <- run_scm(p, env, ctrl_ref)$net_reproduction_ratios

measure <- function(ctrl, mode) {
  stages <- if (mode == "adaptive") 6 else 1   # derivative evals per step
  scm <- run_scm(p, env, ctrl)
  n   <- length(scm$ode_times)
  tibble::tibble(
    mode  = mode,
    work  = n * stages,
    error = abs(scm$net_reproduction_ratios - R0_ref) / R0_ref
  )
}

adaptive_sweep <- purrr::map_dfr(c(1e-2, 1e-3, 1e-4, 1e-5, 1e-6), function(tol) {
  ctrl <- Control()
  ctrl$ode_tol_rel <- tol
  ctrl$ode_tol_abs <- tol
  measure(ctrl, "adaptive")
})

euler_sweep <- purrr::map_dfr(c(2, 1, 0.5, 0.25, 0.125, 0.0625), function(dt) {
  measure(Control(fixed_time_step = dt), "euler")
})

work_precision <- bind_rows(adaptive_sweep, euler_sweep)
```

```{r}
#| label: fig-work-precision
#| fig-cap: "Work-precision diagram. Lower and to the left is better (less work, smaller error). The adaptive RK45 frontier sits well below the fixed-step Euler frontier: for any target accuracy it reaches it with fewer derivative evaluations."
ggplot(work_precision, aes(work, error, colour = mode)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Derivative evaluations (work)",
       y = expression("Relative error in " * R[0]),
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

### Why adaptive stepping dominates

The curves show the same underlying pattern:

- **Forward Euler is first-order.** Halving the step size roughly halves the
  error, so the Euler curve descends slowly — buying an extra digit of accuracy
  costs roughly a *tenfold* increase in work.
- **Adaptive RK45 is high-order *and* adaptive.** It spends small steps only
  where the dynamics demand them (the first weeks, while cohorts diverge) and
  coarsens dramatically once the stand stabilises. It reaches an error floor —
  set by the node-introduction schedule rather than the ODE tolerance — at a
  fraction of the evaluations Euler would need to match it.

To bring daily Euler's accuracy *down to* the adaptive frontier, it would need a
sub-daily step everywhere. That includes the long, smooth tail of the simulation,
where the adaptive solver can take steps of years. This is wasted work, and it
is exactly the inefficiency adaptive stepping avoids.

In practice, fixed-step Euler is useful for **operational-realism comparisons**:
it runs `plant` on the same footing as a daily-step DGVM. It is not the best way
to obtain an accurate or inexpensive solution. For that, retain the default
adaptive solver.

### A run that completes has not necessarily converged

Fixed-step integration removes one of the solver's safeguards. Forward Euler
will march across *anything*, including a discontinuous right-hand side that the
adaptive solver correctly refuses. It may still return numbers, but those
numbers can be artefacts of where the step grid falls rather than an
approximation that improves as the grid is refined.

`plant`'s experimental **PPA hard-step** shading variant
(`ppa_layer_smoothing = 0`, the literal field discretisation) is exactly such a
case: its light profile is discontinuous in height, so the adaptive solver fails
with *"Cannot achieve the desired accuracy"*. Under a fixed step it runs. The
test of whether that result *means* anything is convergence: does it settle as
the step shrinks? We compare it with the smoothed variant
(`ppa_layer_smoothing = 0.3`), which is a well-posed, continuous problem.

```{r}
seed_output <- function(steps_per_year, smoothing) {
  ctrl <- Control(fixed_time_step = 1 / steps_per_year)
  ctrl$shading_model      <- "ppa"
  ctrl$ppa_layer_smoothing <- smoothing
  run_scm(p, env, ctrl)$offspring_production
}

convergence <- purrr::map_dfr(c(91, 182, 365, 730, 1460), function(spy) {
  tibble::tibble(
    steps_per_year = spy,
    `hard step (s = 0)`   = seed_output(spy, 0),
    `smoothed (s = 0.3)`  = seed_output(spy, 0.3))
}) |>
  tidyr::pivot_longer(-steps_per_year, names_to = "variant", values_to = "seed")
```

```{r}
#| label: fig-hard-step-convergence
#| fig-width: 7
#| fig-height: 3.2
#| fig-cap: "Resident seed output as the fixed step shrinks. The smoothed variant converges; the hard step does not — it wanders by an order of magnitude with no limit. Note the independent y-scales: the hard step's values are ~1000x smaller, and erratic. A result that depends on the step size is a property of the grid, not of the model."
ggplot(convergence, aes(steps_per_year, seed, colour = variant)) +
  geom_line() + geom_point(size = 2) +
  facet_wrap(~variant, scales = "free_y") +
  scale_x_log10() +
  labs(x = "Steps per year (1 / fixed_time_step)", y = "Resident seed output",
       colour = NULL) +
  theme_bw() + theme(legend.position = "none")
```

The smoothed variant moves monotonically toward a limit: refining the step buys
accuracy, as it should. The hard step does not converge: every step size gives a
different answer, so no step size gives *the* answer. Such variants must not be
interpreted as fitness or as a solution. The fixed-step solver makes them
*executable*, not *correct*.

### Caveats and scope

- **Mutant fitness / resident replay is not supported under `fixed_time_step`.**
  The mutant pathway replays a resident run by reusing the six cached RK
  sub-step environments per step; forward Euler has a single stage and no such
  cache. Combining `fixed_time_step > 0` with `save_RK45_cache` (or with a pinned
  `use_ode_times` replay) raises a clear error rather than returning a wrong
  fitness.
- **Discontinuous/"hard-step" environment variants** only *run to completion*
  under a fixed grid and do not converge as the step shrinks (see the section
  above) — any apparent result is a grid artefact, not a solution. Do not
  interpret such runs as fitness or as a converged answer.
