---
title: "ODE stepping & control"
---
::: {.eyebrow}
implementation
:::

`plant` casts the entire patch as a system of ordinary differential equations
(ODEs) and steps it through time with classic ODE methods. (For details on the
equations themselves see the [size-structured PDE](../size-structured-pde.qmd)
page.) This page first describes the Runge–Kutta solver and the `Control()`
parameters that govern it, then works through the trade-off between the default
**adaptive** stepper and a **fixed-step** forward-Euler mode using a
work–precision diagram.

# The Runge–Kutta 45 ODE stepper

The `plant` model uses an explicit Runge-Kutta ODE stepper to integrate the
ordinary differential equations that describe the system. Specifically, we use
the Runge-Kutta Cash-Karp 4-5 (RK45) algorithm, with adaptive time step.

In brief, this means that the solver takes a step through time, and then
evaluates the error in the step. Error is estimated by comparing two possible
steps: one using 4 sub-intervals, and one using 5 sub-intervals. By comparing
the step made with more sub-intervals, the algorithm can estimate the
non-linearity in the system and error in the step. If the error is too high, the
solver will attempt to take a smaller step. If the error is well below a target
value, the solver will attempt to take a larger step. Importantly, the 5 step
reuses information from the 4 step and thereby only requires a single extra call
on rates of change, so the RK45 algorithm is much more efficient than
alternatives like the RK48. The math of RK45 stepping has long been established,
and the method is available in most solvers for ODE systems. You can read more
about Runge-Kutta methods and adaptive time steps on
[Wikipedia](https://en.wikipedia.org/wiki/Runge-Kutta_methods).

The RK45 is an explicit method, as opposed to an implicit method. As
[wikipedia explains](https://en.wikipedia.org/wiki/Explicit_and_implicit_methods)
_"Explicit methods calculate the state of a system at a later time from the state
of the system at the current time, while implicit methods find a solution by
solving an equation involving both the current state of the system and the later
one"_. The advantage of explicit methods are that they are simple to implement
and fast, particularly when allowing adaptive time steps. The disadvantage is
that they are less stable than implicit methods when they encounter stiff
systems. As described in wikipedia

* _"[Implicit methods](https://en.wikipedia.org/wiki/Explicit_and_implicit_methods) require an extra computation, and they can be much harder to implement. Implicit methods are used because many problems arising in practice are stiff, for which the use of an explicit method requires impractically small time steps $\Delta t$ to keep the error in the result bounded"_.
* _"[Explicit Runge Kutta methods](https://en.wikipedia.org/wiki/Runge–Kutta_methods) are generally unsuitable for the solution of stiff equations because their region of absolute stability is small; in particular, it is bounded. This issue is especially important in the solution of partial differential equations..... The advantage of implicit Runge–Kutta methods over explicit ones is their greater stability, especially when applied to stiff equations."_

In practice, we've found the RK45 to be extremely stable and fast for solving the
size-density distribution. However, there are circumstances where the system
becomes stiff and the solver struggles. E.g.

- When implementing soil water into plant via the Richards Equation, Isaac
  Towers found the system became stiff.
- When implementing the FF16 model in another solver, Joshi et al 2023
  <https://doi.org/10.1101/2023.08.04.551683> reported that the system became
  stiff when enabling a continuous feedback of seeds into germination.

Any extensions should therefore be wary of stiff behaviours.

# Setup of templated ODE solver in C++

The characteristic method used to solve density requires regular introduction of
new nodes, which entails a resizing of the ODE system as the system progresses.
To make this play nicely with the ODE stepping, Rich FitzJohn designed a custom
ODE solver class based on the RK45 method. The code is built into the model, with
some details ported from the
[GNU Scientific Library](http://www.gnu.org/software/gsl/) [@Galassi-2009].

The main interface is described in the class `plant::ode::Solver` described in
the file `ode_solver.h`. (See folder `inst/include/plant/ode_solver` where this
and related class are implemented.) The class is established using a template
`System` class, meaning you can effectively cast any variety of classes as an ODE
Solver. (In a similar way, you can create a vector of any variety of data types.)

From `ode_solver.h`:

```cpp
template <class System>
class Solver {
public:
  Solver(const System& system, OdeControl control_);
  ...
};
```

## Solving an entire patch

The main solver for our system is the class `SCM`, (in the file `sch.h`). This
creates an ode solver based on patch type. Simultaneously, the class `SCM` owns
an actual instance of the patch, which contains species and environment.

```cpp
patch_type patch;                 # Instance of patch
ode::Solver<patch_type> solver;   # ODE solver
```

All we need to do to advance the system is make calls like the following. These
advance the system, using the patch to estimate rates of change. Moreover, we can
take steps of a specified fixed size, or an adaptive step where the solver
determines the spacing.

```cpp
solver.advance_fixed(patch, e.times);
solver.advance_adaptive(patch, e.time_end());
```

In the background, the solver then passes information (rates of change, states)
between the actual patch (which contains all the biological information informing
rates of change) and ode solver (which uses the RK45 math to calculate future
states from rates of change).

Additionally, we have the following function, which resizes the ODE system after
a node has been introduced.

```cpp
solver.set_state_from_system(patch)
```

For the above to work, the patch class (see `patch.h`) must have some essential
functions used by the ODE system.

```cpp
// * ODE interface
// Caluclate size of ode system (number of equations).
size_t ode_size() const;
// How many auxiallary variables are we tracking. These are being collected but
// are not part of core ode system
size_t aux_size() const;
double ode_time() const;

// Retrieve ode state from patch and save into the ode solver
ode::iterator ode_state(ode::iterator it) const;
// Retrieve ode rates from patch and save into the ode solver
ode::iterator ode_rates(ode::iterator it) const;
// Retrieve auxillary variables and save into the ode solver
ode::iterator ode_aux(ode::iterator it) const;

// Set state of patch, based on estimate of future state estimated by the solver
ode::const_iterator set_ode_state(ode::const_iterator it, double time);
```

Note the use of iterators. These enable state from a complicated data structure
like a patch, to be converted into a vector, as needed by ode solver. And vice
versa.

## Solving individuals and other systems

As templated C++, we can use the `ode::Solver` class to step any "system" that
has the requisite functions. In addition to patches, we also use the ode solver
to run individual plants, given an environment. This is achieved by introducing a
wrapper class `ode::Runner` (in `ode_runner.h`), which according to its author
simplifies _"the difficulty of ownership semantics around the solver and
object."_

Like the `Solver` class, the `Runner` class is templated, meaning we can setup a
variety of systems.

```cpp
template <typename T>
class Runner {
public:
  Runner(T obj_, OdeControl control) : obj(obj_), solver(obj, control) {}
};
```

The class `plant::tools::IndividualRunner` uses this approach to create a system
for stepping individual plants to a given size. This is available via the R
functions `grow_individual_...`, which in turn uses the `OdeRunner`.

The test suite also shows how to use the OdeRunner to step the classic Lorenz
system.

# Controls on ODE stepping

The RK45 ode solver is controlled by a set of parameters that determine how the
solver steps through time.

As a simple example, we can step a system using the default values for ODE
stepping:

```{r}
library(plant)

# configure a patch containing a single species
p0 <- scm_base_parameters("FF16")
p0$max_patch_lifetime <- 5

p <- expand_parameters(trait_matrix(0.0825, "lma"), p0)

# Controls on numerical methods
ctrl <- Control()

# run
out <- run_scm(p, ctrl = ctrl)

# number of steps
length(out$ode_times)
```

The precise times used by the solver are stored in `out$ode_times`, so the length
of this object indicates the number of times used.

The number and timing of steps taken by the solver will vary with the controls on
the ode solver. Looking at the ctrl object, we can extract the items beginning
with `ode_`, which control the ODE solver:

```{r}
ctrl %>% subset(grepl("(?<!n)ode_", names(.), perl = TRUE))
```

These have the following effect

- `ode_step_size_initial`: The initial step size used by the solver in the first
  step.
- `ode_step_size_min`, `ode_step_size_max`: The minimum step size permitted. Any
  attempt by the solver to use a smaller value will be curtailed at this value.
- `ode_tol_abs`, `ode_tol_rel`: When taking a step, the absolute and relative
  error for any single equation that is allowed. If the error is too high, the
  solver will attempt to take a smaller step. If the error is well below this
  value, the solver attempts to expand the step size in the next step.
- `ode_a_y`, `ode_a_dydt`: Weightings on y and dy/dt, in calculation of allowable
  relative error (see `OdeControl::errlevel`).

Varying these will potentially change the number of steps taken by the solver.

However, note that unlike a regular ode solver, the number of steps taken is also
influenced by the timing of node introduction. The ode stepper will only expand
the time step up until the next node introduction event, for any species. So the
node scheduling sets an upper limit on the size of time steps. In the default
setup, the node schedule primarily determines the timesteps in the first months
of the simulation. After that, the ode stepper takes a bigger control.

## Limiting the Minimum and Maximum step size

Let's alter the maximum step size and compare to what we had above.

```{r}
#| eval: false
ctrl2 <- ctrl
ctrl2$ode_step_size_max <- 0.1

# run
out2 <- run_scm(p, ctrl = ctrl2)

# compare
length(out$ode_times)
length(out2$ode_times)

# maximum
f <- function(ode_times, title = "
") {
diff(ode_times) %>% hist(main = title, xlim = c(0,2), breaks  =seq(0, 2, length.out=20), freq = FALSE)}

f(out$ode_times, "original")
f(out2$ode_times, "max 0.1")
```

Now let's alter the minimum step size and compare to what we had above.

```{r}
#| eval: false
ctrl3 <- ctrl
ctrl3$ode_step_size_min <- 0.01
ctrl3$ode_step_size_initial <- 0.01

# run
out3 <- run_scm(p, ctrl = ctrl3)

# compare
length(out$ode_times)
length(out3$ode_times)

f(out$ode_times, "original")
f(out3$ode_times, "min 0.01")
```

## Fixing the step sizes (option 1)

We set exact step sizes by specifying max and min step sizes. But this could
result in other times, depending on the cohort schedule.

```{r}
#| eval: false
ctrl <- Control()
ctrl$ode_step_size_min <- 0.1
ctrl$ode_step_size_max <- 0.1
ctrl$ode_step_size_initial <- 0.1
```

## Fixing the step sizes (option 2)

We can also specify some exact steps, by passing in a vector of ode_times.

```{r}
pc <- p
pc$node_schedule_times[[1]] <- c(0, 0.1, 0.4, 0.8, 1.2)
pc$ode_times <- seq(0, p$max_patch_lifetime, by=0.1)

result2 <- run_scm(pc, ctrl = ctrl, use_ode_times = TRUE)

result2$ode_times
```

## Fixing the step sizes (option 3): forward Euler

The options above still drive the full adaptive RK45 machinery at each step. For
operational-realism comparisons with daily-step DGVMs, `plant` can instead use
plain **forward (explicit) Euler** on a uniform grid — one derivative evaluation
per step, no error estimate — by setting `fixed_time_step` (units: years) on the
`Control`:

```{r}
#| eval: false
ctrl <- Control(fixed_time_step = 1 / 365)  # daily step; 0 = adaptive (default)
run_scm(p, ctrl = ctrl)
```

This is not the way to get an accurate or cheap solution: forward Euler is
first-order, and matching the adaptive solver's accuracy needs far more
evaluations. The [next part](#sec-adaptive-vs-fixed-step) works this trade-off
through with a work–precision diagram. Note that `fixed_time_step` is
incompatible with the mutant-fitness replay path
(`save_RK45_cache` / `use_ode_times`) and errors clearly if combined.

## Future changes

Would it help to have option to pass in ode_times, use if supplied?

- But, current system in NodeSchedule interweaves ode_times and
  node_schedule_times. It's complex code, poorly documented, low value to mess
  with. Suggests leaving as is
- could have option to pass in ode_times, add to pars inside function, use if
  supplied?
- if supplied, do we always want to use ode_times?
- what's the advantage of using previous ode_times?
  - suggest always using them if present?

# Adaptive vs fixed-step {#sec-adaptive-vs-fixed-step}

By default, `plant` integrates the patch with the **adaptive, error-controlled
Cash–Karp Runge–Kutta (RK45)** stepper described above. The solver chooses each
step size to keep the local error within a tolerance: tiny steps early on, when
newly introduced cohorts are diverging rapidly, and much coarser steps later,
once the stand settles.

Many widely used **dynamic global vegetation models (DGVMs)** instead march the
state forward on a *fixed* grid — most commonly a **daily** time step — using
plain forward (explicit) Euler. `plant` can be run this way too, via the
`fixed_time_step` control (option 3 above). This part shows how to enable it and
uses a **work–precision diagram** to make the central trade-off concrete: at
*matched accuracy*, adaptive stepping does far less work.

```{r}
#| label: load-libraries
library(plant)
library(dplyr)
library(ggplot2)
```

## Enabling fixed-step forward Euler

The integration mode is selected by a single control field, `fixed_time_step`
(units: **years**):

- `fixed_time_step = 0` (the default) → adaptive RK45.
- `fixed_time_step > 0` → forward Euler on a uniform grid of that spacing.

A *daily* step is therefore `fixed_time_step = 1 / 365`.

```{r}
p0 <- scm_base_parameters("FF16")
p0$max_patch_lifetime <- 60

p <- expand_parameters(trait_matrix(0.0825, "lma"), p0, birth_rate_list = 1.0)
env <- Environment("FF16")

ctrl_adaptive <- Control()                       # fixed_time_step = 0
ctrl_daily    <- Control(fixed_time_step = 1 / 365)

out_adaptive <- run_scm(p, env, ctrl_adaptive)
out_daily    <- run_scm(p, env, ctrl_daily)
```

Forward Euler does **one** derivative evaluation per step (`y <- y + h * dy/dt`),
whereas each RK45 step evaluates the derivatives six times. The two runs report
slightly different fitness, and use a very different number of steps:

```{r}
tibble::tibble(
  mode    = c("adaptive RK45", "daily Euler"),
  R0      = c(out_adaptive$net_reproduction_ratios,
              out_daily$net_reproduction_ratios),
  n_steps = c(length(out_adaptive$ode_times),
              length(out_daily$ode_times))
)
```

> **Caution — don't read this as "Euler is faster".** A *single* daily step is
> cheaper than a single adaptive step (1 derivative evaluation versus 6), so a
> naive timing race makes Euler look quick. But that comparison holds the *step
> count* fixed, not the *accuracy*. Daily Euler here is visibly less accurate
> than the adaptive run. The honest question is: *how much work does each method
> need to reach the same accuracy?* That is what the work–precision diagram
> below answers — and there adaptive wins comfortably.

## A work–precision diagram

We sweep the adaptive solver over a range of error tolerances, and forward Euler
over a range of step sizes, and for each run record:

- **work**: total derivative evaluations — `6 × n_steps` for RK45, `1 × n_steps`
  for Euler (this is the dominant cost of the inner loop);
- **error**: relative error in the net reproduction ratio $R_0$ against a
  tight-tolerance adaptive **reference**.

```{r}
## Tight-tolerance adaptive reference.
ctrl_ref <- control_accurate()
ctrl_ref$ode_tol_rel <- 1e-8
ctrl_ref$ode_tol_abs <- 1e-8
R0_ref <- run_scm(p, env, ctrl_ref)$net_reproduction_ratios

measure <- function(ctrl, mode) {
  stages <- if (mode == "adaptive") 6 else 1   # derivative evals per step
  scm <- run_scm(p, env, ctrl)
  n   <- length(scm$ode_times)
  tibble::tibble(
    mode  = mode,
    work  = n * stages,
    error = abs(scm$net_reproduction_ratios - R0_ref) / R0_ref
  )
}

adaptive_sweep <- purrr::map_dfr(c(1e-2, 1e-3, 1e-4, 1e-5, 1e-6), function(tol) {
  ctrl <- Control()
  ctrl$ode_tol_rel <- tol
  ctrl$ode_tol_abs <- tol
  measure(ctrl, "adaptive")
})

euler_sweep <- purrr::map_dfr(c(2, 1, 0.5, 0.25, 0.125, 0.0625), function(dt) {
  measure(Control(fixed_time_step = dt), "euler")
})

work_precision <- bind_rows(adaptive_sweep, euler_sweep)
```

```{r}
#| label: fig-work-precision
#| fig-cap: "Work-precision diagram. Lower and to the left is better (less work, smaller error). The adaptive RK45 frontier sits well below the fixed-step Euler frontier: for any target accuracy it reaches it with fewer derivative evaluations."
ggplot(work_precision, aes(work, error, colour = mode)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Derivative evaluations (work)",
       y = expression("Relative error in " * R[0]),
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

## Why adaptive stepping dominates

The two curves tell a consistent story:

- **Forward Euler is first-order.** Halving the step size roughly halves the
  error, so the Euler curve descends slowly — buying an extra digit of accuracy
  costs roughly a *tenfold* increase in work.
- **Adaptive RK45 is high-order *and* adaptive.** It spends small steps only
  where the dynamics demand them (the first weeks, while cohorts diverge) and
  coarsens dramatically once the stand stabilises. It reaches an error floor —
  set by the node-introduction schedule rather than the ODE tolerance — at a
  fraction of the evaluations Euler would need to match it.

To get daily-Euler accuracy *down to* the adaptive frontier you would need a
sub-daily step everywhere, including the long, smooth tail of the simulation
where the adaptive solver is happily taking steps of years. That is wasted work,
and it is exactly the inefficiency adaptive stepping removes.

The practical conclusion: fixed-step Euler is valuable for **operational-realism
comparisons** — running `plant` on the same footing as a daily-step DGVM — but it
is not the way to get an accurate or cheap solution. For that, keep the default
adaptive solver.

## A run that completes is not a run that converged

Fixed-step integration removes the solver's safety net: forward Euler will march
across *anything*, including a discontinuous right-hand side that the adaptive
solver rightly refuses. When that happens the integrator still returns numbers —
but they are an artefact of where the step grid happens to fall, not an
approximation to a solution that gets better as you refine.

`plant`'s experimental **PPA hard-step** shading variant
(`ppa_layer_smoothing = 0`, the literal field discretisation) is exactly such a
case: its light profile is discontinuous in height, so the adaptive solver fails
with *"Cannot achieve the desired accuracy"*. Under a fixed step it runs. The
test of whether that result *means* anything is convergence: does it settle as
the step shrinks? We compare it with the smoothed variant
(`ppa_layer_smoothing = 0.3`), which is a well-posed, continuous problem.

```{r}
seed_output <- function(steps_per_year, smoothing) {
  ctrl <- Control(fixed_time_step = 1 / steps_per_year)
  ctrl$shading_model      <- "ppa"
  ctrl$ppa_layer_smoothing <- smoothing
  run_scm(p, env, ctrl)$offspring_production
}

convergence <- purrr::map_dfr(c(91, 182, 365, 730, 1460), function(spy) {
  tibble::tibble(
    steps_per_year = spy,
    `hard step (s = 0)`   = seed_output(spy, 0),
    `smoothed (s = 0.3)`  = seed_output(spy, 0.3))
}) |>
  tidyr::pivot_longer(-steps_per_year, names_to = "variant", values_to = "seed")
```

```{r}
#| label: fig-hard-step-convergence
#| fig-width: 7
#| fig-height: 3.2
#| fig-cap: "Resident seed output as the fixed step shrinks. The smoothed variant converges; the hard step does not — it wanders by an order of magnitude with no limit. Note the independent y-scales: the hard step's values are ~1000x smaller, and erratic. A result that depends on the step size is a property of the grid, not of the model."
ggplot(convergence, aes(steps_per_year, seed, colour = variant)) +
  geom_line() + geom_point(size = 2) +
  facet_wrap(~variant, scales = "free_y") +
  scale_x_log10() +
  labs(x = "Steps per year (1 / fixed_time_step)", y = "Resident seed output",
       colour = NULL) +
  theme_bw() + theme(legend.position = "none")
```

The smoothed variant marches monotonically toward a limit — refining the step
buys accuracy, as it should. The hard step does not converge to anything: each
step size gives a different answer, so no step size gives *the* answer. This is
why such variants must not be read as fitness or as a solution — the fixed-step
solver makes them *executable*, not *correct*.

## Caveats and scope

- **Mutant fitness / resident replay is not supported under `fixed_time_step`.**
  The mutant pathway replays a resident run by reusing the six cached RK
  sub-step environments per step; forward Euler has a single stage and no such
  cache. Combining `fixed_time_step > 0` with `save_RK45_cache` (or with a pinned
  `use_ode_times` replay) raises a clear error rather than returning a wrong
  fitness.
- **Discontinuous/"hard-step" environment variants** only *run to completion*
  under a fixed grid and do not converge as the step shrinks (see the section
  above) — any apparent result is a grid artefact, not a solution. Do not
  interpret such runs as fitness or as a converged answer.
