---
title: "Reproducing emergent size structure in the FF16 patch"
description: "Re-running the emergence of a size-structured stand from the plant model's introductory paper — published figure vs. our re-run."
date: 2026-06-20
plant-version: "2.1.0"
categories: [reproduction, patch-dynamics, ff16]
reading-time: true
paper:
  authors: "Falster, Brännström, Westoby & Dieckmann"   # TODO verify author list/order
  year: 2016                                            # TODO verify
  title: "Plant: A package for modelling forest trait ecology and evolution"  # TODO verify exact title
  journal: "Methods in Ecology and Evolution"           # TODO verify
  doi: "TODO — fill from the paper"
  figure: "TODO — the figure showing emergent stand structure"
  fidelity: "pending"
---

```{r setup, include=FALSE}
source(here::here("R/version-badge.R"))
source(here::here("R/reproduction.R"))
```

`r plant_version_badge()` `r repro_fidelity_badge()`

`r paper_citation()`

::: {.callout-important}
**Author of this post:** I've left the paper's exact figure number,
parameter values, and DOI as `TODO`. Confirm them against the source
before publishing — the structure below is ready, only the verified
specifics are missing.
:::

## What the original shows

TODO: describe the published panel — typically the time-development of a
patch from establishment, with the size distribution of individuals
emerging from competition for light. State the axes and the headline
result.

## Parameters from the paper

```{r}
#| label: params
#| eval: false
# TODO: transcribe the FF16 strategy parameters used in the figure,
# citing the table/section for each value that differs from defaults.
library(plant)
p0 <- scm_base_parameters("FF16")
# p0$strategy_default$lma <- TODO   # <Author Year>, Table SX
# ... etc.
```

## Re-running it

```{r}
#| label: reproduce
#| eval: false
#| fig-cap: "Our re-run of the patch development. Compare against TODO-figure."
# TODO: minimal path to regenerate. Sketch of the typical call sequence:
# p1 <- expand_parameters(trait_matrix(TODO, "lma"), p0)
# result <- run_scm_collect(p1)
# ... extract the size distribution / state over time and plot ...
```

## Side by side

::: {.repro-compare}
::: {.repro-pane}
**Published** — TODO-figure

> Add the published panel at `assets/papers/ff16-emergence.png`, or if
> licensing disallows redistribution, link it at the DOI above and
> describe it here.
:::
::: {.repro-pane}
**Reproduced** — this post

> The figure rendered above.
:::
:::

## Verdict

TODO: matches / differs / approximate, with reasoning. If defaults have
changed since 2016 (units, a reparameterised trait, a corrected
constant), note it — that's exactly the kind of drift this reproduction
catches. Then update `fidelity:` in the front matter.

```{r}
#| label: session
#| eval: false
plant_session_info()
```
