---
title: "Assimilation & hydraulics"
---
::: {.eyebrow}
theory
:::

::: {.callout-note}
## Prerequisites
This page builds on [the FF16 model](models/ff16.qmd) and
[The maths, gently](maths-primer.qmd). Read those first if the notation here is
unfamiliar.
:::

## The idea first

A leaf takes in carbon dioxide through tiny pores called stomata. When the
stomata open, carbon dioxide enters and water vapour escapes. The lost water
must be replaced by water drawn from the soil through the plant's hydraulic
system.

This creates a trade-off. Opening the stomata further can increase
photosynthesis, but it also increases transpiration and places more strain on
the water-transport pathway. Closing them protects that pathway, but restricts
carbon dioxide and therefore limits photosynthesis.

TF24 represents this trade-off as an optimisation problem. For each possible
leaf xylem water potential, the model calculates a photosynthetic benefit and a
hydraulic cost. It then chooses the water potential at which their difference,
called *profit*, is greatest. This page develops those three pieces in order:
photosynthetic revenue, hydraulic cost, and the numerical search for maximum
profit.

This leaf-level carbon and water balance is central to the
[TF24 strategy](models/tf24.qmd). It determines the carbon assimilation and
transpiration of a unit of leaf area under the current light, atmospheric, and
soil-water conditions. The whole-plant strategy then uses those leaf-level
rates in its growth and water-use calculations.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(plant)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
```

## Introduction

TF24 extends the plant model with processes that link rainfall, soil water,
plant hydraulics, transpiration, and carbon assimilation. This page focuses on
one part of that system: the `Leaf` submodel that chooses a stomatal and
hydraulic operating point. Its answer depends on plant traits and on the
current water availability, vapour pressure deficit, and other environmental
conditions. The supplied leaf temperature adjusts the photosynthetic
biochemistry: `plant` recalculates $V_{c,max}$, $J_{max}$, the CO$_2$
compensation point, Michaelis--Menten terms, and dark respiration from it.

The accompanying [Soil water & root uptake](soil-water.qmd) page describes the
other side of the coupling: how rainfall, drainage, and root uptake determine
the soil-water conditions supplied to this leaf calculation.

Several parameters are especially important for the leaf calculation:

- `vcmax_25`: the maximum rate of carboxylation at 25 °C ($\mu mol~m^{-2}~s^{-1}$)
- `p_50`: the leaf water potential ($\psi_l$) at which 50% of conductivity is lost
- `K_s`: the maximum hydraulic conductivity ($kg~m^{-2}~s^{-1}~MPa^{-1})$
- `c`: the shape parameter for the hydraulic vulnerability curve (unitless)
- `b`: the sensitivity parameter for the hydraulic vulnerability curve (-MPa)
- `psi_crit`: the critical $\psi_l$ at which 95% of conductivity is lost (-MPa)

Together with the other TF24 traits and parameters, these values determine how
the `Leaf` submodel responds to its environment. The implementation lives in
`leaf_model.cpp`; the sections below explain the calculation conceptually.

## Profitmax

Plants vary the aperture of their stomata in response to conditions such as
soil moisture and atmospheric dryness. Several theoretical frameworks can be
used to represent that response. TF24 uses the `profitmax` model, which assumes
that the chosen operating point is the one that maximises profit,
$P~(\mu mol~m^{-2}~s^{-1})$.

::: {.callout-note}
### A second stomatal model lives in the code, but is dormant

The `Leaf` submodel also carries an implementation of the Medlyn *et al.* (2011)
stomatal-conductance model — an empirical $g_1$ formulation (`medlyn_model_gs`,
with settable `g0`/`g1` and a soil-moisture stress factor $\beta$ derived from
the wilting point and field capacity). It is **not** on the default `TF24`
compute path, which optimises $\psi_l$ directly via `profitmax`. It exists as a
comparison/exploration tool, so treat any Medlyn-style $g_1$ behaviour as
opt-in, not the model's default stomatal response.
:::

Profit is the difference between photosynthetic revenue,
$A~(\mu mol~m^{-2}~s^{-1})$, and hydraulic cost,
$C~(\mu mol~m^{-2}~s^{-1})$:

$$
P~(\psi_l) = A~(\psi_l) - C~(\psi_l).
$$ {#eq-profit}

Later code and figures label photosynthetic revenue as $R$ and hydraulic cost as
$H$. These are the same two quantities denoted by $A$ and $C$ in @eq-profit;
only the labels change.

$\psi_l~(-MPa)$ is the water potential of the canopy xylem. In this page's
notation, water stress is expressed as the magnitude of a negative potential:
a larger value means a more negative physical water potential. Changing
$\psi_l$ changes how much water the hydraulic pathway can supply and therefore
the stomatal conductance that can be supported.

Both $A$ and $C$ are therefore functions of $\psi_l$. The model evaluates their
difference over the feasible range of $\psi_l$ and finds the value that gives
the largest $P$.

We begin with the revenue side of the calculation.

## Revenue

Revenue in `profitmax` is the photosynthetic carbon assimilation rate
$A~(\mu mol~m^{-2}~s^{-1})$ supported at a given stomatal conductance. At this
stage, it is the carbon gain before the hydraulic cost is subtracted.

Stomata connect the carbon and water calculations. Fick's first law describes
the flux of $CO_2$ down the concentration gradient from the atmosphere,
$C_a~(Pa)$, to the intercellular spaces in the leaf, $C_i~(Pa)$:

$$
A~(C_i) = \frac{g_c(\psi_{l})(C_a - C_i)}{P_{atm}}.
$$ {#eq-assim-ci}

Here, $P_{atm}~(kPa)$ is atmospheric pressure and $g_c$ is stomatal
conductance to $CO_2$. Water vapour and carbon dioxide diffuse at different
rates, so conductance to $CO_2$ is calculated from conductance to water,
$g_w$, as

$$
g_c(\psi_{l}) = \frac{g_w(\psi_{l})}{1.6},
$$ {#eq-gc}

where the factor 1.6 accounts for the difference in diffusivity between $H_2O$
and $CO_2$ at the stomatal boundary.

For any candidate $\psi_l$, the hydraulic calculation supplies a value of
$g_w$, and therefore of $g_c$. The remaining unknown is $C_i$. This cannot be
read directly from @eq-assim-ci because assimilation also depends on $C_i$
through leaf biochemistry. The model must find a value of $C_i$ at which the
diffusive supply of carbon dioxide and its biochemical use agree.

The biochemical side uses the photosynthesis model of Farquhar et al. (1980).
It calculates a Rubisco-limited rate,
$A_c~(umol~m^{-2}~s^{-1})$, and an electron-transport-limited rate,
$A_j~(umol~m^{-2}~s^{-1})$. Net assimilation follows the more limiting process,
with a smooth transition between them.

The electron-transport-limited rate depends on both light availability and
$C_i$:

$$
A_j=\frac{J(C_i-\Gamma^*)}{4(C_i+2\Gamma^*)},
$$ {#eq-aj}

where $\Gamma^*~(Pa)$ is the $CO_2$ compensation point. The electron transport
rate $J~(umol~m^{-2}~s^{-1})$ carries the effect of irradiance into the
calculation:

$$
J=\frac{aQ+J_{max}-\sqrt{(aQ+J_{max})^2-4caQJ_{max}}}{2c},
$$ {#eq-j}

where $a~(mol~photon~mol^{-1}~electron)$ is the effective quantum yield of
electron transport, $Q~(umol~m^{-2}~s^{-1})$ is photosynthetic photon flux
density (PPFD), $c$ is the unitless curvature of the light-response curve, and
$J_{max}~(umol~m^{-2}~s^{-1})$ is the maximum electron transport rate.

::: {.callout-note}
### Two different curvature parameters use the symbol $c$

The light-response equation above and the hydraulic vulnerability equation
below both use $c$, but they refer to different parameters. In the code, the
light-response curvature is `curv_fact_elec_trans`, while the vulnerability
curve uses `c`.
:::

The Rubisco-limited rate is

$$
A_c=\frac{V_{c,max}(C_i-\Gamma*)}{C_i + K_m},
$$ {#eq-ac}

where $K_m~(Pa)$ is the effective Michaelis--Menten constant. $V_{c,max}$,
$\Gamma^*$, and $K_m$ are all temperature-dependent variables.

The next example compares $A_c$ and $A_j$ across a range of $C_i$ values.

The calculation needs both fixed strategy parameters and environmental or
plant-state values that can change through time. We first create a `Leaf` object
to hold the fixed parameters. Unlike an `Individual` or a patch, this object
represents the physiology of a unit of leaf area rather than a whole plant or
population.

The following executable example uses the `Leaf()` interface in `plant`
2.0.0.9002. The constructor now makes the root-hydraulics and numerical-control
inputs explicit; `hk_s` is not a `Leaf()` argument in this API.

```{r}
sw <- list(
  vcmax_25 = 96,                 #  umol m^-2 s^-1
  c = 2.680147,                  # unitless
  b = 3.898245,                  # -MPa
  psi_crit = 5.870283,           # -MPa
  root_c = 2.680147,             # unitless
  root_b = 3.898245,             # -MPa
  root_psi_crit = 5.870283,      # -MPa
  beta2 = 1.5,                   # exponent for effect of hydraulic risk  (unitless)
  jmax_25 = 157.44,              #  maximum electron transport rate umol m^-2 s^-1
  a = 0.30,                      # quantum yield of photosynthetic electron transport  (mol mol^-1)
  curv_fact_elec_trans = 0.7,    # curvature factor for the light response curve  (unitless)
  curv_fact_colim = 0.99,        # curvature factor for the colimited photosythnthesis equatiom
  GSS_tol_abs = 1e-3,            #  ???
  vulnerability_curve_ncontrol = 100,  # ????
  ci_abs_tol = 1e-3,             # ???
  ci_niter = 1000,
  g1_TF24 = 7.5,                # cost parameter for TF24 profit model umol m^-2 s^-1
  beta_R_H = 3.4e2,             #proportionality constant between minimum horizontal (intraleyer) root hydraulic resistance and C_r^-1 in [MPa * s * (mol C) / (mol H2O)]
  beta_R_V = 9.4e3
)

leaf <- Leaf(
  vcmax_25 = sw$vcmax_25,
  c = sw$c,
  b = sw$b,
  psi_crit = sw$psi_crit,
  root_c = sw$root_c,
  root_b = sw$root_b,
  root_psi_crit = sw$root_psi_crit,
  beta2 = sw$beta2,
  jmax_25 = sw$jmax_25,
  a = sw$a,
  curv_fact_elec_trans = sw$curv_fact_elec_trans,
  curv_fact_colim = sw$curv_fact_colim,
  GSS_tol_abs = sw$GSS_tol_abs,
  vulnerability_curve_ncontrol = sw$vulnerability_curve_ncontrol,
  ci_abs_tol = sw$ci_abs_tol,
  ci_niter = sw$ci_niter,
  g1_TF24 = sw$g1_TF24,
  beta_R_H = sw$beta_R_H,
  beta_R_V = sw$beta_R_V
)
```

Next, `set_physiology` supplies the conditions at one point in time. These
include the light and atmospheric environment, soil water, and size-dependent
properties of the plant. Unlike the fixed parameters passed to `Leaf()`, these
values can change as an individual grows and its environment changes. Here they
are held constant so that the leaf-level relationships can be viewed in
isolation.

```{r}
  psi_soil = c(1)
  soil_depth = c(0.5)
  mass_root_prop_ = c(1)
  area_leaf_ = 0.05
  PPFD = 1000
  leaf_specific_conductance_max = 4.8e-05
  atm_vpd = 2 
  ca = 40 
  sapwood_volume_per_leaf_area = 1.8e-04 
  rho = 608 
  a_bio = 0.0245 
  leaf_temp_ = 25 
  atm_o2_kpa_ = 21 
  atm_kpa_ = 101.3

leaf$set_physiology(area_leaf = area_leaf_, mass_root_prop = mass_root_prop_, rho = 608, a_bio = 0.0245, PPFD = PPFD, psi_soil = psi_soil, soil_depth = soil_depth, leaf_specific_conductance_max = leaf_specific_conductance_max, atm_vpd = atm_vpd, ca = ca, sapwood_volume_per_leaf_area = sapwood_volume_per_leaf_area, leaf_temp = leaf_temp_, atm_o2_kpa = atm_o2_kpa_, atm_kpa = atm_kpa_)

```

We now evaluate the biochemical rates over a sequence of $C_i$ values.

```{r}
C_i = seq(0,40,1)
```

The `Leaf` object uses similar names for calculation methods and stored results.
For example, `leaf$assim_electron_limited` is a function that takes $C_i$ as an
input, whereas `leaf$assim_electron_limited_` stores a calculated result. The
first plot calls the functions directly.

```{r}
tibble(C_i = C_i) %>%
  mutate(A_j = map_dbl(C_i, leaf$assim_electron_limited),
         A_c = map_dbl(C_i, leaf$assim_rubisco_limited)) %>%
  pivot_longer(cols = c(A_j, A_c), names_to = "limit", values_to = "assim") %>%
  ggplot(aes(x = C_i, y = assim))+
  geom_line(aes(colour = limit, group = limit)) +
  theme_classic()
```

Both potential rates increase with $C_i$, but they do not increase at the same
rate. Which process is most limiting can therefore change across the range of
$C_i$. The model combines them with a smooth co-limitation function (see the
paper for details).

```{r}
tibble(C_i = C_i) %>%
  mutate(A_j = map_dbl(C_i, leaf$assim_electron_limited),
         A_c = map_dbl(C_i, leaf$assim_rubisco_limited)) %>%
  mutate(A_net = map_dbl(C_i, leaf$assim_colimited)) %>%
  mutate(A_net_no_R_d = A_net + leaf$R_d_) %>%
  pivot_longer(cols = c(A_j, A_c,A_net, A_net_no_R_d), names_to = "rate", values_to = "assim") %>%
  ggplot(aes(x = C_i, y = assim))+
  geom_line(aes(colour = rate, group = rate)) +
  theme_classic()
```

The `A_net` calculation also subtracts dark respiration. Adding that respiration
term back helps show how the co-limited curve relates to $A_c$ and $A_j$ before
the respiratory deduction.

## Cost

Photosynthesis requires a continuous supply of water from the soil to the
canopy. Water moves along a potential gradient, so the leaf xylem must be at a
more negative physical water potential than the soil. In the positive-magnitude
notation used here, this means that $\psi_l$ must be larger than $\psi_s$.

Making that gradient steeper can supply more water, but it also reduces the
conductance of the xylem pathway as embolism risk increases. The formulation
developed on this page represents the hydraulic pathway with the following
vulnerability curve:

$$
k_l = k_{l,max}e^{-{(\frac{\psi_l}{b})}^c}.
$$ {#eq-vuln-curve}

The example below plots this decline using the parameter values defined above.

```{r}
data <-
  tibble(psi_stem = seq(0, 5, length.out = 100)) %>%
  mutate(
    proportion_conductance = map_dbl(psi_stem, leaf$proportion_of_conductivity),
    conductance = 4.8e-05*proportion_conductance
  )

data %>%
  ggplot(aes(x = psi_stem, y = conductance)) +
  geom_line() +
  theme_classic()
```


$b~(MPa)$ controls the sensitivity of the curve and $c$ controls its shape. The
value of $b$ is calculated as

$$b = \frac{p_{50}}{-log(1-\frac{50}{100}^{\frac{1}{c}})},$$

where $p_{50}$ is the $\psi_l$ at which 50% of hydraulic conductance is lost.
If two points on a hydraulic vulnerability curve are known, they can be used to
calculate $c$. Here, an empirical value for $c$ is obtained from AusTraits.

$p_{50}$ is in turn calculated from an empirical relationship with the
hydraulic conductivity of the xylem pathway. In this relationship, more
conductive stems lose conductivity at a less negative water potential (Liu et
al. 2019):

$$p_{50} = 1.73\frac{K_s}{2}^{-0.72}$$

$k_{l,max}$ is the maximum hydraulic conductance at $\psi_l = 0$:

$$k_{l,max} = \frac{K_s\theta}{h\eta_c},$$

Here, $\theta~(m^2~sapwood~m^{-2}~leaf~area)$ is the Huber value, $h$ is plant
height, and $\eta_c~(unitless)[0,1]$ describes the average relative position of
leaves within the canopy. The [TF24 strategy](models/tf24.qmd) provides more
context for these quantities. $K_s$, $\theta$, and $c$ parameterise the
hydraulic relationships, while the TF24 hyper-parameterisation handles the
remaining parameter links. Plant height is a dynamic state, so $k_{l,max}$ can
change as the plant grows.

For a given soil water potential and candidate leaf water potential, the water
that the pathway can supply per unit leaf area is the integral of conductance
across that potential interval:

$$
E_{supply} = \int^{\psi_l}_{\psi_s}k_l(\psi)\delta\psi 
$$ {#eq-e-supply}

Graphically, $E_{supply}$ is the area under the conductance curve between the
soil water potential, $\psi_s~(-MPa)$, and the leaf xylem water potential,
$\psi_l~(-MPa)$.

```{r}
data %>%
  ggplot(aes(x = psi_stem, y = conductance)) +
  geom_line() +
  theme_classic() +
  geom_area(aes(ifelse(psi_stem > 2 | psi_stem < 1, NA, psi_stem), conductance), fill = "red", alpha= 0.5) +
  geom_vline(xintercept = 2, col = "red", linewidth = 2, linetype = "dashed") +
  geom_vline(xintercept = 1, col = "red", linewidth = 2, linetype = "dashed") +
  annotate("text", x = 1.2, y = 4.8e-05, label = expression(psi[s]), col = "red") +
  annotate("text", x = 2.2, y = 4.8e-05, label = expression(psi[l]), col = "red")
```

Holding $\psi_s$ fixed, the next plot shows how $E_{supply}$ changes as
$\psi_l$ becomes more negative. Supply initially rises as the potential
gradient increases, but the rise tapers because conductance is simultaneously
being lost along the vulnerability curve. Increasing hydraulic strain therefore
produces progressively smaller gains in water supply.

```{r}
data %>%
  mutate(E_supply = map2_dbl(psi_stem, psi_soil, leaf$transpiration_full_integration)) %>%
  ggplot(aes(x = psi_stem, y = E_supply)) +
  geom_vline(xintercept = psi_soil, linetype = "dashed") +
  geom_line() +
  theme_classic()
```

The feasible range ends at a critical value, $\psi_{crit}$, beyond which the
xylem is assumed to fail catastrophically. Here, $\psi_{crit}$ is the
$\psi_l$ at which 95% of conductance has been lost.

The revenue calculation requires stomatal conductance, while the hydraulic
calculation gives water supply. The two are joined by requiring water supplied
through the plant to equal water lost from the leaf:

$$E_{supply} = E_{demand}$$

Atmospheric demand is

$$
E_{demand} = \frac{g_sD}{P_{atm}},
$$ 
{#eq-e-demand}

where $D~(kPa)$ is atmospheric vapour pressure deficit and $g_s$ is stomatal
conductance. The model uses atmospheric VPD rather than leaf-to-atmosphere VPD;
the latter would also depend on differences between leaf and air temperature.

For a fixed $D$, increasing $g_s$ increases $E_{demand}$. Meeting that demand
requires a more negative leaf water potential, which causes further loss of
hydraulic conductance. This closes the central loop: opening stomata can raise
carbon assimilation, but doing so also raises water demand and hydraulic risk.

To compare that risk directly with carbon revenue, the model expresses
hydraulic cost, $H~(\psi_l)$, in the same units as assimilation:
$$
H(\psi_l) = b_1z(1-e^{-{(\frac{\psi_l}{b})}^c})^{b_2}.
$$ {#eq-hydraulic-cost}

Here, $z~(m^3~sapwood~m^{-2}~leaf~area)$ is sapwood volume per leaf area,
$b_1(\mu mol~m^{-2}~s^{-1})$ converts conductivity loss to a carbon cost, and
$b_2~(unitless)$ controls the shape of the cost curve. The term in parentheses
is the proportion of conductivity lost at $\psi_l$, so $H=0$ when
$\psi_l = 0$ and increases as conductivity is lost.

$z$ depends on plant size and the Huber value, and represents the volume of
sapwood supplying each unit of leaf area:

$$z = \frac{\theta*h}{\eta_c}$$

## Profit

We can now put the two sides together. For each candidate $\psi_l$, the model
calculates photosynthetic revenue and hydraulic cost, then subtracts the second
from the first. Revenue tends to benefit from the additional stomatal opening
supported by a steeper water-potential gradient, while hydraulic cost rises as
conductivity is lost. Their difference therefore has a maximum within the
feasible range.

The first example calculates the revenue, cost, and profit curves step by step.


```{r}
f <- function(psi_stem) {
  leaf$set_leaf_states_rates_from_psi_stem(psi_stem, psi_soil)
  R = leaf$assim_colimited_
  H = leaf$hydraulic_cost_TF(psi_stem)
  P = R-H
  tibble(R, H, P)
}

data <-
  tibble(psi_stem = seq(1, 4, length.out = 100)) %>%
  mutate(outcomes = map(psi_stem, f)) %>%
  unnest(outcomes)


data %>%
  pivot_longer(cols = c(P, R, H)) %>%
  ggplot(aes(x = psi_stem, y = value)) +
  geom_line(aes(colour = name, group = name)) +
  theme_classic() +
  geom_vline(xintercept = 2.05, linetype = "dashed")
```

The explicit calculation is useful for seeing where the three curves come from.
In normal use, TF24 performs the same sequence through its built-in
`profit_psi_stem_TF` function.

```{r}
f <- function(psi_stem){
  leaf$profit_psi_stem_TF(psi_stem, psi_soil)
  tibble(H = leaf$hydraulic_cost_, R = leaf$assim_colimited_, P = R-H)
  }

data <-
  tibble(psi_stem = seq(1, 4, length.out = 100)) %>%
  mutate(outcomes = map(psi_stem, f))%>%
  unnest(outcomes)

data  %>%
  pivot_longer(cols = c(H,R,P)) %>%
  ggplot(aes(x = psi_stem, y = value)) +
  geom_line(aes(colour = name, group = name)) +
  theme_classic()
```

## Optimising profit

The leaf submodel ultimately needs two outputs under the current environmental
conditions and plant state: carbon assimilation and water use per unit leaf
area. Obtaining them requires finding the $\psi_l$ that maximises $P$ in
@eq-profit. Once that operating point is known, the corresponding transpiration
$E$ can also be passed to the soil-water calculation.

TF24 uses a golden-section search. The search starts with a lower and upper
bound and repeatedly narrows the interval containing the maximum until that
interval is smaller than a chosen tolerance. Here, the quantity being varied is
$\psi_l$. Its bounds are $\psi_s$ and $\psi_{crit}$: the leaf cannot be less
negative than the soil that supplies it, and it is not allowed to pass the
critical loss-of-conductivity threshold.

The following code first demonstrates the search manually. It evaluates profit
at two interior points, discards the part of the interval that cannot contain
the maximum, and repeats.

```{r}
#golden ratio
gr = (sqrt(5) + 1) / 2;

bound_a = NA
bound_b = NA
bound_c = NA
bound_d = NA

#set lower boundary to psi_soil
bound_a = leaf$psi_soil_;

#set upper boundary to psi_crit
bound_b = sw$psi_crit;

#tolerance value
delta_crit = 0.01;

bound_c = bound_b - (bound_b - bound_a) / gr;
bound_d = bound_a + (bound_b - bound_a) / gr;

while (abs(bound_b - bound_a) > delta_crit) {

  profit_at_c = leaf$profit_psi_stem_TF(bound_c, psi_soil);
  profit_at_d = leaf$profit_psi_stem_TF(bound_d, psi_soil);

  if (profit_at_c > profit_at_d) {
        bound_b = bound_d;
      } else {
        bound_a = bound_c;
      }

      bound_c = bound_b - (bound_b - bound_a) / gr;
      bound_d = bound_a + (bound_b - bound_a) / gr;

      }

(opt_psi_stem = ((bound_b + bound_a) / 2))
(profit = leaf$profit_psi_stem_TF(opt_psi_stem, psi_soil))
(transpiration = leaf$transpiration_)
```

### Validation against the compiled implementation

The final chunk calls the corresponding built-in TF24 `Leaf` optimisation. Its
purpose is to check that the compiled implementation returns the same optimum,
profit, and transpiration as the manual search.

```{r}
leaf$optimise_psi_stem_TF()
leaf$profit_
leaf$transpiration_
```

The complete calculation can therefore be read as a short sequence:

1. choose a candidate leaf xylem water potential $\psi_l$;
2. calculate the water supply and the stomatal conductance it can support;
3. solve for $C_i$ and calculate photosynthetic assimilation;
4. calculate the hydraulic cost associated with lost conductivity;
5. subtract cost from assimilation and search for the $\psi_l$ with the highest
   profit.

TF24 repeats this calculation as plant size and environmental conditions change.
The resulting assimilation contributes to the whole-plant carbon balance, while
transpiration feeds back on the soil-water state described in
[Soil water & root uptake](soil-water.qmd).
