---
title: "The size-structured PDE"
---
::: {.eyebrow}
theory
:::

This page outlines the demographic machinery used to model size-structured
populations in the `plant` package. We outline the system of equations being
solved for individual plants, patches of competing plants, and the age
structure of patches across a metacommunity. These equations were primarily
developed elsewhere, in particular in @Falster-2011 and @Falster-2015,
following general principles laid out in @Deroos-1997, @Kohyama-1993, and
@Moorcroft-2001. They are presented here so that users can understand the full
system of equations being solved within `plant`.

The companion page on [adaptive dynamics](adaptive-dynamics.qmd) builds on this
machinery to define invasion fitness and evolutionary endpoints.

@tbl-definitions provides a list of names and definitions used throughout this
page.

```{r}
#| label: setup
#| include: false
options(rmarkdown.html_vignette.check_title = FALSE)
library(plant)
```

```{r}
#| label: tbl-definitions
#| echo: false
#| output: asis
#| tbl-cap: "Variable names and definitions in the demographic model of the `plant` package."
system.file("docs", "demography-vars.csv", package = "plant") %>% read.csv() %>% dplyr::as_tibble() %>%
knitr::kable(align = "l")
```

# Metacommunity dynamics {#sec-system-dynamics}

The following material is written assuming there exists a physiological sub-model
that takes as inputs a plant's traits $x$, height $H$, and light environment $E_a$,
and returns rates of growth, mortality, and fecundity. The physiological model must
describe how much leaf area the plant contributes to shading other plants in the patch.
Specifically, the physiological model is responsible for calculating the following
variables from @tbl-definitions:
$A_{\rm l}(H)$, $Q(z, H)$, $H_0(x)$, $g(x, H, E_a)$, $f(x, H, E_a)$, $d(x, H, E_a)$, and
$S_{\rm G} (x, H_0, E_{{\rm a}0})$.
All are other variables in @tbl-definitions are calculated in the
demographic model using the equations provided below.

## Individual plants {#sec-individual-plants}

We first consider the dynamics of an individual plant. Throughout, we refer to a
plant as having traits $x$ and size $H$, with the latter given by its height. The
plant grows in a light environment $E$, a function describing the distribution
of light with respect to height within a patch. Ultimately, $E$ depends on the
composition of plants in a patch, and thus on the patch age $a$. To indicate this dependence, we
write $E_a$. Further, let $a_0$be the age in which the plant germinated, and $H_0(x)$ its initial height. The functions $g(x, H, E_a)$, $f(x, H, E_a)$, and
$d(x, H, E_a)$ denote the growth, death, and fecundity rates of the
plant. Then,
$$
H(x, a_0, a) = H_0(x) + \int_{a_0}^{a} g(x, H(x, a_0, a^\prime), E_{a^\prime}) \, {\rm d}a^\prime
$$ {#eq-size}
is the trajectory of plant height,
$$
S_{\rm I} (x, a_0, a) = S_{\rm G} (x, H_0, E_{{\rm a}0}) \, \exp\left(- \int_{a_0}^{a} d(x, H(x, a_0, a^\prime), E_{a^\prime}) \, {\rm d} a^\prime \right)
$$ {#eq-survivalindividual}
is the probability of plant survival within the patch, where $S_{\rm G} (x, H_0, E_{{\rm a}0})$
is the probability a seed germinates successfully, and
$$
\tilde{R}(x, a_0, a) = \int_{a_0}^{a} f(x, H(x, a_0, a^\prime), E_{a^\prime}) \, S_{\rm I} (x, a_0, a^\prime) \, {\rm d} a^\prime
$$ {#eq-tilder1}
is the plant's cumulative seed output, from its birth at patch age
$a = a_0$ until patch age $a$.

The notational complexity required in @eq-size-@eq-tilder1 potentially obscures an important point: these equations are simply the general, non-linear solutions to
integrating growth, mortality, and fecundity over time.

## Patches of competing plants {#sec-patches-of-competing-plants}

We now consider a patch of competing plants. At any age $a$, the
patch is described by the size-density distribution $N(H | x, a)$ of plants
with traits $x$ and height $H$. In a finite-sized patch, $N$ is described by a
collection of points, each indicating the height of one individual, whereas in a very (infinitely) large patch,
$N$ is a continuous function. In either case, the demographic
behaviour of the plants within the patch is given by @eq-size-@eq-tilder1. Integrating the dynamics over time is complicated by-
two other factors: (i) plants interact, thereby altering $E_a$ with
age, and (ii) new individuals may establish, expanding the system of
equations.

In the current version of `plant`, plants interact by shading one another.
Following standards biophysical principles, we let the canopy openness
$E_a(z)$ at height $z$ in a patch of age $a$ decline exponentially
with the total amount of leaf area above $z$,
$$
E_a(z) = \exp \left(-k_{\rm I} \sum_{i = 1}^{K} \int_{0}^{\infty} A_{\rm l}(H) \, Q(z, H) \, N(H | x_i, a) \, {\rm d}H \right),
$$ {#eq-light}
where $A_{\rm l}(H)$ is the total leaf area of a plant with height $H$, $Q(z, H)$ is the fraction of
this leaf area situated above height $z$ for plants of height $H$,
$k_{\rm I}$ is the light extinction coefficient, and $K$ is the number
of species in the metacommunity.

Assuming patches are sufficiently large, the dynamics of $N$ can be modelled
deterministically via the following partial differential equation (PDE)
[@Kohyama-1993; @Deroos-1997; @Moorcroft-2001],
$$
\frac{\partial}{\partial a} N(H | x, a) = - d(x, H, E_a) \, N(H | x, a) - \frac{\partial}{\partial H} \left[g(x, H, E_a) \, N(H | x, a)\right].
$$ {#eq-pde}
See @sec-derivation-size-structured for the derivation of this PDE.

@eq-pde has two boundary conditions. The first links the
flux of individuals across the lower bound $(H_0(x))$ of the size-density
distribution to the rate $Y_{x}$ at which seeds arrive in the patch,
$$
N(H_0 | x, a_0) = \left\{
\begin{array}{ll} \frac{Y_{x} \, S_{\rm G} (x, H_0, E_{{\rm a}0}) }{ g(x, H_0, E_{{\rm a}0}) } & \textrm{if } g(x, H_0, E_{{\rm a}0}) > 0 \\
0 & \textrm{otherwise.}
\end{array} \right.
$$ {#eq-bc1}
The function $S_{\rm G} (x, H_0, E_{{\rm a}0})$ denotes survival through
germination and must be chosen such that
$S_{\rm G} (x, H_0, E_{{\rm a}0}) / g(x, H_0, E_{{\rm a}0}) \rightarrow 0$ as
$g(x, H_0, E_{{\rm a}0}) \rightarrow 0$, to ensure a smooth decline in
initial density as conditions deteriorate [@Falster-2011].

The second boundary condition of @eq-pde specifies the initial size-density
distribution in patches right after a disturbance, i.e., for $a = 0$. Throughout, we consider only
situations starting with an empty patch,
$$
N\left(H|x,0\right) = 0,
$$ {#eq-bc2}
although non--zero distributions could alternatively be specified
[e.g. @Kohyama-1993; @Moorcroft-2001].

## Age structure of patches {#sec-age-structured-distribution}

We now consider the distribution of patch age $a$ in the metacommunity.
With $a$ denoting the time since last disturbance, $P(a)$ denotes the
frequency-density of patch age $a$ and $d_{\rm P}(a)$ is the
age-dependent probability that a patch of age $a$ is transformed into
a patch of age 0 by a disturbance event. Here we focus on situations
where the age structure has reached an equilibrium. See
@sec-derivation-age-structured for the derivation and
non-equilibrium situations. The dynamics of $P$ are given by
[@Mckendrick-1926; @Vonfoerster-1959]
$$
\frac{{\rm d}}{{\rm d} a} P(a) = - d_{\rm P}(a) \, P(a) ,
$$ {#eq-agepde}
with the boundary condition
$$
P(0) = \int_0^\infty d_{\rm P}(a) \, P(a) \, {\rm d} a.
$$ {#eq-agepd2}
The probability that a patch remains undisturbed from patch age $a_0$ to patch age $a$ is
then given by
$$
S_{\rm P} (a_0, a) = \exp\left( - \int_{a_0}^{a} d_{\rm P}(a^\prime) \, {\rm d} a^\prime \right).
$$ {#eq-survivalpatch}
These equations lead to an equilibrium distribution of patch ages,
$$
P(a) = P(0) S_{\rm P} (0, a),
$$ {#eq-agepde3}
where
$$
P(0) = \frac1{\int_0^\infty S_{\rm P} (0, a) {\rm d}a}
$$ {#eq-agepde4}
is the average disturbance frequency of a patch and, at the same time, the frequency-density of patches of
age $0$.

The default approach in `plant` is to assume that $d_{\rm P}(a)$
increases linearly with patch age, which leads to a Weibull distribution for $P$,
specified by a single parameter $\bar{a}$
measuring the mean interval between disturbances (see
@sec-weibull-distribution for details).

## Trait-, size-, and patch-structured metacommunities {#sec-trait-size-patch-structured}

We consider a large area of habitat where: (i) disturbances (such as fires,
storms, landslides, floods, or disease outbreaks) strike patches of the
habitat on a stochastic basis, killing all individuals within the affected
patches; (ii) individuals compete for resources within patches, with a
spatial scale of competitive interactions that renders negligible such interactions between
individuals in adjacent patches; and (iii) there is high
connectivity via dispersal between all patches in the habitat, allowing
empty patches to be quickly re-colonised. Such a system can be modelled
as a metapopulation (for a single species) or metacommunity (for multiple
species). The dynamics of such a metacommunity are described by the
PDEs in @eq-pde-@eq-agepde4.

The offspring production of each species in the metacommunity is given by the rate at which seeds are produced across all patches,
$$
Y_x  = \int_0^{\infty} \int_0^{\infty} P(a) \, S_{\rm D} \, f(x, H, E_a) \, N(H | x, a) \, {\rm d} H \, {\rm d} a,
$$ {#eq-seed-rain}
where $S_{\rm D}$ is the survival probability of seeds during dispersal.

A convenient feature of @eq-pde - @eq-agepde4 is that
the dynamics of a single patch scale up to give the dynamics of the
entire metacommunity. Note that the rate $Y_x$ at which offspring of individuals with traits $x$ arrive from the
disperser pool is constant when the metacommunity is at
equilibrium. Combined with the assumption that all patches have the same
initial (empty) size-density distribution after a disturbance, the assumption of constant birth rates
ensures that all patches show the same temporal behaviour, the only
difference between them being their ages.

To model the temporal dynamics of an archetypal patch, we need only a
value for $Y_x$. The numerical challenge is therefore to find the
right value for $Y_x$, by solving @eq-bc1-@eq-seed-rain simultaneously, for all species with traits $x$ in the metacommunity.

## Emergent properties of metacommunities {#sec-emergent-properties}

Summary statistics of a metacommunity are obtained by integrating
over the size-density distribution, weighting by the frequency-density $P(a)$.
In particular, the average density of individuals per unit ground area across the metacommunity is
$$
\hat{N}(x) = \int_{0}^{\infty} \int_{0}^{\infty} P(a) \, N(H | x, a) \, {\rm d}a \, {\rm d}H,
$$ {#eq-nav}
and the average size-density of plants with height $H$ is
$$
\bar{N}(x, H) = \int_{0}^{\infty}P(a) \, N(H | x, a) \, {\rm d}a.
$$ {#eq-ndav}

Averages for other individual-level quantities can also be calculated.
Denoting by
$w(x, H, E_a)$ a quantity of interest, either a demographic rate
(growth, mortality) or a state (plant height, leaf area, light environment),
the average of $w$ for plants with height $H$ and traits $x$ is
$$
\bar{w}(x, H) = \frac1{\bar{N}(x, H)}\int_{0}^{\infty}P(a) \, N(H | x, a) \, w(x, H, E_a) \, {\rm d}a.
$$ {#eq-wav1}
The average of $w$ across all individuals of the species is
$$
\hat{w}(x) = \frac1{\hat{N}(x) }\int_{0}^{\infty} \int_{0}^{\infty}P(a) \, N(H | x, a) \, w(x, H, E_a) \, {\rm d}a \, {\rm d}H.
$$ {#eq-wav2}
When calculating the average mortality rate, one must decide whether
mortality due to patch disturbance is included. Non-disturbance mortality
is obtained by setting $w(x, H, E_a) = d(x, H, E_a)$, while the total
mortality due to growth processes and disturbance is obtained by setting
$w(x, H, E_a) = d(x, H, E_a) + d_{\rm P}(a) S_{\rm P}(0, a).$

Similarly, we can integrate over the size-density distribution to extract aggregate
features of the vegetation within a patch,
$$
W(a) = \sum_{i = 1}^{N} \int_{0}^{\infty} N(H | x_i, a)\, w(x_i, H, E_a) \, {\rm d}H,
$$ {#eq-wa-a}
and across the entire metacommunity,
$$
\hat{W} = \int_{0}^{\infty} P(a) \, W(a) \, {\rm d}a.
$$ {#eq-wa-b}

For the analysis of invasion fitness and evolutionary endpoints that builds on
these dynamics, see the [adaptive dynamics](adaptive-dynamics.qmd) page.

# Appendices {#sec-appendices}

## Derivation of PDE describing age-structured dynamics {#sec-derivation-age-structured}

We consider patches of habitat that are subject to an intermittent
disturbance, with the age of a patch measuring the time
since the last disturbance. Denoting by $P(a, t)$ the frequency-density of
patch age $a$ at time $t$ and by $d_{\rm P}(a)$ the
age-dependent probability that a patch of age $a$ is transformed into
a patch of age $0$ through disturbance, the dynamics of $P(a, t)$ are given by
$$
\frac{\partial}{\partial t} P(a, t) = -\frac{\partial}{\partial a} P(a, t)-d_{\rm P}(a, t) P(a, t),
$$ {#eq-dp-a}
with boundary condition
$$
P(0, t) = \int^{\infty}_{0}d_{\rm P}(a, t) P(a, t) \, {\rm d}a.
$$ {#eq-p0}

The frequency-density of patches of age $a < a^{\prime}$ is given by
$\int_{0}^{a^{\prime}}P(a, t) \, {\rm d}a$, with
$\int_{0}^{\infty} P(a, t) \, {\rm d}a = 1$. If
$\frac{\partial}{\partial t}d_{\rm P}(a, t) = 0$, then $P(a)$ will
approach an equilibrium solution given by
$$
P(a) = P(0) \, S_{\rm P}(0, a),
$$ {#eq-pa}
where
$$
S_{\rm P}(0, a) = \exp \left( - \int_{0}^{a} d_{\rm P}(a^\prime) \, {\rm d}a^\prime\right)
$$ {#eq-s-p}
is the probability that a patch remains undisturbed for
duration $a$, and
$$
P(0) = \frac1{ \int_{0}^{\infty}S_{\rm P}(0, a) \, {\rm d}a}
$$ {#eq-p0-eq}
is the
frequency-density of patches of age $0$. The rate of disturbance for patches
of age $a$ is given by
$\frac{\partial (1-S_{\rm P}(0, a))}{\partial a} = - \frac{\partial S_{\rm P}(0, a)}{\partial a}$,
while the expected lifetime of patches is
$- \int_0^\infty a \frac{\partial}{\partial a} S_{\rm P}(0, a) \, {\rm d} a = \int_0^\infty S_{\rm P}(0, a) \, {\rm d} a = \frac1{P(0)}$
(first step made using integration by parts).

An equilibrium distribution of patch ages may be achieved under a variety of
conditions, for example, if $d_{\rm P}(a, t)$ depends on patch age $a$ but
not on time $t$. The rate of disturbance
may also depend on features of the vegetation in the patch, rather than on patch age directly, in which case an equilibrium distribution of patch ages can still arise, provided the
vegetation is also assumed to be at equilibrium.

### Exponential distribution {#sec-exponential-distribution}

If the rate $d_{\rm P}$ of patch disturbance is constant with respect to
patch age, then the rates at which patches of age $a$ are
disturbed follow an exponential distribution,
$-\partial S_{\rm P}(0, a)/ \partial a = d_{\rm P} \, \exp(-d_{\rm P} a)$. The
distribution of patch ages is then given by
$$
S_{\rm P}(0, a) = \exp\left(-d_{\rm P} a\right), \, P(0) = d_{\rm P}.
$$ {#eq-s-p0a1}

### Weibull distribution {#sec-weibull-distribution}

If the rate of patch disturbance changes with patch age according to
$d_{\rm P}(a) = \lambda \psi a^{\psi-1}$, then the rates at which
patches of age $a$ are disturbed follow a Weibull distribution,
$-\partial S_{\rm P}(0, a)/ \partial a = \lambda \psi a^{\psi -1}e^{-\lambda a^\psi}$.
$\psi>1$ implies that the probability of disturbance increases with
patch age, while $\psi<1$ implies that it decreases with
patch age. For $\psi = 1$, we obtain the exponential distribution, a special
case of the Weibull distribution. The Weibull distribution results in the following
distribution of patch ages,
$$
S_{\rm P}(0, a) = \exp(-\lambda a^\psi), \, P(0) = \frac{\psi \lambda^{\frac1{\psi}}}{\Gamma\left(\frac1{\psi}\right)},
$$ {#eq-s-p0a2}
where $\Gamma(x) = \int_{0}^{\infty}e^{-t}t^{x-1} \, dt$ is the gamma function. We can
also specify this distribution by the mean  disturbance interval
$\bar{a} = \frac1{P(0)}$. From this, we can calculate the relevant value for
$\lambda = \left(\frac{\Gamma\left(\frac1{\psi}\right)}{\psi \bar{a}}\right)^{\psi}$.

The default in `plant` is to assume $\psi=2$, such that $d_{\rm P}$ increases as
a linear function of patch age. The distribution of patch ages is then specified by a
single parameter, $\bar{a}$.


## Derivation of PDE describing size-structured dynamics {#sec-derivation-size-structured}

To model the metacommunity, we use a PDE describing the dynamics for a thin
slice $\Delta H$. This PDE can be derived as follows, following
@Deroos-1997. Note that in this sub-section the dependencies on the traits $x$ are deliberately not made explicit. Assuming that all rates are constant within the
interval $\Delta H$, the total number of individuals within the
interval spanned by $[H - 0.5\Delta H, H + 0.5\Delta H)$ is
$N(H, a)\Delta H$. The flux of individuals in and out of this size
interval can be expressed as
$$
\begin{array}{ll} &g(H - 0.5 \Delta H, a) \, N(H - 0.5 \Delta H, a) - g(H + 0.5 \Delta H, a) \, N(H + 0.5 \Delta H, a) \\ & - d (H, a) \, N(H, a)\Delta H\\ \end{array}.
$$ {#eq-flux-1}
The first two terms describe the flux in and out of the considered size interval
through growth, while the last term describes losses through mortality.
The change in the density of individuals within this size interval during a time
step $\Delta $a is thus
$$
\begin{array}{ll} N(H, a + \Delta a)\Delta H - N(H, a)\Delta H = &g(H - 0.5 \Delta H, a) \, N(H - 0.5 \Delta H, a)\Delta a \\ & - g(H + 0.5 \Delta H, a) \, N(H + 0.5 \Delta H, a)\Delta a\\& - d (H, a) \, N(H, a)\Delta H\Delta a.
\end{array}
$$ {#eq-flux-2}
By rearranging, we obtain
$$
\begin{array}{ll}
\frac{N(H, a + \Delta a) - N(H, a)}{\Delta a} = & - d (H, a) \, N(H, a) \\
& - \frac{g(H + 0.5 \Delta H, a) \, N(H + 0.5 \Delta H, a) - g(H - 0.5 \Delta H, a) \, N(H - 0.5 \Delta H, a)}{\Delta H}.
\end{array}
$$ {#eq-flux-3}
The left-hand side above corresponds to the derivative of $N$ as $\Delta a\to 0$.
For thin slices, $\Delta H \to 0$, this yields
$$
\frac{\partial}{\partial a} N(H, a) = - d (H, a) \, N(H, a) - \frac{\partial}{\partial H} (g(H, a) \, N(H, a)).
$$ {#eq-pde-app}

To complete the model, this PDE must be supplemented with boundary
conditions that specify the density at the lower end of heights
for all $a$, as well as the initial distribution
$N(H,0)$. The former is derived by integrating the PDE with respect to
$H$ over the interval $(H_{0}, H_{\infty} )$, yielding
$$
\frac{\partial}{\partial a} \int _{H_{0} }^{H_{\infty}}N(H, a) \, {\rm d} H = g(H_{0} , a) \, N(H_{0} , a) - g(H_{\infty}, a) \, N(H_{\infty}, a) - \int _{H_{0} }^{H_{\infty}}d (H, a) \, N(H, a) \, {\rm d} H.
$$ {#eq-partial-a}
The left-hand size above is evidently the rate of change of the total
density of individuals in the population, while the right-hand side is
the population's total death rate. Further, we can assume that $N(H_{\infty}, a) = 0$. Thus,
to balance total births and deaths, $g(H_{0} , a) \, N(H_{0} , a)$ must
equal the population's total birth rate $B$, yielding the boundary condition
$$
g(H_{0} , a) \, N(H_{0} , a) = B.
$$ {#eq-g0}

## Converting density from one size unit to another {#sec-converting-density}

The quantification of size-density depends on the chosen size
unit (@eq-pde). Thus, what if we want to express size-density with respect to
another size unit? A relation between the two corresponding values of size-density can be derived
by noting that the total density of individuals within a given size range
must be equal. We consider a situation in which size-density is expressed in units of size $M$,
but we want it in units of size $H$. First, we require a
one-to-one function that translates the first size unit into the second, $H = \hat{H}(M)$. Then
the following must hold
$$
\int_{M_1}^{M_2} N(M | x, a) \, {\rm d}M = \int_{\hat{H}(M_1)}^{\hat{H}(M_2)} N^\prime(H | x, a) \, {\rm d}H .
$$ {#eq-n-conversion}
For very small size intervals, this equation is equivalent to
$$
\left(M_2 - M_1 \right) \, N(M_1 | x, a) = \left( \hat{H}(M_2) - \hat{H}(M_1)\right) \, N^\prime(\hat{H}(M_1) | x, a).
$$ {#eq-n-conversion2}
Rearranging gives
$$
N^\prime(\hat{H}(M_1) | x, a) = N(M_1 | x, a) \, \frac{M_2 - M_1}{\hat{H}(M_2) - \hat{H}(M_1)}.
$$ {#eq-n-conversion3}
Noting that the second term on the right-hand side is simply the definition of
$\frac{{\rm d} M}{{\rm d} H}$ evaluated at $M_1$, we have
$$
N^\prime(H | x, a) = N(M | x, a) \, \frac{{\rm d} M}{{\rm d} H}.
$$ {#eq-n-conversion4}
