---
title: "The size-structured PDE"
---
::: {.eyebrow}
theory
:::

::: {.callout-note}
## Prerequisites
This page builds on [The big picture](overview.qmd) and [The maths, gently](maths-primer.qmd). Read those first if the notation here is unfamiliar.
:::

## The idea first

Tracking a forest plant by plant would quickly become unmanageable. A
size-structured model takes a different approach: it describes how many plants
occur at each size and follows the resulting distribution through time. Growth
moves plants towards larger sizes, mortality removes them, and recruitment
adds new plants at the smallest size.

Competition links those processes. In the canopy-light strategies, the plants
in a patch create a vertical light profile, and that profile changes their
growth, mortality, and reproduction. Disturbance adds another level: a
landscape contains patches of different ages, each with its own size
distribution. The metacommunity is obtained by combining those patches.

This page develops the equations in that same order: individual life histories,
one competing patch, and then a landscape of patches. The formulation used in
`plant` was developed in @Falster-2011 and @Falster-2015, following general
principles from @Deroos-1997, @Kohyama-1993, and @Moorcroft-2001.

The companion page on [adaptive dynamics](adaptive-dynamics.qmd) builds on this
machinery to define invasion fitness and evolutionary endpoints.

The argument proceeds in layers. We first follow one plant through its life,
then combine many plants into a size distribution within one patch, and finally
combine patches of different ages into a metacommunity. The appendices derive
the two transport equations and show how to convert a density from one size
coordinate to another.

::: {.callout-note}
## How to read size and environment on this page
The notation on this page follows the FF16 and TF24 strategies: $H$ is plant height and
$E_a$ includes the canopy light profile. K93 uses stem diameter instead as its biological size
coordinate and cumulative basal area as its competitive environment, while
TF24 additionally couples physiological rates to soil water. In each case the
strategy supplies the appropriate rates to the same demographic bookkeeping.
:::

@tbl-definitions provides a list of names and definitions used throughout this
page.

```{r}
#| label: setup
#| include: false
options(rmarkdown.html_vignette.check_title = FALSE)
library(plant)
```

```{r}
#| label: tbl-definitions
#| echo: false
#| output: asis
#| tbl-cap: "Variable names and definitions in the demographic model of the `plant` package."
system.file("docs", "demography-vars.csv", package = "plant") %>%
  read.csv() %>%
  dplyr::as_tibble() %>%
  knitr::kable(align = "l")
```

## Metacommunity dynamics {#sec-system-dynamics}

The demographic equations need a physiological sub-model to supply the biology.
For a plant with traits $x$, size $H$, and environmental state $E_a$, that
sub-model returns its growth, mortality, and fecundity rates. In the
canopy-light formulation it also returns the leaf area the plant contributes to
shading. Specifically, the physiological model is responsible for calculating
the following variables from @tbl-definitions:
$A_{\rm l}(H)$, $Q(z, H)$, $H_0(x)$, $g(x, H, E_a)$, $f(x, H, E_a)$, $d(x, H, E_a)$, and
$S_{\rm G} (x, H_0, E_{{\rm a}0})$.
All other variables in @tbl-definitions are calculated by the demographic
model using the equations below. This division is important: changing the
physiological strategy changes the supplied rates, while the demographic
bookkeeping stays the same.

### Individual plants {#sec-individual-plants}

We first follow one plant. Its traits are $x$, its size is represented by height
$H$, and $a$ is the current age of the patch in which it grows. The plant
germinated earlier, when the patch was age $a_0$, at an initial height $H_0(x)$.

The light environment $E$ is a vertical profile through the patch rather than a
single number. Because that profile changes as the patch develops, we write it
as $E_a$. The functions $g(x, H, E_a)$, $f(x, H, E_a)$, and
$d(x, H, E_a)$ are respectively the plant's instantaneous growth, fecundity,
and mortality rates.

Height at patch age $a$ is initial height plus all growth accumulated since
germination:
$$
H(x, a_0, a) = H_0(x) + \int_{a_0}^{a} g(x, H(x, a_0, a^\prime), E_{a^\prime}) \, {\rm d}a^\prime
$$ {#eq-size}

Survival has two parts: the probability of surviving germination, followed by
survival against the mortality hazard accumulated as the plant grows:
$$
S_{\rm I} (x, a_0, a) = S_{\rm G} (x, H_0, E_{{\rm a}0}) \, \exp\left(- \int_{a_0}^{a} d(x, H(x, a_0, a^\prime), E_{a^\prime}) \, {\rm d} a^\prime \right)
$$ {#eq-survivalindividual}

Here $S_{\rm G} (x, H_0, E_{{\rm a}0})$ is the probability that a seed
germinates successfully. Finally, cumulative seed output adds fecundity over
time, but discounts seed production by the probability that the plant is still
alive:
$$
\tilde{R}(x, a_0, a) = \int_{a_0}^{a} f(x, H(x, a_0, a^\prime), E_{a^\prime}) \, S_{\rm I} (x, a_0, a^\prime) \, {\rm d} a^\prime
$$ {#eq-tilder1}

Thus Eqs. [-@eq-size]--[-@eq-tilder1] are three running totals along the same
life history: accumulated growth, accumulated mortality risk, and accumulated
reproduction. Their notation allows every rate to change non-linearly with
traits, plant size, light, and patch age.

### Patches of competing plants {#sec-patches-of-competing-plants}

We now move from one plant to a patch. At patch age $a$,
$N(H | x, a)$ is the size-density distribution of plants with traits $x$ and
height $H$. It is a density with respect to height: integrating it over a
height interval gives the abundance of plants in that interval. A finite patch
can instead be represented by discrete individuals, but for a sufficiently
large patch the model treats $N$ as a continuous function.

Each plant still follows Eqs. [-@eq-size]--[-@eq-tilder1], but two feedbacks now
matter. Existing plants alter the shared light profile $E_a$, and new seedlings
continually enter at the lower end of the size distribution.

For the FF16 and TF24 canopy-light formulation, plants interact by shading one
another. TF24's physiological rates also respond to the soil-water state.
Following standard biophysical principles, we let the canopy openness
$E_a(z)$ at height $z$ in a patch of age $a$ decline exponentially
with the total amount of leaf area above $z$,
$$
E_a(z) = \exp \left(-k_{\rm I} \sum_{i = 1}^{K} \int_{0}^{\infty} A_{\rm l}(H) \, Q(z, H) \, N(H | x_i, a) \, {\rm d}H \right),
$$ {#eq-light}
where $A_{\rm l}(H)$ is the total leaf area of a plant with height $H$, $Q(z, H)$ is the fraction of
this leaf area situated above height $z$ for plants of height $H$,
$k_{\rm I}$ is the light extinction coefficient, and $K$ is the number
of species in the metacommunity.

Read @eq-light from the inside out. For species $i$, the integral totals leaf
area above $z$ across all plant heights; the sum then combines the $K$ species.
Multiplying by $k_{\rm I}$ converts leaf area into optical depth, and the
exponential converts optical depth into the fraction of light that remains.
The result is a single patch light profile that every plant both contributes to
and experiences.

Assuming patches are sufficiently large, the dynamics of $N$ can be modelled
deterministically via the following partial differential equation (PDE)
[@Kohyama-1993; @Deroos-1997; @Moorcroft-2001],
$$
\frac{\partial}{\partial a} N(H | x, a) = - d(x, H, E_a) \, N(H | x, a) - \frac{\partial}{\partial H} \left[g(x, H, E_a) \, N(H | x, a)\right].
$$ {#eq-pde}
See @sec-derivation-size-structured for the derivation of this PDE.
<!-- SY: These display as section numbers, bu the numbers aren't visible in the actual section headings. You could either set number-sections: true in the yaml or use a plain Markdown link e.g. [Infiltration and runoff](#sec-infiltration) and [Drainage](#sec-drainage) define ... -->

The two terms on the right-hand side are a population ledger. The first removes
plants through mortality. The second describes the flow of plants along the
height axis as they grow: plants leave one height interval and enter the next.
It is written as the derivative of the flux $gN$, so changes in both growth
rate and density are handled together.

@eq-pde also needs two boundary conditions. The first supplies recruits at the
smallest height $H_0(x)$. It links the flux of seedlings entering the
size-density distribution to the rate $Y_{x}$ at which seeds arrive in the
patch,
$$
N(H_0 | x, a_0) = \left\{
\begin{array}{ll} \frac{Y_{x} \, S_{\rm G} (x, H_0, E_{{\rm a}0}) }{ g(x, H_0, E_{{\rm a}0}) } & \textrm{if } g(x, H_0, E_{{\rm a}0}) > 0 \\
0 & \textrm{otherwise.}
\end{array} \right.
$$ {#eq-bc1}
The function $S_{\rm G} (x, H_0, E_{{\rm a}0})$ denotes survival through
germination and must be chosen such that
$S_{\rm G} (x, H_0, E_{{\rm a}0}) / g(x, H_0, E_{{\rm a}0}) \rightarrow 0$ as
$g(x, H_0, E_{{\rm a}0}) \rightarrow 0$, to ensure a smooth decline in
initial density as conditions deteriorate [@Falster-2011].

The division by $g$ converts a flux of successful recruits per unit time into a
density per unit height. The special case for $g \leq 0$ prevents seedlings
from accumulating at the boundary when conditions do not allow them to grow
into the size distribution.

The second boundary condition specifies the state of a patch immediately after
disturbance, at $a = 0$. Throughout, we consider situations starting with an
empty patch,
$$
N\left(H|x,0\right) = 0,
$$ {#eq-bc2}
although non--zero distributions could alternatively be specified
[e.g. @Kohyama-1993; @Moorcroft-2001].

### Age structure of patches {#sec-age-structured-distribution}

We now zoom out from one patch to a landscape containing patches of many ages.
Here $a$ is time since the last disturbance, and $P(a)$ is the
frequency-density of patches of age $a$. Integrating $P(a)$ over an age range
gives the fraction of landscape in that range. The function $d_{\rm P}(a)$ is
the instantaneous disturbance rate, or hazard, for a patch of age $a$.

We first focus on an equilibrium patch-age distribution. See
@sec-derivation-age-structured for the time-dependent derivation. At
equilibrium, patches flow steadily toward older ages and leave each age class
when disturbed [@Mckendrick-1926; @Vonfoerster-1959]: <!-- SY: What does a non-equilibrium patch-age distribution look like in contrast? -->
$$
\frac{{\rm d}}{{\rm d} a} P(a) = - d_{\rm P}(a) \, P(a) ,
$$ {#eq-agepde}
with the boundary condition
$$
P(0) = \int_0^\infty d_{\rm P}(a) \, P(a) \, {\rm d} a.
$$ {#eq-agepd2}

The boundary condition says that the flow of newly disturbed, age-zero patches
must equal the total disturbance flow out of all older ages.

The probability that a patch remains undisturbed from patch age $a_0$ to patch age $a$ is
then given by
$$
S_{\rm P} (a_0, a) = \exp\left( - \int_{a_0}^{a} d_{\rm P}(a^\prime) \, {\rm d} a^\prime \right).
$$ {#eq-survivalpatch}

This has the same survival form as the individual-plant equation: cumulative
hazard appears inside a negative exponential. The difference is that the event
is patch disturbance rather than plant death.

These equations lead to an equilibrium distribution of patch ages,
$$
P(a) = P(0) S_{\rm P} (0, a),
$$ {#eq-agepde3}
where
$$
P(0) = \frac1{\int_0^\infty S_{\rm P} (0, a) {\rm d}a}
$$ {#eq-agepde4}
is the average disturbance frequency of a patch and, at the same time, the frequency-density of patches of
age $0$.

In plain language, the abundance of patches of age $a$ equals the rate at which
new patches are created, $P(0)$, multiplied by the chance that a patch survives
undisturbed to age $a$. The normalising expression in @eq-agepde4 makes the
total area under $P(a)$ equal to one.

The default approach in `plant` is to assume that $d_{\rm P}(a)$
increases linearly with patch age, which leads to a Weibull distribution for $P$,
specified by a single parameter $\bar{a}$
measuring the mean interval between disturbances (see
@sec-weibull-distribution for details).

### Trait-, size-, and patch-structured metacommunities {#sec-trait-size-patch-structured}

The one-patch calculation scales up to a metacommunity under three main
assumptions:

- disturbances such as fires, storms, or landslides reset whole patches;
- plants compete within a patch, while competition between neighbouring
  patches is negligible; and
- dispersal connects the patches strongly enough that the same seed pool can
  recolonise them.

With one species this is a metapopulation; with several it is a
metacommunity. Eqs. [-@eq-pde]--[-@eq-agepde4] describe its demographic
dynamics.

The disperser pool links the patch and landscape scales. For each strategy
$x$, its landscape-wide seed rain $Y_x$ is the rate at which viable seeds are
produced across all patches,
$$
Y_x  = \int_0^{\infty} \int_0^{\infty} P(a) \, S_{\rm D} \, f(x, H, E_a) \, N(H | x, a) \, {\rm d} H \, {\rm d} a,
$$ {#eq-seed-rain}
where $S_{\rm D}$ is the survival probability of seeds during dispersal.

Again, read the double integral from the inside out. At a given patch age $a$,
the inner integral adds fecundity over every plant height, weighted by plant
density. The outer integral averages that patch output across patch ages,
weighted by how common each age is. $S_{\rm D}$ retains only seeds that survive
dispersal.

At metacommunity equilibrium, the seed-arrival rate $Y_x$ is constant. If every <!-- SY: What does metacommunity equilibrium mean exactly? -->
disturbance also leaves the same empty initial state, all patches then follow
the same developmental trajectory. Their only difference is how long ago they
were disturbed. This is why one patch trajectory, combined with the patch-age
distribution, can represent the whole metacommunity.

To model the temporal dynamics of an archetypal patch, we need a value for
$Y_x$. But that value is both an input to the patch through @eq-bc1 and an
output of the metacommunity through @eq-seed-rain. At demographic equilibrium,
the two must match. The numerical challenge is therefore to solve
Eqs. [-@eq-bc1]--[-@eq-seed-rain] simultaneously for all strategies $x$ in the
metacommunity.

### Emergent properties of metacommunities {#sec-emergent-properties}

Once $N(H | x, a)$ and $P(a)$ are known, stand- and landscape-level outputs are
weighted integrals. A hat denotes an average over the entire metacommunity,
while a bar denotes an average across patch ages at a specified plant
height.

The average density of individuals with traits $x$ per unit ground area across
the metacommunity is
$$
\hat{N}(x) = \int_{0}^{\infty} \int_{0}^{\infty} P(a) \, N(H | x, a) \, {\rm d}a \, {\rm d}H,
$$ {#eq-nav}
and the average size-density of plants with height $H$ is
$$
\bar{N}(x, H) = \int_{0}^{\infty}P(a) \, N(H | x, a) \, {\rm d}a.
$$ {#eq-ndav}

Averages for other individual-level quantities use the same weighting. Denote by
$w(x, H, E_a)$ a quantity of interest, either a demographic rate
(growth, mortality) or a state (plant height, leaf area, light environment),
the average of $w$ for plants with height $H$ and traits $x$ is
$$
\bar{w}(x, H) = \frac1{\bar{N}(x, H)}\int_{0}^{\infty}P(a) \, N(H | x, a) \, w(x, H, E_a) \, {\rm d}a.
$$ {#eq-wav1}

The numerator adds $w$ across patch ages, weighted by the density of plants to
which it applies; dividing by $\bar{N}(x, H)$ turns that total into an average
for plants at height $H$.

The average of $w$ across all individuals of the species is
$$
\hat{w}(x) = \frac1{\hat{N}(x) }\int_{0}^{\infty} \int_{0}^{\infty}P(a) \, N(H | x, a) \, w(x, H, E_a) \, {\rm d}a \, {\rm d}H.
$$ {#eq-wav2}

Here the same calculation also integrates over height and divides by the total
metacommunity density $\hat{N}(x)$.

When calculating the average mortality rate, one must decide whether
mortality due to patch disturbance is included. Non-disturbance mortality
is obtained by setting $w(x, H, E_a) = d(x, H, E_a)$, while the total
mortality due to growth processes and disturbance is obtained by setting
$w(x, H, E_a) = d(x, H, E_a) + d_{\rm P}(a) S_{\rm P}(0, a).$

Sometimes the desired output is a total rather than a per-plant average. We can
integrate over the size-density distribution to obtain an aggregate feature of
the vegetation within a patch,
$$
W(a) = \sum_{i = 1}^{N} \int_{0}^{\infty} N(H | x_i, a)\, w(x_i, H, E_a) \, {\rm d}H,
$$ {#eq-wa-a}
and across the entire metacommunity,
$$
\hat{W} = \int_{0}^{\infty} P(a) \, W(a) \, {\rm d}a.
$$ {#eq-wa-b}

For the analysis of invasion fitness and evolutionary endpoints that builds on
these dynamics, see the [adaptive dynamics](adaptive-dynamics.qmd) page.

## Appendices {#sec-appendices}

The main text states the model in its working form. The appendices below show
where the transport equations come from and explain one subtle property of any
size density: its numerical value depends on the size coordinate in which it is
expressed.

### Derivation of PDE describing age-structured dynamics {#sec-derivation-age-structured}

We now distinguish two clocks. Patch age $a$ is the time since a patch's last
disturbance, while calendar time $t$ describes change in the whole landscape.
Let $P(a, t)$ be the frequency-density of patches of age $a$ at time $t$, and
let $d_{\rm P}(a, t)$ be their instantaneous disturbance rate. Its dynamics are
given by
$$
\frac{\partial}{\partial t} P(a, t) = -\frac{\partial}{\partial a} P(a, t)-d_{\rm P}(a, t) P(a, t),
$$ {#eq-dp-a}

The first term on the right moves undisturbed patches toward older ages. The
second removes patches of age $a$ when disturbance resets them to age zero.
Those resets supply the boundary condition
$$
P(0, t) = \int^{\infty}_{0}d_{\rm P}(a, t) P(a, t) \, {\rm d}a.
$$ {#eq-p0}

The frequency-density of patches of age $a < a^{\prime}$ is given by
$\int_{0}^{a^{\prime}}P(a, t) \, {\rm d}a$, with
$\int_{0}^{\infty} P(a, t) \, {\rm d}a = 1$. If
$\frac{\partial}{\partial t}d_{\rm P}(a, t) = 0$, then $P(a)$ will
approach an equilibrium solution given by
$$
P(a) = P(0) \, S_{\rm P}(0, a),
$$ {#eq-pa}
where
$$
S_{\rm P}(0, a) = \exp \left( - \int_{0}^{a} d_{\rm P}(a^\prime) \, {\rm d}a^\prime\right)
$$ {#eq-s-p}
is the probability that a patch remains undisturbed for
duration $a$, and
$$
P(0) = \frac1{ \int_{0}^{\infty}S_{\rm P}(0, a) \, {\rm d}a}
$$ {#eq-p0-eq}
is the
frequency-density of patches of age $0$. The rate of disturbance for patches
of age $a$ is given by
$\frac{\partial (1-S_{\rm P}(0, a))}{\partial a} = - \frac{\partial S_{\rm P}(0, a)}{\partial a}$,
while the expected lifetime of patches is
$- \int_0^\infty a \frac{\partial}{\partial a} S_{\rm P}(0, a) \, {\rm d} a = \int_0^\infty S_{\rm P}(0, a) \, {\rm d} a = \frac1{P(0)}$
(first step made using integration by parts).

Thus $1/P(0)$ has two equivalent interpretations at equilibrium: it is the
mean time between disturbances and the reciprocal of the rate at which
age-zero patches appear.

An equilibrium distribution of patch ages may be achieved under a variety of
conditions, for example, if $d_{\rm P}(a, t)$ depends on patch age $a$ but
not on time $t$. The rate of disturbance
may also depend on features of the vegetation in the patch, rather than on patch age directly, in which case an equilibrium distribution of patch ages can still arise, provided the
vegetation is also assumed to be at equilibrium.

#### Exponential distribution {#sec-exponential-distribution}

If the rate $d_{\rm P}$ of patch disturbance is constant with respect to
patch age, then the rates at which patches of age $a$ are
disturbed follow an exponential distribution,
$-\partial S_{\rm P}(0, a)/ \partial a = d_{\rm P} \, \exp(-d_{\rm P} a)$. The
distribution of patch ages is then given by
$$
S_{\rm P}(0, a) = \exp\left(-d_{\rm P} a\right), \, P(0) = d_{\rm P}.
$$ {#eq-s-p0a1}

A constant hazard is memoryless: surviving to an old age does not change a
patch's instantaneous chance of disturbance.

#### Weibull distribution {#sec-weibull-distribution}

If the rate of patch disturbance changes with patch age according to
$d_{\rm P}(a) = \lambda \psi a^{\psi-1}$, then the rates at which
patches of age $a$ are disturbed follow a Weibull distribution,
$-\partial S_{\rm P}(0, a)/ \partial a = \lambda \psi a^{\psi -1}e^{-\lambda a^\psi}$.
$\psi>1$ implies that the probability of disturbance increases with
patch age, while $\psi<1$ implies that it decreases with
patch age. For $\psi = 1$, we obtain the exponential distribution, a special
case of the Weibull distribution. The Weibull distribution results in the following
distribution of patch ages,
$$
S_{\rm P}(0, a) = \exp(-\lambda a^\psi), \, P(0) = \frac{\psi \lambda^{\frac1{\psi}}}{\Gamma\left(\frac1{\psi}\right)},
$$ {#eq-s-p0a2}
where $\Gamma(x) = \int_{0}^{\infty}e^{-t}t^{x-1} \, dt$ is the gamma function. We can
also specify this distribution by the mean  disturbance interval
$\bar{a} = \frac1{P(0)}$. From this, we can calculate the relevant value for
$\lambda = \left(\frac{\Gamma\left(\frac1{\psi}\right)}{\psi \bar{a}}\right)^{\psi}$.

The default in `plant` is to assume $\psi=2$, such that $d_{\rm P}$ increases as
a linear function of patch age. The distribution of patch ages is then specified by a
single parameter, $\bar{a}$.

The shape parameter $\psi$ controls how disturbance risk changes with age;
$\lambda$ then sets the time scale. Fixing $\psi=2$ leaves only the mean
disturbance interval to specify.


### Derivation of PDE describing size-structured dynamics {#sec-derivation-size-structured}

The size-structured PDE is a conservation equation: change inside a thin height
slice equals plants growing in, minus plants growing out, minus deaths. We now
write that ledger explicitly, following @Deroos-1997. To keep the notation
readable, this subsection suppresses the dependence on traits $x$.

Assuming that all rates are constant within a thin interval $\Delta H$, the
total number of individuals within the
interval spanned by $[H - 0.5\Delta H, H + 0.5\Delta H)$ is
$N(H, a)\Delta H$. The flux of individuals in and out of this size
interval can be expressed as
$$
\begin{array}{ll} &g(H - 0.5 \Delta H, a) \, N(H - 0.5 \Delta H, a) - g(H + 0.5 \Delta H, a) \, N(H + 0.5 \Delta H, a) \\ & - d (H, a) \, N(H, a)\Delta H\\ \end{array}.
$$ {#eq-flux-1}
The first two terms describe the flux in and out of the considered size interval
through growth, while the last term describes losses through mortality.
The change in the density of individuals within this size interval during a time
step $\Delta a$ is thus
$$
\begin{array}{ll} N(H, a + \Delta a)\Delta H - N(H, a)\Delta H = &g(H - 0.5 \Delta H, a) \, N(H - 0.5 \Delta H, a)\Delta a \\ & - g(H + 0.5 \Delta H, a) \, N(H + 0.5 \Delta H, a)\Delta a\\& - d (H, a) \, N(H, a)\Delta H\Delta a.
\end{array}
$$ {#eq-flux-2}
By rearranging, we obtain
$$
\begin{array}{ll}
\frac{N(H, a + \Delta a) - N(H, a)}{\Delta a} = & - d (H, a) \, N(H, a) \\
& - \frac{g(H + 0.5 \Delta H, a) \, N(H + 0.5 \Delta H, a) - g(H - 0.5 \Delta H, a) \, N(H - 0.5 \Delta H, a)}{\Delta H}.
\end{array}
$$ {#eq-flux-3}
The left-hand side above corresponds to the derivative of $N$ as $\Delta a\to 0$.
For thin slices, $\Delta H \to 0$, this yields
$$
\frac{\partial}{\partial a} N(H, a) = - d (H, a) \, N(H, a) - \frac{\partial}{\partial H} (g(H, a) \, N(H, a)).
$$ {#eq-pde-app}

This is the same PDE used in the main text: mortality removes density locally,
while the derivative of $gN$ moves density through height space.

To complete the model, this PDE must be supplemented with boundary
conditions that specify the density at the lower end of heights
for all $a$, as well as the initial distribution
$N(H,0)$. The former is derived by integrating the PDE with respect to
$H$ over the interval $(H_{0}, H_{\infty} )$, yielding
$$
\frac{\partial}{\partial a} \int _{H_{0} }^{H_{\infty}}N(H, a) \, {\rm d} H = g(H_{0} , a) \, N(H_{0} , a) - g(H_{\infty}, a) \, N(H_{\infty}, a) - \int _{H_{0} }^{H_{\infty}}d (H, a) \, N(H, a) \, {\rm d} H.
$$ {#eq-partial-a}
The left-hand side above is the rate of change of the total density of
individuals in the population. On the right, the first two terms are fluxes
through the lower and upper size boundaries, and the integral is the total
death rate. We assume that $N(H_{\infty}, a) = 0$. The incoming flux
$g(H_{0} , a) \, N(H_{0} , a)$ must therefore equal the population's total
birth rate $B$, yielding the boundary condition
$$
g(H_{0} , a) \, N(H_{0} , a) = B.
$$ {#eq-g0}

### Converting density from one size unit to another {#sec-converting-density}

The numerical value of a density depends on the unit along its horizontal axis.
For example, a density per kilogram is not numerically identical to a density
per metre, even when both describe the same plants. What must stay unchanged is
the number of individuals represented by a corresponding interval.

Suppose size-density is expressed using size $M$, but we want to express it
using size $H$. We require a one-to-one conversion
$H = \hat{H}(M)$. Equality of abundance over matching intervals then requires
the following:
$$
\int_{M_1}^{M_2} N(M | x, a) \, {\rm d}M = \int_{\hat{H}(M_1)}^{\hat{H}(M_2)} N^\prime(H | x, a) \, {\rm d}H .
$$ {#eq-n-conversion}
For very small size intervals, this equation is equivalent to
$$
\left(M_2 - M_1 \right) \, N(M_1 | x, a) = \left( \hat{H}(M_2) - \hat{H}(M_1)\right) \, N^\prime(\hat{H}(M_1) | x, a).
$$ {#eq-n-conversion2}
Rearranging gives
$$
N^\prime(\hat{H}(M_1) | x, a) = N(M_1 | x, a) \, \frac{M_2 - M_1}{\hat{H}(M_2) - \hat{H}(M_1)}.
$$ {#eq-n-conversion3}
Noting that the second term on the right-hand side is simply the definition of
$\frac{{\rm d} M}{{\rm d} H}$ evaluated at $M_1$, we have
$$
N^\prime(H | x, a) = N(M | x, a) \, \frac{{\rm d} M}{{\rm d} H}.
$$ {#eq-n-conversion4}

The derivative $\mathrm{d}M/\mathrm{d}H$ is the Jacobian of the change of
variables. It stretches or compresses the density exactly enough that
integrating it still returns the same number of plants.
