---
title: "When the Richards equation fights back"
description: "A standalone deSolve diagnosis of the numerical instability we hit adding a Richards-equation soil-water profile to the TF24 strategy — and why the fix is the solver, not the code."
date: 2025-07-29
categories: [tf24, soil-water, numerics, experimental]
reading-time: true
---

```{r setup, include=FALSE}
source(here::here("R/version-badge.R"))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

We've recently been trying to expand the TF24 strategy to handle soil
water, but our initial attempt encountered serious instability after we
implemented the Richards equation. The symptom was unmissable: with
anything but an absurdly small time step the integrator produced
nonsense within a few steps — soil moisture leaping above saturation,
going negative, then collapsing to `NaN`. The same pathology showed up
whether the equation was implemented in C++ or in R, which is the first
clue that the problem is mathematical rather than a coding slip.

This post works through a minimal R reproduction of that instability,
shows the contrast between an explicit and an implicit solver, and ends
with the diagnosis: the discretised Richards equation is **stiff**, and
no amount of careful coding will rescue an explicit Runge–Kutta scheme
from it. The full back-and-forth — including the $\psi$-based form, a
single-bucket fallback, and a delay-equation experiment — lives in
[traitecoevo/plant discussion #444](https://github.com/traitecoevo/plant/discussions/444);
here we keep just enough to stand on its own.

::: {.callout-note}
This is a standalone numerical experiment. It prototypes the soil-water
dynamics directly with `deSolve` and does **not** use the `plant`
package API, so it isn't pinned to a `plant` version — the point is the
solver behaviour, which is independent of the model it will eventually
live in.
:::

```{r}
#| label: libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(deSolve)
```

## The model

We track volumetric soil water content $\theta$ on a vertical grid of
$n$ nodes. Each node exchanges water with its neighbours by Darcy flux,
driven by gradients in the matric potential $\psi$, plus gravity. This
is the *mixed form* of the Richards equation, following
Ireson et al. (2023):

$$\frac{\partial \theta}{\partial t} = -\frac{\partial q}{\partial z},
\qquad
q = -K(\theta)\left(\frac{\partial \psi}{\partial z} - 1\right).$$

The constitutive relations — hydraulic conductivity $K(\theta)$ and
matric potential $\psi(\theta)$ — are the usual power laws:

$$K(\theta) = K_\text{sat}\left(\frac{\theta}{\theta_\text{sat}}\right)^{2n_\psi + 3},
\qquad
\psi(\theta) = a_\psi\left(\frac{\theta}{\theta_\text{sat}}\right)^{-n_\psi}.$$

```{r}
#| label: constitutive
pow <- function(x, y) x^y

soil_K_from_soil_theta <- function(theta, K_sat = 440.628,
                                   soil_moist_sat = 0.453, n_psi = 4.8) {
  K_sat * pow(theta / soil_moist_sat, 2 * n_psi + 3)
}

psi_from_soil_moist <- function(soil_moist, soil_moist_sat = 0.453,
                                a_psi = 8.7, n_psi = 4.8) {
  a_psi * pow(soil_moist / soil_moist_sat, -n_psi)
}

params <- list(
  soil_moist_sat = 0.453,   # saturated water content (m^3/m^3)
  K_sat          = 440.628, # saturated conductivity (units below)
  n_psi          = 4.8,     # pore-size distribution exponent
  a_psi          = 8.7,     # air-entry scaling
  b_infil        = 8.0,     # infiltration shape parameter
  rain           = 1,       # rainfall forcing
  n              = 10       # number of soil nodes
)
```

The right-hand side assembles the inter-node fluxes — an infiltration
flux at the surface that shuts off as the top layer saturates, Darcy
fluxes between interior nodes, and a free-drainage lower boundary
($q = K$ at the bottom) — then differences them to get $d\theta/dt$ at
each node:

```{r}
#| label: rates
rates <- function(t, theta, params) {
  with(as.list(c(theta, params)), {

    psi <- psi_from_soil_moist(theta, soil_moist_sat = soil_moist_sat)
    K   <- soil_K_from_soil_theta(theta, soil_moist_sat = soil_moist_sat)

    q  <- rep(NA_real_, n + 1)
    dz <- rep(2 / n, n)
    dtheta_dt <- rep(NA_real_, n)

    # top boundary: infiltration, shutting off as the top layer fills
    q[1] <- rain * max(0.0, 1 - pow(theta[1] / soil_moist_sat, b_infil))

    # interior boundaries: Darcy flux (Eq. 3 + 11, Ireson et al. 2023)
    for (i in seq_len(n - 1)) {
      dpsi_dz  <- (psi[i + 1] - psi[i]) / dz[i]
      q[i + 1] <- 0.5 * (K[i] + K[i + 1]) * (dpsi_dz - 1)
    }

    # lower boundary: free drainage (Eq. 13, Ireson et al. 2023)
    q[n + 1] <- K[n]

    # divergence of flux gives the rate of change at each node (Eq. 10)
    for (i in seq_len(n)) {
      dq_dz <- (q[i + 1] - q[i]) / dz[i]
      dtheta_dt[i] <- -dq_dz
    }

    list(dtheta_dt)
  })
}
```

A small wrapper integrates the system and returns a long-format tibble,
and a helper plots $\theta$ at each depth over time:

```{r}
#| label: solve-helpers
solve_soil <- function(times, rates, state, method = "rk4", output = "theta") {
  ode(y = state, times = times, func = rates, parms = params, method = method) %>%
    as_tibble() %>%
    pivot_longer(-time, names_to = "depth", values_to = output) %>%
    mutate(depth = factor(as.integer(depth)))
}

plot_solution <- function(solution) {
  solution %>%
    ggplot(aes(x = time, y = theta, colour = depth)) +
    geom_line() +
    labs(x = "time (yr)", y = expression(theta), colour = "node") +
    theme_minimal()
}

n <- 10
state <- rep(0.2533, n)
```

## The instability

Now integrate with the classic explicit Runge–Kutta solver (`rk4`) at a
time step of $10^{-4}$ yr — about an hour, which sounds harmlessly
small:

```{r}
#| label: fig-rk4-fail
#| fig-cap: "Explicit RK4 at Δt = 1e-4 yr. The solution explodes within two steps — moisture overshoots saturation, goes negative, then becomes `NaN` (gaps in the lines)."
#| fig-width: 7
#| fig-height: 4
fail <- solve_soil(times = seq(0, 1e-3, by = 1e-4),
                   rates = rates, state = state, method = "rk4")
plot_solution(fail)
```

It detonates almost immediately. Printing the first few steps for the
top nodes shows the runaway in numbers: a steady start, then values
flung well past the physical range, then `NaN`:

```{r}
#| label: fail-values
fail %>%
  filter(depth %in% c(1, 2, 3)) %>%
  pivot_wider(names_from = depth, values_from = theta, names_prefix = "node ") %>%
  head(4)
```

This is not a bug in the right-hand side — the rates are continuous and
correct. It's the explicit integrator amplifying tiny errors faster than
the dynamics damp them, the hallmark of a **stiff** system.

The reflex fix is to shrink the step. At $10^{-5}$ yr (~5 minutes) RK4
*does* stay stable:

```{r}
#| label: fig-rk4-ok
#| fig-cap: "Explicit RK4 at Δt = 1e-5 yr stays stable — but this is only the first hundredth of a year, and already cost 1000 steps. A full year needs 100,000."
#| fig-width: 7
#| fig-height: 4
ok <- solve_soil(times = seq(0, 1e-2, by = 1e-5),
                 rates = rates, state = state, method = "rk4")
plot_solution(ok)
```

The infiltration front is smooth and physical — but look at the cost.
That figure covers just $1/100$ of a year and already took 1000 steps.
Integrating a single year would need 100,000 explicit steps, and a
forest simulation runs for centuries. This is not a viable solver for
the model.

## The fix: an implicit solver

Stiffness is solved by *implicit* methods, which evaluate the
right-hand side at the end of the step and so remain stable at large
step sizes. `deSolve`'s `lsoda` switches automatically between stiff and
non-stiff modes. Swapping it in — and taking a step of $0.1$ yr, ten
thousand times larger than the one RK4 needed — integrates the whole
year cleanly:

```{r}
#| label: fig-lsoda
#| fig-cap: "Implicit LSODA at Δt = 0.1 yr integrates a full year in ~10 steps with no instability."
#| fig-width: 7
#| fig-height: 4
stable <- solve_soil(times = seq(0, 1, by = 0.1),
                     rates = rates, state = state, method = "lsoda")
plot_solution(stable)
```

Same right-hand side, same parameters, same initial condition — the only
change is the solver, and the instability is gone. That is the clean
experimental proof that the problem was stiffness all along.

## Why this happens

The Richards equation is a nonlinear, degenerate elliptic–parabolic PDE.
As Farthing & Ogden (2017) put it, it is "degenerate" because its
strongly nonlinear coefficients approach zero in parts of the solution
domain. Here the conductivity $K(\theta) \propto \theta^{\,2n_\psi+3}$
spans many orders of magnitude across the profile: with $n_\psi = 4.8$
that exponent is about $12.6$, so a modest moisture gradient produces an
enormous spread of local timescales between nodes. Fast nodes set the
stability limit of an explicit step; slow nodes set how long you must
integrate. The ratio between them *is* the stiffness, and it forces
explicit methods into impractically tiny steps.

## Takeaways for the soil-water work

- The instability is intrinsic to the discretised Richards equation, not
  to our implementation — it reproduced identically in C++ and R.
- Explicit Runge–Kutta is the wrong tool. The model needs an implicit /
  adaptive stiff solver (LSODA, or a BDF-type method) if we keep the
  full Richards profile.
- If a stiff solver proves awkward to embed in the strategy's inner
  loop, the discussion explores cheaper alternatives — a single- or
  two-layer bucket model, or a Kirchhoff-transform linearisation
  (Ross 2003) — that sidestep the stiffness rather than solving it.

The next step for TF24 soil water is to decide between embedding an
implicit solver and adopting a reduced-order bucket model; this
experiment is what ruled out the naive explicit route.

## References

- Farthing, M. W. & Ogden, F. L. (2017). Numerical solution of Richards'
  equation: a review of advances and challenges. *Soil Science Society
  of America Journal*, 81(6), 1257–1269.
- Ireson, A. M. et al. (2023). The finite volume method for solving the
  Richards equation. *Hydrology and Earth System Sciences*.
- Ross, P. J. (2003). Modeling soil water and solute transport — fast,
  simplified numerical solutions. *Agronomy Journal*, 95(6), 1352–1361.
- [traitecoevo/plant discussion #444 — Problem modelling soil water with
  Richards Equation](https://github.com/traitecoevo/plant/discussions/444).
```

