---
# ── REPRODUCTION (draft) ───────────────────────────────────────────────
# This post lives in posts/_drafts/ ON PURPOSE: the paper is in prep, so
# its DOI / journal / year / figure numbers / publication code-ref cannot
# be verified yet. Those fields are TODO, and `fidelity: pending` — which
# is exactly what keeps the reproduction lint from gating it. Move it to
# posts/<date>-<slug>/ only once the paper metadata is real.
title: "Reproducing DunFalster 2026: are cheap canopy-shading models adequate for evolutionary endpoints?"
description: "The heavy analysis behind the canopy-methods paper — demographic equilibrium, invasion-fitness landscapes, the evolutionary attractor, and the PPA grid-artifact — re-run as a six-model adequacy comparison against develop-branch plant."
date: 2026-06-20
# Pinned to the develop branch (this analysis uses the new mean-light /
# crown-centre / box shading models, which are not in the 2.1.0 release).
plant-ref: "develop"
plant-sha: "81cfeab16a049478ca0ff81f1239b78221c49ebb"
categories: [reproduction, canopy, evolution]
reading-time: true
# Structured metadata about the source paper (drives the citation box):
paper:
  authors: "Falster et al."   # in prep; full author list/order TODO
  year: TODO                   # in prep — do not guess a publication year
  title: "TODO — canopy/shading-model adequacy for stand dynamics and evolutionary endpoints"
  journal: "TODO — in prep"
  doi: "TODO — none yet (unpublished)"
  figure: "TODO — paper figure numbers not finalised"
  fidelity: "pending"
  # ── original paper code ──────────────────────────────────────────────
  # The paper's own figure scripts live in the plant repo under
  # vignettes/examples/DunFalster2026/. To re-run them here they would be
  # vendored as a paper-code submodule at paper-code/dunfalster2026-canopy/
  # (CI: git submodule update --init), pinned to the PUBLICATION commit.
  code-repo: "traitecoevo/plant"                                   # scripts under vignettes/examples/DunFalster2026/
  code-ref: "TODO — publication SHA/tag (not yet tagged; paper in prep)"
  code-entrypoint: "vignettes/examples/DunFalster2026/make_figures.R"  # + make_figure_attractor.R, make_figure_gridartifact.R
---

```{r setup, include=FALSE}
source(here::here("R/version-badge.R"))
source(here::here("R/reproduction.R"))
```

`r plant_version_badge()` `r repro_fidelity_badge()` `r paper_code_badge()`

`r paper_citation()`

::: {.callout-important}
**Draft / pending.** This post reproduces the analysis behind the
in-preparation **DunFalster 2026** canopy-methods paper (Falster et al.).
Because the paper is unpublished, its **DOI, journal, year, figure
numbers, and the publication commit (`code-ref`) of the figure scripts
are all `TODO`** — they cannot be verified and have not been guessed. The
fidelity verdict is therefore `pending`. The post stays in `_drafts/` so
the reproduction lint does not gate it; promote it once the paper
metadata is real.
:::

## What the original shows

The DunFalster 2026 paper asks **which canopy-shading simplifications are
safe, and for what purpose**. It puts six shading models through a ladder
of increasingly demanding tests of biological adequacy — stand dynamics,
demographic equilibrium, invasion (mutant) fitness, and finally the
**evolutionary endpoint** (the singular strategy a trait converges to).
Its load-bearing result is that a model can be perfectly adequate for one
rung and badly wrong for the next: the cheap models reproduce stand
*trajectories* fine, yet predict equilibrium stand densities spanning an
order of magnitude, and settle evolution on visibly different optima — a
~15% spread in the attracted leaf-mass-per-area (LMA).

This post carries across the **heavy analysis** deliberately kept out of
the lighter theory page
[canopy models](../../theory/implementation/canopy-models.qmd): the
demographic-equilibrium iteration, the invasion-fitness landscapes, the
evolutionary-attractor / singular-strategy root-finding, the
work-precision / grid-artifact convergence study, and the six-model
adequacy comparison. The transcription is faithful to the source vignette
`vignettes/canopy_methods.Rmd`; the paper's own publication-grade panels
are produced by the scripts under
`vignettes/examples/DunFalster2026/` (see [Re-running the paper's own
figures](#re-running-the-papers-own-figures)).

## The six models

`plant`'s `FF16` strategy offers six shading models via
`control$shading_model`. The paper-to-code naming (from the paper's
`model-name-mapping.md`):

| Paper name | Meaning | `plant` `shading_model` |
|---|---|---|
| Deep Crown | smooth (Yokozawa) shade; assimilation integrated over crown depth | `deep-crown` (default) |
| Mean-Light | smooth shade; light averaged over crown depth, one assimilation call | `mean-light` (new) |
| Crown-Centre | smooth shade; one assimilation call at the crown centre | `crown-centre` (new) |
| PPA (smoothed) | layered stepped light profile, boundaries smoothed (runnable) | `ppa`, `ppa_layer_smoothing > 0` (default 0.3) |
| PPA (hard step) | literal field discretisation; discontinuous, does not run | `ppa`, `ppa_layer_smoothing = 0` |
| Flat Top | box / thin-layer shade (field's usual sense) | `flat-top-box` (hard), `flat-top-soft-box` (smoothed) |

The key isolation the naming serves: `crown-centre` keeps a **smooth**
shade and only flattens the *assimilation* calculation, so it is the
paper's **Crown-Centre** — *not* the field's "Flat Top", which means the
box / thin-layer *shade* (`flat-top-box` / `flat-top-soft-box`).
Crown-Centre and Flat Top share the same crown-centre assimilation and
differ only in the shade cast, so the contrast isolates the shade cast,
not the assimilation rule.

```{r}
#| label: libs
#| message: false
library(plant)
library(dplyr)
library(ggplot2)
library(patchwork)
```

## Shared setup

A single `FF16` species at LMA = 0.0825 (exactly as the source vignette
and `make_figures.R`), run under each of the four smooth-competition
models. Everything except `control$shading_model` is held fixed.

```{r}
#| label: setup-stands
params <- scm_base_parameters("FF16")
patch  <- expand_parameters(trait_matrix(0.0825, "lma"), params)

models <- c("deep-crown", "mean-light", "crown-centre", "ppa")

run_model <- function(model) {
  ctrl <- Control()
  ctrl$shading_model <- model
  run_scm(patch, ctrl = ctrl, collect = TRUE)
}

results <- lapply(models, run_model)
names(results) <- models

species_all <- dplyr::bind_rows(
  lapply(models, function(m) dplyr::mutate(results[[m]]$species, method = m))
)
species_all$method <- factor(species_all$method, levels = models)

# Light-extinction coefficient: E = exp(-k_I * LAI), so LAI = -log(E) / k_I.
k_I <- results[[1]]$p$strategies[[1]]$k_I
```

## Stand dynamics (rung 1)

The least demanding rung: does a stand develop sensibly? Leaf area index
reads straight off the recorded light profile via Beer's law
($\mathrm{LAI} = -\log E(0) / k_I$), needing no integration over the size
distribution.

```{r}
#| label: lai
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Leaf area index through patch age. deep-crown, mean-light and crown-centre track each other; ppa oscillates because its discrete layers create sharp recruitment thresholds."
lai_all <- dplyr::bind_rows(
  lapply(models, function(m) {
    results[[m]]$env$light_availability |>
      dplyr::group_by(time) |>
      dplyr::slice_min(height, n = 1, with_ties = FALSE) |>
      dplyr::ungroup() |>
      dplyr::transmute(method = m, time, lai = -log(light_availability) / k_I)
  })
)
lai_all$method <- factor(lai_all$method, levels = models)

ggplot(lai_all, aes(time, lai, colour = method)) +
  geom_line(linewidth = 1) +
  labs(x = "Patch age (years)", y = "Leaf area index", colour = "Shading model") +
  theme_classic()
```

The three smooth models settle to a steady LAI; `"ppa"` shows large
sustained oscillations — waves of recruitment and self-thinning driven by
the discrete light layers.

## Demographic equilibrium (rung 2)

The next rung asks whether the population closes its own loop sensibly,
and supplies the resident for the fitness analysis. We put the stand at
its **demographic equilibrium**: the seed rain at which the population
exactly replaces itself ($R_0 = 1$). The equilibrium is a fixed point of
the patch's own seed production — put a seed rain in, run the patch, read
the seed rain out, feed it back. It converges in a handful of iterations.

```{r}
#| label: equil-setup
p_fit <- scm_base_parameters("FF16")
p_fit$max_patch_lifetime <- 50
lma_resident <- 0.0825

# Iterate seed-rain-out -> seed-rain-in until it stops moving (R0 = 1).
find_equilibrium_birth_rate <- function(model, n_iter = 6, birth_rate = 1) {
  ctrl <- Control()
  ctrl$shading_model <- model
  for (i in seq_len(n_iter)) {
    resident <- expand_parameters(trait_matrix(lma_resident, "lma"), p_fit,
                                  birth_rate_list = birth_rate)
    birth_rate <- run_scm(resident, ctrl = ctrl)$offspring_production
  }
  birth_rate
}
find_equilibrium_birth_rate("deep-crown")   # converges to ~1.66
```

Crucially the equilibrium **depends on the shading model**: one that lets
plants capture more light supports a denser stand. Running the iteration
for every model gives equilibria spanning an order of magnitude — PPA's
brighter layered light supports more than ten times the seed rain of
deep-crown. These are hardcoded from the paper so the build does not
re-run the slower models (re-derive with `find_equilibrium_birth_rate()`
above):

```{r}
#| label: equil-table
equilibrium_birth_rate <- c(
  "deep-crown"        = 1.662,
  "mean-light"        = 2.178,
  "crown-centre"      = 2.185,
  "ppa"               = 18.835,
  "flat-top-soft-box" = 3.101
)
knitr::kable(data.frame(
  model = names(equilibrium_birth_rate),
  `equilibrium seed rain` = unname(equilibrium_birth_rate),
  check.names = FALSE
))
```

That an order-of-magnitude difference in predicted stand density follows
from the shading assumption alone is the first sign that these models, so
alike in their stand-development trajectories, are not interchangeable.

## Invasion (mutant) fitness (rung 3)

Invasion fitness is far more demanding. The fitness of a rare mutant is
its lifetime reproductive output as it grows through the light cast by an
established resident — so it depends not on the light at one size, but on
how the light *changes as the plant grows*. A fitness landscape is only
useful if it is **smooth**: kinks and jumps leave optima and selection
gradients ill-defined.

We compute invasion fitness for mutants spanning a range of LMA at fine
resolution, under two seedings: (1) every model at the *same* resident
seed rain, isolating the shading model; (2) each model at *its own*
equilibrium, so each casts a self-consistent stand. The resident is run
with `save_RK45_cache = TRUE` so mutants can be introduced into the
fixed landscape with `run_mutant()`; fitness is `log(R0)`.

```{r}
#| label: fitness
#| fig-width: 10
#| fig-height: 3.6
#| fig-cap: "Invasion fitness landscapes under two seedings. The smooth models give smooth landscapes; ppa is jagged, wildly so at its own dense equilibrium."
mutant_lma <- seq(0.07, 0.11, length.out = 60)

fitness_landscape <- function(model, birth_rate) {
  ctrl <- Control()
  ctrl$shading_model <- model
  ctrl$save_RK45_cache <- TRUE                       # required by run_mutant()
  resident <- expand_parameters(trait_matrix(lma_resident, "lma"), p_fit,
                                birth_rate_list = birth_rate)
  scm <- run_scm(resident, ctrl = ctrl)              # caches the light environment
  mutants <- expand_parameters(trait_matrix(mutant_lma, "lma"), resident,
                               birth_rate_list = rep(1, length(mutant_lma)))
  scm$run_mutant(mutants)
  r0 <- scm$net_reproduction_ratios                  # resident first, then mutants
  data.frame(method = model, lma = mutant_lma,
             fitness = log(tail(r0, length(mutant_lma))))
}

common_rain <- equilibrium_birth_rate[["deep-crown"]]
fit_same  <- dplyr::bind_rows(lapply(models, fitness_landscape, birth_rate = common_rain))
fit_equil <- dplyr::bind_rows(lapply(models, function(m)
  fitness_landscape(m, equilibrium_birth_rate[[m]])))

fit_two <- dplyr::bind_rows(
  dplyr::mutate(fit_same,  seeding = "same seed rain"),
  dplyr::mutate(fit_equil, seeding = "each at its own equilibrium")
)
fit_two$method  <- factor(fit_two$method, levels = models)
fit_two$seeding <- factor(fit_two$seeding,
                          levels = c("same seed rain", "each at its own equilibrium"))

ggplot(fit_two, aes(lma, fitness, colour = seeding)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~method, nrow = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey70") +
  labs(x = "Mutant LMA", y = "Invasion fitness, log(R0)", colour = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

`"deep-crown"`, `"mean-light"` and `"crown-centre"` yield smooth
landscapes under both seedings; `"ppa"` is jagged, and at its own (much
denser) equilibrium wildly so, with a non-monotonic dip where small trait
changes flip which layer the mutant ends in. The deeper lesson is about
**continuity of the light profile**, not the cost of the assimilation
calculation: the three smooth models build a continuous light profile and
differ only in how they summarise assimilation, so all three give smooth
fitness. PPA's stepped profile breaks this.

### Getting the competition profile right

Why does crown-centre survive while the box models do not? Crown-centre
approximates a plant's *own* assimilation crudely but still computes the
shade it *casts* from the full smooth Yokozawa crown. The continuity that
fitness needs comes from that competition profile. The box models keep
crown-centre's assimilation but collapse the shade cast towards a thin
layer — mirroring the discrete-layer competition of large vegetation
models (LPJ-GUESS and the like).

```{r}
#| label: compbox
#| fig-width: 7
#| fig-height: 3.6
#| fig-cap: "Shade cast vs height by a 10 m plant. Crown-centre tapers continuously; the hard box is a step; the soft box is a continuous-but-wrong drop."
comp_profile <- function(model, height = 10) {
  s <- FF16_Strategy()
  s$control$shading_model <- model
  ind <- FF16_Individual(s)
  ind$set_state("height", height)
  z <- seq(0, height, length.out = 400)
  data.frame(model = model, z = z,
             competition = sapply(z, function(zz) ind$compute_competition(zz)))
}
comp <- dplyr::bind_rows(
  comp_profile("crown-centre"),
  comp_profile("flat-top-box"),
  comp_profile("flat-top-soft-box")
)

ggplot(comp, aes(competition, z, colour = model)) +
  geom_path() +
  labs(x = "Shade cast (competition) by a 10 m plant", y = "Height (m)",
       colour = "Model") +
  theme_classic()
```

**The hard step does not even run.** Summed over a population of discrete
cohorts, step competition makes the patch light profile discontinuous,
which the adaptive light-environment spline cannot represent — the model
fails before a stand can develop:

```{r}
#| label: box-fail
ctrl_box <- Control()
ctrl_box$shading_model <- "flat-top-box"
tryCatch(
  run_scm(patch, ctrl = ctrl_box),
  error = function(e) conditionMessage(e)
)
```

**The soft box runs, but gives the wrong answer.** Its continuous
competition lets the light environment build and a fitness landscape be
computed, but the shade is mis-distributed through the canopy, so the
landscape is biased relative to deep-crown. The *correct* crown-centre
tracks deep-crown; the soft box does not.

```{r}
#| label: softbox-fitness
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "The soft box runs but is biased: correct crown-centre tracks deep-crown, the soft box does not. Each model at its own equilibrium."
box_models <- c("deep-crown", "crown-centre", "flat-top-soft-box")
fit_compare <- dplyr::bind_rows(lapply(box_models, function(m)
  fitness_landscape(m, equilibrium_birth_rate[[m]])))
fit_compare$method <- factor(fit_compare$method, levels = box_models)

ggplot(fit_compare, aes(lma, fitness, colour = method)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  labs(x = "Mutant LMA", y = "Invasion fitness, log(R0)", colour = "Model") +
  theme_classic()
```

So getting the competition profile right is a prerequisite, not a
refinement. A plant model can tolerate a crude *assimilation*
approximation and still behave well for dynamics and fitness, but the
shade each plant *casts* as a function of height must be both continuous
(or the light environment is ill-defined) and the right shape (or
evolutionary predictions are biased). This is worth flagging as
trait-based DGVMs (FATES and the like) are increasingly pushed towards
eco-evolutionary questions: the trouble is confined to discriminating
mutants that differ only *slightly*, where two near-identical strategies
can fall on opposite sides of a layer boundary and receive incomparable
fitnesses.

## The evolutionary attractor (rung 4)

The most demanding rung. A smooth landscape is not the point in itself —
the reason we want one is to ask **where evolution settles**, and whether
the methods that agree a landscape is smooth also agree on its endpoint.
They do not.

The endpoint is the trait that is its own best reply: the **singular
strategy** $s^\*$, where the selection gradient vanishes. We locate it by
root-solving the gradient. For each candidate resident LMA: (1) find its
demographic-equilibrium seed rain ($R_0 = 1$); (2) the selection gradient
is the slope of mutant $\log R_0$ at mutant = resident (central finite
difference); (3) root-find where the gradient crosses zero. A negative
slope at the crossing means convergence stability; a negative fitness
curvature means uninvadability (an ESS); both together make it a **CSS**.

```{r}
#| label: attractor-machinery
# Demographic-equilibrium seed rain for a resident of given LMA.
equilibrium_birth_rate_at <- function(model, lma, n_iter = 6, birth_rate = 1) {
  ctrl <- Control(); ctrl$shading_model <- model
  for (i in seq_len(n_iter)) {
    resident <- expand_parameters(trait_matrix(lma, "lma"), p_fit,
                                  birth_rate_list = birth_rate)
    birth_rate <- run_scm(resident, ctrl = ctrl)$offspring_production
  }
  birth_rate
}

# Selection gradient d log(R0)/d(LMA) at mutant = resident, resident at its
# own demographic equilibrium. Central finite difference.
sel_gradient <- function(model, lma, delta = 0.0015, n_iter = 6) {
  ctrl <- Control(); ctrl$shading_model <- model; ctrl$save_RK45_cache <- TRUE
  br <- equilibrium_birth_rate_at(model, lma, n_iter)
  resident <- expand_parameters(trait_matrix(lma, "lma"), p_fit,
                                birth_rate_list = br)
  scm <- run_scm(resident, ctrl = ctrl)
  mutants <- expand_parameters(trait_matrix(c(lma - delta, lma + delta), "lma"),
                               resident, birth_rate_list = rep(1, 2))
  scm$run_mutant(mutants)
  f <- log(tail(scm$net_reproduction_ratios, 2))
  (f[2] - f[1]) / (2 * delta)
}

# Curvature d2 log(R0)/d(LMA)^2 at s* (uninvadability test): < 0 => CSS.
fitness_curvature <- function(model, s, delta = 0.0015, n_iter = 6) {
  ctrl <- Control(); ctrl$shading_model <- model; ctrl$save_RK45_cache <- TRUE
  br <- equilibrium_birth_rate_at(model, s, n_iter)
  resident <- expand_parameters(trait_matrix(s, "lma"), p_fit,
                                birth_rate_list = br)
  scm <- run_scm(resident, ctrl = ctrl)
  mutants <- expand_parameters(trait_matrix(c(s - delta, s, s + delta), "lma"),
                               resident, birth_rate_list = rep(1, 3))
  scm$run_mutant(mutants)
  f <- log(tail(scm$net_reproduction_ratios, 3))
  (f[3] - 2 * f[2] + f[1]) / delta^2
}
```

As a live check, we root-find the default `"deep-crown"` attractor over
the LMA interval where its gradient changes sign. A single `uniroot()`
converges in a handful of equilibrium solves:

```{r}
#| label: attractor-live
deep_crown_star <- uniroot(function(x) sel_gradient("deep-crown", x),
                           interval = c(0.100, 0.1025), tol = 1e-5)$root
deep_crown_star                                   # ~0.1007
```

Mapping the gradient over a grid and root-solving + curvature-testing all
five models is the expensive part (a few minutes — every grid point
re-solves an equilibrium and PPA is slow). So, as with the equilibrium
table, we hardcode the precomputed results and plot from them. They are
the paper's own outputs, written by
`examples/DunFalster2026/make_figure_attractor.R`
(`evolutionary-attractor-gradients.csv` and
`evolutionary-attractor-singular-strategies.csv`).

```{r}
#| label: attractor-data
attractor_gradients <- read.csv(strip.white = TRUE, blank.lines.skip = TRUE,
  text = "method,lma,gradient
deep-crown,0.0650,437.8
deep-crown,0.0675,410.9
deep-crown,0.0700,382.6
deep-crown,0.0725,353.4
deep-crown,0.0750,323.2
deep-crown,0.0775,292.2
deep-crown,0.0800,260.7
deep-crown,0.0825,229.0
deep-crown,0.0850,197.1
deep-crown,0.0875,165.2
deep-crown,0.0900,133.4
deep-crown,0.0925,101.8
deep-crown,0.0950,70.5
deep-crown,0.0975,39.5
deep-crown,0.1000,8.9
deep-crown,0.1025,-21.2
deep-crown,0.1050,-50.8
deep-crown,0.1075,-79.8
deep-crown,0.1100,-108.2
deep-crown,0.1125,-136.0
deep-crown,0.1150,-163.2
deep-crown,0.1175,-189.8
deep-crown,0.1200,-215.8
deep-crown,0.1225,-241.3
deep-crown,0.1250,-266.1
mean-light,0.0650,496.5
mean-light,0.0675,466.6
mean-light,0.0700,434.0
mean-light,0.0725,399.1
mean-light,0.0750,362.5
mean-light,0.0775,324.5
mean-light,0.0800,285.4
mean-light,0.0825,245.6
mean-light,0.0850,205.1
mean-light,0.0875,164.3
mean-light,0.0900,123.4
mean-light,0.0925,82.5
mean-light,0.0950,41.8
mean-light,0.0975,1.3
mean-light,0.1000,-38.9
mean-light,0.1025,-78.6
mean-light,0.1050,-117.9
mean-light,0.1075,-156.7
mean-light,0.1100,-195.1
mean-light,0.1125,-233.0
mean-light,0.1150,-270.2
mean-light,0.1175,-306.9
mean-light,0.1200,-343.2
mean-light,0.1225,-379.0
mean-light,0.1250,-414.2
crown-centre,0.0650,571.2
crown-centre,0.0675,531.3
crown-centre,0.0700,487.5
crown-centre,0.0725,440.0
crown-centre,0.0750,389.4
crown-centre,0.0775,336.6
crown-centre,0.0800,281.9
crown-centre,0.0825,226.1
crown-centre,0.0850,169.5
crown-centre,0.0875,112.4
crown-centre,0.0900,54.8
crown-centre,0.0925,-2.5
crown-centre,0.0950,-59.2
crown-centre,0.0975,-115.4
crown-centre,0.1000,-171.1
crown-centre,0.1025,-225.7
crown-centre,0.1050,-279.6
crown-centre,0.1075,-332.4
crown-centre,0.1100,-384.2
crown-centre,0.1125,-434.7
crown-centre,0.1150,-484.2
crown-centre,0.1175,-532.7
crown-centre,0.1200,-579.9
crown-centre,0.1225,-626.2
crown-centre,0.1250,-671.0
flat-top-soft-box,0.0650,717.9
flat-top-soft-box,0.0675,662.2
flat-top-soft-box,0.0700,600.0
flat-top-soft-box,0.0725,532.7
flat-top-soft-box,0.0750,461.6
flat-top-soft-box,0.0775,387.6
flat-top-soft-box,0.0800,312.1
flat-top-soft-box,0.0825,235.5
flat-top-soft-box,0.0850,158.5
flat-top-soft-box,0.0875,81.0
flat-top-soft-box,0.0900,3.7
flat-top-soft-box,0.0925,-73.5
flat-top-soft-box,0.0950,-150.6
flat-top-soft-box,0.0975,-227.7
flat-top-soft-box,0.1000,-304.6
flat-top-soft-box,0.1025,-381.6
flat-top-soft-box,0.1050,-458.8
flat-top-soft-box,0.1075,-536.6
flat-top-soft-box,0.1100,-615.1
flat-top-soft-box,0.1125,-694.2
flat-top-soft-box,0.1150,-774.1
flat-top-soft-box,0.1175,-854.8
flat-top-soft-box,0.1200,-936.9
flat-top-soft-box,0.1225,-1020.0
flat-top-soft-box,0.1250,-1104.0
ppa,0.0650,3461.1
ppa,0.0675,2945.3
ppa,0.0700,3120.4
ppa,0.0725,2647.5
ppa,0.0750,2729.6
ppa,0.0775,2445.4
ppa,0.0800,2650.1
ppa,0.0825,1093.1
ppa,0.0850,271.6
ppa,0.0875,-310.5
ppa,0.0900,-2335.3
ppa,0.0925,-2840.7
ppa,0.0950,-3062.7
ppa,0.0975,-2961.9
ppa,0.1000,-3913.1
ppa,0.1025,-3968.6
ppa,0.1050,-4075.2
ppa,0.1075,-2792.7
ppa,0.1100,-2631.3
ppa,0.1125,-2557.8
ppa,0.1150,-4407.0
ppa,0.1175,-86.9
ppa,0.1200,-4012.5
ppa,0.1225,-461.9
ppa,0.1250,-167.4")

singular_strategies <- read.csv(strip.white = TRUE, blank.lines.skip = TRUE,
  text = "method,lma_star,curvature,class,robust
deep-crown,0.1007,-12477,CSS,TRUE
mean-light,0.0976,-17176,CSS,TRUE
crown-centre,0.0924,-20663,CSS,TRUE
flat-top-soft-box,0.0901,-30828,CSS,TRUE
ppa,0.0859,-250149,CSS,FALSE")

model_order <- c("deep-crown", "mean-light", "crown-centre",
                 "flat-top-soft-box", "ppa")
attractor_gradients$method <- factor(attractor_gradients$method, levels = model_order)
singular_strategies$method <- factor(singular_strategies$method, levels = model_order)
```

```{r}
#| label: tbl-attractor
#| tbl-cap: "Singular strategies. All five are convergence-stable, uninvadable maxima (CSS), yet the attracted LMA spreads ~15%. PPA's endpoint is flagged stepped (set by where the layer boundaries fall)."
knitr::kable(data.frame(
  model                   = as.character(singular_strategies$method),
  `attracted LMA`         = round(singular_strategies$lma_star, 4),
  `equilibrium seed rain` = unname(equilibrium_birth_rate[as.character(singular_strategies$method)]),
  `fitness curvature`     = singular_strategies$curvature,
  classification          = ifelse(singular_strategies$robust,
                                   singular_strategies$class,
                                   paste0(singular_strategies$class, " (stepped)")),
  check.names = FALSE, row.names = NULL))
```

```{r}
#| label: fig-attractor
#| fig-width: 8
#| fig-height: 7.5
#| fig-cap: "(top) Selection gradient vs resident LMA; a zero crossing with negative slope is an attractor (PPA's spikes clipped, |gradient| up to ~4000). (bottom) Where evolution settles — the methods disagree, a ~15% spread. Filled = robust; open = stepped (PPA)."
shape_vals <- c(`TRUE` = 19, `FALSE` = 1)

p_grad <- ggplot(attractor_gradients, aes(lma, gradient, colour = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.9) +
  geom_point(data = singular_strategies, aes(lma_star, 0, shape = robust),
             size = 3) +
  scale_shape_manual(values = shape_vals, guide = "none") +
  coord_cartesian(ylim = c(-400, 400)) +     # PPA's spikes reach ~4000; clip them
  labs(x = "Resident LMA", y = "Selection gradient,  d log(R0) / d(LMA)",
       colour = "Shading model",
       subtitle = "Zero crossing with negative slope = attractor (PPA clipped: |gradient| up to ~4000)") +
  theme_classic()

p_star <- ggplot(singular_strategies, aes(method, lma_star, colour = method)) +
  geom_point(aes(shape = robust), size = 3.5, show.legend = FALSE) +
  geom_text(aes(label = formatC(lma_star, format = "f", digits = 4)),
            vjust = -1.2, size = 3, show.legend = FALSE) +
  scale_shape_manual(values = shape_vals, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.15))) +
  labs(x = NULL, y = "Attracted LMA  (singular strategy)",
       subtitle = "Where evolution settles: the methods disagree (~15% spread)") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))

(p_grad / p_star) + plot_layout(heights = c(1.2, 1))
```

**Every method has a genuine endpoint — but they are different
endpoints.** All five gradients cross zero once, each convergence-stable
and uninvadable (a CSS), yet the attracted LMA marches across the methods:
deep-crown `0.1007`, mean-light `0.0976`, crown-centre `0.0924`,
flat-top-soft-box `0.0901`, ppa `0.0859` — a ~15% spread. **That
disagreement is invisible at the rungs below**: deep-crown, mean-light and
crown-centre agreed on stand dynamics and gave fitness landscapes of the
same *shape*, yet settle on different LMA, because each supports a
different equilibrium *density* (seed rains 1.66 / 2.18 / 2.18 / 3.10 /
18.8), so the realised light environment a mutant grows through differs.
Smoothness is necessary for evolutionary analysis; it is not sufficient
for the answer to be right. **PPA reaches an endpoint, but a fragile
one** — buried in the staircase of its discrete layers, with curvature an
order of magnitude steeper and a location set by where the layer
boundaries fall.

## "Forced to run" is not "valid": the PPA grid artifact

The adaptive solver *refuses* the hard-step variants (discontinuous light
profile → "Cannot achieve the desired accuracy"). A fixed-step
forward-Euler integrator (`control$fixed_time_step`) will instead march
straight across the discontinuity and return numbers — but those numbers
are an artefact of where the step grid falls, not a converged solution.
The paper's `make_figure_gridartifact.R` makes this concrete: a single
resident's seed output **converges** under step refinement for the
smoothed profile but **does not** for the hard step.

This is the heaviest analysis in the paper. The two slow sweeps (gradient
over LMA at three step settings; seed output vs step size) are cached by
the paper script; we transcribe the convergence data here and plot it.
The gradient sweep is left as `eval: false` because re-running it under
fixed-step Euler at fine resolution is slow and the result is the cached
CSV the paper ships.

```{r}
#| label: gridartifact-machinery
#| eval: false
# Transcribed from make_figure_gridartifact.R. Uses the community /
# low-density invasion-fitness method (resident at demographic equilibrium
# + two mutants at tiny birth rate, one multi-species run_scm), which needs
# no RK45 mutant cache and so runs under fixed_time_step.
p_fit_grid <- scm_base_parameters("FF16"); p_fit_grid$max_patch_lifetime <- 50

mk_ctrl <- function(dt, smoothing) {
  ctrl <- Control(); ctrl$shading_model <- "ppa"
  ctrl$ppa_layer_smoothing <- smoothing; ctrl$fixed_time_step <- dt
  ctrl
}
equilibrium_br <- function(ctrl, lma, n_iter = 6, br = 1) {
  for (i in seq_len(n_iter)) {
    res <- expand_parameters(trait_matrix(lma, "lma"), p_fit_grid, birth_rate_list = br)
    br  <- run_scm(res, ctrl = ctrl)$offspring_production
  }
  br
}
# community / low-density selection gradient d log(R0_mutant)/d(LMA)
grad_community <- function(dt, smoothing, lma, delta = 0.0015, eps = 1e-4) {
  ctrl <- mk_ctrl(dt, smoothing)
  br   <- equilibrium_br(ctrl, lma)
  comm <- expand_parameters(
    trait_matrix(c(lma, lma - delta, lma + delta), "lma"), p_fit_grid,
    birth_rate_list = c(br, eps, eps))
  op <- run_scm(comm, ctrl = ctrl)$offspring_production
  f  <- log(op[2:3] / eps)
  (f[2] - f[1]) / (2 * delta)
}

# Convergence of a single resident's seed output as the fixed step shrinks.
dts <- 1 / c(46, 91, 182, 365, 730, 1460)
op_at <- function(dt, smoothing) {
  res <- expand_parameters(trait_matrix(0.1, "lma"), p_fit_grid, birth_rate_list = 1)
  tryCatch(run_scm(res, ctrl = mk_ctrl(dt, smoothing))$offspring_production,
           error = function(e) NA_real_)
}
conv_df <- dplyr::bind_rows(lapply(c(0, 0.3), function(sm) {
  data.frame(smoothing = sm, steps_per_year = 1 / dts,
             offspring = vapply(dts, op_at, numeric(1), smoothing = sm))
}))
```

```{r}
#| label: fig-gridartifact
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Resident seed output vs fixed-step refinement (the paper's cached convergence data). The smoothed PPA converges; the hard step bounces ~10x and does not. Note each panel's own y-scale."
# Cached convergence output from make_figure_gridartifact.R
# (figures/_gridartifact_convergence.csv).
conv_df <- read.csv(strip.white = TRUE, text = "smoothing,steps_per_year,offspring
0,46,0.00867498
0,91,0.01378317
0,182,0.00676893
0,365,0.00949590
0,730,0.03717335
0,1460,0.06593600
0.3,46,25.41085116
0.3,91,25.35579110
0.3,182,25.32100352
0.3,365,25.29680840
0.3,730,25.28289274
0.3,1460,25.27483319")
conv_df$variant <- ifelse(conv_df$smoothing == 0,
                          "PPA hard step (s=0)", "PPA smoothed (s=0.3)")

ggplot(conv_df, aes(steps_per_year, offspring, colour = variant)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  scale_x_log10() +
  facet_wrap(~variant, ncol = 2, scales = "free_y") +
  scale_colour_manual(values = c("PPA hard step (s=0)" = "#0072B2",
                                 "PPA smoothed (s=0.3)" = "#D55E00"),
                      guide = "none") +
  labs(x = "Steps per year  (1 / fixed_time_step, log scale)",
       y = "Resident seed output") +
  theme_classic()
```

"Does not run" understates it: even when *forced* to run under fixed-step
Euler, the PPA hard step has no convergent evolutionary endpoint — its
seed output depends on the arbitrary step grid. The smoothed variant (the
runnable, reported version) is well-behaved.

## Re-running the paper's own figures

There are two re-runs worth distinguishing:

1. **Paper code, today** — the figure scripts from
   `vignettes/examples/DunFalster2026/`, run against current
   dependencies. Tells you whether the paper's own code still executes
   cleanly.
2. **Current `plant`** — the analysis above, transcribed from the source
   vignette and run through today's API (the rendered figures on this
   page). Tells you whether the *model* still produces the published
   behaviour.

Divergence between (1) and (2) is the model-drift signal; divergence
between a published panel and (1) means the paper code itself no longer
reproduces against current deps.

The paper's scripts would be vendored as a paper-code submodule at
`paper-code/dunfalster2026-canopy/`, pinned to the publication commit
(`code-ref` in the front matter — `TODO`, not yet tagged). They write
publication-grade panels (and the cached CSVs this post reads):

- `make_figures.R` — the six-model consequence ladder, the
  competition-profile / soft-box mechanism figure, and the supplementary
  size-distribution and vertical-light-profile panels.
- `make_figure_attractor.R` — the evolutionary-attractor gradient grid
  and singular-strategy table reproduced above.
- `make_figure_gridartifact.R` — the PPA grid-artifact convergence study.

```{r}
#| label: paper-code-rerun
#| eval: false
#| fig-cap: "Paper's own figure scripts, re-run. See code-ref in front matter for the (pending) publication commit."
# Submodule checked out at paper-code/dunfalster2026-canopy/ (CI: git submodule update --init).
# These scripts hardcode absolute output paths and a local devtools::load_all()
# of the plant source tree; adapt the header before running here.
# source(here::here("paper-code/dunfalster2026-canopy", "make_figures.R"))
# source(here::here("paper-code/dunfalster2026-canopy", "make_figure_attractor.R"))
# source(here::here("paper-code/dunfalster2026-canopy", "make_figure_gridartifact.R"))
```

## Side by side

::: {.repro-compare}
::: {.repro-pane}
**Published** — TODO (figure numbers not finalised; paper in prep)

> Add the published panels at `assets/papers/dunfalster2026-canopy.png`
> once the paper exists, or link them at the DOI (also TODO).
:::
::: {.repro-pane}
**Reproduced** — current `plant` (develop)

> The figures rendered above.
:::
:::

## Verdict

`fidelity: pending`. There is no published figure to compare against yet
(the paper is in preparation), so a matches / differs / approximate
verdict cannot be set. What this post *does* establish is that the
analysis runs end-to-end on develop-branch `plant`
(`81cfeab`) and reproduces the paper's headline structure: the
order-of-magnitude spread in equilibrium seed rain, the smooth-vs-jagged
split in invasion fitness, the ~15% disagreement in attracted LMA across
methods, and the non-convergence of the PPA hard step under step
refinement. Re-run as a regression check once the paper is submitted and
its figures and commit can be pinned; then update `fidelity:`.

```{r}
#| label: session
plant_session_info()
```
