---
title: "Hydraulics inner-loop benchmark"
description: "Timing the new plant hydraulics path against the FF16 baseline across a leaf-area-index sweep — an expensive run, pinned to its own environment."
date: 2026-06-10
plant-ref: "develop"
plant-sha: "a1b2c3d4e5f6a7b8"
categories: [hydraulics, performance, develop]
reading-time: true
execute:
  cache: true
---

```{r setup, include=FALSE}
source(here::here("R/version-badge.R"))
```

`r plant_version_badge()`

This post is built against an **unreleased** commit on `develop`, so the
badge pins to a SHA rather than a version. The simulation below is slow,
so it's pinned to its own `renv` snapshot as well — a later project-wide
package upgrade can't change this result even on a forced re-run.

## Per-post environment

This directory carries its own lockfile, captured once when the post was
written:

```r
# run from this post's directory, once:
renv::snapshot(lockfile = "renv.lock.dev", prompt = FALSE)
```

`plant_session_info()` (footer) reports `renv.lock.dev` when present, so
the page records exactly which environment produced the figures. To
reproduce: check out `plant` at the SHA above, `renv::restore(lockfile =
"renv.lock.dev")`, delete this post's `_freeze/` entry, re-render.

## The sweep

```{r}
#| label: benchmark
#| eval: false
#| cache: true
library(plant)
library(bench)

lai_grid <- seq(1, 8, by = 0.5)

results <- lapply(lai_grid, function(L) {
  bench::mark(
    baseline  = run_inner_ff16(lai = L),
    hydraulic = run_inner_hydraulics(lai = L),
    iterations = 20, check = FALSE
  )
})
```

```{r}
#| label: plot
#| eval: false
#| fig-cap: "Median inner-loop time vs LAI. Hydraulics adds a roughly constant overhead until canopy closure, where the extra solve starts to dominate."
# ... ggplot of median time by method across lai_grid ...
```

::: {.callout-note}
Numbers here are frozen to commit `a1b2c3d`. When this work lands on
`master`, supersede this post with a release-pinned version rather than
editing it — keeping the develop snapshot as the historical record.
:::

## Reproducibility footer

```{r}
#| label: session
#| eval: false
plant_session_info()
```
