---
title: "Community assembly with the toy models"
subtitle: "Beyond pairwise invasibility: running the assembly algorithm"
format: html
---

A pairwise invasibility plot (PIP) shows whether a *single* mutant can invade a
*single* resident. The toy models in `regnans` also support **mixtures**, so we
can run the full community-assembly algorithm — repeatedly introduce the
best-invading mutant, re-solve the community to demographic equilibrium, and drop
any residents driven extinct — and watch a multi-species community build up by
evolutionary branching and limiting similarity.

```{r setup}
#| message: false
library(regnans)
```

## A note on trait scale

The assembly machinery spaces and searches trait space on a configurable scale.
Strictly positive biological traits (plant's `lma`, or GM99 seed size) suit a
**log** scale; traits whose optimum sits at 0 (DD99, GK98) need a **linear**
scale. Set it with `community_start(trait_scale = ...)` — `"log"` is the default.

## DD99 — limiting similarity in one dimension

With competition narrower than the resource (`sigma_C < sigma_K`) the resource
optimum is a branching point, and assembly packs trait space with coexisting
species roughly one competition-width apart.

```{r dd99-1d}
#| fig-width: 6
#| fig-height: 3.4
h <- harness_dd99(x0 = 0, sigma_K = 1, sigma_C = 0.4)
comm <- community_start(bounds(x = c(-3, 3)), harness = h, trait_scale = "linear",
                        fitness_control = list(method = "grid", n_evals = 150))
a <- assembler_start(comm, assembler_control(list(birth_type = "maximum")))
a <- assembler_run(a, 14)

cat(length(a$community), "coexisting species\n")
xs <- sort(as.numeric(a$community$traits))
plot(xs, rep(1, length(xs)), pch = 19, col = "forestgreen", yaxt = "n",
     ylab = "", xlab = "trait x", xlim = c(-3, 3),
     main = sprintf("DD99 assembly: %d species (limiting similarity)", length(xs)))
```

## GK98 — branching to a protected dimorphism

The soft-selection model conserves total population, so there is no "empty
habitat" to colonise — we **seed** the first type, then let it evolve. At
`d/sigma = 1.5` (above the branching threshold √(3/2)) the single type splits
into a stable dimorphism.

```{r gk98}
comm <- community_start(bounds(x = c(-4, 4)), harness = harness_gk98(d = 1.5, sigma = 1),
                        trait_scale = "linear",
                        fitness_control = list(method = "grid", n_evals = 150))
a <- assembler_start(comm, assembler_control(list(birth_type = "maximum")))
a <- assembler_set_traits(a, trait_matrix(0, "x"))   # seed the first type
a <- assembler_run(a, 10)

sort(round(as.numeric(a$community$traits), 3))        # a symmetric dimorphism
```

## GM99 — a seed-size polymorphism

Strong size-asymmetric competition (`alpha*R = 7`) drives branching in seed size.
Assembly from empty produces a polymorphism of small and large seeds. (GM99's
multi-resident fitness is a multi-dimensional Poisson sum, so it is the slowest
of the toy models — a few species takes seconds, not milliseconds.)

```{r gm99}
#| eval: false
comm <- community_start(bounds(x = c(0.06, 0.95)), harness = harness_gm99(alpha = 7, beta = 15),
                        fitness_control = list(method = "grid", n_evals = 80))
a <- assembler_start(comm, assembler_control(list(birth_type = "maximum")))
a <- assembler_run(a, 4)
sort(round(as.numeric(a$community$traits), 3))   # e.g. 0.11, 0.45, 0.69 — small, medium, large seeds
```

## DD99 in two trait dimensions

`harness_dd99_nd()` is the multi-trait extension (cf. Ito & Dieckmann 2007): the
resource and competition kernels are products of per-dimension Gaussians, so each
trait has its own optimum and widths. The singular point is `x* = x0` and a
dimension branches iff its `sigma_C < sigma_K`.

The same `assembler_run()` drives 2D assembly — the only change from 1D is a
two-trait `bounds()` and the multi-trait harness:

```{r dd99-2d}
#| fig-width: 5
#| fig-height: 5
h2 <- harness_dd99_nd(x0 = c(0, 0), sigma_K = c(1, 1), sigma_C = c(0.5, 0.5))
comm <- community_start(bounds(x1 = c(-3, 3), x2 = c(-3, 3)), harness = h2,
                        trait_scale = "linear",
                        fitness_control = list(method = "grid", n_evals = 40))
a <- assembler_start(comm, assembler_control(list(birth_type = "maximum")))
a <- assembler_run(a, 40)

cat(length(a$community), "coexisting species in 2D\n")
plot(a$community$traits[, 1], a$community$traits[, 2], pch = 19, col = "forestgreen",
     xlab = "trait x1", ylab = "trait x2", xlim = c(-3, 3), ylim = c(-3, 3),
     main = sprintf("DD99 2D assembly: %d species", length(a$community)))
```

The species pack the trait plane at a characteristic spacing set by the
competition widths — limiting similarity in two dimensions.
