---
title: "GM99 — Geritz, van der Meijden & Metz 1999"
subtitle: "Seed size and seedling competitive ability in safe sites"
format: html
---

## The original paper

Geritz, S.A.H., van der Meijden, E. & Metz, J.A.J. (1999). *Evolutionary Dynamics
of Seed Size and Seedling Competitive Ability.* **Theoretical Population Biology**
55, 324–343.

Plants compete for discrete *safe sites*; seeds arrive at each site by a Poisson
process, and seedlings within a site compete for the site's resources `R` in a
**size-asymmetric lottery** — larger seeds are better competitors. There is a
trade-off: a plant of seed size `x` makes fewer but more competitive (and more
reliably establishing) seeds. The paper shows that strong enough competitive
asymmetry turns the seed-size optimum into a **branching point**, so a community
evolves a polymorphism of small "many cheap seeds" and large "few strong seeds"
strategies. This is the seed-size model implemented in Daniel's MATLAB; it is
distinct from the [GK98](GK98.qmd) soft-selection model.

## Fitness equations

Pre-competitive survival and competitive ability:

$$
s(x) = \max\!\left(0,\; 1 - 2e^{-\beta x}\right),
\qquad
c(x) = e^{\alpha x}.
$$

A focal mutant `x'` lands in a site with $k_j \sim \text{Poisson}(N_j)$ seedlings
of each resident `x_j`; its expected share of the site, averaged over the Poisson
draws, is

$$
g = \mathbb{E}\!\left[\frac{c(x')}{c(x') + \sum_j k_j\,c(x_j)}\right].
$$

Invasion fitness is the mutant's expected number of established offspring (a
lifetime reproductive ratio, **= 1**, i.e. log = 0, at equilibrium):

$$
W(x') = \frac{R}{x'}\, s(x')\, g .
$$

A single resident sits at the density `N*` solving
$\tfrac{R}{x}s(x)\,\tfrac{1-e^{-N}}{N} = 1$. Only the products $\alpha R$ and
$\beta R$ matter. There is **no closed-form singular strategy** (the paper solves
it numerically); branching is driven by competitive asymmetry $\alpha$.

## Example analysis with `regnans`

```{r setup}
#| message: false
library(regnans)
```

The paper's Fig. 5 (bottom row, $\beta R = 15$): low asymmetry $\alpha R = 4.5$
gives a single ESS; high asymmetry $\alpha R = 7$ gives a branching point.

```{r singular}
solve_x <- function(alpha, beta = 15, bounds_x = c(0.06, 0.95)) {
  community_start(bounds(x = bounds_x),
                  harness = harness_gm99(alpha = alpha, beta = beta)) |>
    community_solve_singularity_1D()
}

css    <- solve_x(4.5)   # CSS
branch <- solve_x(7.0)   # branching point
c(css = as.numeric(css$traits), branch = as.numeric(branch$traits))
```

Classify each via the curvature of invasion fitness at the singular strategy
(negative = ESS/CSS, positive = branching):

```{r curvature}
curv <- function(comm, at, d = 1e-4) {
  f <- comm$fitness_function
  (f(at + d) - 2 * f(at) + f(at - d)) / d^2
}
c(css = curv(css, as.numeric(css$traits)),
  branch = curv(branch, as.numeric(branch$traits)))
```

The strong analytic check: a resident at its own equilibrium has lifetime
reproductive ratio exactly 1 (log fitness 0):

```{r invariant}
comm <- community_start(bounds(x = c(0.06, 0.95)),
                        harness = harness_gm99(alpha = 7, beta = 15)) |>
  community_add(trait_matrix(0.645, "x")) |>
  community_demography()
comm$resident_fitness          # ≈ 0
```

### Pairwise invasibility plot (reproducing Fig. 5, lower right)

```{r pip}
#| fig-width: 5
#| fig-height: 5
h  <- harness_gm99(alpha = 7, beta = 15)
xs <- seq(0.06, 0.95, length.out = 81)
S <- sapply(xs, function(res) {
  community_start(bounds(x = range(xs)), harness = h) |>
    community_add(trait_matrix(res, "x")) |>
    community_demography() |>
    (\(comm) comm$fitness_function(xs))()
})
image(xs, xs, t(sign(S)), col = c("white", "grey85", "forestgreen"),
      xlab = "resident seed size", ylab = "mutant seed size",
      main = "GM99 PIP (alpha*R = 7, beta*R = 15)")
abline(0, 1, lty = 2)
```
