---
# ── REPRODUCTION TEMPLATE ──────────────────────────────────────────
# Copy this directory, rename to posts/YYYY-MM-DD-<slug>/, fill in.
# Every field marked TODO must be set from the actual paper — do not
# guess parameter values or figure numbers.
title: "TODO: Reproducing Fig N of <Author Year>"
description: "TODO: one line on what the figure shows and why it matters."
date: 2026-01-01
# Pin to whatever plant state you ran this against:
plant-version: "2.1.0"        # master/released  …OR delete and use:
# plant-ref: "develop"
# plant-sha: "TODO"
categories: [reproduction, TODO-topic]
reading-time: true
# Structured metadata about the source paper (drives the citation box):
paper:
  authors: "TODO"
  year: TODO
  title: "TODO"
  journal: "TODO"
  doi: "TODO"
  figure: "TODO (e.g. Figure 3b)"
  fidelity: "pending"          # pending | matches | differs | approximate
  # ── original paper code (you have these repos) ───────────────────
  # If set, the post can re-run the paper's OWN figure script rather
  # than a hand-transcription. Pin to the publication commit/tag.
  code-repo: "TODO"            # e.g. traitecoevo/<paper-repo> (or "" if none)
  code-ref: "TODO"            # tag or full SHA used for publication
  code-entrypoint: "TODO"      # path to the figure script within that repo
---

```{r setup, include=FALSE}
source(here::here("R/version-badge.R"))
source(here::here("R/reproduction.R"))
```

`r plant_version_badge()` `r repro_fidelity_badge()` `r paper_code_badge()`

`r paper_citation()`

## What the original shows

TODO: 2–3 sentences describing the published figure — the axes, the
result, and why it's a load-bearing result for the model.

## Parameters from the paper

The values below are taken directly from <Author Year> (and its
supplement). Keep them in one visible block so the reproduction is
auditable against the source.

```{r}
#| label: params
#| eval: false
# TODO: transcribe exactly. Cite table/section for each non-default value.
pars <- list(
  # lma   = 0.0825,   # <Author Year>, Table S1
  # rho   = 608,      # ...
)
```

## Re-running it

There are two re-runs worth distinguishing:

1. **Paper code, today** — the figure script from the original repo
   (pinned via `code-ref`), run against current dependencies. Tells you
   whether the *paper's own code* still executes cleanly.
2. **Current `plant`** — the same result through today's model API. Tells
   you whether the *model* still produces the published behaviour.

Divergence between (1) and (2) is the model-drift signal. Divergence
between the published PNG and (1) means the paper code itself no longer
reproduces against current deps.

```{r}
#| label: paper-code-rerun
#| eval: false
#| fig-cap: "Paper's own figure script, re-run. See code-ref in front matter for the pinned commit."
# Submodule checked out at paper-code/<slug>/ (CI: git submodule update --init).
# source(here::here("paper-code/<slug>", "<code-entrypoint>"))
```

```{r}
#| label: reproduce
#| eval: false
#| fig-cap: "Current plant. Compare against TODO-figure of the source."
library(plant)
# TODO: the minimal current-API path that regenerates the figure.
```

## Side by side

::: {.repro-compare}
::: {.repro-pane}
**Published** — TODO-figure of <Author Year>

> Insert the published panel as an image (assets/papers/<slug>.png) or,
> if licensing doesn't permit, link to it at the DOI and describe it.
:::
::: {.repro-pane}
**Reproduced** — current `plant`

> The figure rendered above.
:::
:::

## Verdict

TODO: state plainly whether it **matches**, **differs**, or is
**approximate**, and why. Differences are often informative — a changed
default, a reparameterisation, a bug fixed since publication. Set the
`fidelity:` field in the front matter to match this verdict.

```{r}
#| label: session
#| eval: false
plant_session_info()
```
