---
title: "Emergent properties"
---
::: {.eyebrow}
example
:::

An **emergent property** belongs to the whole patch rather than to any one
plant. It arises from many individuals growing, recruiting, dying, and changing
one another's light environment. This example moves from node-density profiles
to leaf area and growth, then averages patch profiles across a metapopulation.

::: {.callout-note}
A few steps below reach into plant's internal helpers (`plant:::trapezium`,
`plant:::pad_list_to_array`). These are not part of the public API and may
change or disappear between releases. They are kept here exactly as written
because the example depends on them; treat them as implementation detail
rather than something to build on.
:::

We'll work with a single species and an illustrative constant birth rate:

```{r}
#| label: setup-patch
library(plant)

p0 <- scm_base_parameters("FF16")
p <- add_strategies(p0, trait_matrix(0.0825, "lma"), birth_rate = 17.31)
patch <- run_scm(p, refine_schedule = TRUE)$parameters
result <- run_scm(patch, collect = TRUE)
```

```{r}
#| include: false
#| label: sdd-matrix-helper
library(dplyr)
library(tidyr)

# The tidy `species` table returned by `run_scm(collect = TRUE)` has one row
# per node per step. This helper reshapes a single variable into a matrix with
# one row per node and one column per step, padding with `NA` for nodes that
# had not yet been introduced.
sdd_matrix <- function(species_data, var) {
  m <- species_data |>
    dplyr::select(node, step, dplyr::all_of(var)) |>
    tidyr::pivot_wider(names_from = step, values_from = dplyr::all_of(var)) |>
    dplyr::arrange(node)
  as.matrix(m[, -1])
}

```

## Node-density profiles

Patch profiles change continuously, so plotting every recorded age at equal
emphasis is difficult to read. We retain all trajectories in grey and highlight
five representative ages:

```{r}
#| label: highlight-times
closest <- function(t, time) {
  which.min(abs(time - t))
}
last <- function(x) {
  x[[length(x)]]
}

times <- c(5, 10, 20, 40, dplyr::last(result$steps$time))
i <- vapply(times, closest, integer(1), result$steps$time)
blues <- c("#DEEBF7", "#C6DBEF", "#9ECAE1", "#6BAED6",
           "#4292C6", "#2171B5", "#08519C", "#08306B")
cols <- colorRampPalette(blues[-(1:2)])(length(i))

species_1   <- dplyr::filter(result$species, species == "1")
height      <- sdd_matrix(species_1, "height")
log_density <- sdd_matrix(species_1, "log_density")
```

The helper arranges both height and density so rows are nodes and columns are
recorded times. Each column can therefore be read as one patch-age profile:

```{r}
#| label: plot-density-profiles
density <- exp(log_density)
matplot(height, density, type="l", lty=1,
        col=util_colour_set_opacity("black", 0.15),
        xlab="Height (m)", ylab="Density (1 / m / m2)", las=1,
        log="y")
```

Some nodes reach extremely small mathematical densities while their
representative individuals continue to grow. This is a continuous-density
model, so a value below one does not mean that a literal fraction of a plant is
present in a particular physical patch. The highlighted ages make the
consequences easier to see.

```{r}
#| label: plot-density-highlighted
xlim <- c(0, max(height, na.rm=TRUE) * 1.05)
matplot(height, density, type="l", lty=1,
        col=util_colour_set_opacity("black", 0.15),
        xlab="Height (m)", ylab="Density (1 / m / m2)", las=1,
        log="y", xlim=xlim)
matlines(height[, i], density[, i], col=cols, lty=1, type="l")
points(height[1, i], density[1, i], pch=19, col=cols)
text(height[1, i] + strwidth("x"), density[1, i],
     paste0(round(times), c(" years", rep("", length(times) - 1))),
     adj=c(0, 0))
```

At five years, self-thinning creates a dip in density. As the stand develops,
that dip becomes deeper, broader, and shifts towards taller plants because the
low-density nodes still follow the model's
[characteristics](../theory/implementation/characteristic-method.qmd).

A later wave of recruitment creates another density peak near 4 m at 20 years;
by 40 years that cohort has reached roughly 13 m. In old patches, the overall
size-density profile changes more slowly, although a narrow gap remains just
below the canopy top.

## Leaf-area profiles

Which heights contribute most of the patch's leaf area? The collected output
records individual leaf area as `competition_effect`. Multiplying it by node
density gives leaf area density. The canopy model later applies the light
extinction coefficient `k_I` when computing competition; see
[Canopy and light environments](../theory/implementation/canopy-models.qmd).

```{r}
#| label: plot-leaf-area
species_1 <- species_1 |>
  dplyr::mutate(leaf_area = density * competition_effect)
leaf_area <- sdd_matrix(species_1, "leaf_area")

matplot(height, leaf_area, type="l", lty=1, col="lightgrey",
        xlim=xlim, xlab="Height (m)",
        ylab="Leaf area density (m2 / m2 / m)", las=1)
matlines(height[, i], leaf_area[, i], col=cols, lty=1, type="l")
points(height[1, i], leaf_area[1, i], pch=19, col=cols)
text(height[1, i] + strwidth("x"), leaf_area[1, i],
     paste0(round(times), c(" years", rep("", length(times) - 1))),
     adj=c(0, 0))
```

## Height-growth profiles

Where in the size distribution is height growth concentrated? Unlike the fixed
environment in [Your first individual](first-individual.qmd), this patch has a
shared light environment created by all its plants.

The collected output records states rather than instantaneous rates. We
therefore approximate height growth from the change in each node's height
between successive recorded times:

```{r}
#| label: plot-growth-rate
time <- result$steps$time
growth_rate <- t(apply(height, 1, function(hh) c(NA, diff(hh) / diff(time))))

matplot(height, growth_rate, type="l", lty=1, col="lightgrey",
        xlim=xlim, xlab="Height (m)",
        ylab="Height growth rate (m / year)", las=1)
matlines(height[, i], growth_rate[, i], col=cols, lty=1, type="l")
points(height[1, i], growth_rate[1, i], pch=19, col=cols)
text(height[1, i] + strwidth("x"), growth_rate[1, i],
     paste0(round(times), c(" years", rep("", length(times) - 1))),
     adj=c(0, 0))
```

## Average across a metapopulation

The earlier curves each describe a patch of one age. A metapopulation contains
patches of many ages, so its average profile weights every patch age by its
abundance. Before averaging, we must interpolate each patch profile onto the
same set of heights. See the
[size-structured PDE](../theory/size-structured-pde.qmd) for the corresponding
integral.

```{r}
#| label: fixed-heights
hh <- seq_log_range(range(height, na.rm=TRUE), 500)
```

The next block is reusable numerical scaffolding; readers interested mainly in
the ecological result can skip to the resulting average. It interpolates on a
log-log scale and returns zero outside the observed height range.

```{r}
#| label: interpolation-helpers
##' Spline interpolation in log-x space
##' @title Spline interpolation in log-x space
##' @param x,y Vectors giving coordinates of points to be
##' interpolated.  The x points should be naturally on a log scale,
##' and for \code{splinefun_loglog} both x and y should be on a log
##' scale.
##' @param ... Additional parameters passed to
##' @rdname splinefun_log
splinefun_loglog <- function(x, y, ...) {
        f <- splinefun(log(x), log(y), ...)
        function(x) {
                exp(f(log(x)))
        }
}

##' Clamp a function to a fixed value outside a domain
##' @param f A function that takes \code{x} as a first argument.
##' @param r Range of values (vector of length 2)
##' @param value (Single) value to use when out of domain.
##' @return A new function
##' @export
clamp_domain <- function(f, r, value = NA_real_) {
        f <- match.fun(f)
        if (length(r) != 2L) {
                stop("Expected length two range")
        }
        if (any(is.na(r)) || r[[2]] < r[[1]]) {
                stop("Values for range must be finite and not decreasing")
        }
        if (length(value) != 1L) {
                stop("value must be length 1")
        }
        function(x, ...) {
                ret <- rep_len(value, length(x))
                i <- x >= r[[1]] & x <= r[[2]]
                if (any(i)) {
                        ret[i] <- f(x[i])
                }
                ret
        }
}

f <- function(height, density, hout) {
  r <- range(height, na.rm=TRUE)
  clamp_domain(splinefun_loglog(height, density), r, 0)(hout)
}
```

Interpolate every patch-age profile onto the shared height grid:

```{r}
#| label: interpolate-patches
xx <- lapply(seq_along(result$steps$time),
             function(i) f(height[, i], density[, i], hh))
n_hh <- plant:::pad_list_to_array(xx)
```

At each height, integrate across patch ages while weighting by patch abundance:

```{r}
#| label: integrate-metapopulation
trapezium <- plant:::trapezium
n_av <- apply(n_hh, 1,
              function(x) trapezium(result$steps$time, x * result$steps$patch_density))
```

The red line is the resulting metapopulation-average density profile:

```{r}
#| label: plot-metapopulation-average
xlim <- c(0, max(height, na.rm=TRUE) * 1.05)
matplot(height, density, type="l", lty=1,
        col=util_colour_set_opacity("black", 0.15),
        xlab="Height (m)", ylab="Density (1 / m / m2)", las=1,
        log="y", xlim=xlim)
matlines(height[, i], density[, i], col=cols, lty=1, type="l")
points(height[1, i], density[1, i], pch=19, col=cols)
text(height[1, i] + strwidth("x"), density[1, i],
     paste0(round(times), c(" years", rep("", length(times) - 1))),
     adj=c(0, 0))
points(hh, n_av, col="red", type='l')
```
