---
title: "Example analysis"
---
::: {.eyebrow}
example
:::

This worked example follows a complete analysis from model setup to figures.
We will create two TF24 strategies, run them together in one patch, turn the
size distributions into patch totals, and interpolate the output to requested
times and heights.

::: {.callout-note}
## Customising the `TF24` soil-water environment
The [`TF24`](../theory/models/tf24.qmd) model couples plant growth to a
multi-layer soil-water balance. `Environment("TF24")` already provides five soil
layers, an initial moisture state, and constant rainfall. The helper below shows
how to replace those defaults with 15 layers and seasonal rainfall.

<!-- SY: I think a lot of the Guides sections uses jargon-y language and style characteristic of `plant` and modelling, which might be daunting or unhelpful to non-modellers and beginners. Perhaps there could be more explicit discernment between the biology and the model. Maybe the Guides are also a bit hard to grasp unless someone has read the Theory section. It would help if Overstorey had an even broader overview page about why and what we use `plant` for, accessible to modellers and non-modellers alike. 

I think a lot of underlying concepts might get missed by the non-modeller. For example, the simple sentence "The TF24 model couples plant growth to a multi-layer soil-water balance." can be easily understood by a plant modeller, but to a non-modeller, I wonder if they would benefit from something more explicit, even if it's wordier. "Couples" and "balance" feel a bit jargon-y to me. I like something closer to this: "The TF24 model tracks soil water at several depths. Plants draw on that water to grow, and as they use it, the soil dries out, which then limits further growth." -->

TF24's hydraulic calculations make long runs expensive, so this example stops
at a patch age of 25 years. Choose a lifetime appropriate to your scientific
question for a production analysis.
:::

## Set up the analysis

```{r}
#| label: load-libraries
library(plant)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```

Console logging is optional, but it makes progress visible during the slower
TF24 run:

```{r}
#| label: console-logging
plant_log_console()
```

Define the shortened patch lifetime and a function that constructs the same
custom environment every time the model is run:

```{r}
#| label: tf24-env-helper
max_patch_lifetime <- 25

make_tf24_env <- function(max_patch_lifetime, n_layers = 15, theta0 = 0.2) {
  env <- Environment("TF24")
  env$set_soil_number_of_depths(n_layers)
  env$set_soil_water_state(rep(theta0, n_layers))

  # seasonal rainfall around a mean of 1.0
  x <- seq(0, max_patch_lifetime, length.out = 1000)
  y <- 0.25 * sin(2 * pi * x) + 1.0
  env$extrinsic_drivers_set_variable("rainfall", x = x, y = y)

  env
}
```

### Run the smallest version

A minimal run needs base parameters, at least one strategy, and an environment.
The first call refines the node-introduction schedule; the second runs the patch
and collects its state:
<!-- SY: Maybe should explain more why two runs are needed? Are they always necessary? If we're interested in two strategies why are we doing a minimal run and refining the node-introduction schedule with one strategy? I thought there was an argument to refine the schedule in `run_scm`? -->

```{r}
#| label: minimal-run
p0 <- scm_base_parameters("TF24", "TF24_Env")
p0$max_patch_lifetime <- max_patch_lifetime

traits <- trait_matrix(c(0.07), c("lma"))

p1 <- add_strategies(p0, traits)

env <- make_tf24_env(max_patch_lifetime)

results <- run_scm(p1, env = env, refine_schedule = TRUE)$parameters
results <- run_scm(results, env = env, collect = TRUE)
```

For repeated analyses, helper functions keep assumptions in one place. The <!-- SY: What assumptions? -->
first helper starts from TF24 defaults and changes four strategy parameters:

```{r}
#| label: base-parameters-fn
base_parameters <- function() {
  p0 <- scm_base_parameters("TF24", "TF24_Env")
  p0$max_patch_lifetime <- max_patch_lifetime

  p0$strategy_default$pars$hmat <- 15
  p0$strategy_default$pars$rho <- 700
  p0$strategy_default$pars$a_l1 <- 2.17
  p0$strategy_default$pars$a_l2 <- 0.5

  p0
}
```

The run helper below combines three kinds of input:

- biological choices: traits, birth rates, and `B_lf1`;
- location: latitude, used by the TF24 hyperparameterisation; <!-- SY: What does latitude have to do with the model? -->
- a numerical choice: whether to refine the node-introduction schedule.

```{r}
#| label: run-mypatch-fn
run_mypatch <- function(
  traits =  trait_matrix(c(0.07), c("lma")),
  birth_rate = 10,
  B_lf1 = 0.009,
  p0 = base_parameters(),
  optimise_schedule = FALSE,
  latitude = 28.182
  ) {

  hyper_par_fn = make_TF24_hyperpar(B_lf1 = B_lf1, latitude = latitude)
  birth_rate <- rep_len(birth_rate, nrow(traits))
  p1 <- add_strategies(p0, traits, hyper_par_fn, birth_rate = birth_rate)

  env <- make_tf24_env(p0$max_patch_lifetime)

  if(optimise_schedule)
    result <- run_scm(p1, env = env, refine_schedule = TRUE)$parameters
  else
    result <- p1

  # gather outputs at each time step
  run_scm(result, env = env, collect = TRUE)
}
```


### Run two strategies together

Run a patch for two strategies. Here, we provide two LMA values, representing thin
leaves (0.07) and thicker leaves (0.24). To run one species, supply a one-row
trait matrix; `add_strategies()` requires at least one strategy before the patch
can be simulated.

```{r}
#| label: run-two-species
results <-
  run_mypatch(traits=trait_matrix(c(0.07, 0.24), c("lma")))
```

Because `run_mypatch()` calls `run_scm(..., collect = TRUE)`, its result is
already organised for analysis. The `species` table has one row per node at
each recorded step; `steps` and `env` describe patch time and environment.

```{r}
#| label: tidy-results
results_tidy <- results

results_tidy
```

<!-- SY: I think this output is way too long. -->

Each line in the size-distribution plot follows one node through time. New nodes
enter at times chosen by the node schedule, so a recorded step is not the same
as one year. Line transparency represents node density: darker trajectories
contribute more individuals to the patch.

```{r}
#| label: plot-size-distribution
results_tidy$species %>%
  drop_na() %>%
  plot_size_distribution()
```

<!-- SY: Maybe add some interpretation here about what the plot shows, e.g. it stabilises to an overstorey created by species 1 and an understorey around 2 m of species 2. -->

## Calculate patch totals for each strategy

Integration turns a size distribution into one patch-level value per strategy
and time. For example, integrating node density over height gives the total
number of individuals per square metre:

```{r}
#| label: integrate-totals
data_species_tot <-
  results_tidy$species %>% integrate_over_size_distribution()

data_species_tot
```

Plotting that total shows how stem density changes as the patch develops:
<!-- SY: Why stem density, not individual density? -->

```{r}
#| label: plot-total-density
data_species_tot %>%
  ggplot(aes(time, density, colour=species)) +
  geom_line()
```

<!-- SY: Maybe better to label species 1 and 2 with their LMA values? -->

`expand_state()` derives additional quantities, including leaf area and tissue
mass, from the states returned by the solver:

```{r}
#| label: expand-state
results_tidy_expand <- results_tidy %>% expand_state()

results_tidy_expand$species
```

<!-- SY: Difference between patch_density and density? -->

To obtain total leaf area, the helper weights individual leaf area by node
density and integrates over height at each time:

```{r}
#| label: plot-total-leaf-area
data_species_tot <-
  results_tidy_expand$species %>%
  integrate_over_size_distribution()

data_species_tot %>%
  ggplot(aes(time, area_leaf, colour = species)) +
  geom_line()
```


In this run, the thin-leaved strategy initially deploys more total leaf area but
is later overtaken by the thicker-leaved strategy.
<!-- SY: I don't see this being the case? Species 2 always has lower leaf area. -->

The expanded totals also show how above-ground biomass is divided among leaf,
bark, sapwood, and heartwood through time:

```{r}
#| label: plot-biomass-allocation
v <- c("mass_heartwood", "mass_sapwood", "mass_bark", "mass_leaf")

data_long <-
  data_species_tot %>%
  select(time,species, one_of(v)) %>%
  pivot_longer(cols=starts_with("mass"), names_to = "tissue")

data_long$tissue <- factor(data_long$tissue, levels = v)

species_names <- tibble(species = unique(data_long$species),
                      species_name = c("Species_1","Species_2"))

data_long %>%
  left_join(species_names) %>%
  ggplot(aes(time, value, fill=tissue)) +
  geom_area() +
  labs(x = "Patch age (yr)", y = "Above ground mass (kg/m2)") +
  theme_classic() +
  xlim(c(0, 25)) +
  facet_wrap(~species_name)
```

<!-- SY: Maybe explain here that Species 2 has values too small to be plotted? -->

## Interpolate to specific times

The node-introduction schedule sets the times at which new
[characteristics](../theory/implementation/characteristic-method.qmd) enter the
simulation. It can optionally be refined by the
[node-spacing algorithm](../theory/implementation/node-spacing.qmd). The
recorded steps therefore may not fall at the exact times we want to query.
`interpolate_to_times()` estimates the node states at times we are interested in, e.g., 1, 5, and 10 years from an existing run:

```{r}
#| label: interpolate-times
times <- c(1, 5, 10)

tidy_species_data <- results_tidy$species

tidy_species_new <- interpolate_to_times(tidy_species_data, times)

```

The returned rows contain the state variables at those requested times:

```{r}
#| label: interpolated-time-values
tidy_species_new %>% drop_na()
```

To check the interpolation, combine the new rows with the recorded trajectories.
The red points lie on the original curves, showing agreement at the requested
times:

```{r}
#| label: plot-time-interpolation
data_combined <-
  tidy_species_data %>%
  bind_rows(tidy_species_new) %>%
  arrange(species, node, time) %>%
  filter(node %in% seq(1, 101, by=20))

data_combined_long <-
  data_combined  %>%
  select(node, time, step, density, height, species) %>%
  pivot_longer(cols = c("density", "height"))

data_combined_long %>%
  ggplot(aes(time, value, group=node,colour=node)) +
  geom_line() +
  geom_point(data = data_combined_long %>% filter(is.na(step)), col=2) +
  scale_y_log10() +
  xlim(c(0, 20)) +
  facet_grid(name~species, scales="free") +
  theme_classic()
```

## Interpolate to specific heights

The complementary question is: what was the state at a particular height?
`interpolate_to_heights()` estimates values across nodes at each recorded time:

```{r}
#| label: interpolate-heights
heights <- c(1, 5, 10)

tidy_species_data <- results_tidy$species

tidy_species_new <- interpolate_to_heights(tidy_species_data, heights)
```

These rows contain estimates at heights of 1, 5, and 10 metres:

```{r}
#| label: interpolated-height-values
tidy_species_new %>% drop_na()
```

Again, plotting the estimates as red points against the recorded profiles makes
the interpolation easy to assess:

```{r}
#| label: plot-height-interpolation
data_combined <-
  tidy_species_data %>%
  bind_rows(tidy_species_new) %>%
  arrange(species, time, height) %>%
  filter(step %in% unique(round(seq(min(step, na.rm = TRUE),
                                    max(step, na.rm = TRUE), length.out = 2))))

data_combined_long <-
  data_combined  %>%
  select(node, time, step, height, density, mortality, area_heartwood, species) %>%
  pivot_longer(cols = c("density", "mortality", "area_heartwood"))

data_combined_long %>%
    filter(!is.na(node)) %>%
  ggplot() +
  geom_line(aes(height, value, group=step, colour=step)) +
  geom_point(data = data_combined_long %>% filter(is.na(node)), aes(x=height, y=value), col="red")+
  facet_grid(name~species, scales="free") +
  theme_classic()
```
