---
title: "Extrinsic drivers"
---
::: {.eyebrow}
guide
:::

```{r}
library(plant)
```

## Introduction

All models in plant take extrinsic drivers as inputs. At their simplest these are constants. Or they could be variables that change over time. The Extrinsic Drivers class is setup to enable verstaile input.

Extrinsic Drivers are configured as functions over time that are used as inputs in the `plant` solver (SCM). Due to the flexibility of the implementation, the functions can be set up as either constants or time-varying inputs. This document goes through examples of setting up Extrinsic Drivers throughout `plant`.

When handling functions that vary over time, Extrinsic Drivers are created via interpolation. Control points are passed as input, and the
system creates a smooth function that interpolates the control points:

[![Example of polynomial interpolation](../assets/Interpolation_example_polynomial.svg){width="400"}](https://en.wikipedia.org/wiki/Polynomial_interpolation)

Under the hood, every driver is stored under a name (`"rainfall"`, `"ca"`, `"birth_rate"`, ...) as either a constant value or a spline built from `x`/`y` control points — the same two-case `Function` object regardless of which driver it backs. By default a spline will not extrapolate past the range of its control points (evaluating outside that range is an error); this can be relaxed with an extrapolation toggle. The rest of this guide walks through setting drivers of both kinds, starting with `rainfall` and then covering the rest of the set.

## Rainfall

Rainfall is handled inside `TF24_Environment`, and is set upon construction.
Rainfall drives the soil-water balance that the [TF24](../theory/models/tf24.qmd)
strategy couples growth to, via the leaf-level water economy described in
[Assimilation & hydraulics](../theory/assimilation-hydraulics.qmd). If nothing is
passed to the `Environment()` constructor, rainfall will default to a value of
`1`:

```{r}
env <- Environment("TF24")
```

Otherwise, it can either be a constant:

```{r}
env <- Environment("TF24")
env$extrinsic_drivers_set_constant("rainfall", 3.14)
```

... or varying over time, by providing a list with `x` and `y` control points (ie the point pairs are `(x[i], y[i])`):

```{r}
x_pts <- seq(0, 200, 1)
rain <- list(
  x = x_pts,
  y = 1 + sin(x_pts)
)
env <- Environment("TF24")
env$extrinsic_drivers_set_variable("rainfall", rain$x, rain$y)
```

The resulting function of rainfall over time can then be evaluated at any time. Note how the point of evaluation was not one of the control points passed in - we
are able to get rainfall at time `5.634` because of the smooth interpolated function.

```{r}
env$extrinsic_drivers_evaluate("rainfall", 5.634)
```

We can also evaluate the function at a list of times, and get a list of rainfall values at those times as a result:

```{r}
env$extrinsic_drivers_evaluate_range("rainfall", c(5, 1.2, 78.345))
```

## The full climate driver set

`rainfall` is one of a whole family of named drivers that [TF24](../theory/models/tf24.qmd)
(and its acclimation variant, TF24f) register on the environment via
`extrinsic_drivers_set_constant()` at construction time, with the defaults
below:

| Driver         | Meaning                                    | TF24 default |
|----------------|---------------------------------------------|:-------------:|
| `PPFD`         | Photosynthetic photon flux density, i.e. light ($\mu$mol m$^{-2}$ s$^{-1}$) | `1800` |
| `rainfall`     | Rainfall, driving the soil-water balance    | `1` |
| `atm_vpd`      | Atmospheric vapour-pressure deficit (kPa)   | `1` |
| `ca`           | Atmospheric CO$_2$ partial pressure (Pa — so the default `40` is $\approx$ 400 ppm) | `40` |
| `leaf_temp`    | Leaf temperature ($^\circ$C)                | `25` |
| `atm_o2_kpa`   | Atmospheric O$_2$ partial pressure (kPa)    | `21` |
| `atm_kpa`      | Atmospheric pressure (kPa)                  | `100.5` |

Alongside these, `birth_rate` is a further named driver — set per species
rather than on the environment — covered in [Birth rates](#birth-rates) below.

Each of these is configured exactly the same way as `rainfall` above: a
constant with `extrinsic_drivers_set_constant()`, or a time-varying spline
with `extrinsic_drivers_set_variable()`. For example, to hold CO2 fixed at a
different concentration than the TF24 default:

```{r}
#| eval: false
env <- Environment("TF24")
env$extrinsic_drivers_set_constant("ca", 60)
```

... or to drive it with a spline over time, passing `x`/`y` control points the
same way as the sinusoidal rainfall example:

```{r}
#| eval: false
x_pts <- seq(0, 200, 1)
ca_ramp <- list(
  x = x_pts,
  y = 40 + 0.2 * x_pts   # a rising atmospheric CO2 trajectory
)
env <- Environment("TF24")
env$extrinsic_drivers_set_variable("ca", ca_ramp$x, ca_ramp$y)
env$extrinsic_drivers_evaluate("ca", 100)
```

Because every driver shares this same constant-or-spline mechanism, the same
recipe covers any climate-forced scenario: a steadily rising `ca` spline for
a CO2 ramp, a `rainfall` spline that dips and stays low for a drought pulse,
or a `leaf_temp` spline for a warming trajectory. For instance, a drought
pulse superimposed on an otherwise steady rainfall regime might look like:

```{r}
#| eval: false
x_pts <- seq(0, 200, 1)
drought_y <- ifelse(x_pts >= 80 & x_pts <= 120, 0.1, 1)
env <- Environment("TF24")
env$extrinsic_drivers_set_variable("rainfall", x_pts, drought_y)
```

Only the TF24 and TF24f strategies read the atmospheric/leaf drivers
(`atm_vpd`, `ca`, `leaf_temp`, `atm_o2_kpa`, `atm_kpa`) and couple them into
the leaf-level water economy described in
[Assimilation & hydraulics](../theory/assimilation-hydraulics.qmd). FF16 and
K93 are light-only strategies: they read `PPFD` (light) and `birth_rate`, but
setting the other climate drivers on their environment has no effect on
growth.

## Birth rates

Per species birth rates set the seed rain arriving into each patch — the
boundary condition of the [size-structured PDE](../theory/size-structured-pde.qmd)
that the solver integrates. They are set via the `add_strategies()` function. Each object in the list is either a constant number or an embedded list with `x` and `y` control points. There must be the same number of items in the list as there are species:

```{r}
p0 <- scm_base_parameters("TF24")
p0$max_patch_lifetime <- 25      # short patch keeps the hydraulic solve fast

# two species
lmas <- trait_matrix(c(0.0825, 0.125), "lma")

x_pts <- seq(0, 200)
birth_rates <- list(
    species1 = list(x = x_pts, y = 1 + sin(x_pts)),
    species2 = 3.14
  )

p1 <- add_strategies(p0, lmas, TF24_hyperpar,
          keep_existing = FALSE,
          birth_rate = birth_rates)

# TF24 needs a configured soil-water environment: layered soil with an initial
# moisture state and a rainfall driver whose mean stays above zero. A bare
# `Environment("TF24")` would halt with `psi_soil = nan`. See the
# [Example analysis](example-analysis.qmd) guide for the full recipe.
env <- Environment("TF24")
env$set_soil_number_of_depths(15)
env$set_soil_water_state(rep(0.2, 15))
rain_t <- seq(0, p0$max_patch_lifetime, length.out = 1000)
env$extrinsic_drivers_set_variable("rainfall", rain_t, 0.25 * sin(2 * pi * rain_t) + 1.0)

ctrl <- Control()
out <- run_scm(p1, env, ctrl)
```

Evaluating the birth rate can be done with the scm

```{r}
out$patch$species[[1]]$extrinsic_drivers$evaluate("birth_rate", 7)
```
