---
title: "Your first individual"
---
::: {.eyebrow}
guide
:::

This guide begins with the smallest unit in `plant`: one individual. You will
create a plant, inspect its current state, calculate how quickly that state is
changing, and grow it through time. You will then compare plants with different
traits under different light conditions.

We use the [`FF16`](../theory/models/ff16.qmd) model throughout. Its **strategy**
stores the biological parameters for a type of plant; an **individual** is one
plant governed by that strategy. Patch simulations later group individuals of
similar age into nodes and allow them to compete, but those extra layers are not
needed yet.

```{r}
library(plant)
```

## Create an individual

`FF16_Individual()` creates an individual with the default FF16 strategy. You
can also pass it an FF16 strategy with different parameter values. Other model
families have matching constructors, such as `TF24_Individual()`.

```{r}
indv <- FF16_Individual()
```

Printing the object shows the methods and values available on it:

```{r}
indv
```

The most useful starting point is `state()`, which reads the current value of a
state variable. `set_state()` changes a value directly:

```{r}
indv$state("height")
indv$set_state("height", 10)
indv$state("height")

indv$state("fecundity")
indv$state("mortality")
```

Height is the individual's current size. Fecundity and mortality are
cumulative states: they record reproductive output and mortality hazard along
the trajectory. The numerical solver advances these variables together:

```{r}
indv$ode_state
```

Reading a state tells us where the plant is now. To calculate what happens
next, the model also needs an environment. FF16 describes light using canopy
openness, from 0 for no available light to 1 for full light. The helper below
creates a fixed environment with the same openness at every height. The
individual does not alter this demonstration environment.

```{r}
fixed_environment <- function(e = 1.0, height_max = 150.0) {
  env <- Environment("FF16")
  env$set_fixed_environment(e, height_max)
  env
}

env <- fixed_environment(1.0)
```

`compute_rates()` evaluates the plant's carbon budget in that environment and
updates the instantaneous rates of change. The first printout below contains
the initial placeholders; the second contains rates calculated in full light:

```{r}
indv$ode_rates
indv$compute_rates(env)
indv$ode_rates
```

FF16 advances five state variables in total. Their names and current rates are
available directly:

```{r}
indv$ode_names
```

To retrieve one rate by name, use `rate()`:

```{r}
indv$rate("height")
```

Individuals are [R6](https://github.com/wch/R6) reference objects. Assigning an
individual to a new name does not make an independent copy: both names still
point to the same underlying object.

```{r}
indv2 <- indv
indv2$set_state("height", 1)
indv2$state("height")
indv$state("height") # also 1!
```

This behaviour matters when saving or comparing individuals. If you need two
independent plants, construct two objects rather than assigning one to another
name.

## Grow an individual

Setting height directly changes only one state. Normally we want to integrate
all states together so that height, fecundity, mortality, and tissue turnover
describe the same history. `grow_individual_to_time()` provides the simplest
route. Here we request 101 points over 50 years:

```{r}
tt <- seq(0, 50, length.out = 101)
res <- grow_individual_to_time(FF16_Individual(), tt, env)
```

The returned `state` matrix has one row per requested time and one column per
state variable. Plotting height gives the growth trajectory:

```{r}
#| label: fig-internal-growth
#| fig-cap: "Height through time using plant's individual-growth helper."
plot(height ~ tt, res$state, type = "l", las = 1,
     xlab = "Time (years)", ylab = "Height (m)")
```

### Under the hood: integrating the equations directly

Most users can skip this subsection. It demonstrates that the individual's
states and rates can also be passed to a general-purpose ODE solver such as
`deSolve`:

```{r}
#| label: fig-desolve
#| fig-cap: "Height through time, integrating the individual's ODEs directly with deSolve."
# deSolve wrapper of Individual ODEs
derivs <- function(t, y, individual, env) {
  individual$ode_state <- y
  individual$compute_rates(env)
  list(individual$ode_rates)
}

# initialise the individual and its starting state
indv <- FF16_Individual()
y0 <- setNames(indv$ode_state, indv$ode_names)

# solve and plot
yy <- deSolve::lsoda(y0, tt, derivs, indv, env=env)
plot(height ~ time, yy, type="l")
```

The grey points from `deSolve` lie on the trajectory returned by `plant`'s
helper, confirming that both approaches integrate the same equations:

```{r}
#| label: fig-internal-vs-desolve
#| fig-cap: "Internal ODE solver (line) against the deSolve result (grey points)."
plot(height ~ tt, res$state, type="l", las=1,
     xlab="Time (years)", ylab="Height (m)")

points(height ~ time, yy, col="grey", cex=.5)
```

### Grow to a target size

Sometimes the question is not “How large is the plant after 20 years?” but
“When does it reach 15 metres?” `grow_individual_to_size()` integrates until a
sequence of target sizes is reached. `grow_individual_to_height()` is a
convenient height-specific wrapper.

```{r}
# create a sequence from the initial size to height at maturity
indv <- FF16_Individual()
heights <- seq(indv$state("height"), indv$strategy$pars$hmat, length.out=20)
res <- grow_individual_to_size(indv, heights, "height", env)
```

The result contains the time at which each height was reached,

```{r}
res$time
```

the complete state at each of those times,

```{r}
head(res$state)
```

and an individual object for each target. Here the tenth object has the tenth
requested height:

```{r}
res$individual[[10]]
res$individual[[10]]$state("height")
heights[[10]]
```

## Compare traits and environments

We can now ask a biological question: how do leaf traits change a plant's
growth trajectory? We compare two FF16 strategies that differ in leaf mass per
area (LMA). Low LMA describes thinner, cheaper leaves; high LMA describes
thicker, more expensive leaves.

Changing LMA through a trait matrix also updates linked parameters such as leaf
turnover and respiration. This is called **hyperparameterisation**. The
[Strategies and traits](strategies.qmd) guide explains the machinery in more
detail; here we focus on the resulting plants.

Begin with the default FF16 parameters:

```{r}
params <- scm_base_parameters("FF16")
```

Generate one low-LMA and one high-LMA strategy. `param_hyperpar()` selects the
appropriate linking rules for the model stored in `params`:

```{r}
s <- generate_strategy(params, trait_matrix(c(0.0825, 0.2625), "lma"),
                       birth_rate = 1)
s1 <- s[[1]]
s2 <- s[[2]]
```

The resulting strategies show how LMA is linked to leaf turnover (`k_l`) and
dark respiration per unit leaf mass (`r_l`):

```{r}
lapply(s, function(x) x$pars[c("lma", "k_l", "r_l")])
```

Create one individual from each strategy:

```{r}
indv1 <- FF16_Individual(s1)
indv2 <- FF16_Individual(s2)
```

We grow each individual through a sequence of heights. Their starting heights
differ because the same seed mass produces more leaf area, and therefore a
taller seedling, when LMA is lower.

```{r}
#| label: fig-trajectories
#| fig-cap: "Height trajectories of a low-LMA and a high-LMA species in full light."
heights1 <- seq(indv1$state("height"), s1$pars$hmat, length.out=100L)
heights2 <- seq(indv2$state("height"), s2$pars$hmat, length.out=100L)

dat1 <- grow_individual_to_height(indv1, heights1, env,
                             time_max=100, warn=FALSE, filter=TRUE)
dat2 <- grow_individual_to_height(indv2, heights2, env,
                             time_max=100, warn=FALSE, filter=TRUE)

plot(dat1$trajectory[, "time"], dat1$trajectory[, "height"],
     type="l", lty=1,
     las=1, xlab="Time (years)", ylab="Height (m)")
lines(dat2$trajectory[, "time"], dat2$trajectory[, "height"], lty=2)
legend("bottomright", c("Low LMA", "High LMA"), lty=1:2, bty="n")
```

Next compare the same plants at 50% canopy openness. The line type identifies
the LMA strategy, while colour identifies the light environment:

```{r}
#| label: fig-trajectories-light
#| fig-cap: "Height trajectories for both species in high (black) and low (red) light."
env_low <- fixed_environment(0.5)
dat1_low <- grow_individual_to_height(indv1, heights1, env_low,
                                 time_max=100, warn=FALSE, filter=TRUE)
dat2_low <- grow_individual_to_height(indv2, heights2, env_low,
                                 time_max=100, warn=FALSE, filter=TRUE)

cols <- c("black", "#e34a33")
plot(dat1$trajectory[, "time"], dat1$trajectory[, "height"],
     type="l", lty=1,
     las=1, xlab="Time (years)", ylab="Height (m)")
lines(dat2$trajectory[, "time"], dat2$trajectory[, "height"], lty=2)
lines(dat1_low$trajectory[, "time"], dat1_low$trajectory[, "height"],
      lty=1, col=cols[[2]])
lines(dat2_low$trajectory[, "time"], dat2_low$trajectory[, "height"],
      lty=2, col=cols[[2]])
legend("bottomright",
       c("High light", "Low light"), lty=1, col=cols,
       bty="n")
```

### Inspect growth rates

Height growth rate is the slope of a height-through-time trajectory. The solver
integrates this rate to obtain height, so inspecting it helps explain why the
four trajectories diverge.

Select an individual from part-way through the low-LMA, high-light run:

```{r}
indv <- dat1$individual[[50]]
```

Its ODE state records where the individual is,

```{r}
setNames(indv$ode_state, indv$ode_names)
```

while the ODE rates record how those states are changing:

```{r}
setNames(indv$ode_rates, indv$ode_names)
```

An individual saved from a trajectory may retain rates from the solver's last
internal evaluation. Recompute them in the environment before interpreting
them:

```{r}
indv$compute_rates(dat1$env)
setNames(indv$ode_rates, indv$ode_names)
```

They happen to be unchanged here because the demonstration environment is
constant. That need not be true with a varying environment.

The named accessor and the first entry of `ode_rates` return the same height
growth rate:

```{r}
indv$rate("height")
indv$ode_rates[[1]]
```

Collecting that rate across all four trajectories shows how growth changes with
time, trait value, and light:

```{r}
#| label: fig-dhdt-time
#| fig-cap: "Height growth rate through time across traits and light environments."
f <- function(x) x$rate("height")
dhdt1 <- sapply(dat1$individual, f)
dhdt2 <- sapply(dat2$individual, f)
dhdt1_low <- sapply(dat1_low$individual, f)
dhdt2_low <- sapply(dat2_low$individual, f)

plot(dat1$time, dhdt1, type="l", lty=1,
     las=1, xlab="Time (years)", ylab="Height growth rate (m / yr)")
lines(dat2$time, dhdt2, lty=2)
lines(dat1_low$time, dhdt1_low, lty=1, col=cols[[2]])
lines(dat2_low$time, dhdt2_low, lty=2, col=cols[[2]])
legend("topright",
       c("High light", "Low light"), lty=1, col=cols,
       bty="n")
```

Plotting the same rates against height separates the effect of plant size from
the passage of time:

```{r}
#| label: fig-dhdt-height
#| fig-cap: "Height growth rate against height across traits and light environments."
ylim <- c(0, max(dhdt1))
plot(dat1$state[, "height"], dhdt1, type="l", lty=1,
     las=1, xlab="Height (m)", ylab="Height growth rate (m / yr)", ylim=ylim)
lines(dat2$state[, "height"], dhdt2, lty=2)
lines(dat1_low$state[, "height"], dhdt1_low, lty=1, col=cols[[2]])
lines(dat2_low$state[, "height"], dhdt2_low, lty=2, col=cols[[2]])
legend("topright",
       c("High light", "Low light"), lty=1, col=cols,
       bty="n")
```

<!--
TODO(#481): Remove Plant Plus
Over a plant's life, allocation to different structures varies.
This is captured by a set of variables stored within the internals:
e.g., `mass_leaf` `mass_sapwood`.
```{}
indv$internals$mass_leaf
indv$internals$mass_sapwood
```

(these numbers seem a bit off: one of the motivations here is to
develop and use better models of plant allometry.  The
parameterisation used at present are derived from adults and
perform poorly with small plants.  However, based on height / area
relationships [Falster 2011, supporting information], for an 8m
tall plant total leaf areas of 5-10 m are plausible and with an LMA
of 0.08 that implies a total *dry* weight of 400 - 800 g).

Total live dry mass fraction to leaf and stem can be computed as:
```{}
f <- function(p) {
  p_ints <- p$internals
  c(leaf=p_ints$mass_leaf / p_ints$mass_live,
    sapwood=p_ints$mass_sapwood / p_ints$mass_live)
}

cols_part <- c("black", "#045a8d")
matplot(dat1$state[, "height"], t(sapply(dat1$plant, f)),
        type="l", col=cols_part, lty=1, ylim=c(0, 1),
        xlab="Height (m)", ylab="Fractional allocation", las=1)
matlines(dat2$state[, "height"], t(sapply(dat2$plant, f)),
         col=cols_part, lty=2)
legend("topright", c("Sapwood", "Leaf"), lty=1, col=rev(cols_part), bty="n")
```

Relative allocation to leaf mass drops steeply as a plant grows and
is replaced by allocation to sapwood mass.
-->

### Find the light compensation point

For a plant of a fixed size, lower canopy openness reduces height growth:

```{r}
indv <- FF16_Individual()
indv$set_state("height", 10)
indv$compute_rates(fixed_environment(1.0))
indv$rate("height") # in full light

indv$compute_rates(fixed_environment(0.5))
indv$rate("height") # in 1/2 light
```

At sufficiently low light, the individual cannot maintain a positive carbon
balance. For this 10 m plant, height growth has stopped by 25% canopy openness:

```{r}
indv$compute_rates(fixed_environment(0.25))
indv$rate("height")
```

The canopy openness at which whole-plant carbon gain reaches zero is the
**whole-plant light compensation point**.

```{r}
openness <- seq(0, 1, length.out=51)
lcp <- indv$resource_compensation_point()
```

Above this point, height growth increases with canopy openness and gradually
approaches a plateau:

```{r}
#| label: fig-compensation-point
#| fig-cap: "Height growth rate against canopy openness; the filled point marks the light compensation point."
f <- function(x, indv) {
  env <- fixed_environment(x)
  indv$compute_rates(env)
  indv$rate("height")
}
x <- c(lcp, openness[openness > lcp])
plot(x, sapply(x, f, indv), type="l", xlim=c(0, 1),
     las=1, xlab="Canopy openness", ylab="Height growth rate (m / yr)")
points(lcp, 0.0, pch=19)
```

The compensation point is not a single constant for a strategy: it also changes
with plant size. The next plot compares a seedling with a larger individual set
to half its maximum height (`hmat / 2`). The helper `g()` evaluates both plants
across the same openness sequence:

```{r}
#| label: fig-compensation-size
#| fig-cap: "Height growth rate against canopy openness for a seedling and a larger individual."
g <- function(openness, indv) cbind(openness, sapply(openness, f, indv))

indv_seed <- FF16_Individual(s1)
y_seed <- g(openness, indv_seed)

indv_adult <- FF16_Individual(s1)
indv_adult$set_state("height", indv_adult$strategy$pars$hmat / 2)
y_adult <- g(openness, indv_adult)

cols_height <- c("#31a354", "black")
ymax <- max(y_seed[, 2], y_adult[, 2])
plot(y_seed, type="l", col=cols_height[[1]],
     xlim=c(0, 1), ylim=c(0, ymax), las=1,
     xlab="Canopy openness", ylab="Height growth rate (m / yr)")
lines(y_adult, col=cols_height[[2]])
legend("bottomright", c("Seedling", "Larger individual"), lty=1, col=cols_height, bty="n")
```

The compensation point and curve also vary with traits. Here we overlay
the high-LMA species `s2` (dotted) on the low-LMA species `s1` (solid):

```{r}
#| label: fig-compensation-traits
#| fig-cap: "Height growth rate against canopy openness for low-LMA (solid) and high-LMA (dotted) strategies at two sizes."
indv2_seed <- FF16_Individual(s2)
y2_seed <- g(openness, indv2_seed)

indv2_adult <- FF16_Individual(s2)
indv2_adult$set_state("height", indv2_adult$strategy$pars$hmat / 2)
y2_adult <- g(openness, indv2_adult)

ymax <- max(ymax, y2_seed[, 2], y2_adult[, 2])
plot(y_seed, type="l", col=cols_height[[1]],
     xlim=c(0, 1), ylim=c(0, ymax), las=1,
     xlab="Canopy openness", ylab="Height growth rate (m / yr)")
lines(y_adult, col=cols_height[[2]])
lines(y2_seed, col=cols_height[[1]], lty=2)
lines(y2_adult, col=cols_height[[2]], lty=2)
legend("bottomright", c("Seedling", "Larger individual"), lty=1, col=cols_height, bty="n")
```

In this example, the high-LMA strategy has the lower growth rate at most light
levels as a seedling but the higher rate across all light levels at the larger
size. Traits, size, and environment therefore interact: none of these inputs
alone determines the growth trajectory.
