---
title: "Initialising a patch"
---

::: {.eyebrow}
guide
:::

By default a `plant` patch starts empty and grows from seed: nodes (cohorts) are
introduced on a schedule and integrated forward as characteristics of the
[size-structured PDE](../theory/size-structured-pde.qmd). Sometimes you instead
want a patch to *start* with plants already in it — either to **resume** a run
you stopped earlier, or to **seed** a patch with an arbitrary initial size
distribution and watch it evolve.

This guide covers both. In each case, the starting state is attached to a
`Parameters` object—the object that holds the strategies and settings for a
run. When the patch is reset, an ordinary `run_scm(p)` reads that state without
needing a special run argument.

```{r}
#| label: setup
library(plant)
library(dplyr)
library(tidyr)
library(ggplot2)

p0 <- scm_base_parameters("FF16")
p  <- add_strategies(p0, trait_matrix(0.0825, "lma"), birth_rate = 20)
```

## Exporting and resuming a run

Run a patch, collecting a snapshot of the whole population after each step, then
export the state part-way through:

```{r}
#| label: run-and-export
scm <- plant:::SCM("FF16", "FF16_Env")(p, Environment("FF16"), Control())
scm$collect <- TRUE
scm$run()

# export the patch about half-way through the run
k <- floor(length(scm$history) / 2)
state <- export_patch_state(scm, step = k)
```

`export_patch_state()` saves both the plants and the information the solver
needs to continue them. This includes every node's ODE state, the current patch
age, the unused part of the node-introduction schedule, and per-node birth
bookkeeping such as introduction time, initial patch density, and survival at
birth.

`set_initial_state()` writes that onto a `Parameters` object, and `run_scm()`
resumes from it:

```{r}
#| label: resume
p_resumed <- set_initial_state(p, state)
resumed   <- run_scm(p_resumed, collect = TRUE)
```

The resumed run closely reproduces the original. The checkpointed state is
loaded exactly, but the adaptive ODE solver's internal step size is not part of
the checkpoint. Its later steps can therefore differ slightly, so the recovered
fitness agrees to a few parts in $10^5$ rather than bit-for-bit:

```{r}
#| label: compare-fitness
c(original = scm$net_reproduction_ratios,
  resumed  = resumed$net_reproduction_ratios)
```

Because the state lives on the `Parameters` object, the resumed run is fully
self-describing: you could `saveRDS(p_resumed)` and reproduce the continuation in
a fresh session.

## Seeding an arbitrary size distribution

Seeding a size distribution supplies the plants present at patch age zero; it
does not replace recruitment later in the run. Build the initial state from node
heights and densities with `make_initial_state()`. Mortality, fecundity,
accumulated reproduction, and heartwood start at zero. Recruitment then
continues normally for $t > 0$.

As an example, take a log-normal distribution of heights, normalised to a target
total density — the same shape PR #304 used to motivate this feature:

```{r}
#| label: lognormal-distribution
# a log-normal density over heights, scaled to a target total density and
# expressed as node heights + densities
ln_dist <- function(mean_height = 6, log_sd = 0.5, density = 1,
                    min_h = 0.5, max_h = 10, n = 40) {
  h <- seq(min_h, max_h, length.out = n)
  d <- dlnorm(h, log(mean_height), log_sd)
  list(heights = h, densities = d / plant:::trapezium(h, d) * density)
}

dist <- ln_dist(mean_height = 6, density = 1)

tibble(height = dist$heights, density = dist$densities) |>
  ggplot(aes(height, density)) +
  geom_area(fill = "grey80", colour = "black") +
  labs(x = "Height (m)", y = "Density (/m / m2)",
       title = "Seeded initial size distribution (t = 0)") +
  coord_cartesian(expand = FALSE) +
  theme_classic()
```

Seed a patch with it and run as usual. `run_scm(collect = TRUE)` returns the
standard tidy output, so the seeded run drops straight into the usual plotting
and aggregation helpers:

```{r}
#| label: seed-and-run
state0 <- make_initial_state(p, heights = dist$heights,
                             densities = dist$densities)
seeded <- run_scm(set_initial_state(p, state0), collect = TRUE)
```

```{r}
#| label: plot-size-distribution
#| fig-alt: "Node heights over patch age from a seeded log-normal distribution."
plot_size_distribution(seeded$species |> drop_na()) +
  coord_cartesian(expand = FALSE) +
  labs(title = "Patch evolution from a seeded distribution")
```

The fan of lines starting at $t = 0$ is the seeded population growing and being
shaded out; the lines that appear later are recruits arriving on the schedule.

Aggregate properties come from the same tidy table via `FF16_expand_state()` and
`integrate_over_size_distribution()`:

```{r}
#| label: aggregate-dynamics
#| fig-alt: "Leaf area index and stem density over patch age."
totals <- seeded |> FF16_expand_state() |> _$species |>
  integrate_over_size_distribution()

totals |>
  select(time, area_leaf, density) |>
  rename(`leaf area index` = area_leaf, `individuals (/m2)` = density) |>
  pivot_longer(-time) |>
  ggplot(aes(time, value)) +
  geom_line() +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "Patch age (years)", y = NULL) +
  coord_cartesian(expand = FALSE) +
  theme_classic()
```

Seeding the *same* distribution shape at different total densities shows how the
starting density shapes the early dynamics before the patch converges:

```{r}
#| label: density-panel
#| fig-alt: "Size-distribution trajectories for three decreasing initial densities."
densities <- c(1.0, 0.2, 0.05)

panel <- lapply(densities, function(dn) {
  d <- ln_dist(mean_height = 6, density = dn)
  r <- run_scm(set_initial_state(p, make_initial_state(p, d$heights, d$densities)),
               collect = TRUE)
  r$species |> drop_na() |>
    mutate(scenario = sprintf("initial density = %.2f /m2", dn))
}) |> bind_rows()

plot_size_distribution(panel) +
  facet_wrap(~scenario) +
  coord_cartesian(expand = FALSE) +
  labs(title = "Same distribution shape, decreasing initial density")
```

## Plausible initial conditions

Not every combination of starting sizes and densities is biologically or
numerically plausible. Very large plants at very high density can self-shade so
strongly that production becomes negative and node densities become non-finite.
`plant` checks for this and stops with an error:

```{r}
#| label: guard
#| error: true
bad <- make_initial_state(p, heights = seq(8, 12, length.out = 20),
                          densities = rep(50, 20))
run_scm(set_initial_state(p, bad))
```

If this happens, begin with a compatible state from `export_patch_state()`, or
reduce the sizes and densities of the seeded plants. An exported state is the
safest choice for resuming a run because it preserves the full node
bookkeeping.
