---
title: "Patch dynamics"
---
::: {.eyebrow}
guide
:::

This guide moves from one individual to a **patch**: an area in which many
plants grow and compete for resources. You will run a one-species patch, follow
its cohorts through time, inspect its light environment, and then add a second
species.

Rather than simulating every plant separately, `plant` represents cohorts as
**nodes** and follows their changing size and density. These trajectories are
the characteristics of the model's partial differential equation. See the
[size-structured PDE](../theory/size-structured-pde.qmd) being solved and the
[characteristic method](../theory/implementation/characteristic-method.qmd)
used to solve it.

## Set up the patch

First we instantiate the basic parameters of the `FF16` strategy.
`max_patch_lifetime` is the oldest patch age integrated by the solver and also
parameterises the default disturbance distribution. Its default is about 105
years; the corresponding mean disturbance interval is a separate derived value
of 30 years.

```{r}
#| label: setup-params
library(plant)

params <- scm_base_parameters("FF16")
```

```{r}
#| include: false
#| label: helpers
library(dplyr)
library(tidyr)

# The tidy `species` table returned by `run_scm(collect = TRUE)` has one row
# per node per step. This helper reshapes a single variable into a matrix with
# one row per node and one column per step, padding with `NA` for nodes that
# had not yet been introduced.
sdd_matrix <- function(species_data, var) {
  m <- species_data |>
    dplyr::select(node, step, dplyr::all_of(var)) |>
    tidyr::pivot_wider(names_from = step, values_from = dplyr::all_of(var)) |>
    dplyr::arrange(node)
  as.matrix(m[, -1])
}

# Leaf area index, integrated across the size distribution with the trapezium
# rule. `competition_effect` is the leaf area of an individual, so weighting it
# by node density gives leaf area per unit ground area and per unit height.
patch_lai <- function(height, density, competition_effect) {
  fc <- density * competition_effect
  o  <- order(height, decreasing = TRUE)
  h  <- height[o]
  f  <- fc[o]
  sum(diff(-h) * (head(f, -1) + tail(f, -1))) / 2
}
```

## One patch

Add one species, represented by an LMA value of 0.0825:

```{r}
#| label: single-species-patch
patch <- add_strategies(params, trait_matrix(0.0825, "lma"))
```

Run the characteristic solver and collect the state after every recorded step:

```{r}
#| label: run-patch
result <- run_scm(patch, collect = TRUE)
```

The result contains several components. These are the ones used most often:

```{r}
#| label: result-names
names(result)
```

| Component | What it contains |
|---|---|
| `steps` | One row per recorded step, including patch age (`time`) and the abundance of patches of that age (`patch_density`). |
| `species` | One row per node per step, including height, density, and individual leaf area (`competition_effect`). |
| `env` | Resource conditions through time, including the light profile. |
| `offspring_production` | Reproductive output during patch development. |
| `net_reproduction_ratios` | Lifetime reproductive output for each species. |
| `p` | The parameters used for the run, including any refined schedule. |

The result also records `n_spp`, the number of species.

First reshape node heights into a matrix:

```{r}
#| label: extract-heights
t <- result$steps$time
h <- t(sdd_matrix(dplyr::filter(result$species, species == "1"), "height"))
```

here `h` is the height of every node, arranged so that each row is a *time* and
each column is a *node*:

```{r}
#| label: height-dim
dim(h)
```

The time points recorded by the solver are not evenly distributed:

```{r}
#| label: time-points
sample_steps <- unique(round(seq(1, length(t), length.out = 5)))
signif(t[sample_steps], 1)
```

This uneven spacing is deliberate. Early patch development needs higher time
resolution because small differences are amplified by competition. Although
the patch runs for about 105 years, roughly half the recorded steps occur in
the first year. See
[The node-spacing algorithm](../theory/implementation/node-spacing.qmd).

Nodes are introduced at each time-step, except the last. At time step $i$, the
rows of `h` record the height of all previously introduced nodes in the patch:

```{r}
#| label: heights-at-step
h[10, 1:10]
```

with `NA` for nodes that have not yet been introduced:

```{r}
#| label: unintroduced-node
h[10, 11]
```

Similarly, the $j$th column of `h` records the heights of a particular node:

```{r}
#| label: node-trajectory
h[10:19, 10]
```

with `NA` at time steps before it was introduced:

```{r}
#| label: count-na
sum(is.na(h[, 10]))
```

With a single species on this schedule there is one more recorded time step than node
introduction, with the last two steps having the same number of nodes. With
multiple species there can be more time steps than node introductions, as we'll
record data at every node introduction for *either* species when their schedules
differ.

FF16 individuals grow taller through time, but their growth rate depends on
both their size and the light reaching their canopy. Competition therefore
causes nodes introduced at different times to follow different trajectories.

<!-- TODO(#481): Dead link- "(see vignette:plant)" -->

We can plot the trajectories of nodes developing within a patch over time:

```{r}
#| label: plot-trajectories
matplot(t, h, lty=1, col=util_colour_set_opacity("black", 0.25), type="l",
        las=1, xlab="Time (years)", ylab="Height (m)")
```

The light environment — canopy openness as a function of height — is stored over
each time step; see [Canopy & light-environment models](../theory/implementation/canopy-models.qmd)
for how it is computed from the leaf area above. This means we can also plot how
it changes during patch development:

```{r}
#| label: plot-light-environment
light <- result$env$light_availability

# Set limits
xlim <- c(0, 1.1)
ylim <- range(light$height)

# Draw the spline at each step
plot(NA, xlim=xlim, ylim=ylim, las=1, xlab="Canopy openness", ylab="Height (m)")
for (s in unique(light$step)) {
  x <- dplyr::filter(light, step == s)
  lines(x$light_availability, x$height, col = "grey")
}

# Add labels at specific times
blues <- c("#DEEBF7", "#C6DBEF", "#9ECAE1", "#6BAED6",
           "#4292C6", "#2171B5", "#08519C", "#08306B")

times <- c(5, 10, 20, 40, max(result$steps$time))
cols <- colorRampPalette(blues[-(1:2)])(length(times))

for (i in seq_along(times)) {
  s <- result$steps$step[which.min(abs(times[[i]] - result$steps$time))]
  x <- dplyr::filter(light, step == s)
  lines(x$light_availability, x$height, col=cols[[i]])

  y <- max(x$height)
  points(1, y, pch=19, col=cols[[i]])
  text(1 + strwidth("x"), y, paste(round(times[[i]]), "years"),
       adj=c(0, 0))
}
```

Initially, all nodes are short and most of the canopy is open. As the patch
develops, only the tallest plants remain in high light. Shorter plants
experience changing conditions as older plants die and new growth fills the
gaps.

This is best shown by plotting the canopy openness at ground level over time:

```{r}
#| label: plot-ground-openness
ground <- light |>
  dplyr::group_by(time) |>
  dplyr::slice_min(height, n = 1, with_ties = FALSE) |>
  dplyr::ungroup()
plot(ground$time, ground$light_availability, type="l", las=1,
     ylim=c(0, 1), xlab="Time (years)", ylab="Canopy openness")
```

The repeated waves reflect recruitment followed by self-thinning. Mortality is
not instantaneous, so a one-species patch can briefly recruit more plants than
the closed canopy can support.

## Leaf area index in one patch

Leaf area index (LAI) is total leaf area per unit ground area. The collected
output records individual leaf area as `competition_effect`; weighting it by
node density and integrating over height gives patch-level LAI. FF16 multiplies
LAI by the light extinction coefficient `k_I` to obtain the ground-level
optical depth used by its canopy model:

```{r}
#| label: plot-lai
lai <- result$species |>
  dplyr::filter(species == "1") |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    lai = patch_lai(height, density, competition_effect),
    .groups = "drop"
  )
plot(lai$time, lai$lai, type="l", las=1, xlab="Time (years)", ylab="Leaf area index")
```

## Multi-species patches

Now add a second species with higher leaf mass per area (LMA). Both species use
the same light environment, so each changes the conditions experienced by the
other:

```{r}
#| label: two-species-patch
patch_2sp <- add_strategies(patch, trait_matrix(0.2625, "lma"))
```

Then collect the patch-level dynamics:

```{r}
#| label: run-two-species
result_2sp <- run_scm(patch_2sp, collect = TRUE)
```

We can plot the development of each species within the patch:

```{r}
#| label: plot-two-species
t2 <- result_2sp$steps$time
h1 <- t(sdd_matrix(dplyr::filter(result_2sp$species, species == "1"), "height"))
h2 <- t(sdd_matrix(dplyr::filter(result_2sp$species, species == "2"), "height"))

cols <- c("#e34a33", "#045a8d")

# Species 1 - red
matplot(t2, h1, lty=1, col=util_colour_set_opacity(cols[[1]], .25), type="l",
        las=1, xlab="Time (years)", ylab="Height (m)")

# Species 2 - blue
matlines(t2, h2, lty=1, col=util_colour_set_opacity(cols[[2]], .25))
```

To isolate the effect of the competitor, compare the low-LMA species alone
(black) with the same species in the two-species patch (red):

```{r}
#| label: plot-mono-vs-two
# Monoculture patch (black)
matplot(t, h, lty=1, col=util_colour_set_opacity("black", .25), type="l",
        las=1, xlab="Time (years)", ylab="Height (m)")

# Two species patch (red)
matlines(t2, h1, lty=1, col=util_colour_set_opacity(cols[[1]], .25))
```

The additional species has little effect on the *initial* wave because it grows
more slowly and remains shorter. After canopy closure, around year 5 in this
example, later recruits face stronger competition and are slowed, suppressed,
or eliminated.

The dynamics are easier to see when node trajectories are weighted by node
density (some of the lines here represent nodes at close to zero density). Log
density is available as a characteristic variable of all `plant` strategies:

```{r}
#| label: plot-density-weighted
# Relativise the log densities onto (-4, max)
d1 <- t(sdd_matrix(dplyr::filter(result_2sp$species, species == "1"), "log_density"))
d2 <- t(sdd_matrix(dplyr::filter(result_2sp$species, species == "2"), "log_density"))
rel <- function(x, xmin) {
  x[x < xmin] <- xmin
  xmax <- max(x, na.rm=TRUE)
  (x - xmin) / (xmax - xmin)
}
rd1 <- rel(d1, -4)
rd2 <- rel(d2, -4)

# R doesn't seem to offer a way to plot lines that vary in colour, so
# this is quite roundabout using `segments`, shaded by the density at
# the first part of the line segment:
n <- length(t2)
x <- matrix(rep(t2, ncol(h1)), nrow(h1))
col1 <- matrix(util_colour_set_opacity(cols[[1]], rd1), nrow(d1))
col2 <- matrix(util_colour_set_opacity(cols[[2]], rd2), nrow(d2))
plot(NA, xlim=range(t2), ylim=range(h1, na.rm=TRUE),
     las=1, xlab="Time (years)", ylab="Node height (m)")
segments(x[-1, ], h2[-1, ], x[-n, ], h2[-n, ], col=col2[-n, ], lend="butt")
segments(x[-1, ], h1[-1, ], x[-n, ], h1[-n, ], col=col1[-n, ], lend="butt")
```

Now we see that the high LMA species (blue) becomes dominant, excluding the low
LMA species (red) from the canopy.

The changing intensity of competition can be shown as the total leaf area of
each species:

```{r}
#| label: plot-two-species-lai
lai2 <- result_2sp$species |>
  dplyr::group_by(species, step) |>
  dplyr::summarise(
    lai = patch_lai(height, density, competition_effect),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(names_from = species, values_from = lai,
                     names_prefix = "sp", values_fill = 0) |>
  dplyr::right_join(result_2sp$steps, by = "step") |>
  dplyr::arrange(step) |>
  dplyr::mutate(dplyr::across(c(sp1, sp2), ~tidyr::replace_na(.x, 0)),
                total = sp1 + sp2)

# Patch total LAI (dashed)
plot(lai2$time, lai2$total, type="l", las=1, lty=2,
     xlab="Time (years)", ylab="Leaf area index")

# Species 1 LAI (red)
lines(lai2$time, lai2$sp1, col=cols[[1]])

# Species 2 LAI (blue)
lines(lai2$time, lai2$sp2, col=cols[[2]])
```

<!-- TODO(#481): Introduce meta populations:

To find the average value over the metapopulation we weight by patch abundance:

```{r}
#| label: plot-metapopulation-lai
metapopulation <- function(x){
  plant:::trapezium(t2, x*result_2sp$steps$patch_density)
}

lai2_av <- metapopulation(lai2$total)
lai2_1_av <- metapopulation(lai2$sp1)
lai2_2_av <- metapopulation(lai2$sp2)

plot(lai2$time, lai2$total, type="l", las=1, lty=2,
     xlab="Time (years)", ylab="Leaf area index")
lines(lai2$time, lai2$sp1, col=cols[[1]])
lines(lai2$time, lai2$sp2, col=cols[[2]])

axis(4, at=lai2_av,   tck=0.1, lty = 2, labels=NA)
axis(4, at=lai2_1_av, tck=0.1, col.ticks=cols[[1]], labels=NA)
axis(4, at=lai2_2_av, tck=0.1, col.ticks=cols[[2]], labels=NA)
axis(1, at=108, labels="Av")
```
-->

## Appendix - `ggplot` example

```{r}
#| label: ggplot-helper
#' A standard plot for density weighted node trajectories
#'
#' Parameters:
#'  - result: the output of `run_scm(..., collect = TRUE)`
#'  - threshold: lower bound on density
#'
#' Requires:
#'  - ggplot2
#'  - dplyr

patch <- function(result, threshold = 1e-3) {

  # The `species` table is already in long format, with one row per node
  # per step, so we can filter and plot it directly.
  df <- result$species |>
    dplyr::filter(species == "1", density > threshold)

  # Plot with densities
  ggplot(df, aes(x = time, y = height,
                 group = node, alpha = density)) +
    geom_line() +
    labs(x = "Time (years)",
         y = "Height (m)",
         alpha = "Density") +
    theme_bw()
}
```
