---
title: "Running sweeps in parallel"
---
::: {.eyebrow}
workflow
:::

Most `plant` analyses end up as a **sweep**: run the same model over a grid of
parameters, save one result per combination, then compare. A single
[FF16](../theory/models/ff16.qmd) patch takes about a second, a
[TF24](../theory/models/tf24.qmd) patch minutes, and a community assembly tens
of minutes — so the difference between running a grid serially and running it
across cores is the difference between a coffee break and a lost afternoon.

Sweeps are also *embarrassingly parallel*: no job needs anything from another
job. That leaves two questions, and the second is the one that bites. How do you
hand jobs to cores? And how do you know afterwards which results are current?
This guide works through three answers of increasing sophistication, on one grid
throughout:

1. serial `lapply()` — the baseline worth measuring against;
2. `parallel::mclapply()` — one extra line, most of the speedup;
3. a [`targets`](https://books.ropensci.org/targets/) pipeline — for when the
   sweep becomes something you maintain.

If you have not yet run a single patch, start with
[Patch dynamics](patch-dynamics.qmd) and
[Example analysis](example-analysis.qmd); this guide assumes both.

```{r}
#| label: load-libraries
library(plant)
library(ggplot2)
library(dplyr)

theme_set(theme_bw(base_size = 12))
outdir <- file.path(tempdir(), "sweep")
```

## Start with a manifest, not a loop

Every approach here shares one idea: **a table with one row per job**, carrying
the parameters, a human-readable `id`, and the output filename. Build it with a
function rather than typing a CSV, so it can be re-derived, diffed, and
extended:

```{r}
#| label: manifest-fn
sweep_grid <- function(lma = c(0.05, 0.0825, 0.15, 0.25),
                       max_patch_lifetime = c(30, 60),
                       birth_rate = 17.31,
                       dir = outdir) {
  grid <- expand.grid(lma = lma, max_patch_lifetime = max_patch_lifetime,
                      KEEP.OUT.ATTRS = FALSE)
  grid$birth_rate <- birth_rate
  grid$id <- sprintf("lma%g-T%g", grid$lma, grid$max_patch_lifetime)
  grid$file <- file.path(dir, paste0(grid$id, ".rds"))
  grid[order(grid$id), ]
}

# Rows as a list of lists: the shape lapply(), mclapply() and targets all take.
sweep_rows <- function(grid) {
  unname(lapply(split(grid, seq_len(nrow(grid))), as.list))
}

grid <- sweep_grid()
grid |> select(id, lma, max_patch_lifetime, birth_rate)
```

Two rules make everything downstream work. **Every parameter that changes the
result must appear in `id`** — otherwise two jobs write to one filename and the
second silently overwrites the first. And **the filename is derived, never
typed**, so exactly one place knows where a result lives.

## One job, one function, one file

The job is an ordinary function of one manifest row. It refines the
node-introduction schedule, runs the patch, and returns summaries; a thin wrapper
saves the result and returns the filename:

```{r}
#| label: job-fns
run_patch <- function(row) {
  p0 <- scm_base_parameters("FF16")
  p0$max_patch_lifetime <- row$max_patch_lifetime

  p1 <- add_strategies(p0, trait_matrix(row$lma, "lma"),
                       birth_rate = row$birth_rate)

  pars <- run_scm(p1, refine_schedule = TRUE)$parameters
  out <- run_scm(pars, collect = TRUE)

  list(row = row, totals = patch_totals(out),
       emergent = emergent_summary(out))
}

run_job <- function(row) {
  dir.create(dirname(row$file), showWarnings = FALSE, recursive = TRUE)
  saveRDS(run_patch(row), row$file)
  row$file
}
```

```{r}
#| label: summary-fns
#| code-fold: true
#| code-summary: "Summary helpers: patch_totals() and emergent_summary()"
# One row per recorded time: patch totals plus mean size per individual. See
# Example analysis for what integrate_over_size_distribution() does.
patch_totals <- function(out) {
  tot <- integrate_over_size_distribution(FF16_expand_state(out)$species)
  for (v in c("diameter_stem", "area_stem", "area_leaf", "height",
              "mass_above_ground")) {
    tot[[paste0(v, "_av")]] <- tot[[v]] / tot$density
  }
  tot
}

# One row per job: the scalars worth comparing across the grid.
emergent_summary <- function(out) {
  tot <- patch_totals(out)
  n <- nrow(tot)
  data.frame(net_reproduction_ratio = out$net_reproduction_ratios[[1]],
             offspring_production = out$offspring_production[[1]],
             lai_max = max(tot$area_leaf, na.rm = TRUE),
             basal_area_final = tot$area_stem[n],
             density_final = tot$density[n],
             height_av_final = tot$height_av[n])
}
```

Note what the job *saves*. `run_scm(collect = TRUE)` returns roughly 1.4 MB of
solver state per run, while the totals and scalars an analysis actually consumes
come to about 35 kB. Deciding what a job's artefact is — rather than reflexively
saving everything — is the cheapest performance decision in a sweep, because
every later read pays for it.

## The serial baseline

Always measure this first. It tells you whether parallelism is worth any
complexity, and it is the reference your parallel results must match:

```{r}
#| label: serial
rows_small <- sweep_rows(head(grid, 4))

t_serial <- system.time(files_serial <- lapply(rows_small, run_job))
t_serial[["elapsed"]]
```

Four patches on one core, written to the filenames the manifest promised:

```{r}
#| label: serial-files
basename(unlist(files_serial))
```

## `mclapply()`: one line, most of the win

`parallel::mclapply()` is `lapply()` with an `mc.cores` argument. For a sweep
that is genuinely all the change required:

```{r}
#| label: mclapply
workers <- max(1L, min(4L, parallel::detectCores() - 1L))

t_par <- system.time(
  files_par <- parallel::mclapply(sweep_rows(grid), run_job,
                                  mc.cores = workers,
                                  mc.preschedule = FALSE)
)

c(workers = workers,
  serial_4_jobs = t_serial[["elapsed"]],
  parallel_8_jobs = t_par[["elapsed"]])
```

Twice the jobs in less than the serial time for four. Three details matter more
than the speedup itself.

**`mc.preschedule = FALSE` load-balances.** With prescheduling on (the default),
jobs are dealt to workers up front, so one slow job strands a whole worker's
queue behind it. Patch run times vary several-fold with `max_patch_lifetime`, so
turning it off is usually right.

**Forking is why this suits a development build.** `mclapply()` forks the current
session, so workers inherit the already-loaded `plant` namespace and its compiled
library — including a `pkgload::load_all()` dev build. A PSOCK or
`future::multisession` cluster instead spawns fresh sessions that see only the
*installed* package, which quietly runs code you are not testing. The costs of
forking are that it does not exist on Windows (`mc.cores` is ignored, so keep the
serial path working) and that it is unreliable inside the RStudio GUI — run
sweeps from a terminal. On Windows, or when workers must outlive one session,
`mirai` and `future`/`furrr` are the alternatives, at the price of installing
`plant` into the workers.

**`mclapply()` does not stop on error.** A failed job returns a `try-error`
object *in the result list* while everything else succeeds, so a broken sweep can
look complete. Check explicitly, every time:

```{r}
#| label: mclapply-errors
failed <- vapply(files_par, inherits, logical(1), "try-error")
sum(failed)
```

### Reading results back

The manifest doubles as the query interface. Ask what exists:

```{r}
#| label: status
sweep_status <- function(grid) {
  info <- file.info(grid$file)
  grid$exists <- file.exists(grid$file)
  grid$size_kb <- round(info$size / 1e3, 1)
  grid$modified <- info$mtime
  grid
}

sweep_status(grid) |> select(id, exists, size_kb)
```

...then bind the finished jobs into one table and compare across the grid:

```{r}
#| label: collect
collect_emergent <- function(grid, files = grid$file) {
  done <- grid[grid$file %in% files & file.exists(grid$file), ]
  if (nrow(done) == 0L) return(NULL)
  do.call(rbind, lapply(seq_len(nrow(done)), function(i) {
    obj <- readRDS(done$file[i])
    cbind(done[i, c("id", "lma", "max_patch_lifetime")], obj$emergent,
          row.names = NULL)
  }))
}

results <- collect_emergent(grid)
results |> select(id, lma, max_patch_lifetime, net_reproduction_ratio, lai_max)
```

```{r}
#| label: plot-emergent
#| fig-height: 3.5
ggplot(results, aes(lma, net_reproduction_ratio,
                    colour = factor(max_patch_lifetime))) +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "LMA (kg/m2)", y = "Net reproduction ratio",
       colour = "Patch\nlifetime (yr)")
```

Treat those numbers as mechanics rather than biology: `birth_rate` is held at an
illustrative constant and patch lifetimes are shortened for speed, so the net
reproduction ratio sits below 1 everywhere and none of these patches is at
equilibrium. Solving for equilibrium birth rates is a different job — see
[Emergent properties](emergent-properties.qmd) — and the sweep machinery is
identical either way.

### The idiom that gets you into trouble

The natural next step is to make the sweep resumable by skipping finished jobs:

```{r}
#| label: skip-idiom
todo <- sweep_rows(grid[!file.exists(grid$file), ])
length(todo)
```

Zero, because the sweep above already finished all eight — rerun the script and
nothing recomputes. That is exactly the behaviour you want, and exactly where the
trap is.

This is the right instinct and the wrong mechanism. `file.exists()` answers only
"is this file missing?", which leaves three failure modes:

- **A parameter changed that is not in the filename.** Nothing reruns. Your
  output directory now mixes vintages, with no record of which is which.
- **The job function changed.** Same outcome, and worse, because editing
  `run_patch()` invalidates *every* result while all the files still exist.
- **A job crashed mid-write.** A truncated `.rds` exists, so it counts as done.

None of these announce themselves. They surface weeks later as a figure that
cannot be reproduced. Fixing it properly means tracking parameters *and* code.

## `targets`: when the sweep is something you maintain

[`targets`](https://books.ropensci.org/targets/) hashes each job's inputs — the
parameters *and* the body of every function it calls — then reruns exactly the
jobs whose hashes changed. Parallelism comes from a
[`crew`](https://wlandau.github.io/crew/) controller, so it is the same one-line
change as `mc.cores`.

A pipeline is a project directory rather than a script: `_targets.R` declares the
steps, and the functions live in `R/`.

```{r}
#| label: pipeline-listing
#| eval: false
# _targets.R
library(targets)

nworkers <- as.integer(Sys.getenv("NWORKERS", "4"))

tar_option_set(
  packages = c("plant", "dplyr"),
  controller = crew::crew_controller_local(workers = nworkers)
)

tar_source("R")  # sweep_grid(), run_job(), collect_emergent(), ...

list(
  # 1. the manifest
  tar_target(grid, sweep_grid()),

  # 2. one branch per row, run in parallel
  tar_target(job, sweep_rows(grid), iteration = "list"),
  tar_target(job_run, run_job(job), pattern = map(job), format = "file"),

  # 3. collection, depending on the jobs (see below)
  tar_target(emergent, collect_emergent(grid, job_run))
)
```

Then one command, with the worker count taken from the environment so the same
pipeline runs locally and on a cluster:

```bash
NWORKERS=8 Rscript -e 'targets::tar_make()'
```

Three of those lines deserve comment.

`pattern = map(job)` is *dynamic branching*: one branch per manifest row, created
at run time, so extending the grid does not mean editing the pipeline.

`format = "file"` tells `targets` that the target's value is a filename and the
file is the artefact. Results stay as plain `.rds` files at readable paths, and
`targets` tracks them instead of keeping a second hashed copy inside
`_targets/objects/`.

`collect_emergent(grid, job_run)` passes the job target even though the function
could find the files from `grid` alone. **This is the mistake to avoid:** a
collection step that only reads files off disk has no dependency on the jobs that
wrote them, so `targets` sees no reason to rerun it and quietly serves a stale
table. If a step consumes the jobs' output, it must take the job target as an
argument — which is what the `files` argument above is for.

### Watching it decide

Here is the behaviour that justifies the extra machinery, run for real in a
throwaway project directory. Two jobs, then a third parameter value added:

```{r}
#| label: targets-demo
#| eval: !expr requireNamespace("targets", quietly = TRUE)
write_pipeline <- function(lmas) {
  writeLines(sprintf('
    library(targets)
    source("%s")
    tar_option_set(packages = "plant")
    list(
      tar_target(grid, sweep_grid(lma = c(%s), max_patch_lifetime = 30,
                                  dir = "output")),
      tar_target(job, sweep_rows(grid), iteration = "list"),
      tar_target(job_run, run_job(job), pattern = map(job), format = "file"),
      tar_target(emergent, collect_emergent(grid, job_run))
    )', fns, paste(lmas, collapse = ", ")), "_targets.R")
}

job_summary <- function(label) {
  p <- targets::tar_progress()
  b <- p[startsWith(p$name, "job_run_"), ]
  data.frame(run = label, built = sum(b$progress == "completed"),
             skipped = sum(b$progress == "skipped"))
}

# The pipeline needs the functions defined above as a file it can source.
fns <- file.path(tempdir(), "sweep-fns.R")
dump(c("sweep_grid", "sweep_rows", "run_job", "run_patch", "patch_totals",
       "emergent_summary", "collect_emergent"), file = fns)

targets::tar_dir({
  write_pipeline(c(0.05, 0.15))
  targets::tar_make(reporter = "silent")
  first <- job_summary("two jobs, first run")

  write_pipeline(c(0.05, 0.15, 0.25))
  targets::tar_make(reporter = "silent")
  second <- job_summary("one lma value added")

  print(rbind(first, second))
  print(targets::tar_read(emergent)[, c("id", "net_reproduction_ratio")])
})
```

Adding one parameter value ran a single job and skipped two — and the collected
table still grew to three rows, because `emergent` depends on `job_run`. Change
the body of `run_patch()` instead and all three rerun, which is the case
`file.exists()` can never catch.

`targets` also keeps a per-job record. `tar_progress()` distinguishes a job that
*errored* from one that never started, so a crash is a state rather than a
missing file; `tar_outdated()` reports what `tar_make()` would rerun before you
commit to it; and `tar_read(emergent)` retrieves the collected table directly.

## Running on a cluster

The same pipeline runs unchanged on a PBS cluster such as UNSW's Katana. Only the
worker count differs: ask for one node with many cores and pass its core count
through.

```bash
#!/bin/bash
#PBS -l select=1:ncpus=16:mem=64gb
#PBS -l walltime=12:00:00
cd $PBS_O_WORKDIR
module load r/4.5.1
NWORKERS=$(nproc) Rscript -e 'targets::tar_make()'
```

**One large node, not many small ones.** `crew.cluster` can in principle launch
each worker as its own cluster job, but it emits Torque syntax that PBS Pro
installations (Katana among them) reject, so multi-node fan-out needs care.
In practice this is rarely the binding constraint: 16–32 cores on one node covers
most sweeps, and because `targets` resumes rather than restarts, a job that hits
its walltime picks up where it stopped.

## Which one should you use?

| | `mclapply()` | `targets` |
|---|---|---|
| Setup | one line | a project directory |
| Reruns changed jobs | no — `file.exists()` only | yes, on parameters *and* code |
| Failed job | `try-error` in the list, easy to miss | recorded per branch |
| Resume after interruption | manual | automatic |
| Best for | exploration, a grid you run once | a sweep whose results reach a paper |

Start with `mclapply()`. It is the right tool for exploring, and its one-line
cost means there is no excuse for running a grid serially. Move to `targets` when
you catch yourself asking "is this result current?" — because that question has
no reliable answer in the first workflow, and is the one thing the second is
built to answer.
