---
title: "Self-thinning"
---
::: {.eyebrow}
example
:::

After disturbance, many young plants may establish together. As they grow, the
canopy closes and light becomes unevenly distributed. Taller plants retain
access to light, while shaded plants grow slowly and are more likely to die.
The stand therefore shifts from many small individuals to fewer large ones.

This decline in density as average size increases is called **self-thinning**.
On logarithmic axes, many stands follow an approximately straight trajectory
known as the self-thinning line. Its universality across species, environments,
and stand types remains debated.

Here we ask whether this pattern emerges from an
[FF16](../theory/models/ff16.qmd) simulation. It is not imposed on the model:
we calculate it from the growth and mortality represented by the
[size-structured PDE](../theory/size-structured-pde.qmd).

```{r}
#| label: load-libraries
library(plant)
library(dplyr)
library(ggplot2)
```

```{r}
#| label: console-logging
plant_log_console()
```

We begin with one FF16 strategy and rapidly reduce recruitment after
establishment, isolating the development of the initial cohort. Schedule
refinement gives the fast early dynamics more resolution:

```{r}
#| label: grow-stand
p0 <- scm_base_parameters("FF16")

p0$strategy_default$pars$recruitment_decay <- 5
# rapidly reduces later recruitment to isolate the first group of plants

ctrl <- Control()
ctrl$schedule_eps <- 0.005
# more refined scheduling (schedule_eps is a Control field, not a Parameters one)

lma_trait <- trait_matrix(0.1457419, "lma")

p1 <- add_strategies(p0, lma_trait)

stand_results <- run_scm(p1, ctrl = ctrl, refine_schedule = TRUE)$parameters
# refines the node-introduction schedule where the estimated error is largest

out <- run_scm(stand_results, ctrl = ctrl, collect = TRUE)
# run the SCM model
```


The collected `species` table contains one row per node and time. Expanding its
states and integrating over height produces one row of patch totals per time.
Dividing size totals by total density then gives mean size per individual:

```{r}
#| label: expand-patch-state
tidy_stand <- out

stand_expand <- FF16_expand_state(tidy_stand)
# adds in more state variables from our stand output (leaf, bark, sapwood, stem properties to name a few)

patch_total <-
    stand_expand$species %>%
    integrate_over_size_distribution() %>%
    mutate(
      # create average sizes
      across(c(diameter_stem, area_stem, area_leaf, height, mass_above_ground), ~.x/density,
             .names = "{.col}_av"))
# one row per time, containing patch totals and mean individual sizes
```

The self-thinning plot places mean stem diameter on the horizontal axis and
stem density on the vertical axis. Both axes are logarithmic so an approximate
power-law relationship appears as a straight line:

```{r}
#| label: plot-self-thinning
patch_total %>%
    ggplot(aes(diameter_stem_av, density)) +
             geom_line() +
             scale_x_log10() +
             scale_y_log10() +
             xlab("stem diameter (m)") +
             ylab("density " ~(m^-2)) +
             ggtitle("Self-thinning plot") +
             theme_classic(base_size = 18)
```

Leaf area and above-ground biomass provide complementary views of stand
development:

```{r}
#| label: plot-leaf-area-biomass
patch_total %>%
      ggplot(aes(time, area_leaf)) +
               geom_line() +
               xlab("time (yr)") +
               ylab("leaf area " ~(m^2)) +
               ggtitle("Leaf Area over time") +
               theme_classic(base_size = 18)

patch_total %>%
    ggplot(aes(time, mass_above_ground)) +
             geom_line() +
             xlab("time (yr)") +
             ylab("above-ground biomass (kg)") +
             ggtitle("Above-ground Biomass over time") +
             theme_classic(base_size = 18)
```

Finally, compare height distributions at selected patch ages. The helper chooses
the recorded time nearest each requested age, avoiding assumptions about the
number or spacing of solver steps.

```{r}
#| label: plot-size-distributions
target_times <- c(0.5, 2, 5, 9, 14, 28, 56, 98)
recorded_times <- sort(unique(tidy_stand$species$time))
times <- vapply(target_times, function(x) {
  recorded_times[which.min(abs(recorded_times - x))]
}, numeric(1))

# find closest times to a given vector of times for labelling selected size distributions
find_closest <- function(x, at){
  out <- rep("", length(x))

  for(a in at){
    dist <- abs(x-a)
    i <- which(dist == min(dist))
    out[i] <- a
  }
  out
}

data_sub <- tidy_stand$species %>%
  mutate(my_label = find_closest(round(time, 2), at = round(times, 2))) %>%
  na.omit() %>%
  filter(time %in% times)

data_sub2 <- data_sub %>% group_by(time) %>% slice(1) %>% ungroup()

tidy_stand$species %>%
  ggplot(aes(height, density, group=time)) +
  geom_line() +
  geom_line(data = data_sub, col="blue") +
  geom_text(data = data_sub2, aes(label = my_label), nudge_y = 0.05, col="blue") +
  xlab("height (m)") +
  ylab("density " ~(m^-2)) +
  ggtitle("Size distributions over time") +
  theme_classic(base_size = 18) +
  theme(plot.title = element_text(hjust = 0.5))
```

Young patches contain many similarly sized saplings. As competition becomes
asymmetric, density falls and the range of heights broadens because taller
plants retain more light. By the oldest highlighted age, few plants remain and
their height distribution is narrower near the top of the canopy.
