---
title: "Strategies & traits"
---
::: {.eyebrow}
guide
:::

How to inspect and modify the parameters that define a plant strategy, and how
trait values propagate through the hyper-parameterisation. For what the
strategies mean physiologically, see [the model docs](../theory/models/index.qmd).

```{r}
library(plant)
```

There are a large number of parameters to the physiological model,
but all are changeable.  The default strategy is detailed in the
[model docs](../theory/models/index.qmd):

```{r}
s <- FF16_Strategy()
names(s)
```

The strategy object here is a special object of class `r class(s)`.
In contrast with most of the "reference" objects used by `plant`,
this is simply a list with a class attribute.  Some validation will
be done on the parameters every time it is passed into the C++ bits
of the model.

The physiological parameters now live in the nested `pars` element (the
top-level strategy also carries `control`, birth-rate settings, and a few
flags). All of the `pars` values are floating point (decimal) numbers:

```{r}
unlist(s$pars)
```

All of these parameters can be directly changed.  For example, we
can double the leaf mass per unit leaf area (lma) value:

```{r}
s$pars$lma <- s$pars$lma * 2
```

and then from this construct a plant:

```{r}
pl <- FF16_Individual(s)
pl$strategy$pars$lma
```

`lma` affects a few places in the model; see [the source
code](https://github.com/traitecoevo/plant/blob/master/src/ff16_strategy.cpp),
but only really for converting from leaf area to leaf mass (leaf
mass being leaf area multiplied by leaf mass per unit leaf area).
However, as a component of the leaf economic spectrum we imagine
LMA as affecting a number of other components of the model.

We capture this through what we call a "hyper-parameterisation";
additional parameters and functions that mean that changing one
parameter might affect a number of other lower-level parameters.
Our default hyper-parameterisation is via the
`FF16_hyperpar` function:

```{r}
FF16_hyperpar
```

You will rarely need to call this function directly (see below) but
note how setting of lma affects parameters `k_l`, `a_p1` and `r_l`

```{r}
FF16_hyperpar(trait_matrix(0.1, "lma"), s)
```

These are:

  * `k_l`: Turnover rate for leaves
  * `a_p1`: Leaf photosynthesis per area
  * `r_l`: Leaf respiration per mass

This means that it is possible to implement different trade-offs
between parameters relatively easily by modifying the
hyper-parameterisation function in R, rather than having to modify
the underlying physiological model in C++.

The `trait_matrix` function is a simple wrapper that just makes
sure the trait matrix has the right format:

```{r}
trait_matrix(0.1, "lma")
trait_matrix(c(0.1, 500), "rho")
```

To make use of the hyper-parameterisation, the preferred way of
setting parameters is through the utility functions `generate_strategy` and
`add_strategies`.  These take a `Parameters` object:

```{r}
p <- FF16_Parameters()
```

The Parameters object mostly contains information about the patch:

```{r}
names(p)
```

The default hyper-parameterisation function associated with a `Parameters`
object is retrieved with `param_hyperpar()`:

```{r}
identical(param_hyperpar(p), FF16_hyperpar)
```

`max_patch_lifetime` sets the patch disturbance interval (in years). The
physiological parameters (including `k_I`, the light extinction coefficient)
now live on the strategy under `pars`; the `strategy_default` member is the
Strategy that all others are built from:

```{r}
class(p$strategy_default)
```

This is the Strategy object that all others will be built from by
difference.  Running

```{r}
s <- generate_strategy(p, trait_matrix(0.1, "lma"), birth_rate = 1)[[1]]
```

will create a strategy `s` where lma is set but also all the
parameters that *depend* on lma.

The function `generate_strategy` can be used to create a list of
strategies:

```{r}
lma <- trait_matrix(seq(0.1, 0.5, length.out=5), "lma")
FF16_hyperpar(trait_matrix(lma, "lma"), s)

ss <- generate_strategy(p, lma, birth_rate = rep(1, 5))
length(ss)
```

We can then use standard R commands to extract variable from this list

```{r}
sapply(ss, function(x) x$pars$lma)
sapply(ss, function(x) x$pars$k_l)
sapply(ss, function(x) x$pars$a_p1)
sapply(ss, function(x) x$pars$r_l)
```

In addition to the physiological parameters there are a large number
of "control" parameters that affect the behaviour of the various
numerical algorithms used (note that in contrast to the
physiological parameters these have a variety of types). These live in a
separate `Control` object rather than on the `Parameters`:

```{r}
Control()
```

The `Control()` defaults are the pragmatic, fast-ish settings used for
essentially all of plant's runs (`control()` is a lowercase alias for the same
constructor). When you need a high-accuracy run, `control_accurate()` tightens
the ODE and schedule tolerances. We can see which fields it changes:

```{r}
ctrl_accurate <- control_accurate()
ctrl_accurate[unlist(control()) != unlist(ctrl_accurate)]
```
