---
title: "Strategies & traits"
---
::: {.eyebrow}
guide
:::

A strategy contains the biological settings for one type of plant. This guide
shows two ways to create one: change a parameter directly, or provide measured
trait values and let `plant` update the linked parameters. For the equations
behind each strategy, see [Models](../theory/models/index.qmd).

```{r}
library(plant)
```

## Inspect or directly change a strategy

Start with the default FF16 strategy. Printing its names shows that the object
contains biological parameters, numerical controls, recruitment settings, and
other model information:

```{r}
s <- FF16_Strategy()
names(s)
```

The physiological parameters live in the nested `pars` element. They are
ordinary decimal numbers and can be inspected with familiar R tools:

```{r}
unlist(s$pars)
```

For a controlled experiment, you can change one value directly. Here we double
leaf mass per unit leaf area (`lma`):

```{r}
s$pars$lma <- s$pars$lma * 2
```

An individual constructed from this strategy uses the revised value:

```{r}
pl <- FF16_Individual(s)
pl$strategy$pars$lma
```

This direct edit changes only `lma`. That is useful when deliberately varying
one model parameter, but biological traits are often linked. In FF16, LMA
affects leaf mass and seedling allometry directly and is also associated with
leaf turnover and respiration.

## Create strategies from traits

`plant` represents these links with a **hyperparameterisation**: an R function
that translates a small set of input traits into a consistent set of model
parameters. FF16 uses `FF16_hyperpar()`:

```{r}
FF16_hyperpar
```

You will rarely call this function directly. Looking at its output is useful,
however, because it makes the parameter links explicit. Supplying LMA updates
leaf turnover (`k_l`) and leaf respiration per mass (`r_l`). The output also
includes photosynthesis per leaf area (`a_p1`), which is derived from leaf
nitrogen per area (`narea`) rather than LMA itself.

```{r}
FF16_hyperpar(trait_matrix(0.1, "lma"), s)
```

The three returned quantities are:

  * `k_l`: Turnover rate for leaves
  * `a_p1`: Leaf photosynthesis per area
  * `r_l`: Leaf respiration per mass

Because these linking rules live in R, advanced users can define a different
trait trade-off without changing the underlying physiological model in C++.

`trait_matrix()` puts one or more trait values into the row-and-column format
expected by the strategy-building functions:

```{r}
trait_matrix(0.1, "lma")
trait_matrix(c(0.1, 500), "rho")
```

The preferred workflow uses `generate_strategy()` or `add_strategies()`. Both
start from a `Parameters` object, which collects the settings needed for a
patch run and stores the default strategy from which new strategies are built:

```{r}
p <- FF16_Parameters()
```

Its top-level names show the main groups of settings:

```{r}
names(p)
```

`param_hyperpar()` retrieves the linking function associated with the model:

```{r}
identical(param_hyperpar(p), FF16_hyperpar)
```

The `strategy_default` member is the template copied when new strategies are
generated. Physiological parameters, including the light extinction
coefficient `k_I`, live under its `pars` element:

```{r}
class(p$strategy_default)
```

Generating a strategy copies that template, sets LMA, and updates the linked
parameters:

```{r}
s <- generate_strategy(p, trait_matrix(0.1, "lma"), birth_rate = 1)[[1]]
```

The same function can generate several strategies at once. Each row of the
trait matrix becomes one strategy, so the birth-rate vector also has one value
per row:

```{r}
lma <- trait_matrix(seq(0.1, 0.5, length.out=5), "lma")
FF16_hyperpar(trait_matrix(lma, "lma"), s)

ss <- generate_strategy(p, lma, birth_rate = rep(1, 5))
length(ss)
```

Standard R functions can compare the resulting parameter values:

```{r}
sapply(ss, function(x) x$pars$lma)
sapply(ss, function(x) x$pars$k_l)
sapply(ss, function(x) x$pars$a_p1)
sapply(ss, function(x) x$pars$r_l)
```

## Choose numerical accuracy

Strategies describe plants; `Control` settings describe how the numerical
solver runs. The controls include time-step, error-tolerance, and node-schedule
settings, and may be numeric or logical:

```{r}
Control()
```

The `Control()` defaults are appropriate for routine exploration; `control()`
is a lowercase alias for the same constructor. Use `control_accurate()` when a
final result needs tighter numerical tolerances, accepting the additional run
time. The [ODE stepping](../theory/implementation/ode-stepping.qmd) and
[node-spacing](../theory/implementation/node-spacing.qmd) pages explain these
trade-offs. The following comparison shows which defaults change:

```{r}
ctrl_accurate <- control_accurate()
ctrl_accurate[unlist(control()) != unlist(ctrl_accurate)]
```
