---
title: "TF24: hydraulics, and a fast acclimating variant"
---

::: {.eyebrow}
guide
:::

TF24 represents the movement of water from soil, through roots and stems, to
leaves. At each rate calculation it finds the root-collar water potential that
best balances two outcomes: opening stomata increases photosynthesis, but more
negative water potential increases hydraulic risk.

The standard TF24 model solves this optimisation from scratch each time. TF24f
("f" for fast or forecasting) instead carries the operating point forward and
allows it to adjust as conditions change. In practical terms, **TF24 repeatedly
finds the optimum, whereas TF24f tracks it**.

Tracking can reduce repeated calculations and also introduces a finite
acclimation time. This guide explains that approximation, compares the two
models, and shows the parameter that controls tracking speed.

```{r}
#| label: setup
library(plant)
library(dplyr)
library(tidyr)
library(ggplot2)

base <- function(type) {
  scm_base_parameters(type) |>
    add_strategies(trait_matrix(0.0825, "lma"), birth_rate = 20)
}
```

## The idea: let the optimum chase a moving target

As a plant grows and its environment changes, the preferred operating point
moves. TF24f stores its current estimate as an extra state variable and adjusts
that state in the direction that increases carbon profit:

$$\frac{\mathrm{d}\psi}{\mathrm{d}t} = k \,\frac{\partial\,\text{profit}}{\partial\psi}.$$

At the optimum, the profit gradient is zero and the state stops moving. When the
optimum shifts, the gradient moves the state towards its new value. The gain
$k$ controls how quickly that adjustment occurs and therefore how much
acclimation lag remains.

::: {.callout-note}
## How the gradient is calculated

By default, `plant` calculates the gradient exactly with automatic
differentiation and the implicit-function theorem (`use_ad_gradient = TRUE`). A
centred finite-difference calculation is available as a fallback. In R,
`k_acclim` controls the gain, `use_ad_gradient` selects the method, and
`psi_fd_step` matters only on the finite-difference path.
:::

## Run TF24 and TF24f side by side

The following example gives both models the same trait and recruitment inputs.
For TF24f we use an illustrative gain of 2; the package default is 1.

```{r}
#| label: run-both
tf24  <- run_scm(base("TF24"),  collect = TRUE, refine_schedule = FALSE)

p_f <- base("TF24f")
s <- p_f$strategies[[1]]
s$k_acclim <- 2
s$use_ad_gradient <- TRUE
p_f$strategies[[1]] <- s
tf24f <- run_scm(p_f, collect = TRUE, refine_schedule = FALSE)
```

In this example, the two size structures remain very similar:

```{r}
#| label: size-distributions
#| fig-alt: "Node height trajectories for TF24 and TF24f, which closely overlap."
bind_rows(
  tf24$species  |> drop_na() |> mutate(model = "TF24 (optimised)"),
  tf24f$species |> drop_na() |> mutate(model = "TF24f (tracked)")
) |>
  plot_size_distribution() +
  facet_wrap(~model) +
  coord_cartesian(expand = FALSE) +
  labs(title = "Same biology, two ways of resolving the leaf optimum")
```

TF24f records the tracked potential in `opt_root_psi_state`. TF24 reports its
newly optimised value as `opt_root_psi`, with the opposite sign convention in
this output. Plotting them together reveals the acclimation lag: early in life,
the optimum moves faster than the finite-gain state can follow. The gap then
narrows as growth slows.

```{r}
#| label: tracking
#| fig-alt: "Tracked collar potential closely follows TF24's per-step optimum."
node1 <- function(d) d |> filter(node == 1) |> drop_na()
bind_rows(
  node1(tf24$species)  |> transmute(time, psi = -opt_root_psi,       model = "TF24 optimum"),
  node1(tf24f$species) |> transmute(time, psi = opt_root_psi_state, model = "TF24f tracked")
) |>
  ggplot(aes(time, psi, colour = model)) +
  geom_line() +
  labs(x = "Patch age (years)", y = "Root-collar potential (MPa)",
       colour = NULL, title = "The tracked state follows the moving optimum") +
  theme_classic()
```

## Choosing the gain `k`

A larger `k` makes the tracked state respond more quickly, but it also makes the
extra ODE state **stiffer**, which can force the adaptive solver to take many
smaller steps.
A smaller `k` is easier to integrate but produces a longer acclimation lag.
There is therefore no universal best value: compare TF24f against TF24 over the
environmental range and time scale relevant to the analysis.

The default automatic-differentiation path avoids a finite-difference step and
its truncation error. If `use_ad_gradient = FALSE`, TF24f instead uses a centred
finite difference controlled by `psi_fd_step`; that fallback should be treated
as a numerical sensitivity setting and checked explicitly.

## A note on initialisation

Each new seedling starts with its tracked state at the current optimum. Without
this initial full solve, a seedling would begin at a non-physical value and
could temporarily escape the light limitation that should suppress it. After
initialisation, TF24f resumes tracking rather than re-optimising at every step.

## Status

TF24f currently tracks root-collar potential and inherits TF24's selectable
canopy models. It is an approximation with an explicit acclimation lag, not an
exact replacement for TF24.
