---
title: "TF24: carbohydrate storage, buffered growth and mortality"
---

::: {.eyebrow}
guide
:::

Trees can use stored non-structural carbohydrates (NSC) to buffer periods when
photosynthesis does not meet demand. Growth can slow while reserves decline,
and mortality risk can rise gradually rather than responding immediately to
one poor day or season.

TF24 now represents this buffer with an explicit carbon-storage state. Earlier
versions linked growth and mortality directly to instantaneous net production.
During a severe productivity trough, that formulation could produce mortality
rates near $10^{32}$ and overflow the size-density equations (issue #550).

This guide follows carbon into and out of the storage pool, then uses a simple
drought experiment to show its effects. The governing idea is:

> a plant should not grow unless it has ample carbon in storage.

```{r}
#| label: setup
#| message: false
library(plant)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## The carbon flow

Each plant carries a storage pool $S$ (kg of NSC) as an extra ODE state. Net mass
production $P$ is assimilation after respiration and turnover costs. Current
positive production supplies growth and reproduction, but the fraction used is
controlled by how full the reserve pool is. Production not used for growth
refills storage:

$$
\frac{\mathrm{d}S}{\mathrm{d}t} = P - U,
\qquad
U = P^{+}\,G(r),
$$

where $U$ is the carbon actually committed to growth **and** reproduction (split
between them by TF24's usual allocation fractions), $P^{+}$ is a smooth
positive part of net production, and $G(r)$ is a **reserve gate** on the relative
fullness of the pool.

### Capacity and relative reserves

Storage capacity scales with sapwood mass — NSC is held largely in living
sapwood parenchyma:

$$
S_{\max} = a_{\mathrm{st1}}\, m_{s},
\qquad
r = \frac{S}{S_{\max}} \in [0, 1],
$$

with $m_s$ the sapwood mass and $a_{\mathrm{st1}}$ the storable fraction of it
(default `0.10`). $r$ is the plant's relative reserve level, and it is the single
signal that both the growth gate and the mortality term read.

### The reserve gate on growth

Growth proceeds at the **production rate**, scaled by a smooth logistic gate that
is $\approx 0$ when reserves are low and $\approx 1$ when they are ample:

$$
G(r) = \frac{1}{1 + \exp\!\big(-(r - a_{\mathrm{st2}})/w\big)},
\qquad
P^{+} = \tfrac{1}{2}\Big(P + \sqrt{P^{2} + \varepsilon^{2}}\Big).
$$

$a_{\mathrm{st2}}$ is the reserve fraction at which growth is half-on (default
`0.10`), $w$ a fixed width, and $P^{+}$ a differentiable stand-in for
$\max(P, 0)$ that removes the old hard `net > 0` growth cutoff. Carbon not
committed to growth and reproduction stays in $S$ and refills the pool.

A deliberate modelling choice sits here. Growth is gated *at the production rate*,
not metered out of the pool as a discharge $a_{\mathrm{st2}}\,S$ (the literal
"utilisation" of Stefaniak et al. 2026). With capacity tied to sapwood mass, a
seedling's pool is tiny relative to its high specific productivity — a full pool
can fund only a few percent of its production rate — so a discharge-rate limit
would choke establishment and collapse reproduction under light competition.
Gating the *rate* instead preserves healthy growth while still enforcing "no
growth without reserves".

### Reserves-based mortality

The growth-dependent mortality term now reads relative reserves instead of
instantaneous productivity:

$$
d(r) = d_I + a_{\mathrm{dG1}}\,\exp(-a_{\mathrm{dG2}}\, r).
$$

Because $r \in [0,1]$, total mortality is bounded between
$d_I + a_{\mathrm{dG1}} e^{-a_{\mathrm{dG2}}}$ and
$d_I + a_{\mathrm{dG1}}$. Excess mortality is lowest at full reserves and rises
smoothly to the finite maximum $a_{\mathrm{dG1}}$ as the pool empties. The old
runaway is therefore impossible by construction, and mortality changes
gradually as reserves decline.

A newborn seedling is seeded with reserves at a fraction $a_{\mathrm{st3}}$ of its
capacity (default `0.8`), so it can establish rather than dying immediately with
an empty pool.

```{r}
#| label: mortality-bounded
#| fig-alt: "Storage-dependent mortality is bounded between ~0 and a_dG1 as reserves vary 0 to 1."
p <- TF24_Strategy()$pars
tibble(r = seq(0, 1, length.out = 200)) |>
  mutate(mortality = p$d_I + p$a_dG1 * exp(-p$a_dG2 * r)) |>
  ggplot(aes(r, mortality)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = p$d_I + p$a_dG1, linetype = "dashed") +
  annotate("text", x = 0.5, y = p$d_I + p$a_dG1, vjust = -0.5,
           label = "bounded maximum = d_I + a_dG1") +
  labs(x = "Relative reserves  r = S / S_max",
       y = "Instantaneous mortality rate (/yr)",
       title = "Reserves-based mortality is bounded",
       subtitle = "Empty reserves give a finite ~5.5/yr, not the old ~1e32 spike") +
  theme_classic()
```

## Behaviour: growth pauses in drought and resumes on recovery

The clearest way to see the buffer is to drive one plant through a deliberately
simple drought and recovery. The helper below reproduces the C++ storage capacity
($S_{\max} = a_{\mathrm{st1}}\, m_s$) so we can report the relative reserve $r$
alongside height and mortality.

```{r}
#| label: capacity-helper
storage_capacity <- function(s, height) {
  p <- s$pars
  area_leaf    <- (height / p$a_l1)^(1 / p$a_l2)
  eta_c        <- 1 - 2 / (1 + p$eta) + 1 / (1 + 2 * p$eta)
  mass_sapwood <- (area_leaf * p$theta) * height * eta_c * p$rho
  p$a_st1 * mass_sapwood
}

fixed_env <- function(theta) {
  e <- Environment("TF24")
  e$set_soil_number_of_depths(5)
  e$set_soil_water_state(rep(theta, 5))
  e$set_fixed_environment(1.0, 40)   # full light to 40 m
  e
}
```

For transparency, this diagnostic uses a small manual Euler step rather than
the production ODE solver. It follows three years in well-watered soil, a sharp
drought from year 3 to 4.5, and then recovery. Use `run_scm()` for model runs;
this loop is only a teaching probe.

```{r}
#| label: drought-probe
#| fig-alt: "Height growth pauses during drought while reserves deplete and mortality rises to a bounded level; growth resumes after reserves refill."
s   <- TF24_Strategy()
wet <- fixed_env(0.30)
dry <- fixed_env(0.08)

pl <- TF24_Individual(s)
h0 <- pl$state("height")
pl$set_state("storage", s$pars$a_st3 * storage_capacity(s, h0))

dt   <- 0.02
tmax <- 6
rows <- vector("list", round(tmax / dt) + 1)
for (i in seq_along(rows)) {
  t <- (i - 1) * dt
  env <- if (t >= 3 && t < 4.5) dry else wet
  pl$compute_rates(env)
  h <- pl$state("height")
  rows[[i]] <- tibble(
    time      = t,
    height    = h,
    dheight   = pl$rate("height"),
    reserves  = pmax(pl$state("storage"), 0) / storage_capacity(s, h),
    mortality = pl$rate("mortality"),
    soil_water_state = env$get_soil_water_state()[1]
  )
  pl$set_state("height",  h + dt * pl$rate("height"))
  pl$set_state("storage", pl$state("storage") + dt * pl$rate("storage"))
}
traj <- bind_rows(rows)

traj |>
  pivot_longer(c(dheight, reserves, mortality,soil_water_state)) |>
  mutate(name = recode(name,
    dheight   = "Height growth (m/yr)",
    reserves  = "Relative reserves r",
    mortality = "Mortality rate (/yr)",
    soil_water_state = "Soil moisture (m3/m3)")) |>
  ggplot(aes(time, value)) +
  annotate("rect", xmin = 3, xmax = 4.5, ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "brown") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(x = "Time (years)", y = NULL,
       title = "Storage buffers the drought (shaded)",
       subtitle = "Growth pauses and reserves draw down; mortality rises to a bounded rate; growth resumes once reserves refill") +
  theme_classic()
```

The storage model produces three linked responses during the shaded drought:

- **Height growth and reproduction pause.** As reserves fall below the gate
  threshold, their shared growth flux drops smoothly towards zero. The
  turnover-driven conversion of sapwood to heartwood continues independently.
- **Reserves deplete, mortality rises — but stays bounded.** The mortality rate
  climbs toward its finite maximum ($d_I + a_{\mathrm{dG1}} \approx 5.5$/yr), so
  death is gradual over the drought rather than an instantaneous overflow.
- **Recovery is buffered, not instantaneous.** When water returns, the reserve
  gate directs much of the carbon towards refilling the pool. Height growth
  strengthens as reserves recover rather than switching on all at once.

## The growth threshold `a_st2`

`a_st2` is the reserve fraction at which the growth gate is half open. Raising it
requires fuller reserves before growth proceeds strongly; lowering it allows
growth at lower reserve levels. In this particular setup, a well-watered plant
settles near $r \approx 0.45$, so thresholds up to roughly 0.2 mainly affect the
drought response. These values are model outcomes, not universal biological
thresholds. The default is `0.10`.

```{r}
#| label: params
s <- TF24_Strategy()
s$pars[c("a_st1", "a_st2", "a_st3")]
```

| parameter | meaning | default |
|---|---|---|
| `a_st1` | storage capacity per unit sapwood mass (kg NSC / kg) | 0.10 |
| `a_st2` | reserve fraction at which growth is half-on | 0.10 |
| `a_st3` | initial storage at birth, as a fraction of capacity | 0.80 |

## Smoother soil moisture variation

The sharp drought makes the sequence easy to see but is not intended as a
realistic forcing trajectory. The next diagnostic changes soil moisture
smoothly through the same interval. It uses the same manual integration solely
to illustrate how reserve buffering responds to a gradual environmental change.

```{r}
#| label: drought-probe-smooth
#| fig-alt: "Height growth pauses during drought while reserves deplete and mortality rises to a bounded level; growth resumes after reserves refill."
s   <- TF24_Strategy()
wet <- fixed_env(0.30)
dry <- fixed_env(0.08)

pl <- TF24_Individual(s)
h0 <- pl$state("height")
pl$set_state("storage", s$pars$a_st3 * storage_capacity(s, h0))

dt   <- 0.02
tmax <- 6
rows <- vector("list", round(tmax / dt) + 1)
for (i in seq_along(rows)) {
  t <- (i - 1) * dt
  env <- if (t >= 3 && t < 4.5) fixed_env(0.3 + 0.22*(16/9)*(t - 3)*(t - 4.5)) else wet
  pl$compute_rates(env)
  h <- pl$state("height")
  rows[[i]] <- tibble(
    time      = t,
    height    = h,
    dheight   = pl$rate("height"),
    reserves  = pmax(pl$state("storage"), 0) / storage_capacity(s, h),
    mortality = pl$rate("mortality"),
    soil_water_state = env$get_soil_water_state()[1]
  )
  pl$set_state("height",  h + dt * pl$rate("height"))
  pl$set_state("storage", pl$state("storage") + dt * pl$rate("storage"))
}
traj <- bind_rows(rows)

traj |>
  pivot_longer(c(dheight, reserves, mortality,soil_water_state)) |>
  mutate(name = recode(name,
    dheight   = "Height growth (m/yr)",
    reserves  = "Relative reserves r",
    mortality = "Mortality rate (/yr)",
    soil_water_state = "Soil moisture (m3/m3)")) |>
  ggplot(aes(time, value)) +
  annotate("rect", xmin = 3, xmax = 4.5, ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "brown") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(x = "Time (years)", y = NULL,
       title = "Storage buffers the drought (shaded)",
       subtitle = "Growth pauses and reserves draw down; mortality rises to a bounded rate; growth resumes once reserves refill") +
  theme_classic()
```

## Status

The storage formulation keeps mortality bounded and produces the qualitative
growth-and-mortality buffering shown above. Its parameter values are
**prototype defaults, not calibrated estimates**. In particular,
$a_{\mathrm{st1}}$, $a_{\mathrm{st2}}$, $a_{\mathrm{st3}}$, and the
reserve-to-mortality relationship still need to be evaluated against suitable
observations. TF24f inherits the same storage pool.

An independent soil-water singularity at the residual moisture floor was also
fixed while testing this regime (issue #549; PR #554).
