---
title: "Stochastic assembly"
---
::: {.eyebrow}
theory
:::

This vignette runs **stochastic** community assembly with `regnans` on the full
`plant` FF16 model. Where the [max-fitness](assembly_fitmax.qmd) method always
introduces the single best mutant, stochastic assembly draws each invader at
random (here from a Gaussian mutation kernel `vcv`), so the community builds up as
a noisy evolutionary walk rather than a deterministic climb. We show a one-trait
(`lma`) and a two-trait (`lma` + `hmat`) example.

```{r setup}
#| message: false
library(plant)
library(regnans)
library(tidyverse)

model_support <- list(
  p = plant_default_assembly_pars(),
  plant_control = plant_default_assembly_control()
)
```

# One trait: lma

## Assembly

```{r lma-assembly}
control <- assembler_control(
  list(
    run_type = "single_step",
    birth_type = "stochastic",
    vcv = diag(1)
  )
)

community0 <- community_start(bounds(lma = c(0.01, 1)), model_support = model_support)
obj <- assembler_run(assembler_start(community0, control = control), nsteps = 60)

obj$community$traits
obj$community$birth_rate
```

## The assembly walk

`tidy_assembly()` gives one row per resident per step; unnesting the traits shows
the stochastic build-up of the community over assembly steps (point size ∝ birth
rate):

```{r lma-walk}
#| fig-width: 7
#| fig-height: 4
tidy_assembly(obj) |>
  filter(resident) |>
  unnest(traits) |>
  mutate(step = as.integer(step)) |>
  ggplot(aes(step, lma, size = births)) +
  geom_point(alpha = 0.4, colour = "forestgreen") +
  scale_y_log10() +
  labs(x = "assembly step", y = "lma", size = "birth rate") +
  theme_classic()
```

## Fitness landscape

Stochastic runs don't compute a fitness landscape by default, but we can evaluate
it for the endpoint community (on a grid) and confirm the residents (red) sit at
or near fitness zero with no strong invasion peaks left:

```{r lma-landscape}
#| fig-width: 7
#| fig-height: 4
obj$community <- community_fitness_landscape(obj$community, method = "grid", n_evals = 100)
fp <- obj$community$fitness_points
ggplot(fp, aes(lma, fitness)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(data = filter(fp, resident), colour = "red", size = 2) +
  scale_x_log10() +
  theme_classic()
```

# Two traits: lma + hmat

The same method with a two-dimensional mutation kernel assembles a community in
the `lma`–`hmat` plane.

```{r hmat-assembly}
control2 <- assembler_control(
  list(
    run_type = "single_step",
    birth_type = "stochastic",
    compute_viable_fitness = FALSE,
    vcv = diag(2)
  )
)

community0 <- community_start(bounds(lma = c(0.01, 1), hmat = c(0.5, 30)),
                             model_support = model_support)
obj2 <- assembler_run(assembler_start(community0, control = control2), nsteps = 10)

obj2$community$traits
obj2$community$birth_rate
```

The assembled two-trait community (point size ∝ birth rate):

```{r hmat-plot}
#| fig-width: 5.5
#| fig-height: 4.5
as_tibble(obj2$community$traits) |>
  mutate(birth_rate = obj2$community$birth_rate) |>
  ggplot(aes(lma, hmat, size = birth_rate)) +
  geom_point(colour = "forestgreen", alpha = 0.6) +
  scale_x_log10() +
  labs(size = "birth rate") +
  theme_classic()
```
