---
title: "Maximum-fitness community assembly"
---
::: {.eyebrow}
theory
:::

This page uses the **maximum-fitness** method to assemble a community in the full
`plant` FF16 model. At each step, `regnans` finds the mutant trait value with the
highest invasion fitness in the light environment created by the current
residents. It then adds that mutant, returns the community to demographic
equilibrium, and removes any resident strategies that have gone extinct.

The first example allows leaf mass per area (`lma`) to evolve. The second repeats
the workflow for height at maturation (`hmat`). Only one trait varies in each
example; all other FF16 traits retain their default values.

```{r setup}
#| message: false
library(plant)
library(regnans)
library(dplyr)
library(ggplot2)
```

# Leaf mass per area (`lma`)

## Assembly

```{r lma-assembly}
# At each step, introduce the mutant with the highest invasion fitness and
# return the community to demographic equilibrium.
control <- assembler_control(
  list(
    run_type = "to_equilibrium",
    birth_type = "maximum",
    birth_move_tol = 1,
    compute_viable_fitness = FALSE
  )
)

# Start with an empty community using the full plant FF16 model.
community0 <-
  community_start(
    bounds = bounds(lma = c(0.01, 2)),
    fitness_control = list(method = "bayesopt", n_evals = 10, n_init = 5),
    demography_control = demographic_step_control(
      list(equilibrium_solver_name = "equilibrium_iteration", equilibrium_nsteps = 10)),
    model_support = list(
      p = plant_default_assembly_pars(max_patch_lifetime = 30),
      plant_control = plant::control())
  )

obj <- assembler_run(assembler_start(community0, control = control), nsteps = 10)
```

## Analysis

Inspect the trait values and birth rates of the resident strategies that remain
after assembly:

```{r lma-result}
obj$community$traits
obj$community$birth_rate
```

Next, plot invasion fitness across the allowed `lma` range. Resident strategies
are shown as points. At the end of a successful assembly run, they should lie at
fitness peaks near zero, with no positive peak at which another mutant could
invade:

```{r lma-landscape}
community_plot_fitness_landscape(obj$community)
```

`tidy_assembly()` returns one row for every strategy at every assembly step. It
keeps the traits, birth rates, and fitness landscape for each step in list
columns, which is useful for making custom plots of the assembly history:

```{r lma-history}
tidy_assembly(obj)
```

# Height at maturation (`hmat`)

The same method can be applied to height at maturation. Here `hmat` varies while
the other FF16 traits remain fixed at their defaults.

```{r hmat-assembly}
community0 <-
  community_start(
    bounds = bounds(hmat = c(0.5, 30)),
    fitness_control = list(method = "bayesopt", n_evals = 10, n_init = 5),
    demography_control = demographic_step_control(
      list(equilibrium_solver_name = "equilibrium_iteration", equilibrium_nsteps = 10)),
    model_support = list(
      p = plant_default_assembly_pars(max_patch_lifetime = 30),
      plant_control = plant::control())
  )

obj_h <- assembler_run(assembler_start(community0, control = control), nsteps = 10)

obj_h$community$traits
obj_h$community$birth_rate
```

```{r hmat-landscape}
community_plot_fitness_landscape(obj_h$community)
```

<!-- SY: Explain this figure more? -->