---
title: "Community assembly with simple models"
subtitle: "From one successful mutant to a community of coexisting strategies"
format: html
---

A pairwise invasibility plot (PIP) asks whether one mutant can invade a community
containing one resident strategy. Community assembly goes further by allowing
several resident strategies to coexist.

At each assembly step, the algorithm:

1. finds the mutant with the highest invasion fitness;
2. adds that mutant to the resident community;
3. returns the community to demographic equilibrium; and
4. removes any resident strategies whose populations fall to zero.

Repeating these steps shows how evolutionary branching and competition can
build a multi-strategy community. The small models in `regnans` make this process
quick enough to explore interactively.

```{r setup}
#| message: false
library(regnans)
```

## A note on trait scale

The assembly algorithm must search across possible trait values. The scale used
for that search matters:

- use a **log scale** for strictly positive traits, such as `plant`'s leaf mass
  per area (`lma`) or seed size in GM99;
- use a **linear scale** for traits that can be zero or negative, as in DD99 and
  GK98.

Set the scale with `community_start(trait_scale = ...)`. The default is `"log"`.

## DD99 — limiting similarity in one dimension

When competition covers a narrower trait range than the resource distribution
(`sigma_C < sigma_K`), the resource optimum is a branching point. Repeated
branching fills trait space with coexisting strategies. Strategies that are too
similar compete too strongly to coexist, so the survivors are spaced apart.
This spacing is called **limiting similarity**.

```{r dd99-1d}
#| fig-width: 6
#| fig-height: 3.4
h <- harness_dd99(x0 = 0, sigma_K = 1, sigma_C = 0.4)
comm <- community_start(bounds(x = c(-3, 3)), harness = h, trait_scale = "linear",
                        fitness_control = list(method = "grid", n_evals = 150))
a <- assembler_start(comm, assembler_control(list(birth_type = "maximum")))
a <- assembler_run(a, 14)

cat(length(a$community), "coexisting species\n")
xs <- sort(as.numeric(a$community$traits))
plot(xs, rep(1, length(xs)), pch = 19, col = "forestgreen", yaxt = "n",
     ylab = "", xlab = "trait x", xlim = c(-3, 3),
     main = sprintf("DD99 assembly: %d species (limiting similarity)", length(xs)))
```

## GK98 — branching to a protected dimorphism

Soft selection keeps the model's total population fixed, so an empty community
cannot be colonised in the usual way. We instead **seed** the community with one
strategy and allow it to evolve. At `d/sigma = 1.5`, which is above the branching
threshold $\sqrt{3/2}$, the original strategy splits into two coexisting
strategies. This stable pair is called a **protected dimorphism**.

```{r gk98}
comm <- community_start(bounds(x = c(-4, 4)), harness = harness_gk98(d = 1.5, sigma = 1),
                        trait_scale = "linear",
                        fitness_control = list(method = "grid", n_evals = 150))
a <- assembler_start(comm, assembler_control(list(birth_type = "maximum")))
a <- assembler_set_traits(a, trait_matrix(0, "x"))   # seed the first type
a <- assembler_run(a, 10)

sort(round(as.numeric(a$community$traits), 3))        # a symmetric dimorphism
```

## GM99 — a seed-size polymorphism

Strong size-asymmetric competition (`alpha*R = 7`) drives branching in seed
size. Starting from an empty community produces a mixture of small- and
large-seeded strategies. GM99 evaluates a multidimensional Poisson sum when
several residents are present, so this example is slower than the other simple
models. Assembling even a few strategies can take several seconds.

```{r gm99}
#| eval: false
comm <- community_start(bounds(x = c(0.06, 0.95)), harness = harness_gm99(alpha = 7, beta = 15),
                        fitness_control = list(method = "grid", n_evals = 80))
a <- assembler_start(comm, assembler_control(list(birth_type = "maximum")))
a <- assembler_run(a, 4)
sort(round(as.numeric(a$community$traits), 3))   # e.g. 0.11, 0.45, 0.69 — small, medium, large seeds
```

## DD99 in two trait dimensions

`harness_dd99_nd()` extends DD99 to more than one trait. The resource and
competition curves are products of separate Gaussian curves for each trait, so
every trait has its own optimum and curve widths. The joint singular strategy is
`x* = x0`. Evolution branches along any trait for which `sigma_C < sigma_K`.

The same `assembler_run()` function handles two-dimensional assembly. We only
need to provide bounds for two traits and use the multi-trait harness:

```{r dd99-2d}
#| fig-width: 5
#| fig-height: 5
h2 <- harness_dd99_nd(x0 = c(0, 0), sigma_K = c(1, 1), sigma_C = c(0.5, 0.5))
comm <- community_start(bounds(x1 = c(-3, 3), x2 = c(-3, 3)), harness = h2,
                        trait_scale = "linear",
                        fitness_control = list(method = "grid", n_evals = 40))
a <- assembler_start(comm, assembler_control(list(birth_type = "maximum")))
a <- assembler_run(a, 40)

cat(length(a$community), "coexisting species in 2D\n")
plot(a$community$traits[, 1], a$community$traits[, 2], pch = 19, col = "forestgreen",
     xlab = "trait x1", ylab = "trait x2", xlim = c(-3, 3), ylim = c(-3, 3),
     main = sprintf("DD99 2D assembly: %d species", length(a$community)))
```

The resulting strategies are separated across the trait plane. The competition
widths determine their characteristic spacing, extending limiting similarity to
two dimensions.
