---
title: "max_fitness assembly"
---
::: {.eyebrow}
theory
:::

This vignette runs community assembly with the **max-fitness** method in
`regnans`: at each step the single best-invading mutant (the trait that maximises
invasion fitness in the current community's light environment) is introduced, the
community is re-solved to demographic equilibrium, and any residents driven
extinct are dropped. We show a one-trait (`lma`) and a one-trait (`hmat`) example
on the full `plant` FF16 model.

```{r setup}
#| message: false
library(plant)
library(regnans)
library(dplyr)
library(ggplot2)
```

# One trait: lma

## Assembly

```{r lma-assembly}
# controls on the assembler: keep introducing the max-fitness mutant and
# re-solving to equilibrium
control <- assembler_control(
  list(
    run_type = "to_equilibrium",
    birth_type = "maximum",
    birth_move_tol = 1,
    compute_viable_fitness = FALSE
  )
)

# start from an empty community on the full plant FF16 model
community0 <-
  community_start(
    bounds = bounds(lma = c(0.01, 2)),
    fitness_control = list(method = "bayesopt", n_evals = 10, n_init = 5),
    demography_control = demographic_step_control(
      list(equilibrium_solver_name = "equilibrium_iteration", equilibrium_nsteps = 10)),
    model_support = list(
      p = plant_default_assembly_pars(max_patch_lifetime = 30),
      plant_control = plant::control())
  )

obj <- assembler_run(assembler_start(community0, control = control), nsteps = 10)
```

## Analysis

The resident traits and birth rates of the assembled community:

```{r lma-result}
obj$community$traits
obj$community$birth_rate
```

The fitness landscape of the assembled community — residents (points) sit at the
peaks/zeros, and no mutant can invade (fitness ≤ 0 everywhere):

```{r lma-landscape}
community_plot_fitness_landscape(obj$community)
```

`tidy_assembly()` returns a tidy record of the run — one row per strategy per
step, carrying each step's traits, births, and fitness landscape in list columns
for custom plotting:

```{r lma-history}
tidy_assembly(obj)
```

# One trait: height (`hmat`)

The same method on a different trait — height at maturation.

```{r hmat-assembly}
community0 <-
  community_start(
    bounds = bounds(hmat = c(0.5, 30)),
    fitness_control = list(method = "bayesopt", n_evals = 10, n_init = 5),
    demography_control = demographic_step_control(
      list(equilibrium_solver_name = "equilibrium_iteration", equilibrium_nsteps = 10)),
    model_support = list(
      p = plant_default_assembly_pars(max_patch_lifetime = 30),
      plant_control = plant::control())
  )

obj_h <- assembler_run(assembler_start(community0, control = control), nsteps = 10)

obj_h$community$traits
obj_h$community$birth_rate
```

```{r hmat-landscape}
community_plot_fitness_landscape(obj_h$community)
```
