---
title: "Stochastic assembly"
---
::: {.eyebrow}
theory
:::

This page runs **stochastic** community assembly with `regnans` and the full
`plant` FF16 model. The [maximum-fitness method](assembly_fitmax.qmd) always adds
the mutant with the highest invasion fitness. Stochastic assembly instead draws
a new mutant at random from a Gaussian mutation distribution, controlled by
`vcv`.

The sequence of mutants therefore resembles a noisy evolutionary walk rather
than a direct climb towards the highest-fitness strategy. The first example
allows leaf mass per area (`lma`) to vary. The second allows both `lma` and height
at maturation (`hmat`) to vary.

```{r setup}
#| message: false
library(plant)
library(regnans)
library(tidyverse)

model_support <- list(
  p = plant_default_assembly_pars(),
  plant_control = plant_default_assembly_control()
)
```

# Leaf mass per area (`lma`)

## Assembly

```{r lma-assembly}
control <- assembler_control(
  list(
    run_type = "single_step",
    birth_type = "stochastic",
    vcv = diag(1)
  )
)

community0 <- community_start(bounds(lma = c(0.01, 1)), model_support = model_support)
obj <- assembler_run(assembler_start(community0, control = control), nsteps = 60)

obj$community$traits
obj$community$birth_rate
```

## The assembly walk

`tidy_assembly()` gives one row for each resident strategy at each step. After
unnesting the trait values, we can plot how the community changes through time.
Larger points represent strategies with higher birth rates:

```{r lma-walk}
#| fig-width: 7
#| fig-height: 4
tidy_assembly(obj) |>
  filter(resident) |>
  unnest(traits) |>
  mutate(step = as.integer(step)) |>
  ggplot(aes(step, lma, size = births)) +
  geom_point(alpha = 0.4, colour = "forestgreen") +
  scale_y_log10() +
  labs(x = "assembly step", y = "lma", size = "birth rate") +
  theme_classic()
```

## Fitness landscape

Stochastic runs do not calculate a complete fitness landscape by default. We can
evaluate it on a grid for the final community. Resident strategies are marked in
red; they should lie at or near zero fitness, with no large positive invasion
peaks remaining:

```{r lma-landscape}
#| fig-width: 7
#| fig-height: 4
obj$community <- community_fitness_landscape(obj$community, method = "grid", n_evals = 100)
fp <- obj$community$fitness_points
ggplot(fp, aes(lma, fitness)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(data = filter(fp, resident), colour = "red", size = 2) +
  scale_x_log10() +
  theme_classic()
```

# Leaf mass per area and height at maturation

With a two-dimensional mutation distribution, both `lma` and `hmat` can change
at each step. The community is then assembled across a two-dimensional trait
space.

```{r hmat-assembly}
control2 <- assembler_control(
  list(
    run_type = "single_step",
    birth_type = "stochastic",
    compute_viable_fitness = FALSE,
    vcv = diag(2)
  )
)

community0 <- community_start(bounds(lma = c(0.01, 1), hmat = c(0.5, 30)),
                             model_support = model_support)
obj2 <- assembler_run(assembler_start(community0, control = control2), nsteps = 10)

obj2$community$traits
obj2$community$birth_rate
```

The final plot shows the assembled strategies in the `lma`-`hmat` trait space.
Larger points represent strategies with higher birth rates:

```{r hmat-plot}
#| fig-width: 5.5
#| fig-height: 4.5
as_tibble(obj2$community$traits) |>
  mutate(birth_rate = obj2$community$birth_rate) |>
  ggplot(aes(lma, hmat, size = birth_rate)) +
  geom_point(colour = "forestgreen", alpha = 0.6) +
  scale_x_log10() +
  labs(size = "birth rate") +
  theme_classic()
```
