---
title: "DD99 — Dieckmann & Doebeli 1999"
subtitle: "Competition for a Gaussian resource, and the origin of evolutionary branching"
format: html
---

## The original paper

Dieckmann, U. & Doebeli, M. (1999). *On the origin of species by sympatric
speciation.* **Nature** 400, 354–357.

A single quantitative trait `x` (e.g. beak size) sets how an individual uses a
resource. Carrying capacity is highest at an intermediate phenotype `x0` and
competition is strongest between similar phenotypes. The paper's central result is
that this simple ingredient — competition that weakens with phenotypic distance —
makes the resource optimum `x0` a **fitness minimum** once the population sits
there, so the population splits into two diverging branches. It is the canonical
demonstration of **evolutionary branching**, and (with sexual reproduction added)
of sympatric speciation.

## Fitness equations

Carrying capacity and competition kernel:

$$
K(x) = K_0 \exp\!\left(-\frac{(x-x_0)^2}{2\sigma_K^2}\right),
\qquad
C(d) = \exp\!\left(-\frac{d^2}{2\sigma_C^2}\right).
$$

For a rare mutant `y` in a resident community $\{x_i, N_i\}$ at demographic
equilibrium, the invasion fitness (a per-capita growth **rate**, zero at the
resident) is

$$
s(y) = r\left(1 - \frac{\sum_i N_i\, C(y - x_i)}{K(y)}\right).
$$

A single resident sits at $N^\* = K(x)$. The singular strategy is $x^\* = x_0$, and
it is an **evolutionary branching point iff $\sigma_C < \sigma_K$** (competition
narrower than the resource), otherwise an ESS.

## Example analysis with `regnans`

```{r setup}
#| message: false
library(regnans)
```

Build the harness with the paper's branching parameters ($\sigma_C = 0.4 <
\sigma_K = 1$) and solve for the singular strategy:

```{r singular}
h <- harness_dd99(r = 1, K0 = 500, x0 = 0, sigma_K = 1, sigma_C = 0.4)

out <- community_start(bounds(x = c(-2, 2)), harness = h) |>
  community_solve_singularity_1D()

as.numeric(out$traits)        # x* — expected 0 (= x0)
```

Is it a branching point? Check the curvature of invasion fitness at the resident
(positive = fitness minimum = disruptive selection):

```{r curvature}
resident_at <- function(h, x) {
  community_start(bounds(x = c(-2, 2)), harness = h) |>
    community_add(trait_matrix(x, "x")) |>
    community_demography()
}
curv <- function(comm, at, d = 1e-4) {
  f <- comm$fitness_function
  (f(at + d) - 2 * f(at) + f(at - d)) / d^2
}

curv(resident_at(h, 0), 0)                                   # > 0  → branching
curv(resident_at(harness_dd99(sigma_C = 1.5), 0), 0)        # < 0  → ESS
```

### Pairwise invasibility plot

Sign of $s(y)$ over (resident `x`, mutant `y`) — green where a mutant can invade.
The two off-diagonal invasion wedges around `x0 = 0` are the branching signature.

```{r pip}
#| fig-width: 5
#| fig-height: 5
xs <- seq(-2, 2, length.out = 81)
S <- sapply(xs, function(res) {
  resident_at(h, res)$fitness_function(xs)   # vector over all mutants
})                                            # S[mutant, resident]

image(xs, xs, t(sign(S)), col = c("white", "grey85", "forestgreen"),
      xlab = "resident trait x", ylab = "mutant trait y",
      main = "DD99 PIP (sigma_C = 0.4 < sigma_K = 1)")
abline(0, 1, lty = 2)
```
