---
title: "DD99: competition and evolutionary branching"
subtitle: "How local competition can split one strategy into two"
format: html
---

## What the model represents

A single quantitative trait `x`, such as beak size, determines which part of a
continuous resource an individual uses. The environment supports the largest
population at an intermediate trait value `x0`, while individuals with similar
trait values compete most strongly with one another.

These two effects can pull evolution in different directions. The resource
distribution first draws the population towards `x0`. Once the population is
there, however, strong competition among similar individuals can favour mutants
on either side. The central strategy then becomes a fitness minimum and one
lineage can split into two diverging strategies. This is **evolutionary
branching**.

The model is based on Dieckmann and Doebeli (1999), *On the origin of species by
sympatric speciation*.

## Fitness equations

The model uses two Gaussian curves:

$$
K(x) = K_0 \exp\!\left(-\frac{(x-x_0)^2}{2\sigma_K^2}\right),
\qquad
C(d) = \exp\!\left(-\frac{d^2}{2\sigma_C^2}\right).
$$

For a rare mutant `y` in a resident community $\{x_i, N_i\}$ at demographic
equilibrium, the invasion fitness (a per-capita growth **rate**, zero at the
resident) is

$$
s(y) = r\left(1 - \frac{\sum_i N_i\, C(y - x_i)}{K(y)}\right).
$$

A single resident reaches the equilibrium population size $N^\* = K(x)$. The
singular strategy is $x^\* = x_0$. Its outcome depends on the widths of the two
curves:

- if $\sigma_C < \sigma_K$, competition acts over a narrower trait range than
  the resource distribution, and $x_0$ is an evolutionary branching point;
- otherwise, $x_0$ is an evolutionarily stable strategy (ESS).

## Example analysis with `regnans`

```{r setup}
#| message: false
library(regnans)
```

Choose parameters that produce branching ($\sigma_C = 0.4 < \sigma_K = 1$),
then solve for the singular strategy:

```{r singular}
h <- harness_dd99(r = 1, K0 = 500, x0 = 0, sigma_K = 1, sigma_C = 0.4)

out <- community_start(bounds(x = c(-2, 2)), harness = h) |>
  community_solve_singularity_1D()

as.numeric(out$traits)        # x* — expected 0 (= x0)
```

Finding the singular strategy does not tell us whether it is stable. We also
check the curvature of invasion fitness at the resident. Positive curvature
means the resident lies at a fitness minimum, so selection is disruptive and
branching can occur:

```{r curvature}
resident_at <- function(h, x) {
  community_start(bounds(x = c(-2, 2)), harness = h) |>
    community_add(trait_matrix(x, "x")) |>
    community_demography()
}
curv <- function(comm, at, d = 1e-4) {
  f <- comm$fitness_function
  (f(at + d) - 2 * f(at) + f(at - d)) / d^2
}

curv(resident_at(h, 0), 0)                                   # > 0  → branching
curv(resident_at(harness_dd99(sigma_C = 1.5), 0), 0)        # < 0  → ESS
```

### Pairwise invasibility plot

A pairwise invasibility plot shows the sign of $s(y)$ for each resident `x` and
mutant `y`. Green marks combinations for which the mutant can invade. The two
green wedges on either side of `x0 = 0` show that nearby mutants can invade the
singular resident, which is the signature of branching.

```{r pip}
#| fig-width: 5
#| fig-height: 5
xs <- seq(-2, 2, length.out = 81)
S <- sapply(xs, function(res) {
  resident_at(h, res)$fitness_function(xs)   # vector over all mutants
})                                            # S[mutant, resident]

image(xs, xs, t(sign(S)), col = c("white", "grey85", "forestgreen"),
      xlab = "resident trait x", ylab = "mutant trait y",
      main = "DD99 PIP (sigma_C = 0.4 < sigma_K = 1)")
abline(0, 1, lty = 2)
```
