---
title: "GK98: adaptation across different habitats"
subtitle: "How soft selection can maintain two strategies"
format: html
---

## What the model represents

This model follows one trait across several habitats, each with a different
optimal trait value. Selection within each habitat favours individuals close to
its local optimum. Density regulation is **soft**: every habitat contributes a
fixed share of the next generation, regardless of how many individuals survive
there. This prevents the most productive habitat from dominating the whole
population.
<!-- SY: I don't really understand soft selection -->

When habitat optima are close together, evolution settles on one compromise
strategy. When they are far enough apart, the same compromise attracts gradual
evolution but can be invaded after it is reached. The population then branches
into two strategies adapted to different parts of the habitat distribution.

The model comes from Geritz et al. (1998), *Evolutionarily singular strategies
and the adaptive growth and branching of the evolutionary tree* [@Geritz-1998].

## Fitness equations

There are three habitats ($m=3$), with optima $\mu = (-d, 0, d)$. Habitat $j$
has carrying capacity $K_j$ and therefore contributes the fraction <!-- SY: Add carrying capacity to a glossary? -->
$c_j = K_j/\sum_h K_h$ of the next generation. Survival within each habitat is
a Gaussian curve with width $\sigma$:

$$
f_j(x) = \exp\!\left(-\frac{(x-\mu_j)^2}{2\sigma^2}\right).
$$

For a single resident `x`, the invasion fitness of a mutant `y` (a log per-capita
growth, which is zero at the resident) is

$$
s_x(y) = \log \sum_{j} c_j \, \frac{f_j(y)}{f_j(x)}.
$$

The singular strategy is the capacity-weighted mean optimum,
$x^\* = \sum_j c_j \mu_j$. It equals zero when the three habitats have equal
capacities. Evolution always moves towards this strategy, but what happens on
arrival depends on the separation of habitat optima: <!-- SY: On arrival of a mutant? -->

- if $d/\sigma > \sqrt{3/2} \approx 1.2247$, it is a branching point;
- below that threshold, it is a continuously stable strategy (CSS): evolution
  approaches it and nearby mutants cannot invade it.

## Example analysis with `regnans`

```{r setup}
#| message: false
library(regnans)
```

```{r singular}
h <- harness_gk98(d = 1.5, sigma = 1)   # d/sigma = 1.5 > sqrt(1.5) → branching

out <- community_start(bounds(x = c(-4, 4)), harness = h, trait_scale = "linear") |>
  community_solve_singularity()

as.numeric(out$traits)   # x* — expected 0 (symmetric three-patch)
```

We can recover the branching threshold numerically. The following code finds
where the curvature of invasion fitness at `x* = 0` changes sign:

```{r threshold}
curv0 <- function(d, sigma = 1, h = 1e-4) {
  comm <- community_start(bounds(x = c(-4, 4)),
                          harness = harness_gk98(d = d, sigma = sigma)) |>
    community_add(trait_matrix(0, "x")) |>
    community_demography()
  f <- comm$fitness_function
  (f(h) - 2 * f(0) + f(-h)) / h^2
}

uniroot(curv0, c(1.0, 1.5))$root   # ≈ sqrt(3/2)
sqrt(3 / 2)
```

### Pairwise invasibility plots either side of the threshold

The first plot uses closely spaced habitat optima and has one stable strategy.
The second uses more widely spaced optima and shows the invasion pattern around
a branching point: the second contour folds back on itself at the singular
strategy, so some residents can be invaded from three directions. The fold
appears exactly when $d/\sigma$ exceeds $\sqrt{3/2}$, the branching
condition.
<!-- SY: I think this page needs more hand holding. The first plot is a ESS because at resident x = 0, there is no mutant x that has positive fitness (i.e. is in the green areas of the graph). In the second plot, why can residents be invaded from three directions? What is the second contour and the fold? Where is $\sqrt{3/2}$? -->

```{r pip-css}
#| fig-width: 5
#| fig-height: 5
pip_at <- function(d) {
  community_start(bounds(x = c(-4, 4)), trait_scale = "linear",
                  harness = harness_gk98(d = d, sigma = 1)) |>
    community_pip()
}
plot(pip_at(1.0)) + ggplot2::ggtitle("d/sigma = 1.0 (CSS)")
```

```{r pip-branching}
#| fig-width: 5
#| fig-height: 5
plot(pip_at(1.5)) + ggplot2::ggtitle("d/sigma = 1.5 (branching)")
```
