---
title: "GK98 — Geritz, Kisdi, Meszéna & Metz 1998"
subtitle: "Soft selection across patches, and the adaptive branching of the evolutionary tree"
format: html
---

## The original paper

Geritz, S.A.H., Kisdi, É., Meszéna, G. & Metz, J.A.J. (1998). *Evolutionarily
singular strategies and the adaptive growth and branching of the evolutionary
tree.* **Evolutionary Ecology** 12, 35–57.

This is the foundational adaptive-dynamics paper — it defines invasion fitness,
the selection gradient, pairwise invasibility plots, and the classification of
**evolutionarily singular strategies** (ESS, convergence stability, branching
points). Its worked example is a haploid **Levene soft-selection** model: a single
trait under stabilising selection in each of several patches with different
optima, and *soft* density regulation (each patch contributes a fixed share to the
next generation). It is not a seed-size model — that is the separate
[GM99](GM99.qmd) paper.

## Fitness equations

Three patches ($m=3$) with optima $\mu = (-d, 0, d)$, equal capacities $K_j$
(weights $c_j = K_j/\sum_h K_h$), and Gaussian within-patch survival of width
$\sigma$:

$$
f_j(x) = \exp\!\left(-\frac{(x-\mu_j)^2}{2\sigma^2}\right).
$$

For a single resident `x`, the invasion fitness of a mutant `y` (a log per-capita
growth, zero at the resident) is

$$
s_x(y) = \log \sum_{j} c_j \, \frac{f_j(y)}{f_j(x)}.
$$

The singular strategy is the capacity-weighted mean optimum,
$x^\* = \sum_j c_j \mu_j$ (= 0 for the symmetric default). It is **always
convergence stable**, and is an evolutionary **branching point iff
$d/\sigma > \sqrt{3/2} \approx 1.2247$** (a continuously stable strategy below
that threshold).

## Example analysis with `regnans`

```{r setup}
#| message: false
library(regnans)
```

```{r singular}
h <- harness_gk98(d = 1.5, sigma = 1)   # d/sigma = 1.5 > sqrt(1.5) → branching

out <- community_start(bounds(x = c(-4, 4)), harness = h) |>
  community_solve_singularity_1D()

as.numeric(out$traits)   # x* — expected 0 (symmetric three-patch)
```

The branching threshold $d/\sigma = \sqrt{3/2}$ is recovered numerically by
finding where the invasion-fitness curvature at `x* = 0` changes sign:

```{r threshold}
curv0 <- function(d, sigma = 1, h = 1e-4) {
  comm <- community_start(bounds(x = c(-4, 4)),
                          harness = harness_gk98(d = d, sigma = sigma)) |>
    community_add(trait_matrix(0, "x")) |>
    community_demography()
  f <- comm$fitness_function
  (f(h) - 2 * f(0) + f(-h)) / h^2
}

uniroot(curv0, c(1.0, 1.5))$root   # ≈ sqrt(3/2)
sqrt(3 / 2)
```

### Pairwise invasibility plots either side of the threshold

```{r pip}
#| fig-width: 8
#| fig-height: 4.2
pip <- function(h, xs) {
  sapply(xs, function(res) {
    community_start(bounds(x = range(xs)), harness = h) |>
      community_add(trait_matrix(res, "x")) |>
      community_demography() |>
      (\(comm) comm$fitness_function(xs))()
  })
}
xs <- seq(-4, 4, length.out = 81)

op <- par(mfrow = c(1, 2))
for (d in c(1.0, 1.5)) {
  S <- pip(harness_gk98(d = d, sigma = 1), xs)
  image(xs, xs, t(sign(S)), col = c("white", "grey85", "forestgreen"),
        xlab = "resident x", ylab = "mutant y",
        main = sprintf("d/sigma = %.1f (%s)", d,
                       if (d > sqrt(1.5)) "branching" else "CSS"))
  abline(0, 1, lty = 2)
}
par(op)
```
