---
title: "JJ12 — Johansson & Jonzén 2012"
subtitle: "Evolution of migratory arrival time, and a tragedy of the commons"
format: html
---

## The original paper

Johansson, J. & Jonzén, N. (2012). *Game theory sheds new light on ecological
responses to current climate change when phenology is historically mismatched.*
**Ecology Letters** 15, 881–888.

Migratory birds compete for a fixed number of breeding territories. Arriving early
is costly (harsh conditions) but improves the chance of securing a territory;
arriving at the seasonal optimum maximises reproduction. Because securing a
territory is a *competitive* advantage, individuals are selected to arrive
**earlier than the date that would maximise population output** — an evolutionary
tragedy of the commons, and a baseline mismatch that matters when climate change
shifts the optimal date. This write-up uses the simplified, fully analytic form of
the model from Brännström, Johansson & von Festenberg (2013, *Games* 4:304–328,
§4), which makes it an exact benchmark for numerical solvers.

## Fitness equations

Competitive ability (declines with arrival time `x`) and reproduction (Gaussian
about the seasonal optimum `x_opt`):

$$
C(x) = e^{-a x},
\qquad
R(x) = R_0\,\exp\!\left(-\frac{(x-x_{\mathrm{opt}})^2}{2\sigma^2}\right).
$$

With year-to-year survival `p`, a single resident reaches
$n^\* = K R(x)/(1-p)$. The invasion fitness of a mutant `x'` (a discrete-time
geometric growth rate; $w = 1$, i.e. log = 0, at the resident) is

$$
w(x') = (1-p)\,\frac{C(x')\,R(x')}{C(x)\,R(x)} + p .
$$

The selection gradient is linear and decreasing, so there is a single
**continuously stable strategy** (no branching) with the closed form

$$
x^\* = x_{\mathrm{opt}} - a\,\sigma^2 ,
$$

which lies *before* the population optimum `x_opt` whenever `a > 0`.

## Example analysis with `regnans`

```{r setup}
#| message: false
library(regnans)
```

```{r singular}
a <- 0.125; x_opt <- 0; sigma <- 10
h <- harness_jj12(a = a, x_opt = x_opt, sigma = sigma, R0 = 1, K = 5, p = 0.5)

out <- community_start(bounds(x = c(-40, 40)), harness = h) |>
  community_solve_singularity_1D()

c(numeric = as.numeric(out$traits),
  analytic = x_opt - a * sigma^2)     # should match: -12.5
```

The numerically-solved singular strategy matches the closed form exactly — this
is the model's value as a test oracle. It arrives `a*sigma^2` = `r 0.125*100`
units *before* the seasonal optimum (the tragedy of the commons).

### Selection gradient and convergence

The gradient is positive below `x*` and negative above it, so evolution converges
to `x*` from either side:

```{r gradient}
#| fig-width: 5.5
#| fig-height: 4
xs <- seq(-30, 10, length.out = 60)
grad <- sapply(xs, function(x) {
  community_start(bounds(x = c(-40, 40)), harness = h) |>
    community_add(trait_matrix(x, "x")) |>
    community_demography() |>
    community_selection_gradient() |>
    (\(comm) comm$selection_gradient)()
})

plot(xs, grad, type = "l", lwd = 2,
     xlab = "resident arrival time x", ylab = "selection gradient")
abline(h = 0, lty = 3); abline(v = x_opt - a * sigma^2, col = "forestgreen", lty = 2)
legend("topright", legend = "x* = x_opt - a sigma^2", col = "forestgreen",
       lty = 2, bty = "n")
```

### Mismatch grows with the competitive advantage `a`

```{r mismatch}
sapply(c(0, 0.04, 0.125), function(a) {
  h <- harness_jj12(a = a, x_opt = 0, sigma = 10)
  as.numeric(community_start(bounds(x = c(-40, 40)), harness = h) |>
               community_solve_singularity_1D() |>
               (\(comm) comm$traits)())
})   # 0, -4, -12.5  — earlier arrival as competition for territories intensifies
```
