---
title: "JJ12: the evolution of migratory arrival time"
subtitle: "How competition can favour arrival before the seasonal optimum"
format: html
---

## What the model represents

Migratory birds compete for a fixed number of breeding territories. Arriving
early can be costly because conditions are poor, but it also improves an
individual's chance of securing a territory. Arriving at the seasonal optimum
would maximise reproduction if there were no competition.

Competition changes that result. Each individual gains by arriving slightly
earlier than its competitors, so evolution favours an arrival date before the
date that would maximise total population output. This difference between the
individual and population optima is an evolutionary **tragedy of the commons**.

The ecological setting follows Johansson and Jonzén (2012), *Game theory sheds
new light on ecological responses to current climate change when phenology is
historically mismatched*. The equations below use the simplified analytic form
described by @Brannstrom-2013b, which gives us an exact answer against which to
check the numerical solver.

## Fitness equations

Competitive ability declines with arrival time `x`, while reproduction follows
a Gaussian curve centred on the seasonal optimum `x_opt`:

$$
C(x) = e^{-a x},
\qquad
R(x) = R_0\,\exp\!\left(-\frac{(x-x_{\mathrm{opt}})^2}{2\sigma^2}\right).
$$

With year-to-year survival `p`, a single resident reaches
$n^\* = K R(x)/(1-p)$. The invasion fitness of a mutant `x'` (a discrete-time
geometric growth rate; $w = 1$, i.e. log = 0, at the resident) is

$$
w(x') = (1-p)\,\frac{C(x')\,R(x')}{C(x)\,R(x)} + p .
$$

The selection gradient is a decreasing straight line, so the model has one
**continuously stable strategy** and no evolutionary branching. Its exact value
is

$$
x^\* = x_{\mathrm{opt}} - a\,\sigma^2 ,
$$

which lies *before* the population optimum `x_opt` whenever `a > 0`.

## Example analysis with `regnans`

```{r setup}
#| message: false
library(regnans)
```

```{r singular}
a <- 0.125; x_opt <- 0; sigma <- 10
h <- harness_jj12(a = a, x_opt = x_opt, sigma = sigma, R0 = 1, K = 5, p = 0.5)

out <- community_start(bounds(x = c(-40, 40)), harness = h) |>
  community_solve_singularity_1D()

c(numeric = as.numeric(out$traits),
  analytic = x_opt - a * sigma^2)     # should match: -12.5
```

The numerical solution should match the exact result. That known answer makes
this model a useful check on the solver. With these parameters, the evolved
strategy arrives `a*sigma^2` = `r 0.125*100` time units before the seasonal
optimum.

### Selection gradient and convergence

The gradient is positive for arrival times below `x*` and negative for arrival
times above it. Small evolutionary steps therefore move the population towards
`x*` from either side:

```{r gradient}
#| fig-width: 5.5
#| fig-height: 4
xs <- seq(-30, 10, length.out = 60)
grad <- sapply(xs, function(x) {
  community_start(bounds(x = c(-40, 40)), harness = h) |>
    community_add(trait_matrix(x, "x")) |>
    community_demography() |>
    community_selection_gradient() |>
    (\(comm) comm$selection_gradient)()
})

plot(xs, grad, type = "l", lwd = 2,
     xlab = "resident arrival time x", ylab = "selection gradient")
abline(h = 0, lty = 3); abline(v = x_opt - a * sigma^2, col = "forestgreen", lty = 2)
legend("topright", legend = "x* = x_opt - a sigma^2", col = "forestgreen",
       lty = 2, bty = "n")
```

### Mismatch grows with the competitive advantage `a`

Increasing `a` strengthens the benefit of arriving before competitors. The
evolved arrival date therefore moves farther ahead of the seasonal optimum:

```{r mismatch}
sapply(c(0, 0.04, 0.125), function(a) {
  h <- harness_jj12(a = a, x_opt = 0, sigma = 10)
  as.numeric(community_start(bounds(x = c(-40, 40)), harness = h) |>
               community_solve_singularity_1D() |>
               (\(comm) comm$traits)())
})   # 0, -4, -12.5  — earlier arrival as competition for territories intensifies
```
