---
title: "Finding evolutionary attractors"
format: html
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

An evolutionary attractor is a set of trait values towards which small
evolutionary steps move a resident population. At the attractor, the selection
gradient is zero: a small increase or decrease in the trait does not have an
immediate fitness advantage.

This page shows how to calculate selection gradients for a one-species `plant`
community. We first vary leaf mass per area (`lma`) alone, then vary `lma` and
height at maturation (`hmat`) together.

## Set up the model

Load the required packages and use the default parameters for community
assembly:

```{r, setup}
#| message: false
library(plant)
library(regnans)
library(parallel)
library(tidyverse)

model_support <- list(
  p = plant_default_assembly_pars(),
  plant_control = plant_default_assembly_control()
)

```

# One trait: leaf mass per area (`lma`)

## Selection gradients

We begin by calculating the selection gradient at one value of `lma`. A positive
gradient favours mutants with slightly larger `lma`; a negative gradient favours
slightly smaller values.

```{r}
trait <- 0.04045

# Start with an empty community.
community0 <-
  community_start(
    bounds(lma = c(0.01, 2)),
    fitness_control = list(method = "grid", n_evals = 20),
    model_support = model_support
  )

community_eq <-
  community0 %>%
  # Add one resident strategy with the chosen trait value.
  community_add(trait_matrix(trait, "lma"), birth_rate = 200) %>%
  # Solve for its demographic equilibrium.
  community_demography() %>%
  # Calculate the local slope of invasion fitness.
  community_selection_gradient()

community_eq$resident_fitness
community_eq$selection_gradient
```

The resident fitness is close to zero because its population has been brought to <!-- SY: Why is it not exactly zero? -->
demographic equilibrium. The positive selection gradient means that a mutant
with slightly larger `lma` would have an advantage.

We can see the same result in the local fitness landscape. The resident is shown
in red, and the slope at that point is positive:

```{r}
community_eq <- community_eq %>%
  community_fitness_landscape()

data <- 
  community_eq$fitness_points

ggplot(data, aes(lma, fitness)) +
  geom_line() +
  geom_point(data = filter(data, resident), col="red") +
  scale_x_log10()
```

One gradient describes only one resident trait value. To locate the attractor,
we repeat the calculation across a range of `lma` values. The attractor is where
the gradient crosses zero.

The helper function below adds a resident, solves its demography, and calculates
its selection gradient:

```{r}
f <- function(x, community) {
  community %>%
    community_add(trait_matrix(x, "lma")) %>%
    community_demography() %>%
    community_selection_gradient() 
}

# Create a sequence of trait values and solve each resident community.

data_communities <- 
  tibble(traits = seq_log_range(c(0.03, 0.15), length.out = 9)) %>%
  mutate(
    # Solve one equilibrium community for each trait value.
    communities = parallel::mclapply(
      traits, function(x) f(x, community0),
      mc.cores = base::max(1L, parallel::detectCores() - 2L, na.rm = TRUE)
    ),
    # Extract the selection gradient from each solution.
    selection_gradient = map_dbl(communities, ~ .x$selection_gradient)
  )

# Plot the gradient against the resident trait value.
plot_sg <- 
  ggplot(data_communities, aes(traits, selection_gradient)) +
  geom_line(col="red") +
  geom_point(col="red") +
  geom_abline(intercept = 0, linetype = "dashed") +
  ylim(-10000, 10000) +
  scale_x_log10()

plot_sg
```

## Using a root solver to find the 1D attractor

The sign-change plot helps us see the attractor, but a root finder gives a more
precise estimate. `community_solve_singularity_1D()` wraps the demographic and
selection-gradient calculations and searches for the trait value where the
gradient equals zero:

```{r}
community_attractor <- community0 %>%
 community_solve_singularity_1D(bounds = c(0.05, 0.15), tol = 1e-4)

community_attractor$traits
```

Add the root finder's solution to the selection-gradient plot:

```{r}
plot_sg +
  geom_point(col="red",
    data = tibble(
      traits = community_attractor$traits, selection_gradient = community_attractor$selection_gradient 
    )
  )
```


# Two traits: `lma` and `hmat`

## Selection gradients

With two evolving traits, the selection gradient is a vector with one component
for each trait. The following example calculates that vector for a resident with
`lma = 0.04045` and `hmat = 10`:

```{r}
traits <- c(0.04045, 10)

# Start with an empty two-trait community.
community0 <-
  community_start(
    bounds(lma = c(0.01, 2), hmat = c(0.01, 100)),
    fitness_control = list(method = "grid"),
    model_support = model_support
  )

community_eq <-
  community0 %>%
  # Add one resident with the chosen lma and hmat values.
  community_add(trait_matrix(traits, c("lma", "hmat")), birth_rate = 200) %>%
  # Solve for its demographic equilibrium.
  community_demography() %>%
  # Calculate one selection-gradient component for each trait.
  community_selection_gradient()

community_eq$resident_fitness
community_eq$selection_gradient
```

Resident fitness is again close to zero at demographic equilibrium. The two
selection-gradient values describe the locally favoured direction of change in
`lma` and `hmat`.

To see how that direction changes across trait space, we calculate the gradient
on a grid. The two-dimensional attractor lies where both gradient components are
zero. We can reuse a helper function similar to the one-trait version:

```{r}
f <- function(x, community) {
  community %>%
    community_add(trait_matrix(x, c("lma", "hmat"))) %>%
    community_demography() %>%
    community_selection_gradient() 
}

# Create a grid of trait combinations and solve each resident community.
data_communities <- 
  expand_grid(
    lma = seq_log_range(c(0.03, 0.05), length.out = 5),
    hmat = seq_log_range(c(5, 15), length.out = 5)
  ) %>%
  mutate(i = seq_len(n())) %>%
  nest(.by = i, .key = "traits") %>%
  mutate(
    # Solve one equilibrium community for each trait combination.
    communities = parallel::mclapply(
      traits, function(x) f(as.matrix(x), community0),
      mc.cores = base::max(1L, parallel::detectCores() - 2L, na.rm = TRUE)
    ))

data_communities2 <-
  data_communities %>%
    mutate(
      selection_gradient_lma = map_dbl(communities, ~ .x$selection_gradient[1]),
      selection_gradient_hmat = map_dbl(communities, ~ .x$selection_gradient[2])
    ) %>% 
    unnest(traits)

# Plot the selection-gradient vector field. ggquiver draws the two gradient
# components as an arrow at each grid point. Both trait axes use log coordinates.
library(ggquiver)

plot_sg <-
  data_communities2 %>%
    ggplot(aes(x = log(lma), y = log(hmat), u = selection_gradient_lma, v = selection_gradient_hmat)) +
    geom_point() +
    geom_quiver(vecsize = 0.4)

plot_sg
```

Each arrow shows the locally favoured direction of evolutionary change. The
arrows converge near the joint attractor, where both components approach zero.
`regnans` does not yet provide a two-dimensional equivalent of
`community_solve_singularity_1D()`, so this example locates the attractor
approximately from the vector field.
<!-- SY: Are all arrows supposed to face the right? Is that what is meant by there's only a 1D solver? -->

