---
title: "Solving for attractors"
format: html
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette provides an example of calculating attractors for 1 spp communities, using 1,2 or more traits. 


First, some basic setup

```{r, setup}
#| message: false
library(plant)
library(regnans)
library(parallel)
library(tidyverse)

# set baseline parameters
p <- plant_default_assembly_pars()

model_support <- list(
  p = plant_default_assembly_pars(),
  plant_control = plant_default_assembly_control()
)

```

# Example for 1 trait (lma)

## Selection gradients

Solving for the 1D attractor is possible with the community objects. 

Let's say you want to calculate the selection gradient at the following trait value. 

```{r}
trait <- 0.04045

# setup the assembler to start with empty community
community0 <-
  community_start(
    bounds(lma = c(0.01, 2)),
    fitness_control = list(method = "grid", n_evals = 20),
    model_support = model_support
  )

community_eq <-
  # Empty community
  community0 %>%
  # Add a strategy specified by it's trait value
  community_add(trait_matrix(trait, "lma"), birth_rate = 200) %>%
  # solve equilibrium density
  community_demography() %>%
  # calculate fitness gradient
  community_selection_gradient()

community_eq$resident_fitness
community_eq$selection_gradient
```

In this example fitness is near zero (as it should be), because the community is at equilibrium. The selection gradient is positive meaning, selection is to increase trait value. 

Let's visualise this by calculating a fitness landscape:

```{r}
community_eq <- community_eq %>%
  community_fitness_landscape()

data <- 
  community_eq$fitness_points

ggplot(data, aes(lma, fitness)) +
  geom_line() +
  geom_point(data = filter(data, resident), col="red") +
  scale_x_log10()
```

So let's plot the selection gradient across a range of trait values. The attractor is where the gradient crosses zero.

I'll make a worker function that runs the above routines.

```{r}
f <- function(x, community) {
  community %>%
    community_add(trait_matrix(x, "lma")) %>%
    community_demography() %>%
    community_selection_gradient() 
}

# Create a data frame with range of trait values and solve

data_communities <- 
  # trait matrix
  tibble(traits = seq_log_range(c(0.03, 0.15), length.out = 9)) %>%
  mutate(
    # solve communities
    communities = parallel::mclapply(
      traits, function(x) f(x, community0),
      mc.cores = parallel::detectCores() - 2
    ),
    # extract gradient
    selection_gradient = map_dbl(communities, ~ .x$selection_gradient)
  )

# plot
plot_sg <- 
  ggplot(data_communities, aes(traits, selection_gradient)) +
  geom_line(col="red") +
  geom_point(col="red") +
  geom_abline(intercept = 0, linetype = "dashed") +
  ylim(-10000, 10000) +
  scale_x_log10()

plot_sg
```

## Using a root solver to find the 1D attractor

We've wrapped up the above routines to enable solving of the 1D attractor

```{r}
community_attractor <- community0 %>%
 community_solve_singularity_1D(bounds = c(0.05, 0.15), tol = 1e-4)

community_attractor$traits
```

Let's add this to our previous plot
```{r}
plot_sg +
  geom_point(col="red",
    data = tibble(
      traits = community_attractor$traits, selection_gradient = community_attractor$selection_gradient 
    )
  )
```


# 2D Example: Attractor for lma and hmat

## Selection gradients

Solving for the 2D attractor is possible with the community objects. 

Let's say you want to calculate the selection gradient at the following trait values. 

```{r}
traits <- c(0.04045, 10)

# setup the assembler to start with empty community
community0 <-
  community_start(
    bounds(lma = c(0.01, 2), hmat = c(0.01, 100)),
    fitness_control = list(method = "grid"),
    model_support = model_support
  )

community_eq <-
  # Empty community
  community0 %>%
  # Add a strategy specified by it's trait value
  community_add(trait_matrix(traits, c("lma", "hmat")), birth_rate = 200) %>%
  # solve equilibrium density
  community_demography() %>%
  # calculate fitness gradient
  community_selection_gradient()

community_eq$resident_fitness
community_eq$selection_gradient
```

In this example fitness is near zero (as it should be), because the community is at equilibrium. The selection gradient is positive meaning, selection is to increase trait value. 

So let's plot the selection gradient across a range of trait values. The attractor is where the gradient crosses zero.

I'll make a worker function that runs the above routines.

```{r}
f <- function(x, community) {
  community %>%
    community_add(trait_matrix(x, c("lma", "hmat"))) %>%
    community_demography() %>%
    community_selection_gradient() 
}

# Create a data frame with range of trait values and solve
data_communities <- 
  # trait matrix
  expand_grid(
    lma = seq_log_range(c(0.03, 0.05), length.out = 5),
    hmat = seq_log_range(c(5, 15), length.out = 5)
  ) %>%
  mutate(i = seq_len(n())) %>%
  nest(.by = i, .key = "traits") %>%
  mutate(
    # solve communities
    communities = parallel::mclapply(
      traits, function(x) f(as.matrix(x), community0),
      mc.cores = parallel::detectCores() - 2
    ))

data_communities2 <-
  data_communities %>%
    mutate(
      selection_gradient_lma = map_dbl(communities, ~ .x$selection_gradient[1]),
      selection_gradient_hmat = map_dbl(communities, ~ .x$selection_gradient[2])
    ) %>% 
    unnest(traits)

# Plot the selection-gradient vector field. ggquiver draws the (u, v) gradient
# arrows at each grid point; the trait axes are shown on the log scale.
library(ggquiver)

plot_sg <-
  data_communities2 %>%
    ggplot(aes(x = log(lma), y = log(hmat), u = selection_gradient_lma, v = selection_gradient_hmat)) +
    geom_point() +
    geom_quiver(vecsize = 0.4)

plot_sg
```

The arrows point toward the joint attractor — where both trait gradients vanish.
A wrapped 2D root solver (the analogue of `community_solve_singularity_1D`) is not
yet available in `regnans`; for now the 2D attractor is read off the vector field.
