---
title: "FF16"
---

::: {.eyebrow}
model
:::

::: {.callout-note}
## Prerequisites
This page builds on [The big picture](../overview.qmd) and [The maths,
gently](../maths-primer.qmd). If the notation below is unfamiliar, those pages
provide a gentler introduction to the model and its equations.
:::

## The idea first

FF16 turns three kinds of information into demographic rates:

- the plant's traits, such as leaf mass per area and wood density;
- its current size; and
- the light available at each height in its crown.

From these inputs, the model calculates how quickly the plant grows, how many
seeds it produces, and how likely it is to die. All three rates share the same
starting point: a carbon budget. Leaves capture carbon through photosynthesis,
while respiration and tissue turnover use some of that income. The carbon that
remains is **net biomass production**. FF16 then divides this production
between growth and reproduction, and uses poor production as a signal of
increased mortality risk.

This shared carbon budget is important. Growth, reproduction, and mortality
are not three unrelated assumptions; they are different consequences of the
same plant physiology. Size and traits affect both sides of the budget. For
example, a larger plant can carry more leaves and capture more light, but it
also has more living tissue to maintain.

The rest of this page develops that idea mathematically. The model was
developed elsewhere, particularly in @Falster-2011. The equations are collected
here so that users can see the complete FF16 system implemented in `plant`, not
as a claim that the model originated in this documentation.

```{r}
#| echo: false
library(plant)
```

## Introduction

A physiological model in `plant` connects an individual plant to the wider
population model. It takes the individual's size, light environment, and
physiological parameters as inputs, and returns three **vital rates**: growth,
mortality, and fecundity.

In FF16, net biomass production links these inputs to all three outputs
(@fig-phys). Physiological parameters determine how the plant earns carbon,
what it costs to maintain its tissues, and how it divides the remaining
production. Changing those parameters can therefore represent differences
among species. The [Tables](#tables) section defines the variables and lists
the default parameter values used by the implementation.

```{r}
#| label: fig-phys
#| echo: false
#| out-width: '100%'
#| fig-cap: "**Physiological model in `plant`, giving demographic rates on the basis of its traits, size, and light environment, as functions of net mass production.** The dashed arrow towards mortality indicates that, although the mortality rate is assumed to depend on mass production, no mass is actually allocated there. Figure adapted from @Falster-2011 and @Falster-2015."

"../../assets/figures/models/FF16-schematic-phys.png" %>%
knitr::include_graphics()
```

## Growth {#growth}

### Leaf photosynthesis {#leaf-photosynthesis}

We begin with photosynthesis by a small patch of leaf. Let $x$ denote the
plant's traits, $z$ a height within its crown, and $E(z)$ the light available at
that height. The gross photosynthetic rate per unit leaf area is $p(x,E)$,
which FF16 represents with a saturating light-response curve:
$$
p(x, E(z)) = \alpha_{{\rm p1}} \frac{ E(z)}{E(z) + \alpha_{{\rm p2}}},
$$ {#eq-photosynthesis}
Here, $\alpha_{{\rm p1}}$ is the maximum annual photosynthetic rate, and
$\alpha_{{\rm p2}}$ is the light level at which the leaf reaches half that
maximum. Both parameters are derived from a more detailed leaf-level model.

Leaves at different heights receive different amounts of light. To obtain a
whole-plant average, the model weights the photosynthetic rate at each height
by the fraction of the plant's leaf area found there:
$$
\bar{p}(x, H, E_{a}) = \int_0^H p(x, E_a(z)) \, q(z, H) \, {\rm d}z,
$$ {#eq-photosynthesis-av}
where $H$ is plant height and $q(z,H)$ describes the vertical distribution of
leaf area (@eq-crown3). The result, $\bar p$, is the mean photosynthetic rate
per unit leaf area across the whole crown.

### Mass production {#standard-model-for-mass-production}

The next step is to turn photosynthesis into net biomass production. The
accounting is much like a budget: total photosynthesis is the income, while
maintenance respiration and tissue turnover are expenses
[@Makela-1997; @Thornley-2000; @Falster-2011]. Their balance gives
${\rm d}B/{\rm d}t$:
$$
\underbrace{\strut \frac{{\rm d}B}{{\rm d}t}}_\textrm{net biomass production}
= \underbrace{\strut \alpha_{\rm bio} }_\textrm{mass per C} \,
\underbrace{\strut \alpha_{\rm y} }_\textrm{yield}
\big( \underbrace{\strut A_{\rm l} \, \bar{p}}_\textrm{photosynthesis} -
\underbrace{\strut \, \sum_{i = {\rm l}, {\rm b}, {\rm s}, {\rm r}}{M_i \, r_i}}_\textrm{respiration}\big)
- \underbrace{\strut \sum_{i = {\rm l}, {\rm b}, {\rm s},  {\rm r}}{M_i \, k_i}}_\textrm{turnover}.
$$ {#eq-dbdt}
The tissue subscripts are $l$ for leaves, $b$ for bark, $s$ for sapwood, and
$r$ for roots. For each tissue, $M_i$ is its mass, $r_i$ its maintenance
respiration rate, and $k_i$ its turnover rate. The other terms make two
conversions:

- $\alpha_{\rm y}$ is the fraction of assimilated carbon retained after growth
  respiration; and
- $\alpha_{\rm bio}$ converts retained carbon into biomass.

Total photosynthesis equals the mean rate per unit leaf area, $\bar p$, times
total leaf area, $A_{\rm l}$. Leaf area is related to leaf mass by
$A_{\rm l}=M_{\rm l}/\phi$, where $\phi$ is leaf mass per area. The total living
mass that the plant must support is
$M_{\rm a}=M_{\rm l}+M_{\rm b}+M_{\rm s}+M_{\rm r}$.

### Height growth {#height-growth-rate}

The demographic model tracks individuals by height, so it needs to translate
net biomass production into height growth. FF16 makes this conversion in three
conceptual steps: production builds living tissue, living tissue supports leaf
area, and leaf area is related to height. The chain rule combines these steps
[@Falster-2011]:
$$
g(x, H, E_{a}) = \frac{{\rm d}H}{{\rm d}t} = \frac{{\rm d}H}{{\rm d}A_{\rm l}}
\times \frac{{\rm d}A_{\rm l} }{{\rm d}M_{\rm a}}
\times \frac{{\rm d}M_{\rm a}}{{\rm d}B}
\times \frac{{\rm d}B}{{\rm d}t}.
$$ {#eq-dhdt}

Read from right to left, the four factors ask:

1. How much net biomass does the plant produce?
2. What fraction of that production becomes new living tissue rather than
   reproductive tissue?
3. How much additional leaf area can that living tissue support?
4. How much taller does the plant become for that increase in leaf area?

The final factor, ${\rm d}H/{\rm d}A_{\rm l}$, captures plant architecture.
Some species deploy relatively more leaf area, whereas others gain relatively
more height from the same increase in leaf area.

The term ${\rm d}A_{\rm l}/{\rm d}M_{\rm a}$ represents the marginal cost of
deploying more leaf area. A plant must build not only the new leaves, but also
the sapwood, bark, and roots needed to support them. The inverse of those
combined construction costs is
$$
\frac{{\rm d}A_{\rm l}}{{\rm d}M_{\rm a}}
= \bigg( \frac{{\rm d}M_{\rm l}}{{\rm d}A_{\rm l}} + \frac{{\rm d}M_{\rm s}}{{\rm d}A_{\rm l}} + \frac{{\rm d}M_{\rm b}}{{\rm d}A_{\rm l}} + \frac{{\rm d}M_{\rm r}}{{\rm d}A_{\rm l}}\bigg)^{-1}.
$$ {#eq-daldmt}

The remaining conversion, ${\rm d}M_{\rm a}/{\rm d}B$, is the fraction of net
biomass production allocated to growth rather than reproduction. FF16 assumes
that young plants allocate nearly all production to growth, then shift more
production towards reproduction as they approach their maturation height:
$$
\frac{{\rm d}M_{\rm a}}{{\rm d}B}(H) = 1 -
\frac{\alpha_{{\rm f1}}} {1 + \exp\left(\alpha_{{\rm f2}} \left(1 - H / H_{\rm mat}\right)\right)},
$$ {#eq-allocation}
Here, $\alpha_{{\rm f1}}$ is the maximum fraction allocated to reproduction
(between 0 and 1), $\alpha_{{\rm f2}}$ controls how abruptly allocation changes,
and $H_{\rm mat}$ sets the height around which the transition occurs
[@Falster-2011].

### Diameter growth

Although height is the size variable used in the population model, FF16 can
also calculate stem diameter. Total stem basal area, $A_{\rm st}$, is the sum
of sapwood, bark, and heartwood cross-sectional areas ($A_{\rm s}$,
$A_{\rm b}$, and $A_{\rm h}$). Its rate of change is therefore
$$
\frac{{\rm d}A_{\rm st}}{{\rm d}t} = \frac{{\rm d}A_{\rm b}}{{\rm d}t} + \frac{{\rm d}A_{\rm s}}{{\rm d}t} + \frac{{\rm d}A_{\rm h}}{{\rm d}t}.
$$ {#eq-dA-dst}

Basal-area growth uses the same production, allocation, and construction terms
as height growth. It also includes the conversion of sapwood to heartwood:
$$
\frac{{\rm d}A_{\rm st}}{{\rm d}t} = \left(\frac{{\rm d}A_{\rm s}}{{\rm d}A_{\rm l}} + \frac{{\rm d}A_{\rm b}}{{\rm d}A_{\rm l}}\right) \times
\frac{{\rm d}A_{\rm l}}{{\rm d}M_{\rm a}} \times \frac{{\rm d}M_{\rm a}}{{\rm d}B} \times \frac{{\rm d}B}{{\rm d}t} + \frac{{\rm d}A_{\rm h}}{{\rm d}t} .
$$ {#eq-dast}
Finally, the geometric relationship between stem area and diameter converts
basal-area growth into diameter growth:
$$
\frac{{\rm d}D}{{\rm d}t} = \left(\pi \, A_{\rm st}\right)^{-0.5} \, \frac{{\rm d}A_{\rm st}}{{\rm d}t}.
$$ {#eq-dDdt}

# Functional-balance model for allocation {#a-functional-balance-model-for-allocation}

Here we describe an allometric model linking to a plant's height its various other size dimensions
required by most ecologically realistic vegetation models
(i.e., the masses of leaves, sapwood, bark, and fine
roots). This approach allows us to track only the
plant's height, while still accounting for the mass needs to build
leaves, roots, and stems. The growth rates of various tissues can then
also be derived (@tbl-allometry).

## Leaf area {#leaf-area}

Based on empirically observed allometry [@Falster-2011], we assume an allometric power-law
scaling relationship between the accumulated leaf area of a plant and
its height,
$$
H = \alpha_{{\rm l1}} \, \left(A_{\rm l} \, {\rm m^{-2}}\right)^{\alpha_{{\rm l2}}}.
$$ {#eq-ha}
This relationship is normalised around a leaf area of 1m${^2}$.

## Vertical distribution of leaf area {#vertical-distribution-of-leaf-area}

We follow the model of @Yokozawa-1995 describing the vertical
distribution of leaf area within the crowns of individual plants. This
model can account for a variety of canopy profiles through a single
parameter $\eta$. Setting $\eta = 1$ results in a conical canopy, as
seen in many conifers, while higher values, e.g., $\eta = 12$ , give a
top-heavy canopy profile similar to those seen among angiosperms. We denote
by $A_{{\rm s},z}$ the sapwood area at height $z$ within the plant, by $q(z, H)$ the vertical distribution of leaf area of
leaf area with respect to height $z$, and by $Q(z, H)$ the cumulative fraction of a
plant's leaves above height $z$. As defined previously, $A_{\rm s}$ is the sapwood
area at the base of the plant. Following @Yokozawa-1995, we
assume a relationship between $A_{{\rm s},z}$ and height such that
$$
\frac{A_{{\rm s},z}}{A_{\rm s}} = \left(1 - \left(\frac{z}{H}\right)^\eta\right)^2.
$$ {#eq-crown1}

We also assume that each unit of leaf area is supported by a fixed area $\theta$ of
sapwood [in agreement with the pipe model; @Shinozaki-1964], so that the total canopy
area of a plant relates to its basal sapwood area $A_{\rm s}$,
$$
A_{\rm s} = \theta \, A_{\rm l}.
$$ {#eq-crown2}
The pipe model is assumed to hold within individual plants, as well as
across plants of different size. It follows that,

$$
Q(z, H) = \left(1-\left(\frac{z}{H}\right)^\eta\right)^2.
$$ {#eq-crown1b}

Differentiating with respect to $z$ then yields a solution for the
probability density of leaf area as a function of height,
$$
q(z, H) = 2\frac{\eta}{H}\left(1-\left(\frac{z}{H}\right)^{\eta}\right) \left(\frac{z}{H}\right)^{\eta-1}.
$$ {#eq-crown3}

## Sapwood mass {#mass-of-sapwood}

Integrating $A_{{\rm s},z}$ yields the total mass of
sapwood in a plant,
$$
M_{\rm s}  = \rho \int_0^H A_{{\rm s},z} \, {\rm d}z = \rho A_{\rm s} H \eta_{\rm c} ,
$$ {#eq-ms1}
where $\eta_{\rm c} = 1 - \frac{2}{1 + \eta} + \frac{1}{1 + 2\eta}$
[@Yokozawa-1995]. Substituting from @eq-crown2 into @eq-ms1 then gives an expression for sapwood mass as a function of
leaf area and height,
$$
M_{\rm s} = \rho \, \eta_{\rm c} \, \theta \, A_{\rm l}  \, H.
$$ {#eq-ms2}

## Bark mass {#bark-mass}

Bark and phloem tissue are modelled using an analogue of the pipe model,
leading to a similar equation as that for sapwood mass (@eq-ms2). The cross-section area of bark per unit leaf area is
assumed to be a constant fraction $\alpha_{{\rm b1}}$ of sapwood area per unit leaf
area such that
$$
M_{\rm b} = \alpha_{{\rm b1}} M_{\rm s}.
$$ {#eq-mb}

## Root mass {#root-mass}

Also consistent with the pipe model, we assume a fixed ratio of
root mass per unit leaf area,
$$
M_{\rm r} = \alpha_{{\rm r1}} \, A_{\rm l}.
$$ {#eq-mr}
Even though nitrogen and water uptake are not modelled explicitly,
imposing a fixed ratio of root mass to leaf area ensures that
approximate costs of root production are included in calculations of
carbon budget.

# Seed production {#seed-production}

The rate of seed production, $f(x, H, E_{a})$, is a direct
function of the mass allocated to reproduction,
$$
f(x, H, E_{a}) = \frac{(1 - \frac{{\rm d}M_{\rm a}}{{\rm d}B}) \times \frac{{\rm d}B}{{\rm d}t}}{
\omega + \alpha_{{\rm f3}}},
$$ {#eq-fecundity}
where $\omega$ is the mass of the seed and $\alpha_{{\rm f3}}$ is the cost
per seed of accessories, such as fruits, flowers, and dispersal
structures. The function $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ is the fraction
of $\frac{{\rm d}B}{{\rm d}t}$ that is allocated to growth (from @eq-allocation, while
$1-\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ gives the fraction allocated
to reproduction).

# Mortality {#mortality}

Instantaneous rates of plant mortality are given by the sum of a growth-independent and a growth-dependent rate
[@Falster-2011; @Moorcroft-2001],
$$
d(x, H, E_{a}) = d_{{\rm I}}(x, H) + d_{{\rm G}}(x, H, E_{a}).
$$ {#eq-mortality}
The growth-independent rate is taken to be constant, independent of
plant performance, but potentially varying with species traits. The
growth-dependent rate is assumed to decline exponentially with the rate
of mass production per unit leaf area,
$$
d_{{\rm G}}(x, H, E_{a}) = \alpha_{{\rm dG1}} \exp(-\alpha_{{\rm dG2}} X),
$$ {#eq-mortality-gd}
where $X = {\rm d}B/{\rm d}t / A_{\rm l}$. This
relationship allows for plants to increase in mortality as their growth rate
approaches zero, while allowing for species to differ in
the parameters $\alpha_{{\rm dG1}}$ and $\alpha_{{\rm dG2}}$.

We also require a function $S_{\rm G}  (x^\prime, H_0, E_{{\rm a}0})$ for plant survival
through germination. For the demographic model to behave smoothly,
$S_{\rm G}  (x^\prime, H_0, E_{{\rm a}0}) / g(x, H_0, E_{{\rm a}0})$ should approach zero as
$g(x, H_0, E_{{\rm a}0})$ approaches zero. Following @Falster-2011,
we use the function
$$
S_{\rm G} (x^\prime, H_0, E_{{\rm a}0}) = \frac1{1 + X^2},
$$ {#eq-pi1}
where $X = \alpha_{{\rm d0}} \frac{A_{\rm l}}{{\rm d}B / {\rm d}t}$ and $\alpha_{{\rm d0}}$ is a
constant. @eq-pi1 is consistent with @eq-mortality-gd, as both cause survival to decline with mass production.

# Hyper-parameterisation of physiological model via traits {#traits}

The `FF16` physiological model includes default values for all needed parameters
(@tbl-params-core). Species are known to vary considerably
in many of these parameters, such as $\phi$, $\rho$, $\nu$, and $\omega$; so by varying parameters
one can account for species differences. When altering a parameter in the model, however,
one must also consider whether there are trade-offs linking parameters.

`plant` allows for the
hyper-parameterisation of the `FF16` physiological model via plant functional traits: this enables
simultaneous variation in multiple parameters in accordance with an assumed trade-off. In the
`FF16` physiological model, we implement the relationships described below. For more
details, see `make_FF16_hyperpar.R`.

## Leaf mass per unit area

The trait leaf mass per unit area, denoted by $\phi$, directly influences growth by changing
${\rm d}A_{\rm l} / {\rm d}M_{\rm a}$. In addition, we
link $\phi$ to the rate of leaf turnover,
based on a widely observed scaling relationship from @Wright-2004,
$$
k_{\rm l} = \beta_{{\rm kl1}} \, \left(\frac{\phi}{\phi_0}\right)^{-\beta_{{\rm kl2}}}.
$$ {#eq-k-l}

This relationship is normalised around $\phi_0$, the global mean of $\phi$. This
allows us to vary $\beta_{{\rm kl1}}$ and $\beta_{{\rm kl2}}$ without displacing the relationship from the
observed mean.

We also vary the mass-based leaf respiration rate so that it stays constant per unit leaf area and
varies with $\phi$ and nitrogen per unit leaf area $\nu$, as empirically observed by @Wright-2004,
$$
r_{\rm l} = \frac{\beta_{{\rm lf4}}\, \nu}{\phi}.
$$ {#eq-rl}


## Wood density

The trait wood density, denoted by $\rho$, directly influences growth by changing
${\rm d}A_{\rm l} / {\rm d}M_{\rm a}$. In addition, we allow for
$\rho$ to influence the rate of growth-independent mortality,
$$
d_{\rm I} = \beta_{{\rm dI1}} \, \left(\frac{\rho}{\rho_0}\right) ^ {-\beta_{{\rm dI2}}},
$$ {#eq-di}

and also the rate of sapwood turnover,
$$
k_{\rm s} = \beta_{{\rm ks1}} \, \left(\frac{\rho}{\rho_0}\right)^ {-\beta_{{\rm ks2}}},
$$ {#eq-ks}

As for $\phi$, these relationships are normalized around $\rho_0$, the global mean of $\rho$.
By default, $\beta_{{\rm kI2}}$ and $\beta_{{\rm ks2}}$ are set to zero, so these linkages
only become present when these parameters are set to something other than their default values.

The rate of sapwood respiration per unit volume is assumed to be constant, so
sapwood respiration per unit mass varies as
$$
r_{\rm s}  = \frac{\beta_{{\rm rs1}}}{\rho},
$$ {#eq-rs}

where $\beta_{{\rm rs1}}$ is a default rate per volume of sapwood.
Similarly, the rate of bark respiration per unit mass varies as
$$
r_{\rm b} = \frac{\beta_{{\rm rb1}}}{\rho},
$$ {#eq-rb}

with $\beta_{{\rm rb1}} = 2 \beta_{{\rm rs1}}$.

## Seed mass

Effects of the trait seed mass, denoted by $\omega$, are naturally embedded in the equation determining
fecundity (@eq-fecundity) and the initial height of seedlings. In addition,
we let the accessory cost per seed be a multiple of seed size,
$$
\alpha_{{\rm f3}} = \beta_{{\rm f1}} \omega,
$$ {#eq-a-f3}
as empirically observed [@Henery-2001].

## Nitrogen per unit leaf area

Photosynthesis per unit leaf area and respiration rates per unit leaf mass (or area) are assumed to vary with leaf nitrogen per unit area, $\nu$. The calculation of respiration rates is already described above. To calculate the average annual photosynthesis for a leaf, we integrate the instantaneous rate per unit leaf area over the annual solar trajectory, using a rectangular-hyperbolic photosynthesis light response curve,
$$
p(\nu, E) = \frac1{365 \textrm{d}}\int_0^{365 \textrm{d}}\frac{Y(t) + A_{\rm max} - \sqrt{(Y(t) + A_{\rm max})^2 - 4 \beta_{{\rm lf2}} Y(t) A_{\rm max} }}{2 \beta_{{\rm lf2}}} {\rm d}t,
$$ {#eq-ptss}
where

- $A_{\rm max}$ is the maximum photosynthetic capacity of the leaf,
- $\beta_{{\rm lf2}}$ is the curvature of the light response curve,
- $Y(t) = \beta_{{\rm lf3}} I(t)$ is the initial yield of the light response curve, with $\beta_{{\rm lf3}}$ being the quantum yield parameter,
- $I(t) = k_{\rm I} \, I_0(t)\, E$ is the intensity of light on the leaf surface, and
- $I_0(t)$ is light incident on a surface perpendicular to the sun's rays directly above the canopy at time $t$.

The profile of $I_0(t)$ is given by a solar model adapted from @TerSteege-1997.

We allow for the maximum photosynthetic capacity of the leaf to vary with leaf nitrogen per unit area, as
$$
A_{\rm max} = \beta_{{\rm lf1}} \, \left(\frac{\nu}{\nu_0}\right)^{\beta_{{\rm lf5}}},
$$ {#eq-amax}
where $\beta_{{\rm lf1}}$ and $\beta_{{\rm lf5}}$ are constants. The relationships is normalized around $\nu_0$, the global mean of leaf nitrogen per unit area.

Values of $p(\nu, E)$ are calculated across a range of values of $E$, and then an expression of the form in @eq-photosynthesis is fitted to extract the parameters
$\alpha_{{\rm p1}}$ and $\alpha_{{\rm p2}}$, such that these become functions of $\nu$.

# Tables {#tables}

```{r}
#| label: functions
#| echo: false
extract_strategy_parameter_values <- function(vars){

  s <- c(FF16_Strategy(),
         lapply(as.list(args(make_FF16_hyperpar)), eval))

  out <- rep(NA_character_, length(vars))

  # Now extract values from plant
  to_code <- vars != "" & vars %in% names(s)
  code <- vars[to_code]

  val <- s[code]
  err <- sapply(val, is.null)
  if (any(err)) {
    message("Missing parameters: ", paste(code[err], collapse=", "))
  }

  oo <- options(scipen=999)
  on.exit(options(oo))
  out[to_code] <- unlist(lapply(val, prettyNum))
  out[is.na(out)] <- ""

  out
}
```

```{r}
#| label: tbl-definitions
#| echo: false
#| tbl-cap: 'Key variables of the `FF16` physiological model. For mass ($M$), respiration ($r$), and turnover ($k$) variables, subscripts refer to any of the following tissues: ${\rm l}$ = leaves, ${\rm b}$ = bark, ${\rm s}$ = sapwood,${\rm r}$ = roots, ${\rm a}$ = all living tissue. For area $A$ variables, subscripts refer to any of the following: ${\rm l}$ = leaves, ${\rm st}$ = total stem cross-section, ${\rm s}$ = sapwood cross-section, ${\rm b}$ = bark cross-section, ${\rm h}$ = heartwood cross-section'
system.file("docs", "demography-vars.csv", package = "plant") %>%
read.csv() %>% dplyr::as_tibble() %>%
# TODO(#481): can we get groups as headers spanning multiple columns?
knitr::kable(align = "l")
```

```{r}
#| label: tbl-allometry
#| echo: false
#| tbl-cap: 'Equations of the allometric growth model implemented within the `FF16` physiological model, based on functional-balance assumptions (see text for details). In (a) we list  key assumptions of allocation model between specific variation with leaf area, under "Function". From these assumptions, allocation functions are derived for tissue areas and tissue masses (b). Also the growth rate of each tissue type can be expressed as a function of the growth rate of leaf area. For mass ($M$), respiration ($r$), and turnover ($k$) variables, subscripts refer to any of the following tissues: ${\rm l}$ = leaves, ${\rm b}$ = bark, ${\rm s}$ = sapwood, ${\rm r}$ = roots, ${\rm a}$ = all living tissue. For area $A$ variables, subscripts refer to any of the following: ${\rm l}$ = leaves, ${\rm st}$ = total stem cross-section, ${\rm s}$ = sapwood cross-section, ${\rm b}$ = bark cross-section, ${\rm h}$ = heartwood cross-section'
system.file("docs/FF16", "FF16-eqns.csv", package = "plant") %>%
  read.csv() %>% dplyr::as_tibble() %>%
  knitr::kable(align = "l")
```

```{r}
#| label: tbl-params-core
#| echo: false
#| tbl-cap: 'Core parameters of the `FF16` physiological model.'
x <- system.file("docs/FF16", "FF16-params-core.csv", package = "plant") %>%
  read.csv() %>% dplyr::as_tibble()

x$Value <- extract_strategy_parameter_values(x$Symbol)
knitr::kable(x,  align = "l")
```


```{r}
#| label: tbl-params-hyper
#| echo: false
#| tbl-cap: 'Parameters for hyper-parameterisation of the `FF16` physiological model.'
x <- system.file("docs/FF16", "FF16-params-hyper.csv", package = "plant") %>%
  read.csv() %>% dplyr::as_tibble()

x$Value <- extract_strategy_parameter_values(x$Code)
knitr::kable(x,  align = "l")
```

# References
