---
title: "FF16"
---

::: {.eyebrow}
model
:::

::: {.callout-note}
## Prerequisites
This page builds on [The big picture](../overview.qmd) and [The maths,
gently](../maths-primer.qmd). If the notation below is unfamiliar, those pages
provide a gentler introduction to the model and its equations.
:::

## The idea first

FF16 turns three kinds of information into demographic rates:

- the plant's traits, such as leaf mass per area and wood density;
- its current size; and
- the light available at each height in its crown.

From these inputs, the model calculates how quickly the plant grows, how many
seeds it produces, and how likely it is to die. All three rates share the same
starting point: a carbon budget. Leaves capture carbon through photosynthesis,
while respiration and tissue turnover use some of that income. The carbon that
remains is **net biomass production**. FF16 then divides this production
between growth and reproduction, and uses poor production as a signal of
increased mortality risk.

This shared carbon budget is important. Growth, reproduction, and mortality
are not three unrelated assumptions; they are different consequences of the
same plant physiology. Size and traits affect both sides of the budget. For
example, a larger plant can carry more leaves and capture more light, but it
also has more living tissue to maintain.

The rest of this page develops that idea mathematically. The model was
developed elsewhere, particularly in @Falster-2011. The equations are collected
here so that users can see the complete FF16 system implemented in `plant`.

```{r}
#| echo: false
library(plant)
```

## Introduction

A physiological model in `plant` connects an individual plant to the wider
population model. It takes the individual's size, light environment, and
physiological parameters as inputs, and returns three **vital rates**: growth,
mortality, and fecundity.

In FF16, net biomass production links these inputs to all three outputs
(@fig-phys). Physiological parameters determine how the plant earns carbon,
what it costs to maintain its tissues, and how it divides the remaining
production. Changing those parameters can therefore represent differences
among species. The [Tables](#tables) section defines the variables and lists
the default parameter values used by the implementation.

```{r}
#| label: fig-phys
#| echo: false
#| out-width: '100%'
#| fig-cap: "**Physiological model in `plant`, giving demographic rates on the basis of its traits, size, and light environment, as functions of net mass production.** The dashed arrow towards mortality indicates that, although the mortality rate is assumed to depend on mass production, no mass is actually allocated there. Figure adapted from @Falster-2011 and @Falster-2015."

"../../assets/figures/models/FF16-schematic-phys.png" %>%
knitr::include_graphics()
```

## Growth {#growth}

### Leaf photosynthesis {#leaf-photosynthesis}

We begin with photosynthesis by a small patch of leaf. Let $x$ denote the
plant's traits, $z$ a height within its crown, and $E(z)$ the light available at
that height. The gross photosynthetic rate per unit leaf area is $p(x,E)$,
which FF16 represents with a saturating light-response curve:
$$
p(x, E(z)) = \alpha_{{\rm p1}} \frac{ E(z)}{E(z) + \alpha_{{\rm p2}}},
$$ {#eq-photosynthesis}
Here, $\alpha_{{\rm p1}}$ is the maximum annual photosynthetic rate, and
$\alpha_{{\rm p2}}$ is the light level at which the leaf reaches half that
maximum. Both parameters are derived from a more detailed leaf-level model.

<!-- SY: Could be nice to add graphs of the curves in the equations -->

Leaves at different heights receive different amounts of light. To obtain a
whole-plant average, the model weights the photosynthetic rate at each height
by the fraction of the plant's leaf area found there:
$$
\bar{p}(x, H, E_{a}) = \int_0^H p(x, E_a(z)) \, q(z, H) \, {\rm d}z,
$$ {#eq-photosynthesis-av}
where $H$ is plant height and $q(z,H)$ describes the vertical distribution of
leaf area (@eq-crown3). The result, $\bar p$, is the mean photosynthetic rate
per unit leaf area across the whole crown.

### Mass production {#standard-model-for-mass-production}

The next step is to turn photosynthesis into net biomass production. The
accounting is much like a budget: total photosynthesis is the income, while
maintenance respiration and tissue turnover are expenses
[@Makela-1997; @Thornley-2000; @Falster-2011]. Their balance gives
${\rm d}B/{\rm d}t$:
$$
\underbrace{\strut \frac{{\rm d}B}{{\rm d}t}}_\textrm{net biomass production}
= \underbrace{\strut \alpha_{\rm bio} }_\textrm{mass per C} \,
\underbrace{\strut \alpha_{\rm y} }_\textrm{yield}
\big( \underbrace{\strut A_{\rm l} \, \bar{p}}_\textrm{photosynthesis} -
\underbrace{\strut \, \sum_{i = {\rm l}, {\rm b}, {\rm s}, {\rm r}}{M_i \, r_i}}_\textrm{respiration}\big)
- \underbrace{\strut \sum_{i = {\rm l}, {\rm b}, {\rm s},  {\rm r}}{M_i \, k_i}}_\textrm{turnover}.
$$ {#eq-dbdt}
The tissue subscripts are $l$ for leaves, $b$ for bark, $s$ for sapwood, and
$r$ for roots. For each tissue, $M_i$ is its mass, $r_i$ its maintenance
respiration rate, and $k_i$ its turnover rate. The other terms make two
conversions:

- $\alpha_{\rm y}$ is the fraction of assimilated carbon retained after growth
  respiration; and
- $\alpha_{\rm bio}$ converts retained carbon into biomass.

Total photosynthesis equals the mean rate per unit leaf area, $\bar p$, times
total leaf area, $A_{\rm l}$. Leaf area is related to leaf mass by
$A_{\rm l}=M_{\rm l}/\phi$, where $\phi$ is leaf mass per area. The total living
mass that the plant must support is
$M_{\rm a}=M_{\rm l}+M_{\rm b}+M_{\rm s}+M_{\rm r}$.

### Height growth {#height-growth-rate}

The demographic model tracks individuals by height, so it needs to translate
net biomass production into height growth. FF16 makes this conversion in three
conceptual steps: production builds living tissue, living tissue supports leaf
area, and leaf area is related to height. The chain rule combines these steps
[@Falster-2011]:
$$
g(x, H, E_{a}) = \frac{{\rm d}H}{{\rm d}t} = \frac{{\rm d}H}{{\rm d}A_{\rm l}}
\times \frac{{\rm d}A_{\rm l} }{{\rm d}M_{\rm a}}
\times \frac{{\rm d}M_{\rm a}}{{\rm d}B}
\times \frac{{\rm d}B}{{\rm d}t}.
$$ {#eq-dhdt}

Read from right to left, the four factors ask:

1. How much net biomass does the plant produce?
2. What fraction of that production becomes new living tissue rather than
   reproductive tissue?
3. How much additional leaf area can that living tissue support?
4. How much taller does the plant become for that increase in leaf area?

The final factor, ${\rm d}H/{\rm d}A_{\rm l}$, captures plant architecture.
Some species deploy relatively more leaf area, whereas others gain relatively
more height from the same increase in leaf area.

The term ${\rm d}A_{\rm l}/{\rm d}M_{\rm a}$ represents the marginal cost of
deploying more leaf area. A plant must build not only the new leaves, but also
the sapwood, bark, and roots needed to support them. The inverse of those
combined construction costs is
$$
\frac{{\rm d}A_{\rm l}}{{\rm d}M_{\rm a}}
= \bigg( \frac{{\rm d}M_{\rm l}}{{\rm d}A_{\rm l}} + \frac{{\rm d}M_{\rm s}}{{\rm d}A_{\rm l}} + \frac{{\rm d}M_{\rm b}}{{\rm d}A_{\rm l}} + \frac{{\rm d}M_{\rm r}}{{\rm d}A_{\rm l}}\bigg)^{-1}.
$$ {#eq-daldmt}

The remaining conversion, ${\rm d}M_{\rm a}/{\rm d}B$, is the fraction of net
biomass production allocated to growth rather than reproduction. FF16 assumes
that young plants allocate nearly all production to growth, then shift more
production towards reproduction as they approach their maturation height:
$$
\frac{{\rm d}M_{\rm a}}{{\rm d}B}(H) = 1 -
\frac{\alpha_{{\rm f1}}} {1 + \exp\left(\alpha_{{\rm f2}} \left(1 - H / H_{\rm mat}\right)\right)},
$$ {#eq-allocation}
Here, $\alpha_{{\rm f1}}$ is the maximum fraction allocated to reproduction
(between 0 and 1), $\alpha_{{\rm f2}}$ controls how abruptly allocation changes,
and $H_{\rm mat}$ sets the height around which the transition occurs
[@Falster-2011].

### Diameter growth

Although height is the size variable used in the population model, FF16 can
also calculate stem diameter. Total stem basal area, $A_{\rm st}$, is the sum
of sapwood, bark, and heartwood cross-sectional areas ($A_{\rm s}$,
$A_{\rm b}$, and $A_{\rm h}$). Its rate of change is therefore
$$
\frac{{\rm d}A_{\rm st}}{{\rm d}t} = \frac{{\rm d}A_{\rm b}}{{\rm d}t} + \frac{{\rm d}A_{\rm s}}{{\rm d}t} + \frac{{\rm d}A_{\rm h}}{{\rm d}t}.
$$ {#eq-dA-dst}

Basal-area growth uses the same production, allocation, and construction terms
as height growth. It also includes the conversion of sapwood to heartwood:
$$
\frac{{\rm d}A_{\rm st}}{{\rm d}t} = \left(\frac{{\rm d}A_{\rm s}}{{\rm d}A_{\rm l}} + \frac{{\rm d}A_{\rm b}}{{\rm d}A_{\rm l}}\right) \times
\frac{{\rm d}A_{\rm l}}{{\rm d}M_{\rm a}} \times \frac{{\rm d}M_{\rm a}}{{\rm d}B} \times \frac{{\rm d}B}{{\rm d}t} + \frac{{\rm d}A_{\rm h}}{{\rm d}t} .
$$ {#eq-dast}
<!-- SY: Where is the conversion of sapwood to heartwood? -->

Finally, the geometric relationship between stem area and diameter converts
basal-area growth into diameter growth:
$$
\frac{{\rm d}D}{{\rm d}t} = \left(\pi \, A_{\rm st}\right)^{-0.5} \, \frac{{\rm d}A_{\rm st}}{{\rm d}t}.
$$ {#eq-dDdt}

## Functional-balance model for allocation {#a-functional-balance-model-for-allocation}

The previous section used tissue masses and areas even though the population
model tracks only plant height. FF16 connects these quantities through an
allometric **functional-balance model**. Given a plant's height, the model
infers how much leaf, sapwood, bark, and fine-root tissue it needs to maintain
a workable plant architecture.
<!-- SY: What does 'functional-balance' mean? -->

This approach keeps the population model one-dimensional while still charging
the plant for all the tissue needed to support new leaves. It also allows the
growth rates of individual tissues to be derived from height growth
(@tbl-allometry).

### Leaf area {#leaf-area}

The first relationship links total leaf area to height. Following empirical
allometry [@Falster-2011], FF16 uses a power law:
$$
H = \alpha_{{\rm l1}} \, \left(A_{\rm l} \, {\rm m^{-2}}\right)^{\alpha_{{\rm l2}}}.
$$ {#eq-ha}
The factor ${\rm m^{-2}}$ makes the quantity inside the power dimensionless, so
the relationship is normalised at a leaf area of 1 m${^2}$.

### Vertical distribution of leaf area {#vertical-distribution-of-leaf-area}

Total leaf area is not enough to calculate photosynthesis: the model also needs
to know where those leaves sit within the crown. FF16 follows
@Yokozawa-1995 and describes crown shape with a single parameter, $\eta$.
Setting $\eta=1$ gives a conical canopy, as seen in many conifers. A larger
value, such as $\eta=12$, produces a more top-heavy crown like those of many
angiosperms.

Let $A_{{\rm s},z}$ be sapwood area at height $z$, and let $A_{\rm s}$ be
sapwood area at the base of the plant. Following @Yokozawa-1995, their
relationship is
$$
\frac{A_{{\rm s},z}}{A_{\rm s}} = \left(1 - \left(\frac{z}{H}\right)^\eta\right)^2.
$$ {#eq-crown1}

The **pipe model** assumes that each unit of leaf area is supported by a fixed
area, $\theta$, of sapwood [@Shinozaki-1964]. Basal sapwood area is therefore
proportional to total leaf area:
<!-- SY: Is there a good figure of the pipe model? -->
$$
A_{\rm s} = \theta \, A_{\rm l}.
$$ {#eq-crown2}
FF16 applies this relationship both within an individual and across plants of
different sizes. <!-- SY: What does this mean? --> As a result, the fraction of a plant's leaves above height
$z$, denoted $Q(z,H)$, is

$$
Q(z, H) = \left(1-\left(\frac{z}{H}\right)^\eta\right)^2.
$$ {#eq-crown1b}

Differentiating this cumulative relationship gives $q(z,H)$, the density that
describes how leaf area is distributed with height:
$$
q(z, H) = 2\frac{\eta}{H}\left(1-\left(\frac{z}{H}\right)^{\eta}\right) \left(\frac{z}{H}\right)^{\eta-1}.
$$ {#eq-crown3}
<!-- SY: Again, would be cool to have graphs of these equations throughout! -->

### Sapwood mass {#mass-of-sapwood}

Sapwood mass depends on its cross-sectional area throughout the full height of
the stem. Integrating $A_{{\rm s},z}$ over height and multiplying by wood
density, $\rho$, gives
$$
M_{\rm s}  = \rho \int_0^H A_{{\rm s},z} \, {\rm d}z = \rho A_{\rm s} H \eta_{\rm c} ,
$$ {#eq-ms1}
where $\eta_{\rm c} = 1 - \frac{2}{1 + \eta} + \frac{1}{1 + 2\eta}$ is a
factor determined by crown shape [@Yokozawa-1995]. Substituting the pipe-model
relationship from @eq-crown2 expresses sapwood mass directly in terms of leaf
area and plant height:
$$
M_{\rm s} = \rho \, \eta_{\rm c} \, \theta \, A_{\rm l}  \, H.
$$ {#eq-ms2}

### Bark mass {#bark-mass}

FF16 treats bark and phloem with a relationship analogous to the pipe model.
Their mass is a fixed fraction, $\alpha_{{\rm b1}}$, of sapwood mass:
$$
M_{\rm b} = \alpha_{{\rm b1}} M_{\rm s}.
$$ {#eq-mb}

### Root mass {#root-mass}

Root mass is also proportional to leaf area:
$$
M_{\rm r} = \alpha_{{\rm r1}} \, A_{\rm l}.
$$ {#eq-mr}
FF16 does not explicitly model nitrogen or water uptake. This relationship
nevertheless ensures that building and maintaining roots contributes to the
plant's carbon budget.

## Seed production {#seed-production}

Once the plant begins reproducing, the fraction of production not allocated to
growth is available for seeds. Dividing that reproductive production by the
total cost per seed gives the seed-production rate, $f(x,H,E_a)$:
$$
f(x, H, E_{a}) = \frac{(1 - \frac{{\rm d}M_{\rm a}}{{\rm d}B}) \times \frac{{\rm d}B}{{\rm d}t}}{
\omega + \alpha_{{\rm f3}}},
$$ {#eq-fecundity}
Here, $\omega$ is seed mass and $\alpha_{{\rm f3}}$ is the additional cost of
flowers, fruits, and dispersal structures associated with each seed. From
@eq-allocation, ${\rm d}M_{\rm a}/{\rm d}B$ is the fraction allocated to growth,
so $1-{\rm d}M_{\rm a}/{\rm d}B$ is the fraction allocated to reproduction.

## Mortality {#mortality}

FF16 represents mortality as a continuous hazard: the instantaneous risk of
death at a particular size and in a particular environment. The total hazard
has two parts [@Falster-2011; @Moorcroft-2001]:
$$
d(x, H, E_{a}) = d_{{\rm I}}(x, H) + d_{{\rm G}}(x, H, E_{a}).
$$ {#eq-mortality}
The growth-independent component, $d_{\rm I}$, is constant with respect to
current plant performance, although it may vary among trait combinations. The
growth-dependent component, $d_{\rm G}$, falls exponentially as net biomass
production per unit leaf area increases:
$$
d_{{\rm G}}(x, H, E_{a}) = \alpha_{{\rm dG1}} \exp(-\alpha_{{\rm dG2}} X),
$$ {#eq-mortality-gd}
where $X=({\rm d}B/{\rm d}t)/A_{\rm l}$. Plants with little production relative
to their leaf area therefore face a higher mortality hazard. The parameters
$\alpha_{{\rm dG1}}$ and $\alpha_{{\rm dG2}}$ allow the level and sensitivity of
this response to differ among species.

Germination and establishment require a separate survival function, <!-- SY: Why the use of x' (prime)? -->
$S_{\rm G}(x^\prime,H_0,E_{{\rm a}0})$, evaluated for a new plant of initial
height $H_0$ in its initial environment $E_{{\rm a}0}$. For the demographic
model to remain well behaved, the ratio
$S_{\rm G}(x^\prime,H_0,E_{{\rm a}0})/g(x,H_0,E_{{\rm a}0})$ must approach zero
as initial growth approaches zero. Following @Falster-2011, FF16 uses
$$
S_{\rm G} (x^\prime, H_0, E_{{\rm a}0}) = \frac1{1 + X^2},
$$ {#eq-pi1}
where $X = \alpha_{{\rm d0}} \frac{A_{\rm l}}{{\rm d}B / {\rm d}t}$ and
$\alpha_{{\rm d0}}$ is a constant. Because $X$ is large when production per
unit leaf area is small, establishment survival falls towards zero for plants
with very poor initial production. This is consistent with the
growth-dependent mortality rule in @eq-mortality-gd: both penalise poor carbon
balance.

## Hyper-parameterisation of physiological model via traits {#traits}

FF16 has default values for every physiological parameter
(@tbl-params-core), but species can differ substantially in quantities such as
leaf mass per area ($\phi$), wood density ($\rho$), leaf nitrogen per area
($\nu$), and seed mass ($\omega$). These measurable quantities can therefore be
used as functional traits that distinguish strategies.

A trait often affects more than one model parameter. For example, changing
leaf mass per area alters both the cost of constructing leaf area and the rate
at which leaves turn over. Treating those effects independently could create a
plant strategy that violates an assumed biological trade-off.

To keep linked parameters consistent, `plant` **hyper-parameterises** FF16:
trait values are passed through a set of relationships that update several
physiological parameters together. Those relationships are described below.
Their implementation is in the `make_FF16_hyperpar()` function.

### Leaf mass per unit area

Leaf mass per unit area, $\phi$, measures how much dry mass is needed to build
a unit of leaf area. A larger $\phi$ therefore increases the construction cost
captured by ${\rm d}A_{\rm l}/{\rm d}M_{\rm a}$. It is also linked to leaf
turnover through the widely observed relationship reported by @Wright-2004:
$$
k_{\rm l} = \beta_{{\rm kl1}} \, \left(\frac{\phi}{\phi_0}\right)^{-\beta_{{\rm kl2}}}.
$$ {#eq-k-l}

Here, $\phi_0$ is the global mean of $\phi$. Normalising the relationship at
$\phi_0$ allows $\beta_{{\rm kl1}}$ and $\beta_{{\rm kl2}}$ to change its level
and slope without shifting it away from the observed mean.

Leaf respiration is specified per unit leaf mass, but is constrained to remain
constant per unit leaf area for a given leaf nitrogen content. It therefore
varies with both $\phi$ and nitrogen per unit leaf area, $\nu$, following
@Wright-2004:
$$
r_{\rm l} = \frac{\beta_{{\rm lf4}}\, \nu}{\phi}.
$$ {#eq-rl}


### Wood density

Wood density, $\rho$, affects how much biomass is needed to build a given
volume of stem. It therefore changes the construction-cost term
${\rm d}A_{\rm l}/{\rm d}M_{\rm a}$. FF16 can also link wood density to the
growth-independent mortality rate,
$$
d_{\rm I} = \beta_{{\rm dI1}} \, \left(\frac{\rho}{\rho_0}\right) ^ {-\beta_{{\rm dI2}}},
$$ {#eq-di}

and to the sapwood turnover rate,
$$
k_{\rm s} = \beta_{{\rm ks1}} \, \left(\frac{\rho}{\rho_0}\right)^ {-\beta_{{\rm ks2}}},
$$ {#eq-ks}

where $\rho_0$ is the global mean wood density. As with $\phi$, normalising at
this mean keeps the relationships anchored while their other parameters vary.
By default, $\beta_{{\rm dI2}}$ and $\beta_{{\rm ks2}}$ are zero, so wood density
does not affect these two rates unless the corresponding exponents are changed.

FF16 assumes that sapwood respiration is constant per unit volume. A unit of <!-- SY: I was wondering why sapwood respiration would decrease with increasing wood density. I found a paper by Westerband et al. (2022) called "nitrogen concentration and physical properties are key drivers of woody tissue respiration" which seems like a great reference for this section. -->
mass occupies less volume when wood is denser, so respiration per unit mass is
$$
r_{\rm s}  = \frac{\beta_{{\rm rs1}}}{\rho},
$$ {#eq-rs}

where $\beta_{{\rm rs1}}$ is the default respiration rate per unit sapwood
volume. Bark respiration per unit mass follows the same logic:
$$
r_{\rm b} = \frac{\beta_{{\rm rb1}}}{\rho},
$$ {#eq-rb}

with $\beta_{{\rm rb1}} = 2 \beta_{{\rm rs1}}$.

### Seed mass

Seed mass, $\omega$, directly affects fecundity through @eq-fecundity and also
determines a seedling's initial height. FF16 assumes that the accessory cost of
producing flowers, fruits, and dispersal structures scales with seed mass:
$$
\alpha_{{\rm f3}} = \beta_{{\rm f1}} \omega,
$$ {#eq-a-f3}
as observed empirically by @Henery-2001.

### Nitrogen per unit leaf area

Leaf nitrogen per unit area, $\nu$, affects both photosynthesis and respiration.
The respiration relationship was introduced above. For photosynthesis, FF16
first calculates the instantaneous response of a leaf to light and then
averages that response over an annual solar trajectory. The light-response
curve is
$$
p(\nu, E) = \frac1{365 \textrm{d}}\int_0^{365 \textrm{d}}\frac{Y(t) + A_{\rm max} - \sqrt{(Y(t) + A_{\rm max})^2 - 4 \beta_{{\rm lf2}} Y(t) A_{\rm max} }}{2 \beta_{{\rm lf2}}} {\rm d}t,
$$ {#eq-ptss}
The terms in this expression are:

- $A_{\rm max}$ is the maximum photosynthetic capacity of the leaf,
- $\beta_{{\rm lf2}}$ is the curvature of the light response curve,
- $Y(t) = \beta_{{\rm lf3}} I(t)$ is the initial yield of the light response curve, with $\beta_{{\rm lf3}}$ being the quantum yield parameter,
- $I(t) = k_{\rm I} \, I_0(t)\, E$ is the intensity of light on the leaf surface, and
- $I_0(t)$ is light incident on a surface perpendicular to the sun's rays directly above the canopy at time $t$.

The time-varying profile $I_0(t)$ comes from a solar model adapted from
@TerSteege-1997.

Maximum photosynthetic capacity increases with leaf nitrogen per unit area
according to
$$
A_{\rm max} = \beta_{{\rm lf1}} \, \left(\frac{\nu}{\nu_0}\right)^{\beta_{{\rm lf5}}},
$$ {#eq-amax}
where $\beta_{{\rm lf1}}$ and $\beta_{{\rm lf5}}$ are constants, and $\nu_0$ is
the global mean leaf nitrogen per unit area used to normalise the relationship.

Finally, the model evaluates $p(\nu,E)$ across a range of light levels and fits
the simpler curve in @eq-photosynthesis. This fitting step turns
$\alpha_{{\rm p1}}$ and $\alpha_{{\rm p2}}$ into functions of leaf nitrogen.
The rest of FF16 can then use the compact photosynthesis equation while
retaining the effect of $\nu$ derived from the detailed annual calculation.

<!-- SY: Need a link to the assimilation and hydraulics page? Or a what next to read? -->

## Tables {#tables}

The following tables collect the notation, allometric relationships, and
parameter values used above. They are generated directly from the installed
`plant` package so that the displayed defaults match the implementation.

```{r}
#| label: functions
#| echo: false
extract_strategy_parameter_values <- function(vars) {
  s <- c(FF16_Strategy(), lapply(as.list(args(make_FF16_hyperpar)), eval))

  out <- rep(NA_character_, length(vars))

  # Now extract values from plant
  to_code <- vars != "" & vars %in% names(s)
  code <- vars[to_code]

  val <- s[code]
  err <- sapply(val, is.null)
  if (any(err)) {
    message("Missing parameters: ", paste(code[err], collapse = ", "))
  }

  oo <- options(scipen = 999)
  on.exit(options(oo))
  out[to_code] <- unlist(lapply(val, prettyNum))
  out[is.na(out)] <- ""

  out
}
```

```{r}
#| label: tbl-definitions
#| echo: false
#| tbl-cap: 'Key variables of the `FF16` physiological model. For mass ($M$), respiration ($r$), and turnover ($k$) variables, subscripts refer to any of the following tissues: ${\rm l}$ = leaves, ${\rm b}$ = bark, ${\rm s}$ = sapwood,${\rm r}$ = roots, ${\rm a}$ = all living tissue. For area $A$ variables, subscripts refer to any of the following: ${\rm l}$ = leaves, ${\rm st}$ = total stem cross-section, ${\rm s}$ = sapwood cross-section, ${\rm b}$ = bark cross-section, ${\rm h}$ = heartwood cross-section'
system.file("docs", "demography-vars.csv", package = "plant") %>%
  read.csv() %>%
  dplyr::as_tibble() %>%
  knitr::kable(align = "l")
```
<!-- SY: I believe this table contains more than just key variables of the FF16 physiological model, such as the metacommunity / patch state variables -->

```{r}
#| label: tbl-allometry
#| echo: false
#| tbl-cap: 'Equations of the allometric growth model implemented within the `FF16` physiological model, based on functional-balance assumptions (see text for details). In (a) we list  key assumptions of allocation model between specific variation with leaf area, under "Function". From these assumptions, allocation functions are derived for tissue areas and tissue masses (b). Also the growth rate of each tissue type can be expressed as a function of the growth rate of leaf area. For mass ($M$), respiration ($r$), and turnover ($k$) variables, subscripts refer to any of the following tissues: ${\rm l}$ = leaves, ${\rm b}$ = bark, ${\rm s}$ = sapwood, ${\rm r}$ = roots, ${\rm a}$ = all living tissue. For area $A$ variables, subscripts refer to any of the following: ${\rm l}$ = leaves, ${\rm st}$ = total stem cross-section, ${\rm s}$ = sapwood cross-section, ${\rm b}$ = bark cross-section, ${\rm h}$ = heartwood cross-section'
system.file("docs/FF16", "FF16-eqns.csv", package = "plant") %>%
  read.csv() %>% dplyr::as_tibble() %>%
  knitr::kable(align = "l")
```

```{r}
#| label: tbl-params-core
#| echo: false
#| tbl-cap: 'Core parameters of the `FF16` physiological model.'
x <- system.file("docs/FF16", "FF16-params-core.csv", package = "plant") %>%
  read.csv() %>% dplyr::as_tibble()

x$Value <- extract_strategy_parameter_values(x$Symbol)
knitr::kable(x,  align = "l")
```

<!-- SY: value column is empty, so should fill or remove -->

```{r}
#| label: tbl-params-hyper
#| echo: false
#| tbl-cap: 'Parameters for hyper-parameterisation of the `FF16` physiological model.'
x <- system.file("docs/FF16", "FF16-params-hyper.csv", package = "plant") %>%
  read.csv() %>% dplyr::as_tibble()

x$Value <- extract_strategy_parameter_values(x$Code)
knitr::kable(x,  align = "l")
```

## References
