---
title: "K93"
subtitle: "Kohyama 1993 — the forest architecture hypothesis for the stable coexistence of species"
---
::: {.eyebrow}
model
:::

::: {.callout-note}
## Prerequisites
This page builds on [The big picture](../overview.qmd). Read it first if the notation here is unfamiliar.
:::

## The idea first

K93 is easiest to understand as a size-based demographic model. For each tree,
the model asks two main questions:

1. How large is the tree?
2. How much basal area is contributed by trees larger than it?

The second quantity is a measure of competition. A small tree beneath many
larger neighbours experiences more competition than a large tree with little
basal area above it. K93 uses the tree's diameter and this competition measure
to calculate its growth, mortality, and fecundity.

All species use the same *form* of these equations, but they need not behave
identically. Species can have different coefficient values, and those values
change how strongly size and competition affect their vital rates. The
three-species example near the end of this page illustrates this distinction.

This is a different starting point from the trait-based FF16 model. FF16 links
traits such as leaf mass per area and wood density to a mechanistic carbon
budget, then derives vital rates from that budget. K93 instead describes vital
rates directly as functions of observed size and competition. In that sense,
K93 is a phenomenological model of forest demography rather than a mechanistic
model of plant physiology.

The model was introduced in @Kohyama-1993. This page does not present the
equations as new results; it explains how `plant` represents and solves them.
Within the package, K93 is implemented through the same broad interface as the
other physiological models: an individual's current state and environment are
used to calculate its growth, mortality, and fecundity rates.

```{r}
#| echo: false
#| label: setup
library(plant)
```

## Model description

The equations below share one important environmental quantity:
$B(t,a,x)$, the cumulative basal area of trees larger than a focal tree of size
$x$. A large value of $B$ means that the focal tree has many, or very large,
competitors above it. The [competition equation](#size-structured-competition)
defines this quantity in full.

### Growth

Eq. 10 of @Kohyama-1993:

$$
G(t, a, x) = x \cdot \Bigg(b_0 - b_1 \cdot \ln{x} - b_2 \cdot B(t, a, x)\Bigg)
$$ {#eq-k93-growth}

The growth rate $G$ is proportional to current diameter $x$. Inside the
parentheses, $b_0$ sets the baseline growth rate, $b_1$ controls how growth
changes with size, and $b_2$ controls the reduction caused by competition.
As the tree becomes larger, the $b_1\ln{x}$ term reduces its proportional
growth rate. More basal area above the tree also reduces growth.

### Fecundity

Eq. 12 of @Kohyama-1993:

$$
R(t, a, x_0) = d_0 \cdot B_0 \cdot \exp{(-d_1 \cdot B(t, a, x_0))}
$$ {#eq-k93-fecundity}

Fecundity $R$ increases with the focal tree's basal area, $B_0$, and decreases
exponentially as competition increases. The coefficient $d_0$ sets reproductive
output per unit basal area, while $d_1$ determines how strongly competition
suppresses that output.

In the `plant` implementation, $B_0$ is interpreted as the basal area of one
individual. Integrating these individual contributions across the stand gives
the same total reproductive output as Eq. 9 of @Kohyama-1993.

### Mortality

Eq. 11 of @Kohyama-1993:

$$
\mu(t, a, x) = - c_0 + c_1 \cdot B(t, a, x)
$$ {#eq-k93-mortality}

Competition-dependent mortality $\mu$ rises linearly with basal area above the
focal tree. The implementation does not allow this component to become
negative, so values below zero are set to zero. Mortality associated with
large-scale disturbance is handled separately by the patch disturbance regime.

### Size structured competition

Eq. 8 of @Kohyama-1993:

$$
B(t, a, x) = \frac{\pi}{4 \cdot s(t, a)} \int_{x}^{x_{max}} y^2 \cdot f(t, a, y) \, dy
$$ {#eq-k93-competition}

This integral sums the basal area of trees with diameters between the focal
tree's diameter, $x$, and the largest diameter in the stand, $x_{max}$. The
function $f(t,a,y)$ describes the density of trees of size $y$, while $s(t,a)$
provides the patch-level scaling used in the original formulation. Because the
calculation includes only trees at least as large as the focal tree, small trees
generally experience more competition than large ones.

The `plant` implementation adds a smoothing parameter, $\eta = 12.0$, so that
the competitive effect can be interpolated with a continuous function.

### Dispersal and establishment

The wider `plant` framework includes dispersal and establishment effects. They
were not part of the original K93 formulation, so both are set to $1.0$. This
choice makes them neutral: neither process changes the number of recruits.

## Examples

The examples move from one individual to a whole patch. They use the
package's generic state names, so a column or state labelled `height` should be
read as diameter at breast height (DBH) for K93.

### Individuals

First, create a K93 individual and inspect its state variables:

```{r}
#| label: k93-individual
p <- K93_Individual()

# 3 parameters
p$ode_size

# size, mortality and fecundity
p$ode_names

# initialised at 2 cm DBH
p$ode_state

# set to fixed state
p$ode_state <- c(50, 0, 0)
```

The size characteristic is currently labelled `height` by the shared package
interface, but K93 interprets it as DBH in centimetres.

An individual alone is not enough to calculate rates: the model also needs an
environment. Here we create a fixed environment, then ask the individual to
calculate the instantaneous rates of change in its state variables:

```{r}
#| label: fixed-environment
e <- 0.75
env <- K93_Environment()
env$set_fixed_environment(e, height_max = 300)

p$compute_rates(env)

p$ode_rates
```

In a fixed environment, every individual experiences the same competitive
conditions, regardless of its size. The generic `plant` environment represents
resource availability on a proportional scale from zero to one. Values near one
represent high availability; values near zero represent strong limitation.

K93 itself describes the environment as cumulative basal area above the focal
individual (@eq-k93-competition). The package maps that competition measure onto
the generic zero-to-one scale. We can reverse the mapping to recover cumulative
basal area from the fixed environmental value:

```{r}
#| label: growth-rate
# ind. parameters
s <- p$strategy

# backtransform environment to cumulative basal area
k_I <- K93_Strategy()$pars$k_I
basal_area_above <- function(e) -log(e) / k_I

# growth has both size and competition dependent terms
x <- p$ode_state[1]
x * (s$pars$b_0 - s$pars$b_1 * log(x) -
       s$pars$b_2 * basal_area_above(e))
```

The extinction coefficient $k_I$ controls the mapping between environmental
availability and cumulative basal area. Its default value is $0.01$, and it is
stored in the strategy's biological parameters as
`K93_Strategy()$pars$k_I`.

Fecundity is proportional to an individual's basal area. The original model
expresses reproduction at the level of the whole forest; here, we calculate one
individual's contribution to that total:

```{r}
#| label: fecundity
# basal area
basal_area <- pi / 4 * x^2

s$pars$d_0 * basal_area * exp(-s$pars$d_1 * basal_area_above(e))
```

As in the original paper, mortality has two parts. Competition-dependent
mortality acts on individuals, while competition-independent mortality comes
from the patch disturbance regime. The disturbance regime describes how the
risk of patch-ending disturbance changes as the patch ages. The following code
shows its mean interval. It also reports the death probability implied by the
individual's accumulated competition-dependent mortality state. That
probability is still zero here because we calculated instantaneous rates but
did not advance the individual through time:

```{r}
#| label: disturbance-mortality
# average disturbance frequency
disturbance <- Weibull_Disturbance_Regime(K93_Parameters()$max_patch_lifetime)
disturbance$mean_interval()

# current death probability implied by the accumulated hazard
p$mortality_probability
```

Next, isolate the competition-dependent mortality rate. With the default
extinction coefficient, the rate is small. Reducing $k_I$ means that the same
environmental value corresponds to more basal area above the tree, and hence to
stronger competition:

```{r}
#| label: competition-mortality
# small, but matches the mortality rate computed above
max(-s$pars$c_0 + s$pars$c_1 * basal_area_above(e), 0)

# but increasing intensity of competition increases the rate of mortality
k_I <- 1e-6
-s$pars$c_0 + s$pars$c_1 * basal_area_above(e)
```

We can confirm this result by rebuilding the individual with the modified
strategy and calculating its rates again:

```{r}
#| label: rebuild-individual
s$pars$k_I <- 1e-6
p <- K93_Individual(s)
p$ode_state <- c(50, 0, 0)
p$compute_rates(env)

p$ode_rates[2]
```

### Patches

To move from an individual to a patch, we add its strategy to a parameter
object and run the size-structured community model with `run_scm()`. Setting
`collect = TRUE` retains the state at each output time. We then reshape the
result and plot the DBH trajectory of every cohort:

```{r}
#| label: plot-patch
library(ggplot2)
library(dplyr)

class(s)
par <- K93_Parameters()
par$strategies[[1]] <- s

results <- run_scm(par, collect = TRUE)

# `K93_expand_state()` returns a tidy data frame with one row per node per
# output time, so we can plot the height of each cohort directly.
tidy <- K93_expand_state(results)$species

ggplot(tidy, aes(time, height, group = node)) +
  geom_line(colour = util_colour_set_opacity("black", 0.25)) +
  labs(x = "Time (years)", y = "Size (DBH cm)") +
  theme_bw()
```

The run stops at about 105 years because that is the default
`max_patch_lifetime` for K93. At that point, the simulated patch is still
developing. Extending its maximum lifetime to 700 years allows the oldest
cohorts to approach their maximum size of 136 cm DBH, as in Fig. 2 of
@Kohyama-1993:

```{r}
#| label: plot-patch-extended
par$max_patch_lifetime <- 700
results <- run_scm(par, collect = TRUE)

tidy <- K93_expand_state(results)$species

ggplot(tidy, aes(time, height, group = node)) +
  geom_line(colour = util_colour_set_opacity("black", 0.25)) +
  labs(x = "Time (years)", y = "Size (DBH cm)") +
  theme_bw()
```

The convenience function `scm_base_parameters()` supplies faster numerical
control settings. These settings make it practical to explore the longer run:

```{r}
#| label: plot-patch-fast
fast_par <- scm_base_parameters("K93")
fast_par$strategies[[1]] <- s
fast_par$max_patch_lifetime <- 700

results <- run_scm(fast_par, collect = TRUE)

tidy <- K93_expand_state(results)$species

ggplot(tidy, aes(time, height, group = node)) +
  geom_line(colour = util_colour_set_opacity("black", 0.25)) +
  labs(x = "Time (years)", y = "Size (DBH cm)") +
  theme_bw()
```

The previous lines give every cohort the same visual weight, even when few
individuals remain in it. Mapping line transparency to cohort density makes the
changing abundance of trees visible and helps reveal the formation of canopy
gaps:

```{r}
#| label: plot-density-weighted
tidy %>%
  filter(density > 1e-7) %>%
  ggplot(aes(time, height, group = node, alpha = density)) +
  geom_line() +
  labs(x = "Time (years)",
       y = "Size (DBH cm)",
       alpha = "Density") +
  theme_bw()
```

Finally, reproduce the three-species setup described by Kohyama. Each row of
the trait matrix supplies a different set of vital-rate coefficients. The
species therefore share the K93 equation structure but differ in their growth,
mortality, and fecundity responses:

```{r}
#| label: plot-three-species
sp <- trait_matrix(c(0.042, 0.063, 0.052,
                     8.5e-3, 0.014, 0.015,
                     2.2e-4, 4.6e-4, 3e-4,
                     0.008, 0.008, 0.008,
                     1.8e-4, 4.4e-4, 5.1e-4,
                     1.4e-4, 2.5e-3, 8.8e-3,
                     0.044, 0.044, 0.044),
                   c("b_0", "b_1", "b_2",
                     "c_0", "c_1", "d_0", "d_1"))

fast_par2 <- scm_base_parameters("K93")
fast_par2$max_patch_lifetime <- 700
fast_par2$strategy_default$pars$k_I <- 1e-6

sp_par <- add_strategies(fast_par2, sp, birth_rate = rep(1, 3))

sp_results <- run_scm(sp_par, collect = TRUE)

sp_names <- c("D. racemosum", "I. anisatum", "E. japonica")

sp_tidy <- K93_expand_state(sp_results)$species %>%
  mutate(species = sp_names[as.integer(species)]) %>%
  filter(density > 1e-7)

ggplot(sp_tidy, aes(time, height, colour = species,
                    group = paste(species, node), alpha = density)) +
  geom_line() +
  labs(x = "Time (years)",
       y = "Size (DBH cm)",
       alpha = "Density") +
  theme_bw()
```

## What to take away

K93 reduces a complex forest to a clear demographic picture: trees change in
diameter, and larger trees suppress smaller ones through cumulative basal area.
Species differ through the coefficients in their vital-rate equations, not
through an explicit carbon budget or a set of physiological traits. The
`plant` package then embeds those individual rules in its shared framework for
cohorts, patches, recruitment, and disturbance.
