---
title: "K93"
subtitle: "Kohyama 1993 — the forest architecture hypothesis for the stable coexistence of species"
---
::: {.eyebrow}
model
:::

This document outlines the `K93` physiological model used in the `plant`
package. This model was first described, and has primarily been developed,
elsewhere, in particular in @Kohyama-1993. The model's equations are presented
here not as original findings, but rather so that users can understand the full
system of equations being solved within `plant`.

The purpose of a physiological model in `plant` is to take a plant's current
size, light environment, and physiological parameters as inputs, and return its
growth, mortality, and fecundity rates. In the `K93` physiological model, these
vital rates are all estimated using simple phenomenological equations, as
functions of diameter and competition.

```{r}
#| echo: false
#| label: setup
library(plant)
```

# Model description

## Growth

Eq. 10 of @Kohyama-1993:

$$
G(t, a, x) = x \cdot \Bigg(b_0 - b_1 \cdot \ln{x} - b_2 \cdot B(t, a, x)\Bigg)
$$ {#eq-k93-growth}

## Fecundity

Eq. 12 of @Kohyama-1993:

$$
R(t, a, x_0) = d_0 \cdot B_0 \cdot \exp{(-d_1 \cdot B(t, a, x_0))}
$$ {#eq-k93-fecundity}

*$B_0$ re-interpreted as an individual basal area, integrated over the whole
stand, providing equivalent total reproductive output as Eq. 9.*

## Mortality

Eq. 11 of @Kohyama-1993:

$$
\mu(t, a, x) = - c_0 + c_1 \cdot B(t, a, x_0)
$$ {#eq-k93-mortality}

*Growth independent mortality provided by disturbance regime.*

## Size structured competition

Eq. 8 of @Kohyama-1993:

$$
B(t, a, x) = \frac{\pi}{4 \cdot s(t, a)} \int_{x}^{x_{max}} y^2 \cdot f(t, a, y) \, dy
$$ {#eq-k93-competition}

*Added additional smoothing parameter to interpolate competitive effect using a
continuous function. $\eta = 12.0$.*

## Dispersal and establishment

Dispersal and establishment are fixtures of the `plant` framework, but not used
in the original formulation of Kohyama. We set them both to $1.0$.

# Examples

## Individuals

Instantiate a K93 individual then examine the state of each characteristic:

```{r}
#| label: k93-individual
p <- K93_Individual()

# 3 parameters
p$ode_size

# size, mortality and fecundity
p$ode_names

# initialised at 2 cm DBH
p$ode_state

# set to fixed state
p$ode_state <- c(50, 0, 0)
```

*Note the size characteristic is currently hardcoded as 'height'.*

An environment is required to compute the rate of characteristic change (i.e.
gradients):

```{r}
#| label: fixed-environment
e <- 0.75
env <- K93_Environment()
env$set_fixed_environment(e, height_max = 300)

p$compute_rates(env)

p$ode_rates
```

Here, a fixed environment means that individuals of any height experience the
same competitive conditions. A `plant` environment is expressed on a
proportional scale (between zero and one) describing the availability of
resources. Environment values near one mean that individuals receive their full
allocation of resources (e.g. light in an open canopy, with no shading), while
values near zero mean that few resources are available.

The Kohyama model describes the environment in terms of cumulative basal area of
surrounding trees larger than the focal individual (@eq-k93-competition). We back
transform the environmental value into cumulative basal area such that the
smallest individuals receive the largest competitive effects from their
neighbours:

```{r}
#| label: growth-rate
# ind. parameters
s <- p$strategy

# backtransform environment to cumulative basal area
k_I <- K93_Strategy()$k_I
basal_area_above <- function(e) -log(e) / k_I

# growth has both size and competition dependent terms
x <- p$ode_state[1]
x * (s$b_0 - s$b_1 * log(x) - s$b_2 * basal_area_above(e))
```

*The extinction coefficient $k_I$ is set to 0.01 by default (it lives on the
strategy, `K93_Strategy()$k_I`).*

Fecundity is proportional to basal area — in the original formulation this is
calculated for the whole forest, but here we calculate individual contributions:

```{r}
#| label: fecundity
# basal area
basal_area <- pi / 4 * x^2

s$d_0 * basal_area * exp(-s$d_1 * basal_area_above(e))
```

Like in the original paper, mortality is split into competition dependent and
independent processes, the latter described by the average disturbance
interval/frequency of the patch and is initially zero, but will increase as the
patch develops over time.

```{r}
#| label: disturbance-mortality
# average disturbance frequency
disturbance <- Weibull_Disturbance_Regime(K93_Parameters()$max_patch_lifetime)
disturbance$mean_interval()

# cumulative hazard
p$mortality_probability
```

We focus here on competition dependent mortality of an individual:

```{r}
#| label: competition-mortality
# small, but matches the mortality rate computed above
max(-s$c_0 + s$c_1 * basal_area_above(e), 0)

# but increasing intensity of competition increases the rate of mortality
k_I <- 1e-6
-s$c_0 + s$c_1 * basal_area_above(e)
```

which we confirm by rebuilding our K93_Individual with the lower extinction
coefficient (`k_I` lives on the strategy):

```{r}
#| label: rebuild-individual
s$k_I <- 1e-6
p <- K93_Individual(s)
p$ode_state <- c(50, 0, 0)
p$compute_rates(env)

p$ode_rates[2]
```

## Patches

First we need to add a species to the model parameters, then we run the
characteristic solver `scm`, collecting the results at each timestep. Finally, we
plot the height of each cohort of individuals over time:

```{r}
#| label: plot-patch
library(ggplot2)
library(dplyr)

class(s)
par <- K93_Parameters()
par$strategies[[1]] <- s

results <- run_scm(par, collect = TRUE)

# `K93_expand_state()` returns a tidy data frame with one row per node per
# output time, so we can plot the height of each cohort directly.
tidy <- K93_expand_state(results)$species

ggplot(tidy, aes(time, height, group = node)) +
  geom_line(colour = util_colour_set_opacity("black", 0.25)) +
  labs(x = "Time (years)", y = "Size (DBH cm)") +
  theme_bw()
```

The run stops at around 105 years because that is the default
`max_patch_lifetime` for `K93` — the patch is still developing when it ends.
Increasing the maximum patch lifetime extends the simulation to 700 years and
allows the first cohorts to reach their maximum size of 136cm DBH, matching
Fig. 2 of @Kohyama-1993:

```{r}
#| label: plot-patch-extended
par$max_patch_lifetime <- 700
results <- run_scm(par, collect = TRUE)

tidy <- K93_expand_state(results)$species

ggplot(tidy, aes(time, height, group = node)) +
  geom_line(colour = util_colour_set_opacity("black", 0.25)) +
  labs(x = "Time (years)", y = "Size (DBH cm)") +
  theme_bw()
```

*Note: we obtain more interesting results using the fast control parameters
provided by the convenience function `scm_base_parameters`.*

```{r}
#| label: plot-patch-fast
fast_par <- scm_base_parameters("K93")
fast_par$strategies[[1]] <- s
fast_par$max_patch_lifetime <- 700

results <- run_scm(fast_par, collect = TRUE)

tidy <- K93_expand_state(results)$species

ggplot(tidy, aes(time, height, group = node)) +
  geom_line(colour = util_colour_set_opacity("black", 0.25)) +
  labs(x = "Time (years)", y = "Size (DBH cm)") +
  theme_bw()
```

Weighting each cohort by its density shows how gap-formation can occur:

```{r}
#| label: plot-density-weighted
tidy %>%
  filter(density > 1e-7) %>%
  ggplot(aes(time, height, group = node, alpha = density)) +
  geom_line() +
  labs(x = "Time (years)",
       y = "Size (DBH cm)",
       alpha = "Density") +
  theme_bw()
```

The Kohyama model also describes a patch with three species:

```{r}
#| label: plot-three-species
sp <- trait_matrix(c(0.042, 0.063, 0.052,
                     8.5e-3, 0.014, 0.015,
                     2.2e-4, 4.6e-4, 3e-4,
                     0.008, 0.008, 0.008,
                     1.8e-4, 4.4e-4, 5.1e-4,
                     1.4e-4, 2.5e-3, 8.8e-3,
                     0.044, 0.044, 0.044),
                   c("b_0", "b_1", "b_2",
                     "c_0", "c_1", "d_0", "d_1"))

fast_par2 <- scm_base_parameters("K93")
fast_par2$max_patch_lifetime <- 700
fast_par2$strategy_default$k_I <- 1e-6

sp_par <- expand_parameters(sp, fast_par2, birth_rate_list = rep(1, 3))

sp_results <- run_scm(sp_par, collect = TRUE)

sp_names <- c("D. racemosum", "I. anisatum", "E. japonica")

sp_tidy <- K93_expand_state(sp_results)$species %>%
  mutate(species = sp_names[as.integer(species)]) %>%
  filter(density > 1e-7)

ggplot(sp_tidy, aes(time, height, colour = species,
                    group = paste(species, node), alpha = density)) +
  geom_line() +
  labs(x = "Time (years)",
       y = "Size (DBH cm)",
       alpha = "Density") +
  theme_bw()
```
