---
title: "Scientific versioning of models"
subtitle: "Telling a scientific change apart from a software change"
---
::: {.eyebrow}
theory
:::

`plant` has a single software version — its package `Version` — but the
*scientific* models it ships evolve on a different clock than the code. `FF16`
and `K93` encode published theory [@Falster-2015; @Kohyama-1993] and rarely
change, while `TF24` and its fast variant `TF24f` are under active development. A
new `plant` release might be a pure refactor, a performance tweak, or a genuine
change to the equations — the package version moves in all three cases, so on its
own it cannot tell you whether a simulation's *results* would change.

That distinction matters most when you archive a run. What you are really
archiving is a *result of the science*, and the question that matters months
later is not "which release of the package produced this?" but "**is the science
still the same, so this archived run is still valid?**" Re-running every archived
simulation on every release would be wasteful; never re-running them would be
wrong.

The **scientific version** exists to answer exactly that question. It is a small
integer, carried per model, independent of the package `Version`, that changes
*only* when a model's equations or default parameters change the simulation
output for identical inputs — never for refactors, performance work, or interface
changes.

## Reading a model's version

```{r setup}
#| message: false
library(plant)
```

`model_version()` returns the scientific version as a string, and `model_id()`
pairs it with the model name — the identifier used to label archived runs:

```{r read-version}
model_version("FF16")   # scientific version
model_id("FF16")        # name-and-version identifier
```

The versions the installed `plant` currently ships are:

```{r all-ids}
sapply(c("FF16", "K93", "TF24", "TF24f"), model_id)
```

`FF16` and `K93` are scientifically frozen at `v1`. `TF24` began at `v2` rather
than `v1` because a published result used the pre-versioning "v1" science; the
integer has advanced since as its defaults have been recalibrated. `TF24f`
carries a compound version (see below).

## The two versions, and why they are separate

|  | changes when… | example |
|---|---|---|
| Package `Version` | *anything* ships — bug fixes, refactors, performance, interface changes | renaming an argument |
| Scientific version | the simulation output changes for identical inputs — an equation or a default parameter moves | recalibrating a default trait value |

A refactor that leaves every number identical bumps the package `Version` but
**not** the scientific version. That is the whole point: an archived run stays
valid across such releases, and downstream tools can tell.

## Compound versions: `TF24f`

Most models carry a single integer. The exception is a model defined *relative to
another*. `TF24f` ("f" for fast) is a fast approximation of `TF24` — same
biology, cheaper evaluation — so its version has two components,
`"<TF24 version>.<approximation revision>"`:

```{r tf24f}
model_id("TF24f")
```

- The **major** component auto-tracks `TF24`'s scientific version. This is the
  safe direction: any scientific change to `TF24` also invalidates `TF24f`,
  automatically, with no second edit. Because `TF24f` is built on `TF24`, a
  `TF24` change usually shifts `TF24f`'s output too, and an unnecessary rerun is
  always safe whereas a missed one silently serves stale results.
- The **minor** component (`approximation_revision`) tracks changes to the
  approximation itself — a change that affects `TF24f` but not `TF24`.

So a `TF24` science bump advances the major component while leaving the minor
component in place, and tightening the approximation advances the minor
component alone.

## Where the number lives, and the bump rule

There is a single source of truth, and it is in C++, sitting next to the
equations it versions — the `scientific_version` constant on each strategy class:

```cpp
// inst/include/plant/models/ff16_strategy.h
static constexpr int scientific_version = 1;
```

`TF24f` instead declares `approximation_revision`; the compound string is
assembled in `src/strategy_version.cpp`, which the R accessors call through. R
holds no duplicated copy of the number.

**The rule for bumping it:** increment the constant when, and only when, a change
moves simulation output for identical inputs — a changed equation or a changed
default parameter. Do *not* bump it for refactors, performance work, or interface
changes. When you add a new strategy, add its `scientific_version` constant to the
model header and a dispatch arm in `src/strategy_version.cpp`.

## The drift guard

Reviewer discipline is fallible, so a test backs the rule for the common,
mechanical case — a default parameter that quietly changes without a version
bump. `tests/testthat/test-model-version.R` snapshots each model's *scientific
surface*: its `model_id` together with the default strategy (biological `pars`,
numerical `control`, and birth-rate defaults), each value formatted to full
precision.

If a default changes but the version constant does **not**, the `model_id` in the
snapshot stays the same while the surrounding values differ — so the snapshot
test fails. The developer is then forced to choose: bump the version (if the
change is scientific) or accept the new snapshot with
`testthat::snapshot_accept()` (if it is not). The guard does not catch *pure
equation* changes that leave every default untouched — those still rely on
reviewer discipline and the bump convention in the header comments.

## How downstream tools use it

The reason the version is a stable, machine-readable string is so that tools
built on top of `plant` can key their caches on it.
[`logpile`](https://github.com/traitecoevo/logpile) archives `plant` simulations
in a content-addressed cache: a run is identified by a hash of its resolved
inputs, and the model's scientific version is one of those inputs. So:

- **Reuse.** A software-only `plant` release leaves the scientific version
  untouched, the hash is unchanged, and archived results are reused — you don't
  pay to recompute simulations whose science hasn't moved.
- **Rerun.** When a model's scientific version is bumped, the hash changes and
  the affected simulations are recomputed. `FF16`/`K93` runs stay put; `TF24`
  runs re-run when `TF24`'s science advances.

You never type the version yourself. `logpile` reads it from the installed
`plant`, so you can name a model bare and let `plant` supply the version:

```r
logpile::resolve_request(list(model_id = "TF24"))$model_id
#> "TF24@v3"
```

Pinning a version that disagrees with the installed `plant` is a hard error
rather than a silent stale-cache hit — a safeguard against accidentally reading
results from a different scientific version than you think you're running.

## For developers

See `plant`'s `agents.md` (§7) for the bump rules; family-wide conventions live
in [`plant-meta`](https://github.com/traitecoevo/plant-meta).

# References
