---
title: "Scientific versioning of models"
subtitle: "Telling a scientific change apart from a software change"
---
::: {.eyebrow}
theory
:::

There are two different kinds of change to keep track of when you use `plant`.

The first is a change to the **software**. The package has a `Version`, just like
any other R package. That version may move because an argument was renamed, some
code was made faster, a bug was fixed, or the model itself changed.

The second is a change to the **scientific model**. This happens when an equation
or a default parameter changes in a way that changes the simulation result for
the same inputs. `FF16` and `K93` encode established published models
[@Falster-2015; @Kohyama-1993] and therefore change rarely. `TF24` and its fast
variant, `TF24f`, are still being developed more actively.

The package version alone cannot distinguish these cases. A performance
improvement and a new growth equation may both produce a new release, even
though only the second change should alter the scientific result.

This matters when simulations are expensive and their outputs are archived. If
only the software implementation changes, an existing result can usually be
reused. If the model's science changes, that result may need to be recomputed.
Re-running everything after every software release would waste time, but never
re-running anything could leave you using results from an older model.

To make that decision explicit, each model has its own **scientific version**.
This version is independent of the package `Version`. It advances only when a
change to the model's equations or default parameters changes its output for
identical inputs. Refactors, performance work, and interface changes do not
advance it.

## Reading a model's version

```{r setup}
#| message: false
library(plant)
```

Use `model_version()` when you need only the scientific version. It returns a
string because some models have compound versions. Use `model_id()` when you
want the model name and version together in an identifier suitable for labelling
an archived run:

```{r read-version}
model_version("FF16")   # scientific version
model_id("FF16")        # name-and-version identifier
```

To inspect all four models in the installed copy of `plant`:

```{r all-ids}
sapply(c("FF16", "K93", "TF24", "TF24f"), model_id)
```

`FF16` and `K93` are scientifically frozen at `v1`, meaning that scientific
changes to them should be exceptional. `TF24` began at `v2` rather than `v1`
because a published result used the pre-versioning “v1” science. Its version has
advanced as the model has developed. `TF24f` uses a compound version, explained
below.

## The two versions, and why they are separate

| Version | Changes when… | Example |
|---|---|---|
| Package `Version` | *anything* ships — bug fixes, refactors, performance, interface changes | renaming an argument |
| Scientific version | the simulation output changes for identical inputs — an equation or a default parameter moves | recalibrating a default trait value |

For example, suppose a refactor makes a simulation run faster but leaves every
number unchanged. The package `Version` advances because new software was
released. The model's scientific version does not advance because its scientific
meaning and output have not changed.

Now suppose a default mortality parameter is recalibrated. The package version
advances, and so does the scientific version of the affected model. An archived
run made with the old default can then be recognised as belonging to the earlier
science.

## Compound versions: `TF24f`

Most models need only one number. `TF24f` is different because it is defined
relative to another model. It is a fast approximation of `TF24`: it shares the
underlying biology but uses a cheaper way to track the leaf's operating point.

Its version therefore has two components:

`"<TF24 version>.<approximation revision>"`

```{r tf24f}
model_id("TF24f")
```

- The **major** component automatically follows the scientific version of
  `TF24`. Because `TF24f` is built on `TF24`, a scientific change to `TF24` may
  also change `TF24f`'s output. Following the parent version ensures that
  archived `TF24f` results are reconsidered after such a change.
- The **minor** component (`approximation_revision`) tracks changes to the
  approximation itself. These changes affect `TF24f` but not `TF24`.

For example, a scientific change to `TF24` advances the first component while
leaving the second in place. A change only to the fast approximation advances
the second component.

## Where the number lives, and the bump rule

The scientific version has one source of truth. It is declared in C++, beside
the strategy implementation that it describes, as the `scientific_version`
constant on each strategy class:

```cpp
// inst/include/plant/models/ff16_strategy.h
static constexpr int scientific_version = 1;
```

`TF24f` instead declares `approximation_revision`. The code in
`src/strategy_version.cpp` combines this revision with the current `TF24`
version. The R functions read the result from the compiled code, so there is no
second copy of the version number to keep in sync.

The rule is deliberately narrow: increment the scientific version when, and
only when, a changed equation or default parameter changes simulation output
for identical inputs. Do not increment it for refactors, performance work, or
interface changes.

When adding a new strategy, developers must add a `scientific_version` constant
to its model header and a matching dispatch arm in
`src/strategy_version.cpp`.

## The drift guard

The version still has to be changed by a person, so `plant` includes a test to
catch one common mistake: changing a default without changing the scientific
version. `tests/testthat/test-model-version.R` records a snapshot of each
model's *scientific surface*. The snapshot contains the `model_id` and the
default strategy, including biological parameters (`pars`), numerical settings
(`control`), and birth-rate defaults. Numeric values are stored at full
precision.

If a default changes while the version stays fixed, the snapshot changes but
the `model_id` does not, and the test fails. The developer must then decide
whether to bump the scientific version or, if the change is genuinely
non-scientific, deliberately accept the new snapshot.

This guard has a limit: it cannot detect a change to an equation when all
defaults remain unchanged. Those changes still depend on code review and the
versioning rule documented beside the strategy.

## How downstream tools use it

The version is machine-readable so downstream tools can treat it as part of a
simulation's identity. [`logpile`](https://github.com/traitecoevo/logpile), for
example, archives `plant` simulations in a content-addressed cache. It hashes a
resolved set of inputs, including the versioned model identifier.

- **After a software-only change**, the scientific version is unchanged. With
  the same inputs, the request identity is unchanged and the archived result
  can be reused.
- **After a scientific change**, the model identifier changes. The request is
  therefore different, so a result from the earlier scientific version is not
  mistaken for the new one.

At present, `logpile` expects an explicit versioned model identifier rather than
a bare name such as `"TF24"`. Generate that identifier from the installed copy
of `plant`, then include it in the request:

```r
tf24_id <- plant::model_id("TF24")
logpile::resolve_request(list(model_id = tf24_id))$model_id
#> "TF24@v3"
```

This makes the link between the request and the installed scientific model
explicit. If you are reproducing an older archived run, keep its older
identifier with the archive rather than silently relabelling it as current.

## For developers

See `plant`'s `agents.md` (§7) for the maintenance rules. Family-wide
conventions live in
[`plant-meta`](https://github.com/traitecoevo/plant-meta).

# References
